### 2015-12-31 scripts/figures.Rmd
initial work on figures

---
title: "Figures"
author: "JT McCrone"
date: "December 30, 2015"
output: html_document
---
```{r,settings_and_lib}
require(ggplot2)
require("plyr")
require("Biostrings")
require(reshape2)
require("pROC")
require(gridExtra)
theme_set(new = theme_classic(base_size=24))
```

## 1) Miseq out of the box
### ROC skematic
```{r, roc_skematic}
set.seed(42) # make the figures reproducible
example<-data.frame(Condition=c(rep(x = c(TRUE,FALSE),each=20)),test=c(runif(20,-10,-4),runif(20,-6,0))) # Make a dataframe of T and F positives with random p values that favors a lower distribution for the TP
ex.roc<-roc(example$Condition~example$test) # make an ROC based on this data

cord<-coords(ex.roc,x="all") # get the coordinates of the ROC
cord.l<-melt(cord) # melt the coodinates for ease of plotting
cord.l<- mutate(cord.l,Id=c(0,head(as.numeric(rownames(cord.l))%/%3,-1))) # Add an Id so that the specificity and sensitivity at a given threshold can be groupled together

# Set an id column to group the threshold,specificity, and sensitivity together with the same number by integer division. I adjust with c(0, head ..., -1) since 3%/%3=1 but it should be grouped with the 0's.
roc.df<-dcast(cord.l,Id~Var1) # dcast into columns of threshold, specificity and sensitivity
roc.df<-roc.df[order(roc.df$sensitivity),] # So that the points connect nicely

# Plot the roc to be used later
roc.p<-ggplot(roc.df,aes(x=1-specificity,y=sensitivity))+geom_step(color="black",size=2)+xlab("1-Specificity")+ylab("Sensitivity")+theme(axis.title.y=element_text(vjust=1))

# png("../../Talks/figures/roc.ex.png")
# roc.p
# dev.off()

# plot the point distribution to be used later
points.p<-ggplot(example,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center")+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks = element_blank(), axis.text.x = element_blank())

#+scale_fill_brewer(palette="Set1")#+theme(text = element_text(size=30))

#This function takes in a specificity and adds lines indicating where this specificity is found on the ROC plotted above
plot.roc<-function(x){ 
  above.sense.cut<-subset(roc.df,specificity>=x,select=c(threshold,specificity,sensitivity)) # get the points where specificity is above the cutoff
  sense.min<-min(above.sense.cut$specificity) # get the minimum of these to get as close to desired sensitivity as possible
  sense.cut<-subset(above.sense.cut,specificity==sense.min) # get all rows where sensitivity is at this point
  spec.max<-max(sense.cut$sensitivity) # but there can be different FPR for each TPR so here we maximize specificity
  output<-subset(sense.cut,sensitivity==spec.max)
  lines<-data.frame(sense=c(output$sensitivity,output$sensitivity,0), # make a data frame with the coordinates for the lines whicha re to be added
                    spec=c(0,1-output$specificity,1-output$specificity))
  p1<-roc.p+geom_line(data=lines,aes(x=spec,y=sense),size=1.5,lty=2)+xlab("1-Specificity")+ylab("Sensitivity") # plot

  p1<-p1+theme(axis.title.y = element_text(vjust=1.2))+theme(axis.title.x = element_text(vjust=0))
  print(p1)
}

#This function takes in a specificity and adds a line to show where this specificity is reached in the p value distribution above

plot.pv<-function(x){
  above.sense.cut<-subset(roc.df,specificity>=x,select=c(threshold,specificity,sensitivity)) # get the points where sensitivity is above the cutoff
  sense.min<-min(above.sense.cut$specificity) # get the minimum of these to get as close to desired sensitivity as possible
  sense.cut<-subset(above.sense.cut,specificity==sense.min) # get all rows where sensitivity is at this point
  spec.max<-max(sense.cut$sensitivity) # but there can be different FPR for each TPR so here we maximize specificity
  output<-subset(sense.cut,sensitivity==spec.max)
  p2<-points.p+geom_hline(yintercept=output$threshold,
                          size=1.5,lty=2)
  p2<-p2+scale_x_continuous(breaks=c(1),labels=c(""))
  p2<-p2+xlab("")+ylab("Log(p value)")
  p2<-p2+theme(axis.title.y = element_text(vjust=1.2))
  print(p2)
}

pdf("../results/figures/pv.1.pdf")
plot.pv(0.35)
dev.off()

pdf("../results/figuresROC.1.pdf")
plot.roc(0.35)
m dev.off()

pdf("../results/figures/pv.2.pdf")
plot.pv(0.95)
dev.off()

pdf("../results/figures/ROC.2.pdf")
plot.roc(0.95)
dev.off()

```
Now it might also be useful to include a figure that has 2 cutoffs and outputs the plots side by side.

```{r,roc_skematic_onplot}
#Helper function that gets the threshold and sensitivity from and roc at a given speceificity
helper.roc<-function(spec,roc.df){
  above.sense.cut<-subset(roc.df,specificity>=spec,select=c(threshold,specificity,sensitivity)) # get the points where specificity is above the cutoff
  sense.min<-min(above.sense.cut$specificity) # get the minimum of these to get as close to desired sensitivity as possible
  sense.cut<-subset(above.sense.cut,specificity==sense.min) # get all rows where sensitivity is at this point
  spec.max<-max(sense.cut$sensitivity) # but there can be different FPR for each TPR so here we maximize specificity
  output<-subset(sense.cut,sensitivity==spec.max)
}

plot.ex<-function(spec1,spec2,roc.df,distribution){ 
  cut1<-helper.roc(spec1,roc.df)#dataframe with the threshold,sensitivity and specificity at the given specificity
  cut2<-helper.roc(spec2,roc.df)
  output<-rbind(cut1,cut2)
  lines<-data.frame(sense=c(output$sensitivity,output$sensitivity,0,0), # make a data frame with the coordinates for the lines whicha re to be added
                    spec=c(0,0,1-output$specificity,1-output$specificity), # really 1-specificity
                    group=rep(c(spec1,spec2),times=3))

  #plot the roc
  roc.p<-ggplot(roc.df,aes(x=1-specificity,y=sensitivity))+geom_step(color="black",size=2)+xlab("1-Specificity")+ylab("Sensitivity")+theme(axis.title.y=element_text(vjust=1))
  roc.p<-roc.p+geom_line(data=lines,aes(x=spec,y=sense,color=as.factor(group)),size=c(1.5),lty=c(2))+scale_color_brewer(palette="Set1")

 roc.p<-roc.p+theme(axis.title.y = element_text(vjust=1.2))+theme(axis.title.x = element_text(vjust=0))+theme(legend.position="none")

 #plot the points and cut offs
 points.p<-ggplot(distribution,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center")+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks = element_blank(), axis.text.x = element_blank(),axis.title.y = element_text(vjust=1.2))

 points.p<-points.p+geom_hline(yintercept=output$threshold,
                          size=1.5,lty=2,color=c("#e41a1c","#377eb8")) # these are the first 2 colors in set 1

 grid.arrange(points.p,roc.p,ncol=2)

}
plot.ex(0.35,0.95,roc.df,example)

pdf("../results/figures/roc.example.pdf")
plot.ex(0.35,0.95,roc.df,example)
dev.off()

```
### Experimental setup diagram

### DeepSNV ROC

```{r,Miseq_out_of_the_box}
find_freq<-function(Id){ # helper function to get the expected frequency from the sample name
  if(Id=="Covaris_5"){
    x=0.05} else if ( Id=="Covaris_25"){
    x=0.025} else if ( Id=="Covaris_125"){
    x=0.0125} else if ( Id=="Covaris_063"){
    x=0.0063} else if ( Id=="Covaris_016"){
    x=0.0016}
}
find_freq.v<-Vectorize(find_freq) # vectorized version of the helper function above

sum.df<-read.csv("../data/process/2014-5-30/Variants/all.sum.csv",stringsAsFactors=F,comment.char = '#')

true_snv<-read.csv("../data/reference/PR8_WSN33.csv",comment.char = '#') # read in a csv of the differences in PR8 and WSN33 as determined by sanger
true_snv<-subset(true_snv,Ref!="-" & Allele.1!="-") # remove any indels as we are not concerned with them here
mutate(true_snv,mutant=paste0(Name,"_",Ref,Ref.Pos,Allele.1))->true_snv #make a column that has names the variants as segment_ReferencePositionVariant

PR8_var<-read.csv("../data/reference/PR8_variants.csv",comment.char = "#") # read in differences between plasmid control on viral PR8 used in dilutions

mutate(PR8_var,mutant=paste0(Name,"_",Ref,Ref.Pos,Allele.1))->PR8_var #make a column that has names the variants as segment_ReferencePositionVariant
sum.df<-mutate(sum.df,category=mutation %in% true_snv$mutant) # add Column for True and false variants to variant calls
sum.df$category[sum.df$mutation %in% PR8_var$mutant]<-"PR8" # id the variants found in PR8
sum.df<-mutate(sum.df,exp.freq=find_freq.v(Id)) # add the expected frequency of the true variants

# Get information about the reference genome from the fasta file
reference.fasta<-"../data/reference/pHW2000_PR8-N_A.fa"
segments <- fasta.seqlengths(reference.fasta)
regions.bed <- data.frame(chr = gsub("[ ].*","", names(segments)), start=1, stop=segments, row.names=NULL)
regions.bed<-mutate(regions.bed,chr=as.character(chr))

possible_vars<-sum(regions.bed$stop)*3 # three possible variants/ position
id<-function(x){
  x<-strsplit(x,".",fixed=T)[[1]][1]
}

prior.seg.length<-c()

for(k in 1:length(regions.bed$chr)){ 
  prior.seg.length[k]<-sum(regions.bed$stop[1:k])  # the end positions of each segment relative to one sequence not including the trimming step
}

prior.seg.length<-c(0,prior.seg.length)

mutate(regions.bed,concat.start=start+prior.seg.length[match(chr,regions.bed$chr)],concat.stop=concat.start+stop)->regions.bed

melt(subset(regions.bed,select=-c(stop,start)),id.vars="chr")->regions.l

seg<-function (x) {
  x <- strsplit(x, ".", fixed = T)[[1]][2]
}

## These functions make an ROC by allowing pROC to calculate the positions based on only the variants that are provided, and then we adjust the senesitivity for with what we know.  For example if only 3 TP are found in one sample pROC will call that 100% Sensitivity but we will adjust it to 3/possible_tp and so forth.
sum_roc<-function(x){
  sum.df<-subset(x,category %in% c(TRUE,FALSE),select=c(category,p.val))# get the TRUE and false variant calls and p.vals
  if(length(which(sum.df$category==T))>0){ # filter out cases where there aren't any TP found
    if(length(which(sum.df$category==F))==0){
      sum.df<-rbind(sum.df,data.frame(category=F,p.val=1))
    }  
    roc(sum.df$category~sum.df$p.val,plot=F,CI=T)
  }
}

roc_df<-function(roc_analysis){ # get the coordinates and cut offs for all the points in the ROC object
  roc_analysis<-roc_analysis[unlist(lapply(roc_analysis,is.null))==F]
  all<-lapply(roc_analysis,coords,x="all")
  all.long<-lapply(all, melt) 
  all.long<-lapply(all.long, function(x){ 
    mutate(x, 
           Id=c(0,head(as.numeric(rownames(x))%/%3,-1))
    )
  }) # Set an id column to group the threshold,specificity, and sensitivity together with the same number by integer division. I adjust with c(0, head ..., -1) since 3%/%3=1 but it should be grouped with the 0's.
  roc.ls<-lapply(all.long,function(x) dcast(x,Id~Var1)) # dcast into columns of threshold, specificity and sensitivity
  roc.df<-do.call(rbind,roc.ls) # combine into data frame
  roc.df$samp<-unlist(lapply(rownames(roc.df),id)) # add sample id column

  mutate(roc.df, exp.freq=find_freq.v(samp))-> roc.df
}  
tp<-unique(true_snv$mutant)
possible_tp<-length(tp)
adjust.coords<-function(roc.df,sum.df){ # adjsut the sensitivity and specificity called by pROC  
  samp<-roc.df$samp[1]
  samp.df<-subset(sum.df,Id==samp)
  sense.factor<-length(which(samp.df$category==T))/possible_tp
  TN.samp<-length(which(samp.df$category==F))
  mutate(roc.df,adj.sensitivity=sensitivity*sense.factor,FP=(TN.samp-TN.samp*specificity),TP=sense.factor*possible_tp,adj.specificity=(possible_vars-possible_tp-FP)/((possible_vars-possible_tp)))
}

roc.ls<-dlply(sum.df,~Id,sum_roc)

roc.df<-roc_df(roc.ls,"all")

roc.df.adj<-ddply(roc.df,~samp,adjust.coords,sum.df)
roc.df.adj<-roc.df.adj[order(roc.df.adj$adj.sensitivity),]
roc.df.adj<-mutate(roc.df.adj,exp.freq=as.factor(find_freq.v(samp)))
roc.df.adj$exp.freq <- factor(roc.df.adj$exp.freq, levels = rev(levels(roc.df.adj$exp.freq)))
miseq.roc<-ggplot(roc.df.adj,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.001))

miseq.roc

roc_df.one<-function(roc_analysis,thr){ # get the coordinates and cut offs for all the points in the ROC object
  roc_analysis<-roc_analysis[unlist(lapply(roc_analysis,is.null))==F]
  all<-lapply(roc_analysis,coords,x=thr,input="thr")

  #roc.ls<-lapply(all.long,function(x) dcast(x,Id~Var1)) # dcast into columns of threshold, specificity and sensitivity
  roc.df<-as.data.frame(do.call(rbind,all)) # combine into data frame
  roc.df$samp<-rownames(roc.df) # add sample id column

  mutate(roc.df, exp.freq=find_freq.v(samp))-> roc.df
}

roc.df.adj<-ddply(roc.df,~samp,adjust.coords,sum.df)
roc.df.adj<-roc.df.adj[order(roc.df.adj$exp.freq,decreasing = T),]

knitr::kable(subset(roc.df.adj,select=c(exp.freq,TP,FP)))

```

### 2016-01-01 scripts/figures.Rmd
Miseq ROC work done

@@
 
-sum.df<-read.csv("../data/process/2014-5-30/Variants/all.sum.csv",stringsAsFactors=F,comment.char = '#')
+sum.df<-read.csv("../data/process/2014-5-30/one.sided/Variants/all.sum.csv",stringsAsFactors=F,comment.char = '#')
 
@@
 mutate(true_snv,mutant=paste0(Name,"_",Ref,Ref.Pos,Allele.1))->true_snv #make a column that has names the variants as segment_ReferencePositionVariant
-
+true_snv<-rename(true_snv,c("Name"="chr","Ref.Pos"="pos")) # rename the columns for use later
 PR8_var<-read.csv("../data/reference/PR8_variants.csv",comment.char = "#") # read in differences between plasmid control on viral PR8 used in dilutions
@@
 mutate(PR8_var,mutant=paste0(Name,"_",Ref,Ref.Pos,Allele.1))->PR8_var #make a column that has names the variants as segment_ReferencePositionVariant
+PR8_var<-rename(PR8_var,c("Name"="chr","Ref.Pos"="pos")) # rename the columns for use later  
 sum.df<-mutate(sum.df,category=mutation %in% true_snv$mutant) # add Column for True and false variants to variant calls
@@
 
-# Get information about the reference genome from the fasta file
-reference.fasta<-"../data/reference/pHW2000_PR8-N_A.fa"
-segments <- fasta.seqlengths(reference.fasta)
-regions.bed <- data.frame(chr = gsub("[ ].*","", names(segments)), start=1, stop=segments, row.names=NULL)
-regions.bed<-mutate(regions.bed,chr=as.character(chr))
+# Get information about the regions we can investigate based on the primer sites
 
-possible_vars<-sum(regions.bed$stop)*3 # three possible variants/ position
+regions.bed <-read.csv("../data/reference/pr8_noprimingsites.bed.csv",stringsAsFactors = F,comment.char = "#")
+regions.bed<-mutate(regions.bed,length=stop-start-1) #because the numbers in the csv are the inner most binding sites of the primers and represent the last positiions that cannot be interogated.
+
+possible_vars<-sum(regions.bed$length)*3 # three possible variants/ position
+
+# limit the PR8 and the true positive data.frames to those not excluded by the priming sites
+
+primer.cut<-function(x){ # a helper function to do this
+  chr<-unique(x$chr)
+  start<-regions.bed$start[match(x$chr,regions.bed$chr)]
+  stop<-regions.bed$stop[match(x$chr,regions.bed$chr)]
+
+  subset(x,pos>start & pos<stop)
+}
+sum.df<-ddply(sum.df,~chr,primer.cut) # interogating only the sites within primer location removes 244 called variants for two sided and 406 for one sided test
+true_snv<-ddply(true_snv,~chr,primer.cut) # interogating only the sites within primer location removes 8 potential true positive (491)
+PR8_var<-ddply(PR8_var,~chr,primer.cut) # interogating only the sites within primer location removes 12 potential PR8 differences (15)
+
 id<-function(x){
@@
 }
-
-prior.seg.length<-c()
-
-for(k in 1:length(regions.bed$chr)){ 
-  prior.seg.length[k]<-sum(regions.bed$stop[1:k])  # the end positions of each segment relative to one sequence not including the trimming step
-}
-
-prior.seg.length<-c(0,prior.seg.length)
-
-mutate(regions.bed,concat.start=start+prior.seg.length[match(chr,regions.bed$chr)],concat.stop=concat.start+stop)->regions.bed
-
-melt(subset(regions.bed,select=-c(stop,start)),id.vars="chr")->regions.l
 
@@
 possible_tp<-length(tp)
-adjust.coords<-function(roc.df,sum.df){ # adjsut the sensitivity and specificity called by pROC  
+adjust.coords<-function(roc.df,sum.df){ # adjust the sensitivity and specificity called by pROC  
   samp<-roc.df$samp[1]
@@
 
-roc.df<-roc_df(roc.ls,"all")
+roc.df<-roc_df(roc.ls)
 
@@
 roc.df.adj$exp.freq <- factor(roc.df.adj$exp.freq, levels = rev(levels(roc.df.adj$exp.freq)))
-miseq.roc<-ggplot(roc.df.adj,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.001))
+miseq.roc<-ggplot(roc.df.adj,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.001))
 
 miseq.roc
+
+pdf("../results/figures/miseq_roc.one.sided.pdf")
+miseq.roc
+dev.off()
 
@@
 
+roc.df<-roc_df.one(roc.ls,0.001)
 roc.df.adj<-ddply(roc.df,~samp,adjust.coords,sum.df)
@@
 
-knitr::kable(subset(roc.df.adj,select=c(exp.freq,TP,FP)))
+knitr::kable(subset(roc.df.adj,select=c(exp.freq,adj.sensitivity,TP,adj.specificity,FP)))
 

### 2016-01-02 scripts/figures.Rmd
Hiseq roc and figs

@@
 plot.roc(0.35)
-m dev.off()
+dev.off()
 
@@
 
+```{r,roc_helper_runctions}
+id<-function(x){
+  x<-strsplit(x,".",fixed=T)[[1]][1]
+}
+
+seg<-function (x) {
+  x <- strsplit(x, ".", fixed = T)[[1]][2]
+}
+
+sum_roc<-function(x){
+  sum.df<-subset(x,category %in% c(TRUE,FALSE),select=c(category,p.val))# get the TRUE and false variant calls and p.vals
+  if(length(which(sum.df$category==T))>0){ # filter out cases where there aren't any TP found
+    if(length(which(sum.df$category==F))==0){
+      sum.df<-rbind(sum.df,data.frame(category=F,p.val=1))
+    }  
+    roc(sum.df$category~sum.df$p.val,plot=F,CI=T)
+  }
+}
+
+roc_df<-function(roc_analysis){ # get the coordinates and cut offs for all the points in the ROC object
+  roc_analysis<-roc_analysis[unlist(lapply(roc_analysis,is.null))==F]
+  all<-lapply(roc_analysis,coords,x="all")
+  all.long<-lapply(all, melt) 
+  all.long<-lapply(all.long, function(x){ 
+    mutate(x, 
+           Id=c(0,head(as.numeric(rownames(x))%/%3,-1))
+    )
+  }) # Set an id column to group the threshold,specificity, and sensitivity together with the same number by integer division.
+I adjust with c(0, head ..., -1) since 3%/%3=1 but it should be grouped with the 0's.
+  roc.ls<-lapply(all.long,function(x) dcast(x,Id~Var1)) # dcast into columns of threshold, specificity and sensitivity
+  roc.df<-do.call(rbind,roc.ls) # combine into data frame
+  roc.df$samp<-unlist(lapply(rownames(roc.df),id)) # add sample id column
+
+  mutate(roc.df, exp.freq=find_freq.v(samp))-> roc.df
+}
+
+adjust.coords<-function(roc.df,sum.df,possible_tp,possible_vars){ # adjust the sensitivity and specificity called by pROC  
+  samp<-roc.df$samp[1]
+  samp.df<-subset(sum.df,Id==samp)
+  sense.factor<-length(which(samp.df$category==T))/possible_tp
+  TN.samp<-length(which(samp.df$category==F))
+  mutate(roc.df,adj.sensitivity=sensitivity*sense.factor,FP=(TN.samp-TN.samp*specificity),TP=sense.factor*possible_tp,adj.specificity=(possible_vars-possible_tp-FP)/((possible_vars-possible_tp)))
+}
+roc_df.one<-function(roc_analysis,thr){ # get the coordinates and cut offs for all the points in the ROC object
+  roc_analysis<-roc_analysis[unlist(lapply(roc_analysis,is.null))==F]
+  all<-lapply(roc_analysis,coords,x=thr,input="thr")
+
+  #roc.ls<-lapply(all.long,function(x) dcast(x,Id~Var1)) # dcast into columns of threshold, specificity and sensitivity
+  roc.df<-as.data.frame(do.call(rbind,all)) # combine into data frame
+  roc.df$samp<-rownames(roc.df) # add sample id column
+
+  mutate(roc.df, exp.freq=find_freq.v(samp))-> roc.df
+} 
+######### Miseq roc functions ##########
+miseq.roc<-function(sum.df,possible_tp,possible_vars){ # A function to run the roc calculations and adjustments
+  roc.ls<-dlply(sum.df,~Id,sum_roc)
+
+  roc.df<-roc_df(roc.ls)
+
+  roc.df.adj<-ddply(roc.df,~samp,adjust.coords,sum.df,possible_tp,possible_vars)
+  roc.df.adj<-roc.df.adj[order(roc.df.adj$adj.sensitivity),]
+  roc.df.adj<-mutate(roc.df.adj,exp.freq=as.factor(find_freq.v(samp)))
+  roc.df.adj$exp.freq <- factor(roc.df.adj$exp.freq, levels = rev(levels(roc.df.adj$exp.freq)))
+  return(roc.df.adj)
+}
+miseq.roc.table<-function(sum.df,cut.off,possible_tp,possible_vars){
+  roc.ls<-dlply(sum.df,~Id,sum_roc)
+  roc.df<-roc_df.one(roc.ls,cut.off)
+  roc.df.adj<-ddply(roc.df,~samp,adjust.coords,sum.df,possible_tp,possible_vars)
+  roc.df.adj<-roc.df.adj[order(roc.df.adj$exp.freq,decreasing = T),]
+  roc.table<-subset(roc.df.adj,select=c(exp.freq,adj.sensitivity,TP,adj.specificity,FP))
+  return(roc.table)
+}
+
+######### Hiseq roc functions ##########
+hiseq.roc<-function(sum.df,possible_tp,possible_vars){ # A function to run the roc calculations and adjustments
+  roc.ls<-dlply(sum.df,~Id,sum_roc)
+
+  roc.df<-roc_df(roc.ls)
+
+  roc.df.adj<-ddply(roc.df,~samp,adjust.coords,sum.df,possible_tp,possible_vars)
+  roc.df.adj<-roc.df.adj[order(roc.df.adj$adj.sensitivity),]
+
+  mutate(roc.df.adj, gc = as.numeric(sub(pat,"\\1",samp)),
+       exp.freq = sub(pat,"\\2",samp)
+  )-> roc.df.adj
+  mutate(roc.df.adj,exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->roc.df.adj
+  roc.df.adj$exp.freq<-as.factor(roc.df.adj$exp.freq)
+  roc.df.adj$exp.freq <- factor(roc.df.adj$exp.freq, levels = rev(levels(roc.df.adj$exp.freq)))
+  return(roc.df.adj)
+}
+
+hiseq.roc.table<-function(sum.df,cut.off,possible_tp,possible_vars){
+  roc.ls<-dlply(sum.df,~Id,sum_roc)
+  roc.df<-roc_df.one(roc.ls,cut.off)
+  roc.df.adj<-ddply(roc.df,~samp,adjust.coords,sum.df,possible_tp,possible_vars)
+mutate(roc.df.adj, gc = as.numeric(sub(pat,"\\1",samp)),
+       exp.freq = sub(pat,"\\2",samp)
+)-> roc.df.adj
+mutate(roc.df.adj,exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->roc.df.adj
+roc.df.adj<-roc.df.adj[order(roc.df.adj$exp.freq,decreasing = T),]
+roc.table<-subset(roc.df.adj,select=c(gc,exp.freq,adj.sensitivity,TP,adj.specificity,FP))
+return(roc.table)
+}
+```
+
 ```{r,Miseq_out_of_the_box}
@@
 
-id<-function(x){
-  x<-strsplit(x,".",fixed=T)[[1]][1]
-}
-
-seg<-function (x) {
-  x <- strsplit(x, ".", fixed = T)[[1]][2]
-}
-
 ## These functions make an ROC by allowing pROC to calculate the positions based on only the variants that are provided, and then we adjust the senesitivity for with what we know.
 For example if only 3 TP are found in one sample pROC will call that 100% Sensitivity but we will adjust it to 3/possible_tp and so forth.
-sum_roc<-function(x){
-  sum.df<-subset(x,category %in% c(TRUE,FALSE),select=c(category,p.val))# get the TRUE and false variant calls and p.vals
-  if(length(which(sum.df$category==T))>0){ # filter out cases where there aren't any TP found
-    if(length(which(sum.df$category==F))==0){
-      sum.df<-rbind(sum.df,data.frame(category=F,p.val=1))
-    }  
-    roc(sum.df$category~sum.df$p.val,plot=F,CI=T)
-  }
-}
 
-roc_df<-function(roc_analysis){ # get the coordinates and cut offs for all the points in the ROC object
-  roc_analysis<-roc_analysis[unlist(lapply(roc_analysis,is.null))==F]
-  all<-lapply(roc_analysis,coords,x="all")
-  all.long<-lapply(all, melt) 
-  all.long<-lapply(all.long, function(x){ 
-    mutate(x, 
-           Id=c(0,head(as.numeric(rownames(x))%/%3,-1))
-    )
-  }) # Set an id column to group the threshold,specificity, and sensitivity together with the same number by integer division.
-I adjust with c(0, head ..., -1) since 3%/%3=1 but it should be grouped with the 0's.
-  roc.ls<-lapply(all.long,function(x) dcast(x,Id~Var1)) # dcast into columns of threshold, specificity and sensitivity
-  roc.df<-do.call(rbind,roc.ls) # combine into data frame
-  roc.df$samp<-unlist(lapply(rownames(roc.df),id)) # add sample id column
-
-  mutate(roc.df, exp.freq=find_freq.v(samp))-> roc.df
-}  
 tp<-unique(true_snv$mutant)
 possible_tp<-length(tp)
-adjust.coords<-function(roc.df,sum.df){ # adjust the sensitivity and specificity called by pROC  
-  samp<-roc.df$samp[1]
-  samp.df<-subset(sum.df,Id==samp)
-  sense.factor<-length(which(samp.df$category==T))/possible_tp
-  TN.samp<-length(which(samp.df$category==F))
-  mutate(roc.df,adj.sensitivity=sensitivity*sense.factor,FP=(TN.samp-TN.samp*specificity),TP=sense.factor*possible_tp,adj.specificity=(possible_vars-possible_tp-FP)/((possible_vars-possible_tp)))
-}
 
-roc.ls<-dlply(sum.df,~Id,sum_roc)
+m.roc.df<-miseq.roc(sum.df,possible_tp,possible_vars)
 
-roc.df<-roc_df(roc.ls)
-
-roc.df.adj<-ddply(roc.df,~samp,adjust.coords,sum.df)
-roc.df.adj<-roc.df.adj[order(roc.df.adj$adj.sensitivity),]
-roc.df.adj<-mutate(roc.df.adj,exp.freq=as.factor(find_freq.v(samp)))
-roc.df.adj$exp.freq <- factor(roc.df.adj$exp.freq, levels = rev(levels(roc.df.adj$exp.freq)))
-miseq.roc<-ggplot(roc.df.adj,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.001))
+miseq.roc<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.001))
 
@@
 
-pdf("../results/figures/miseq_roc.one.sided.pdf")
-miseq.roc
-dev.off()
+# pdf("../results/figures/miseq_roc.one.sided.pdf")
+# miseq.roc
+# dev.off()
 
-roc_df.one<-function(roc_analysis,thr){ # get the coordinates and cut offs for all the points in the ROC object
-  roc_analysis<-roc_analysis[unlist(lapply(roc_analysis,is.null))==F]
-  all<-lapply(roc_analysis,coords,x=thr,input="thr")
+m.roc.table<-miseq.roc.table(sum.df,0.001,possible_tp,possible_vars)
+knitr::kable(m.roc.table)
 
-  #roc.ls<-lapply(all.long,function(x) dcast(x,Id~Var1)) # dcast into columns of threshold, specificity and sensitivity
-  roc.df<-as.data.frame(do.call(rbind,all)) # combine into data frame
-  roc.df$samp<-rownames(roc.df) # add sample id column
+rm(list = setdiff(ls(), lsf.str())) # remove all variables except the functions
+```
 
-  mutate(roc.df, exp.freq=find_freq.v(samp))-> roc.df
-}
+## 2) Hiseq out of the box
 
-roc.df<-roc_df.one(roc.ls,0.001)
-roc.df.adj<-ddply(roc.df,~samp,adjust.coords,sum.df)
-roc.df.adj<-roc.df.adj[order(roc.df.adj$exp.freq,decreasing = T),]
+### ROC
 
-knitr::kable(subset(roc.df.adj,select=c(exp.freq,adj.sensitivity,TP,adj.specificity,FP)))
+```{r,hiseq_roc}
+sum.df<-read.csv("../data/process/2015-6-23/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#") # reading in the variants from the pipeline output
+
+# Getting genome lenght and segment information from the wsn33 fasta file
+reference.fasta<-"../data/reference/wsn33_wt_plasmid.fa" 
+segments <- fasta.seqlengths(reference.fasta)
+regions.bed <- data.frame(chr = gsub("[ ].*","", names(segments)), start=12, stop=segments-13, row.names=NULL) # the univeral primers are 12 and 13 bp long
+regions.bed<-mutate(regions.bed,chr=as.character(chr))
+
+# The samples are named in the following format log(genome copy/ul)_expected frequency of variants We'll use that information here to get the gc# and exp.freq for each variant called
+
+pat<-"([0-9]+)_([0-9]+)" 
+mutate(sum.df, gc = sub(pat,"\\1",Id),
+       exp.freq = sub(pat,"\\2",Id)
+)-> sum.df
+mutate(sum.df,exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->sum.df
+
+# Id the variants as T or F 
+true_snv<-read.csv("../data/reference/mutant_id.csv",stringsAsFactor=F) # get the T variants
+sum.df<-mutate(sum.df,category=mutation %in% true_snv$mutant) # add Column for True and false variants
+wt1_mut<-subset(sum.df,Id=="WT1",select=mutation)
+wt2_mut<-subset(sum.df,Id=="WT2",select=mutation)
+
+wt_mut<-intersect(wt1_mut$mutation,wt2_mut$mutation)
+
+length(wt_mut)
+sum.df$category[sum.df$mutation %in% wt_mut]<-"wt"
+sum.df<-subset(sum.df,!(grepl("WT",Id))) # remove the WT samples
+
+possible_tp<-20 # there are 20 possible true positives in the mix
+regions.bed<-mutate(regions.bed,length=stop-start-1) #because the numbers in the csv are the inner most binding sites of the primers and represent the last positiions that cannot be interogated.
+
+possible_vars<-sum(regions.bed$length)*3 # three possible variants/ position
+
+h.roc.df<-hiseq.roc(sum.df,possible_tp,possible_vars)
+
+Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1));Hiseq.roc
+
+h.table.df<-hiseq.roc.table(sum.df,0.001,possible_tp,possible_vars)
+
+knitr::kable(subset(h.table.df,gc==5))
+```
+
+## 3) Phred and MapQ filters
+
+```{r, phred_mapq filter}
+ggplot(subset(sig.df,gc==5 & category %in% c(T,F)),aes(x=Phred,y=MapQ,color=category))+geom_point()
+
+sig.df<-subset(sum.df,p.val<0.001)
+
+ggplot(subset(sig.df,gc==5 & category %in% c(T,F)),aes(x=Phred,y=MapQ,color=category))+geom_point()
+
+cut.df<-subset(sum.df,MapQ>30 & Phred>35)
+
+h.cut.df<-hiseq.roc(cut.df,possible_tp,possible_vars)
+
+ggplot(subset(h.cut.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+
+h.cut.table<-hiseq.roc.table(cut.df,0.001,possible_tp,possible_vars)
+
+knitr::kable(subset(h.cut.table,gc==4))
+```
+
+## 4) Read position
+
+```{r,read_pos}
+ggplot(sig.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
+ggplot(subset(sig.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
+
+ggplot(cut.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
+ggplot(subset(cut.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
+
+sig.cut.df<-subset(cut.df,p.val<0.001)
+
+ggplot(sig.cut.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
+ggplot(subset(sig.cut.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
+
+ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')
 

### 2016-01-02 scripts/figures.Rmd
read position figs

@@
 
+cut.read.df<-subset(cut.df,Read_pos>32 & Read_pos<94)
+
+h.cut.read.df<-hiseq.roc(cut.read.df,possible_tp,possible_vars)
+
+ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+
+h.cut.read.table<-hiseq.roc.table(cut.read.df,0.001,possible_tp,possible_vars)
+
+knitr::kable(subset(h.cut.read.table,gc==5))
+
 ```

### 2016-01-02 scripts/figures.Rmd
makes all figures

--- 2016-01-02 scripts/figures.Rmd
@@
 ```{r, phred_mapq filter}
-ggplot(subset(sig.df,gc==5 & category %in% c(T,F)),aes(x=Phred,y=MapQ,color=category))+geom_point()
+sig.df<-subset(sum.df,p.val<0.001)
 
-sig.df<-subset(sum.df,p.val<0.001)
+ggplot(subset(sig.df,category %in% c(T,F)),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=0.5)+facet_wrap(~gc)
 
@@
 
-## 4) Read position
+## Figure 4
 
+### Read Position
 ```{r,read_pos}
@@
 
+cut.read.df<-subset(cut.df,Read_pos>32 & Read_pos<94)
+
+h.cut.read.df<-hiseq.roc(cut.read.df,possible_tp,possible_vars)
+
+ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+
+h.cut.read.table<-hiseq.roc.table(cut.read.df,0.001,possible_tp,possible_vars)
+knitr::kable(subset(h.cut.read.table,gc==5))
 ```
+### Lofreq out of the box
+```{r,lofreq}
+
+lofreq.df<-read.csv("../data/process/2015-6-23/lofreq/all.removed.mapq_vcf.vcf_csv.csv")
+
+lofreq.df$p.val<-lofreq.df$QUAL # to make the roc based on the quality scores given by lofreq
+
+lofreq.roc<-hiseq.roc(cut.df,possible_tp,possible_vars)
+
+ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+
+#lofreq claims no further monkey bussiness is needed to we'll just make a table with all its output
+ddply(lofreq.df,~Id,summarize,gc=gc[1],exp.freq=exp.freq[1],TP=length(which(category==T)),FP=length(which(category==F)))->lofreq.table
+lofreq.table<-lofreq.table[order(lofreq.table$exp.freq,decreasing = T),]
+
+knitr::kable(subset(lofreq.table,gc==5,select=c(exp.freq,TP,FP)))
+
+```
+### Frequency accuracy
+```{r, frequency_accuracy}
+# Here we'll just be using
+
+## Frequency
+cut.read.sig.df<-subset(cut.read.df,p.val<0.01)
+medians<-ddply(subset(cut.read.sig.df,category==T ),~exp.freq+gc, summarize, median=median(freq.var))
+
+# ac_freq<-ggplot(subset(sum.df,category==T),aes(x=exp.freq,y=freq.var,fill=as.factor(exp.freq)))+facet_wrap(~gc)
+
+#cut.read.sig.df$exp.freq<-as.factor(cut.read.sig.df$exp.freq)
+#cut.read.sig.df$exp.freq <- factor(cut.read.sig.df$exp.freq, levels = rev(levels(cut.read.sig.df$exp.freq)))
+
+ac_freq<-ggplot(subset(cut.read.sig.df,category==T & gc=="5"),aes(x=exp.freq,y=freq.var,fill=factor(exp.freq,levels=rev(levels(as.factor(exp.freq))))))
+
+ac_freq<-ac_freq+geom_dotplot(binaxis = "y", stackdir = "center",dotsize=1)+scale_fill_brewer(palette="Set1")
+
+ac_freq<-ac_freq+scale_y_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.05,0.1))#+theme(axis.text.y = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
+ac_freq<-ac_freq+scale_x_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.050,0.1))#+theme(axis.text.x = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
+ac_freq<-ac_freq+geom_abline(intercept = 0, slope = 1,linetype=2,size=1)
+ac_freq<-ac_freq+xlab("Expected Frequency")+ylab("Measured Frequency")+theme(legend.position="none")
+
+ac_freq<-ac_freq+geom_point(data=subset(medians,gc=="5"),mapping=aes(x=exp.freq,y=median),shape=95,size=30)
+ac_freq
+```
+
+## Figure 5 Lower inputs
+
+```{r,roc lower_inputs}
+ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+
+ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+```
+
+### Duplicates
+
+```{r,duplicates}
+sum.df<-read.csv("../data/process/2015-11-14/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#")
+pat<-"([0-9]+)_([0-9]+)"
+mutate(sum.df, exp.freq = sub(pat,"\\1",Id),
+       dup = as.numeric(sub(pat,"\\2",Id)),
+)-> sum.df
+mutate(sum.df,exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->sum.df
+
+sum.df<-subset(sum.df,Id!="PC") # remove the plasmid control that was run
+
+sum.df<-mutate(sum.df,category=mutation %in% true_snv$mutant) # add Column for True and false variants
+sum.df$category[sum.df$mutation %in% wt_mut]<-"wt"
+
+cut.df<-subset(sum.df,Phred>35 & MapQ>30 & Read_pos<94 & Read_pos>32)
+
+dups<-ddply(cut.df,~exp.freq+mutation,summarize,freq1=freq.var[1],freq2=freq.var[2],p.val1=p.val[1],p.val2=p.val[2],category=category[1],Id=Id[1])
+dups$freq2[is.na(dups$freq2)]<-0
+dups$p.val2[is.na(dups$p.val2)]<-1
+
+dups<-mutate(dups,p.val=pmax(p.val1,p.val2),Id=gsub(pat,'\\1',Id),gc=4)
+
+dups.roc<-hiseq.roc(dups,possible_tp,possible_vars)
+
+ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+
+dups.table<-hiseq.roc.table(dups,0.001,possible_tp,possible_vars)
+knitr::kable(subset(dups.table))
+
+```

### 2016-01-03 scripts/figures.Rmd
ordinations

@@
 ---
-```{r,settings_and_lib}
+
+Here are the figures (or figure panels) we are thinking of using in the Benchmarking paper.
+I will include my comments and decriptions on each and perhaps over time this will develop into the results section
+```{r,settings_and_lib,echo=F,message=F,}
 require(ggplot2)
@@
 theme_set(new = theme_classic(base_size=24))
+require(knitr)
+opts_chunk$set(fig.align = "center", warning = FALSE, echo = FALSE, fig.width = 10, 
+    fig.height = 7,message=F)
 ```
 
-## 1) Miseq out of the box
+## Figure 1
+
+The accuracy of variants callers is most appropiately described by the sensitivity and specificity of a the caller at a given qualtiy threshold.
+The sensitivity of a caller is the proportion of true, known variants identified by the algorith.
+The specificty is the the propotion of false variants corrected ignored.
+There is a trade off between sensitivity and specificity.
+One may be able to correctly identify all known variants from a data; however, if in doing the the algorithm identifies a large number of false positives then a sensitivity of 1 is of little use in practive.
+Conversly it is possible to increase the stringency of a caller to the point that it does not identify any erroneous variants however, if in doing so all the true variants are also removed, perfect specificity is of no use.
+A reciever operator curve (ROC) is a common way of visualizing the trade off between sensitivity and specificity.
+The DeepSNV algorithm returns a p value for each called variant.
+Figure 1 A depicts the relationship between various p value thresholds, and the resulting ROC using a hypothetical data set.
+
 ### ROC skematic
@@
 
-# Plot the roc to be used later
-roc.p<-ggplot(roc.df,aes(x=1-specificity,y=sensitivity))+geom_step(color="black",size=2)+xlab("1-Specificity")+ylab("Sensitivity")+theme(axis.title.y=element_text(vjust=1))
-
-# png("../../Talks/figures/roc.ex.png")
-# roc.p
-# dev.off()
-
-# plot the point distribution to be used later
-points.p<-ggplot(example,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center")+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks = element_blank(), axis.text.x = element_blank())
-
-#+scale_fill_brewer(palette="Set1")#+theme(text = element_text(size=30))
-
-#This function takes in a specificity and adds lines indicating where this specificity is found on the ROC plotted above
-plot.roc<-function(x){ 
-  above.sense.cut<-subset(roc.df,specificity>=x,select=c(threshold,specificity,sensitivity)) # get the points where specificity is above the cutoff
-  sense.min<-min(above.sense.cut$specificity) # get the minimum of these to get as close to desired sensitivity as possible
-  sense.cut<-subset(above.sense.cut,specificity==sense.min) # get all rows where sensitivity is at this point
-  spec.max<-max(sense.cut$sensitivity) # but there can be different FPR for each TPR so here we maximize specificity
-  output<-subset(sense.cut,sensitivity==spec.max)
-  lines<-data.frame(sense=c(output$sensitivity,output$sensitivity,0), # make a data frame with the coordinates for the lines whicha re to be added
-                    spec=c(0,1-output$specificity,1-output$specificity))
-  p1<-roc.p+geom_line(data=lines,aes(x=spec,y=sense),size=1.5,lty=2)+xlab("1-Specificity")+ylab("Sensitivity") # plot
-
-  p1<-p1+theme(axis.title.y = element_text(vjust=1.2))+theme(axis.title.x = element_text(vjust=0))
-  print(p1)
-}
-
-#This function takes in a specificity and adds a line to show where this specificity is reached in the p value distribution above
-
-plot.pv<-function(x){
-  above.sense.cut<-subset(roc.df,specificity>=x,select=c(threshold,specificity,sensitivity)) # get the points where sensitivity is above the cutoff
-  sense.min<-min(above.sense.cut$specificity) # get the minimum of these to get as close to desired sensitivity as possible
-  sense.cut<-subset(above.sense.cut,specificity==sense.min) # get all rows where sensitivity is at this point
-  spec.max<-max(sense.cut$sensitivity) # but there can be different FPR for each TPR so here we maximize specificity
-  output<-subset(sense.cut,sensitivity==spec.max)
-  p2<-points.p+geom_hline(yintercept=output$threshold,
-                          size=1.5,lty=2)
-  p2<-p2+scale_x_continuous(breaks=c(1),labels=c(""))
-  p2<-p2+xlab("")+ylab("Log(p value)")
-  p2<-p2+theme(axis.title.y = element_text(vjust=1.2))
-  print(p2)
-}
-
-pdf("../results/figures/pv.1.pdf")
-plot.pv(0.35)
-dev.off()
-
-pdf("../results/figuresROC.1.pdf")
-plot.roc(0.35)
-dev.off()
-
-pdf("../results/figures/pv.2.pdf")
-plot.pv(0.95)
-dev.off()
-
-pdf("../results/figures/ROC.2.pdf")
-plot.roc(0.95)
-dev.off()
-
-```
-Now it might also be useful to include a figure that has 2 cutoffs and outputs the plots side by side.
-
-```{r,roc_skematic_onplot}
 #Helper function that gets the threshold and sensitivity from and roc at a given speceificity
@@
 
+The accuracy of DeepSNV algorithm has been extensively benchmarked using a region of the HIV genome (?kb); however, we wanted to determine its accuracy in calling variants accross the entire influenza A genome ( 8 segments totally roughly 13kb).
+To this end we diluted one lab strain WSN33 into another PR8 at decreasing frequencies.
+491 single nucleotide variants were confirmed between these strains using sanger sequencing of plaque isolated samples.
+
 ### DeepSNV ROC 
-
 ```{r,roc_helper_runctions}
@@
   TN.samp<-length(which(samp.df$category==F))
-  mutate(roc.df,adj.sensitivity=sensitivity*sense.factor,FP=(TN.samp-TN.samp*specificity),TP=sense.factor*possible_tp,adj.specificity=(possible_vars-possible_tp-FP)/((possible_vars-possible_tp)))
+  mutate(roc.df,adj.sensitivity=sensitivity*sense.factor,FP=(TN.samp-TN.samp*specificity),TP=adj.sensitivity*possible_tp,adj.specificity=(possible_vars-possible_tp-FP)/((possible_vars-possible_tp)))
 }
@@
 
+The initial ROC of DeepSNV applied the viral dilutions above.
+Note for these analyses we only inlcude variants with p<0.05.
+I will extend this in the future.
+Doing so will extend the ROCs but will not change the results as we set a cut off of 0.01 at most in all the tables.
+
+> Uniform ROC axis?
+
 ```{r,Miseq_out_of_the_box}
@@
 
-m.roc.table<-miseq.roc.table(sum.df,0.001,possible_tp,possible_vars)
+m.roc.table<-miseq.roc.table(sum.df,0.01,possible_tp,possible_vars)
 knitr::kable(m.roc.table)
@@
 
-## 2) Hiseq out of the box
+## Figure 2
+
+### Hiseq expermintal setup
+
+To test our accuracy under conditions that more closely resemble those found in patient derived samples, 20 single point mutants were mixed and diluted into wild type decreasing frequencies.
+This mixtures were then diltued into viral media to yeild a spectrum of known variants at a range of input levels.
 
 ### ROC
+Out of the box here is how we do for the $10^5$ sample.
 
@@
 
-h.table.df<-hiseq.roc.table(sum.df,0.001,possible_tp,possible_vars)
+h.table.df<-hiseq.roc.table(sum.df,0.01,possible_tp,possible_vars)
 
@@
 
+It is common practice to account for base and mapping quality scores when calling variants.
+While DeepSNV is agnostic towards mapping qualty our initial analysis masked any base that had a phred score less than 30.
+Here we show that by focusing our analysis to varaints that have average phred >35 and average mapping quality >30 we can increase specificity without decrease sensitivity.
+
+> The first plot here is just to show you the distributions.
+p<0.01.
+We usually only look at this after the read cut off.
+But here you can see the cut offs to quite a bit, but do leave a bit to do.
+the cut offs of 30 and 35 can be adjusted as well.
+There must be a nice way to find the cutoffs algorithmically.
+
 ```{r, phred_mapq filter}
-sig.df<-subset(sum.df,p.val<0.001)
+sig.df<-subset(sum.df,p.val<0.01)
 
-ggplot(subset(sig.df,category %in% c(T,F)),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=0.5)+facet_wrap(~gc)
+ggplot(subset(sig.df,category %in% c(T,F)),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=0.5)+geom_hline(yintercept=30)+geom_vline(xintercept=35)+facet_wrap(~gc)
 
-ggplot(subset(sig.df,gc==5 & category %in% c(T,F)),aes(x=Phred,y=MapQ,color=category))+geom_point()
+#ggplot(subset(sig.df,gc==5 & category %in% c(T,F)),aes(x=Phred,y=MapQ,color=category))+geom_point()
 
@@
 
-h.cut.table<-hiseq.roc.table(cut.df,0.001,possible_tp,possible_vars)
+h.cut.table<-hiseq.roc.table(cut.df,0.01,possible_tp,possible_vars)
 
-knitr::kable(subset(h.cut.table,gc==4))
+knitr::kable(subset(h.cut.table,gc==5))
 ```
+
+This is still for p<0.01 and 10^5 genomes/ul viral transport media.
 
@@
 ### Read Position
+
+> Again the first plot is just to show the distribution is the significant variants that pass the map and prhed cutoff for all samples run.
+
 ```{r,read_pos}
-ggplot(sig.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
-ggplot(subset(sig.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
+# ggplot(sig.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
+# ggplot(subset(sig.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
+# 
+# ggplot(cut.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
+# ggplot(subset(cut.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
 
-ggplot(cut.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
-ggplot(subset(cut.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
-
-sig.cut.df<-subset(cut.df,p.val<0.001)
+sig.cut.df<-subset(cut.df,p.val<0.01)
 
 ggplot(sig.cut.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
-ggplot(subset(sig.cut.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
+# ggplot(subset(sig.cut.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
 
@@
 
-h.cut.read.table<-hiseq.roc.table(cut.read.df,0.001,possible_tp,possible_vars)
+h.cut.read.table<-hiseq.roc.table(cut.read.df,0.01,possible_tp,possible_vars)
 knitr::kable(subset(h.cut.read.table,gc==5))
 ```
+
+Same as above just with an added read cut off at 32 and 94 (middle 50%). p<0.01
+
 ### Lofreq out of the box
+
 ```{r,lofreq}
@@
 ```
+
 ### Frequency accuracy
+
+For p<0.01 with phred, mapq and read position cut off.
 ```{r, frequency_accuracy}
@@
 ## Figure 5 Lower inputs
-
-```{r,roc lower_inputs}
+### 10^4
+```{r,roc_lower_inputs4}
 ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
 
+knitr::kable(subset(h.cut.read.table,gc==4))
+```
+```{r,roc_lower_inputs3}
 ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+
+knitr::kable(subset(h.cut.read.table,gc==3))
 ```
@@
 sum.df<-read.csv("../data/process/2015-11-14/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#")
+#sum.df<-read.csv("~/Box Sync/Lab_work/NGS/Benchmarking/data/2015-11-14/all.sum.csv")
 pat<-"([0-9]+)_([0-9]+)"
@@
 )-> sum.df
-mutate(sum.df,exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->sum.df
 
 sum.df<-subset(sum.df,Id!="PC") # remove the plasmid control that was run
+mutate(sum.df,Id=paste(4,exp.freq,sep="_"),exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->sum.df
 
@@
 
-dups<-mutate(dups,p.val=pmax(p.val1,p.val2),Id=gsub(pat,'\\1',Id),gc=4)
+dups<-mutate(dups,p.val=pmax(p.val1,p.val2))
+
+dups<-subset(dups,p.val!=1)
 
@@
 
-dups.table<-hiseq.roc.table(dups,0.001,possible_tp,possible_vars)
+dups.table<-hiseq.roc.table(dups,0.01,possible_tp,possible_vars)
 knitr::kable(subset(dups.table))
@@
 ```
+
+These are variants that pass mapq,phred, and read cut and p<0.01 in both samples.10^4
+
+This is so much better than before. Is there really that much variability between runs?
+
+> tables are right but does the ROC make sense?
+
+### Ordinations
+```{r}
+# No wt variants for this work. I'll start with the out of the box data p<0.01
+
+sig.df<-subset(sig.df,category!="wt" & freq.var>0.01 & exp.freq>0.005) # no need to look at the samples were we expect things to be less than 1%
+
+# at this point I'm just taking the counts on the forward direction and assuming that everyone has the same coverage this will be adjusted later to take a subsample
+
+dcast(sig.df,Id~mutation,value.var ="freq.var",fill = 0)->counts
+
+sample.names<-counts$Id
+row.names(counts)<-sample.names
+counts<-(subset(counts,select=-c(Id)))
+vare.dist<-vegdist(counts);vare.dist
+
+NMDS<-metaMDS(vare.dist,trymax = 100)
+stressplot(NMDS)
+
+plot(NMDS)
+orditorp(NMDS,display="sites",cex=1.25,air=0.01)
+
+cut.read.sig.df<-subset(cut.read.sig.df,category!="wt" & freq.var>0.01 & exp.freq>0.005) # no need to look at the samples were we expect things to be less than 1%
+
+# at this point I'm just taking the counts on the forward direction and assuming that everyone has the same coverage this will be adjusted later to take a subsample
+
+dcast(cut.read.sig.df,Id~mutation,value.var ="freq.var",fill = 0)->counts
+
+sample.names<-counts$Id
+row.names(counts)<-sample.names
+counts<-(subset(counts,select=-c(Id)))
+vare.dist<-vegdist(counts);vare.dist
+
+NMDS<-metaMDS(vare.dist,trymax = 100)
+stressplot(NMDS)
+
+plot(NMDS)
+orditorp(NMDS,display="sites",cex=1.25,air=0.01)
+
+```

### 2016-01-04 scripts/figures.Rmd
still need table work

@@
 }
+
 roc_df.one<-function(roc_analysis,thr){ # get the coordinates and cut offs for all the points in the ROC object
@@
 
-miseq.roc<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.001))
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.001))
 
-miseq.roc
+miseq.roc.p
 
@@
 
+ggplot(subset(sum.df,p.val<0.01),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)+geom_hline(yintercept=c(63,188))
+
+ggplot(sum.df,aes(x=Read_pos,fill=category))+geom_histogram(position="dodge")
+
+```
+
+### With read position cut off 63 and 188
+
+```{r}
+read.df<-subset(sum.df,Read_pos>63 & Read_pos<188)
+m.roc.df<-miseq.roc(read.df,possible_tp,possible_vars)
+
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.001))
+
+miseq.roc.p
+
+# pdf("../results/figures/miseq_roc.one.sided.pdf")
+# miseq.roc
+# dev.off()
+
+m.roc.table<-miseq.roc.table(read.df,0.01,possible_tp,possible_vars)
+knitr::kable(m.roc.table)
+
+ggplot(subset(sum.df,p.val<0.01),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=0.5)
+
 rm(list = setdiff(ls(), lsf.str())) # remove all variables except the functions
+
 ```
@@
 ```
+
+### 10^3
 ```{r,roc_lower_inputs3}
@@
 ### Ordinations
-```{r}
+```{r,eval=F}
 # No wt variants for this work. I'll start with the out of the box data p<0.01

### 2016-01-05 scripts/figures.Rmd
standard roc axis

@@
 
+# The ROC data only includes points there the sensitivity changes.
+At this point I have limitted our analysis to variants with p<0.05.
+So the lines in some plots stop short.
+For the final draft I will not apply this cut off.
+Even in that future case this step may be need as the ROC data only includes points where the sensitivity changes.
+For now   This function extents the lines by adding a data point at the specificity of 0.002 and appropriate sensitivity.
+At the most it hurts us as we do not increase sensitivity and it might , but I doubt it.
+Anyway it is of no concern as we only care about the p=0.01 case which is well contained before this extension is needed
+fill_in_plots<-function(x){
+  spec<-min(x$specificity[which(x$specificity>0.998)])
+  sense<-max(x$sensitivity[x$specificity==spec]) # incase there is a step right here
+  y<-data.frame(threshold=NA,Id=NA,specificity=0.9981,sensitivity=sense,samp=unique(x$samp))
+}
+
 ######### Hiseq roc functions ##########
@@
   roc.df.adj<-roc.df.adj[order(roc.df.adj$adj.sensitivity),]
-
+######START HERE WITH THE FILL IN  AND DO IT FOR THE MISEQ ABOVE##############
   mutate(roc.df.adj, gc = as.numeric(sub(pat,"\\1",samp)),
@@
 
-miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.001))
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
 
@@
 
-miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.001))
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.002)
 
@@
 
-Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1));Hiseq.roc
+Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002));Hiseq.roc
 
@@
 
-ggplot(subset(h.cut.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+ggplot(subset(h.cut.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
 
@@
 
-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
 
@@
 
-ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
 
@@
 ```{r,roc_lower_inputs4}
-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
 
@@
 ```{r,roc_lower_inputs3}
-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
 
@@
 
-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
+ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
 

### 2016-01-05 scripts/figures.Rmd
fill in plots

@@
 ```{r,roc_helper_runctions}
+find_freq<-function(Id){ # helper function to get the expected frequency from the sample name
+  if(Id=="Covaris_5"){
+    x=0.05} else if ( Id=="Covaris_25"){
+    x=0.025} else if ( Id=="Covaris_125"){
+    x=0.0125} else if ( Id=="Covaris_063"){
+    x=0.0063} else if ( Id=="Covaris_016"){
+    x=0.0016}
+}
+
+find_freq.v<-Vectorize(find_freq) # vectorized version of the helper function above
+
 id<-function(x){
@@
   x <- strsplit(x, ".", fixed = T)[[1]][2]
+}
+
+fill_in_plots<-function(x){
+  spec<-min(x$adj.specificity[which(x$adj.specificity>0.998)])
+  sense<-max(x$adj.sensitivity[x$adj.specificity==spec]) # incase there is a step right here
+
+  y<-data.frame(threshold=x$threshold[1],Id=x$Id[1],adj.specificity=0.9980001,adj.sensitivity=sense,samp=unique(x$samp),exp.freq=x$exp.freq[1],FP=x$FP[1],TP=x$TP[1],sensitivity=x$sensitivity[1],specificity=x$specificity[1])
+
+  if("gc" %in% names(x)){
+    y$gc=x$gc[1]
+  }
+  return(y)
 }
@@
   roc.df.adj<-ddply(roc.df,~samp,adjust.coords,sum.df,possible_tp,possible_vars)
+
+  roc.df.adj<-mutate(roc.df.adj,exp.freq=as.factor(find_freq.v(samp)))
+  fills<-ddply(roc.df.adj,~samp,fill_in_plots) # take out these lines to not fill in the plots
+  roc.df.adj<-rbind(fills,roc.df.adj) # here too
   roc.df.adj<-roc.df.adj[order(roc.df.adj$adj.sensitivity),]
-  roc.df.adj<-mutate(roc.df.adj,exp.freq=as.factor(find_freq.v(samp)))
   roc.df.adj$exp.freq <- factor(roc.df.adj$exp.freq, levels = rev(levels(roc.df.adj$exp.freq)))
@@
 Anyway it is of no concern as we only care about the p=0.01 case which is well contained before this extension is needed
-fill_in_plots<-function(x){
-  spec<-min(x$specificity[which(x$specificity>0.998)])
-  sense<-max(x$sensitivity[x$specificity==spec]) # incase there is a step right here
-  y<-data.frame(threshold=NA,Id=NA,specificity=0.9981,sensitivity=sense,samp=unique(x$samp))
-}
 
@@
   roc.df.adj<-roc.df.adj[order(roc.df.adj$adj.sensitivity),]
-######START HERE WITH THE FILL IN  AND DO IT FOR THE MISEQ ABOVE##############
   mutate(roc.df.adj, gc = as.numeric(sub(pat,"\\1",samp)),
@@
   )-> roc.df.adj
+  fills<-ddply(roc.df.adj,~samp,fill_in_plots) # take out these lines to not fill in the plots
+  roc.df.adj<-rbind(fills,roc.df.adj) # here too
+  roc.df.adj$exp.freq<-as.character(roc.df.adj$exp.freq) # so that teh calculations below work since it is a factor here
   mutate(roc.df.adj,exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->roc.df.adj
@@
 )-> roc.df.adj
+
 mutate(roc.df.adj,exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->roc.df.adj
@@
 ```{r,Miseq_out_of_the_box}
-find_freq<-function(Id){ # helper function to get the expected frequency from the sample name
-  if(Id=="Covaris_5"){
-    x=0.05} else if ( Id=="Covaris_25"){
-    x=0.025} else if ( Id=="Covaris_125"){
-    x=0.0125} else if ( Id=="Covaris_063"){
-    x=0.0063} else if ( Id=="Covaris_016"){
-    x=0.0016}
-}
-find_freq.v<-Vectorize(find_freq) # vectorized version of the helper function above
 

### 2016-01-05 scripts/figures.Rmd
touched up ROC fixed roc function

@@
 plot.ex(0.35,0.95,roc.df,example)
-
-pdf("../results/figures/roc.example.pdf")
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.ex.pdf")
 plot.ex(0.35,0.95,roc.df,example)
@@
     }  
-    roc(sum.df$category~sum.df$p.val,plot=F,CI=T)
+    roc(sum.df$category~sum.df$p.val,plot=F,CI=T,direction=">")
   }
@@
 
-roc_df.one<-function(roc_analysis,thr){ # get the coordinates and cut offs for all the points in the ROC object
-  roc_analysis<-roc_analysis[unlist(lapply(roc_analysis,is.null))==F]
+roc_df.one<-function(roc_anal,thr){ # get the coordinates and cut offs for all the points in the ROC object
+  roc_analysis<-roc_anal[unlist(lapply(roc_anal,is.null))==F]
   all<-lapply(roc_analysis,coords,x=thr,input="thr")
@@
 
-# pdf("../results/figures/miseq_roc.one.sided.pdf")
-# miseq.roc
-# dev.off()
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/miseq_roc.one.sided.pdf")
+ miseq.roc
+ dev.off()
 
@@
 
-ggplot(subset(sum.df,p.val<0.01),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=0.5)
+ggplot(subset(sum.df,p.val<0.01),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)
 
@@
 
+sum.df<-subset(sum.df,exp.freq>0.001) # remove this to include exp.freq 0.01%
 # Id the variants as T or F 
@@
 
-Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002));Hiseq.roc
+Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002));Hiseq.roc
+
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/hiseq.roc.pdf")
+Hiseq.roc
+dev.off()
+
+Hiseq.roc.4<-ggplot(subset(h.roc.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002));Hiseq.roc
+
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/hiseq.roc.4.pdf")
+Hiseq.roc.4
+dev.off()
 
@@
 knitr::kable(subset(h.table.df,gc==5))
+
+knitr::kable(subset(h.table.df,gc==4))
 ```
@@
 
-ggplot(subset(sig.df,category %in% c(T,F)),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=0.5)+geom_hline(yintercept=30)+geom_vline(xintercept=35)+facet_wrap(~gc)
+mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)+geom_hline(yintercept=20)+geom_vline(xintercept=30)+scale_x_continuous(limits=c(30,40))+scale_color_hue(name="",labels=c("False posivite","True positive"))
+
+mapq_phred.normal
+
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/mapq_phred.normal.pdf")
+mapq_phred.normal
+dev.off()
+
+mapq_phred.stingy<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)+scale_x_continuous(limits=c(30,40))+geom_hline(yintercept=30)+geom_vline(xintercept=35)+scale_color_hue(name="",labels=c("False positive","True positive"))
+
+mapq_phred.stingy
+
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/mapq_phred.stingy.pdf")
+mapq_phred.stingy
+dev.off()
 
@@
 
-ggplot(subset(h.cut.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+mapq_phred.roc<-ggplot(subset(h.cut.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+
+mapq_phred.roc
+
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/mapq_phred.rocpdf")
+mapq_phred.roc
+dev.off()
 
@@
 
-ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')
+read.pos<-ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')+scale_fill_hue(name="",labels=c("False positive","True positive"))+xlab("Average position on read")+ylab("Number of variants")
+read.pos
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/read.pos.pdf")
+read.pos
+dev.off()
 
@@
 
-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+
+read.roc
+
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/read.roc.pdf")
+read.roc
+dev.off()
 
@@
 ac_freq
+
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/freq.pdf")
+ac_freq
+dev.off()
+
 ```
@@
 ```{r,roc_lower_inputs4}
-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+
+roc.4
+
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.4.pdf")
+roc.4
+dev.off()
 
@@
 ```{r,roc_lower_inputs3}
-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.3.pdf")
+roc.3
+dev.off()
 
@@
 
-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+
+dups.roc.p
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/dups.roc.pdf")
+dups.roc.p
+dev.off()
 
@@
 
-This is so much better than before. Is there really that much variability between runs?
-
-> tables are right but does the ROC make sense?
-
-### Ordinations
-```{r,eval=F}
-# No wt variants for this work. I'll start with the out of the box data p<0.01
-
-sig.df<-subset(sig.df,category!="wt" & freq.var>0.01 & exp.freq>0.005) # no need to look at the samples were we expect things to be less than 1%
-
-# at this point I'm just taking the counts on the forward direction and assuming that everyone has the same coverage this will be adjusted later to take a subsample
-
-dcast(sig.df,Id~mutation,value.var ="freq.var",fill = 0)->counts
-
-sample.names<-counts$Id
-row.names(counts)<-sample.names
-counts<-(subset(counts,select=-c(Id)))
-vare.dist<-vegdist(counts);vare.dist
-
-NMDS<-metaMDS(vare.dist,trymax = 100)
-stressplot(NMDS)
-
-plot(NMDS)
-orditorp(NMDS,display="sites",cex=1.25,air=0.01)
-
-cut.read.sig.df<-subset(cut.read.sig.df,category!="wt" & freq.var>0.01 & exp.freq>0.005) # no need to look at the samples were we expect things to be less than 1%
-
-# at this point I'm just taking the counts on the forward direction and assuming that everyone has the same coverage this will be adjusted later to take a subsample
-
-dcast(cut.read.sig.df,Id~mutation,value.var ="freq.var",fill = 0)->counts
-
-sample.names<-counts$Id
-row.names(counts)<-sample.names
-counts<-(subset(counts,select=-c(Id)))
-vare.dist<-vegdist(counts);vare.dist
-
-NMDS<-metaMDS(vare.dist,trymax = 100)
-stressplot(NMDS)
-
-plot(NMDS)
-orditorp(NMDS,display="sites",cex=1.25,air=0.01)
-
-```
+This is so much better than before. 
+### Concequences for diversity measurements

### 2016-01-06 scripts/figures.Rmd
tweek some plots

@@
 fill_in_plots<-function(x){
-  spec<-min(x$adj.specificity[which(x$adj.specificity>0.998)])
+  spec<-min(x$adj.specificity[which(x$adj.specificity>0.995)])
   sense<-max(x$adj.sensitivity[x$adj.specificity==spec]) # incase there is a step right here
 
-  y<-data.frame(threshold=x$threshold[1],Id=x$Id[1],adj.specificity=0.9980001,adj.sensitivity=sense,samp=unique(x$samp),exp.freq=x$exp.freq[1],FP=x$FP[1],TP=x$TP[1],sensitivity=x$sensitivity[1],specificity=x$specificity[1])
+  y<-data.frame(threshold=x$threshold[1],Id=x$Id[1],adj.specificity=0.9950001,adj.sensitivity=sense,samp=unique(x$samp),exp.freq=x$exp.freq[1],FP=x$FP[1],TP=x$TP[1],sensitivity=x$sensitivity[1],specificity=x$specificity[1])
 
@@
 
-miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
 
@@
 
-miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.002)
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.005)
 
@@
 
-Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002));Hiseq.roc
+Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005));Hiseq.roc
 
@@
 
-Hiseq.roc.4<-ggplot(subset(h.roc.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002));Hiseq.roc
+Hiseq.roc.4<-ggplot(subset(h.roc.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005));Hiseq.roc
 
@@
 
-mapq_phred.roc<-ggplot(subset(h.cut.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+mapq_phred.roc<-ggplot(subset(h.cut.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
 
@@
 
-read.pos<-ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')+scale_fill_hue(name="",labels=c("False positive","True positive"))+xlab("Average position on read")+ylab("Number of variants")
+read.pos<-ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')+scale_fill_hue(name="",labels=c("False positive","True positive"))+xlab("Average position on read")+ylab("Number of variants")+theme(legend.position=c(.5,.7))
 read.pos
@@
 
-read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
 
@@
 
-ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
 
@@
 ```{r,roc_lower_inputs4}
-roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
 
@@
 ```{r,roc_lower_inputs3}
-roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
 
@@
 
-dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.002))
+dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
 

### 2016-01-18 scripts/figures.Rmd

@@
   #plot the roc
-  roc.p<-ggplot(roc.df,aes(x=1-specificity,y=sensitivity))+geom_step(color="black",size=2)+xlab("1-Specificity")+ylab("Sensitivity")+theme(axis.title.y=element_text(vjust=1))
+  roc.p<-ggplot(roc.df,aes(x=1-specificity,y=sensitivity))+geom_step(color="black",size=2)+xlab("False Positive Rate")+ylab("Sensitivity")+theme(axis.title.y=element_text(vjust=1))
   roc.p<-roc.p+geom_line(data=lines,aes(x=spec,y=sense,color=as.factor(group)),size=c(1.5),lty=c(2))+scale_color_brewer(palette="Set1")
@@
 plot.ex(0.35,0.95,roc.df,example)
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.ex.pdf")
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.ex.pdf",width = 12,height = 8)
 plot.ex(0.35,0.95,roc.df,example)
@@
 
-miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab(bquote("False Positive Rate (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
 
@@
 
-miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.005)
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("False Positive Rate")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.005)
 
@@
 
-Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005));Hiseq.roc
+Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("False Positive Rate (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5));Hiseq.roc
 
@@
 
-Hiseq.roc.4<-ggplot(subset(h.roc.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005));Hiseq.roc
+Hiseq.roc.4<-ggplot(subset(h.roc.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("False Positive Rate")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005));Hiseq.roc.4
 
@@
 
-mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)+geom_hline(yintercept=20)+geom_vline(xintercept=30)+scale_x_continuous(limits=c(30,40))+scale_color_hue(name="",labels=c("False posivite","True positive"))
+mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)+geom_hline(yintercept=20)+geom_vline(xintercept=30)+scale_x_continuous(limits=c(30,40))+scale_color_hue(name="",labels=c("False posivite","True positive"))+ylab("Mapping Quality")+xlab("Base Qualtiy (Phred) ")
 
@@
 
-mapq_phred.stingy<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)+scale_x_continuous(limits=c(30,40))+geom_hline(yintercept=30)+geom_vline(xintercept=35)+scale_color_hue(name="",labels=c("False positive","True positive"))
+mapq_phred.stingy<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)+scale_x_continuous(limits=c(30,40))+geom_hline(yintercept=30)+geom_vline(xintercept=35)+scale_color_hue(name="",labels=c("False positive","True positive"))+ylab("Mapping Quality")+xlab("Base Qualtiy (Phred) ")
 
@@
 
-mapq_phred.roc<-ggplot(subset(h.cut.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+mapq_phred.roc<-ggplot(subset(h.cut.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("False Positive Rate (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
 
@@
 
-read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("False Positive Rate (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
 
@@
 
-ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("False Positive Rate (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
 
@@
 ```{r,roc_lower_inputs4}
-roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("False Positive Rate (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
 
@@
 ```{r,roc_lower_inputs3}
-roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("False Positive Rate (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
 
@@
 
-dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("False Positive Rate (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
 
@@
 
+exp.meas.freq<-ddply(cut.df,~exp.freq+mutation+category, function(x){
+                                                                                      freq1=x$freq.var[x$dup=="1"]
+                                                                                      freq2=x$freq.var[x$dup=="2"]
+                                                                                      if(length(freq1)==0) freq1=0
+                                                                                      if(length(freq2)==0) freq2=0
+                                                                                      out=data.frame(freq1=freq1,freq2=freq2)})
+
+ggplot(subset(exp.meas.freq),aes(x=freq1,y=freq2,color=category))+geom_point(aes(size=3))+scale_y_log10()+scale_x_log10()+xlab("Frequency in sample 1")+ylab("Frequency in sample 2")+scale_color_hue(name="",labels=c("False positive","True positive"))+theme(legend.position=c(0.5,0.95))+geom_abline(intercept=0,slope=1,lty=2)->dup.freq
+dup.freq
+
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/dup.freq.pdf")
+dup.freq
+dev.off()
+
 ```

### 2016-01-19 scripts/figures.Rmd
starting to put boxes together

--- 2016-01-06 scripts/figures.Rmd
@@
 require(gridExtra)
-theme_set(new = theme_classic(base_size=24))
+theme_set(new = theme_classic())
 require(knitr)
@@
  #plot the points and cut offs
- points.p<-ggplot(distribution,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center")+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks = element_blank(), axis.text.x = element_blank(),axis.title.y = element_text(vjust=1.2))
+ points.p<-ggplot(distribution,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center")+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks = element_blank(), axis.text.x = element_blank(),axis.title.y = element_text(vjust=1.2))+xlab("")
 
@@
 
- grid.arrange(points.p,roc.p,ncol=2)
+ plot_grid(points.p, roc.p, labels=c("A", "B"), ncol = 2, nrow = 1)
 
@@
 plot.ex(0.35,0.95,roc.df,example)
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.ex.pdf")
-plot.ex(0.35,0.95,roc.df,example)
-dev.off()
+# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.ex.pdf",width = 12,height = 8)
+# plot.ex(0.35,0.95,roc.df,example)
+# dev.off()
 
@@
 
-miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
 
@@
 
-Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005));Hiseq.roc
+Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5));Hiseq.roc
 
@@
 
-Hiseq.roc.4<-ggplot(subset(h.roc.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005));Hiseq.roc
+Hiseq.roc.4<-ggplot(subset(h.roc.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005));Hiseq.roc.4
 
@@
 
-mapq_phred.roc<-ggplot(subset(h.cut.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+mapq_phred.roc<-ggplot(subset(h.cut.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
 
@@
 
-read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
 
@@
 ```{r,roc_lower_inputs4}
-roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
 
@@
 ```{r,roc_lower_inputs3}
-roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
 
@@
 
-dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
 

### 2016-01-19 scripts/figures.Rmd
new color scheme

@@
 require(gridExtra)
+require(cowplot)
+require(gtable)
+require(png)
 theme_set(new = theme_classic())
@@
 opts_chunk$set(fig.align = "center", warning = FALSE, echo = FALSE, fig.width = 10, 
-    fig.height = 7,message=F)
+    fig.height = 7,message=F,cache=T)
+
+rhg_cols<-c("#08519cff","#3182b4ff","#6baed6ff","#b1bdc5ff","#787878ff")
+
 ```
@@
 ```
+
 ### Experimental setup diagram
@@
   roc.df.adj<-mutate(roc.df.adj,exp.freq=as.factor(find_freq.v(samp)))
-  fills<-ddply(roc.df.adj,~samp,fill_in_plots) # take out these lines to not fill in the plots
-  roc.df.adj<-rbind(fills,roc.df.adj) # here too
+  #fills<-ddply(roc.df.adj,~samp,fill_in_plots) # take out these lines to not fill in the plots
+  #roc.df.adj<-rbind(fills,roc.df.adj) # here too
   roc.df.adj<-roc.df.adj[order(roc.df.adj$adj.sensitivity),]
@@
   )-> roc.df.adj
-  fills<-ddply(roc.df.adj,~samp,fill_in_plots) # take out these lines to not fill in the plots
-  roc.df.adj<-rbind(fills,roc.df.adj) # here too
+  #fills<-ddply(roc.df.adj,~samp,fill_in_plots) # take out these lines to not fill in the plots
+  #roc.df.adj<-rbind(fills,roc.df.adj) # here too
   roc.df.adj$exp.freq<-as.character(roc.df.adj$exp.freq) # so that teh calculations below work since it is a factor here
@@
 
+## Figure 2
 ```{r,Miseq_out_of_the_box}
@@
 
-miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position=c(0.8,.5))
 
-miseq.roc.p
+#miseq.roc.p
 
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/miseq_roc.one.sided.pdf")
- miseq.roc
- dev.off()
+# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/miseq_roc.one.sided.pdf")
+#  miseq.roc.p
+# dev.off()
 
 m.roc.table<-miseq.roc.table(sum.df,0.01,possible_tp,possible_vars)
-knitr::kable(m.roc.table)
 
-ggplot(subset(sum.df,p.val<0.01),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)+geom_hline(yintercept=c(63,188))
+#knitr::kable(m.roc.table)
+m.roc.table<-rename(m.roc.table,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))
 
-ggplot(sum.df,aes(x=Read_pos,fill=category))+geom_histogram(position="dodge")
+m.roc.table$Frequency<-c("5.0%","2.5%","1.25%","0.63%","0.016%")
+m.roc.table$Sensitivity<-round(m.roc.table$Sensitivity,digits=3)
+m.roc.table$Specificity<-round(m.roc.table$Specificity,digits=3)
+
+tbl<-tableGrob(m.roc.table,rows=NULL,theme=ttheme_minimal()) # Make table
+
+readPNG("../results/pictures/miseq_setup.png")->img # read in set up png
+g <- rasterGrob(img, interpolate=TRUE)
+qplot(1:1, 1:1, geom="blank") +
+annotation_custom(g, xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
+geom_blank()+xlab("")+ylab("")+
+theme(line = element_blank(),
+text = element_blank(),
+line = element_blank(),
+title = element_blank())->miseq.setup # empty plot with setup png a background
+
+#plot_grid(miseq.setup,miseq.roc.p, tbl, labels=c("A", "B","C"), ncol = 3, nrow = 1)
+# Figure 2
+ggdraw() +
+  draw_plot(miseq.setup, 0, .5, .5, .5) +
+  draw_plot(miseq.roc.p, .5, .5, .5, .5) +
+  draw_plot(tbl, 0, 0, 1, .5) +
+  draw_plot_label(c("A", "B", "C"), c(0, 0.5, 0), c(1, 1, 0.5), size = 15)
+
+#ggplot(subset(sum.df,p.val<0.01),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)+geom_hline(yintercept=c(63,188))
+
+#ggplot(sum.df,aes(x=Read_pos,fill=category))+geom_histogram(position="dodge")
 
@@
 
+```{r,Miseq_read.cut,eval=F}
 ### With read position cut off 63 and 188
-
-```{r}
 read.df<-subset(sum.df,Read_pos>63 & Read_pos<188)
@@
 
-miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.005)
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.005)
 
@@
 
-## Figure 2
-
 ### Hiseq expermintal setup
@@
 
-### ROC
+### Figure 3
 Out of the box here is how we do for the $10^5$ sample.
@@
 
-Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5));Hiseq.roc
+Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
 
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/hiseq.roc.pdf")
-Hiseq.roc
-dev.off()
-
-Hiseq.roc.4<-ggplot(subset(h.roc.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005));Hiseq.roc.4
-
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/hiseq.roc.4.pdf")
-Hiseq.roc.4
-dev.off()
+# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/hiseq.roc.pdf")
+# Hiseq.roc
+# dev.off()
 
@@
 
-knitr::kable(subset(h.table.df,gc==5))
+#knitr::kable(subset(h.table.df,gc==5))
 
-knitr::kable(subset(h.table.df,gc==4))
+h.table.df<-rename(h.table.df,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))
+
+h.table.df<-subset(h.table.df,gc==5,select=-c(gc))
+h.table.df$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
+h.table.df$Sensitivity<-round(h.table.df$Sensitivity,digits=3)
+h.table.df$Specificity<-round(h.table.df$Specificity,digits=3)
+
+tbl<-tableGrob(h.table.df,rows=NULL,theme=ttheme_minimal()) # Make table
+
+readPNG("../results/pictures/hiseq_setup.png")->img # read in set up png
+g <- rasterGrob(img, interpolate=TRUE)
+qplot(1:1, 1:1, geom="blank") +
+annotation_custom(g, xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
+geom_blank()+xlab("")+ylab("")+
+theme(line = element_blank(),
+text = element_blank(),
+line = element_blank(),
+title = element_blank())->hiseq.setup
+
+ggdraw() +
+  draw_plot(hiseq.setup, 0, .5, .5, .5) +
+  draw_plot(Hiseq.roc, .5, .5, .5, .5) +
+  draw_plot(tbl, 0, 0, 1, .5) +
+  draw_plot_label(c("A", "B", "C"), c(0, 0.5, 0), c(1, 1, 0.5), size = 15)
+
+# 
+# knitr::kable(subset(h.table.df,gc==4))
 ```
@@
 
-mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)+geom_hline(yintercept=20)+geom_vline(xintercept=30)+scale_x_continuous(limits=c(30,40))+scale_color_hue(name="",labels=c("False posivite","True positive"))+ylab("Mapping Quality")+xlab("Base Qualtiy (Phred) ")
+mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)+geom_hline(yintercept=20,linetype=2)+geom_vline(xintercept=30,linetype=2)+scale_x_continuous(limits=c(30,40))+scale_color_hue(name="",labels=c("False posivite","True positive"))+ylab("Mean mapping quality")+xlab("Mean base qualtiy (Phred) ")
 
-mapq_phred.normal
+#mapq_phred.normal
 
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/mapq_phred.normal.pdf")
-mapq_phred.normal
-dev.off()
+# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/mapq_phred.normal.pdf")
+# mapq_phred.normal
+# dev.off()
 
-mapq_phred.stingy<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)+scale_x_continuous(limits=c(30,40))+geom_hline(yintercept=30)+geom_vline(xintercept=35)+scale_color_hue(name="",labels=c("False positive","True positive"))+ylab("Mapping Quality")+xlab("Base Qualtiy (Phred) ")
-
-mapq_phred.stingy
-
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/mapq_phred.stingy.pdf")
-mapq_phred.stingy
-dev.off()
+# mapq_phred.stingy<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)+scale_x_continuous(limits=c(30,40))+geom_hline(yintercept=30)+geom_vline(xintercept=35)+scale_color_hue(name="",labels=c("False positive","True positive"))+ylab("Mapping Quality")+xlab("Base Qualtiy (Phred) ")
+# 
+# mapq_phred.stingy
+# 
+# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/mapq_phred.stingy.pdf")
+# mapq_phred.stingy
+# dev.off()
 
@@
 
-h.cut.df<-hiseq.roc(cut.df,possible_tp,possible_vars)
-
-mapq_phred.roc<-ggplot(subset(h.cut.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
-
-mapq_phred.roc
-
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/mapq_phred.rocpdf")
-mapq_phred.roc
-dev.off()
-
-h.cut.table<-hiseq.roc.table(cut.df,0.01,possible_tp,possible_vars)
-
-knitr::kable(subset(h.cut.table,gc==5))
+# h.cut.df<-hiseq.roc(cut.df,possible_tp,possible_vars)
+# 
+# mapq_phred.roc<-ggplot(subset(h.cut.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
+# 
+# mapq_phred.roc
+# 
+# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/mapq_phred.rocpdf")
+# mapq_phred.roc
+# dev.off()
+# 
+# h.cut.table<-hiseq.roc.table(cut.df,0.01,possible_tp,possible_vars)
+# 
+# knitr::kable(subset(h.cut.table,gc==5))
 ```
@@
 
-ggplot(sig.cut.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
+#ggplot(sig.cut.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
 # ggplot(subset(sig.cut.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
@@
 read.pos<-ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')+scale_fill_hue(name="",labels=c("False positive","True positive"))+xlab("Average position on read")+ylab("Number of variants")+theme(legend.position=c(.5,.7))
-read.pos
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/read.pos.pdf")
-read.pos
-dev.off()
+# read.pos
+# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/read.pos.pdf")
+# read.pos
+# dev.off()
 
@@
 
-read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
+read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
 
-read.roc
-
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/read.roc.pdf")
-read.roc
-dev.off()
+# read.roc
+# 
+# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/read.roc.pdf")
+# read.roc
+# dev.off()
 
 h.cut.read.table<-hiseq.roc.table(cut.read.df,0.01,possible_tp,possible_vars)
-knitr::kable(subset(h.cut.read.table,gc==5))
+#knitr::kable(subset(h.cut.read.table,gc==5))
+
+h.cut.read.table.5<-subset(h.cut.read.table,gc==5,select=-c(gc))
+rename(h.cut.read.table.5,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))->h.cut.read.table.5
+h.cut.read.table.5$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
+h.cut.read.table.5$Sensitivity<-round(h.cut.read.table.5$Sensitivity,digits=3)
+h.cut.read.table.5$Specificity<-round(h.cut.read.table.5$Specificity,digits=3)
+
+tbl<-tableGrob(h.cut.read.table.5,rows=NULL,theme=ttheme_minimal()) # Make table
+
+ggdraw() +
+  draw_plot(mapq_phred.normal, 0, .5, .5, .5) +
+  draw_plot(read.pos, .5, .5, .5, .5) +
+  draw_plot(read.roc, 0, 0, 0.5, .5) +
+  draw_plot(tbl, 0.5, 0, 0.5, .5)+
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(1, 1, 0.5,0.5), size = 15)
+
 ```
@@
 
-ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
+ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))+theme(legend.position="none")
 
@@
 
-ac_freq<-ac_freq+geom_dotplot(binaxis = "y", stackdir = "center",dotsize=1)+scale_fill_brewer(palette="Set1")
+ac_freq<-ac_freq+geom_dotplot(binaxis = "y", stackdir = "center",dotsize=1)+scale_color_manual(values=rhg_cols)
 
@@
 ## Figure 5 Lower inputs
-### 10^4
+
 ```{r,roc_lower_inputs4}
-roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
+roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
 
-roc.4
+# roc.4
+# 
+# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.4.pdf")
+# roc.4
+# dev.off()
+# 
+# knitr::kable(subset(h.cut.read.table,gc==4))
 
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.4.pdf")
-roc.4
-dev.off()
+roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))+theme(legend.position="none")
 
-knitr::kable(subset(h.cut.read.table,gc==4))
+# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.3.pdf")
+# roc.3
+# dev.off()
+# 
+# 
+# knitr::kable(subset(h.cut.read.table,gc==3))
+
+h.cut.read.table.4<-subset(h.cut.read.table,gc==4,select=-c(gc))
+rename(h.cut.read.table.4,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))->h.cut.read.table.4
+h.cut.read.table.4$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
+h.cut.read.table.4$Sensitivity<-round(h.cut.read.table.4$Sensitivity,digits=3)
+h.cut.read.table.4$Specificity<-round(h.cut.read.table.4$Specificity,digits=3)
+
+tbl.4<-tableGrob(h.cut.read.table.4,rows=NULL,theme=ttheme_minimal()) # Make table
+
+h.cut.read.table.3<-subset(h.cut.read.table,gc==3,select=-c(gc))
+rename(h.cut.read.table.3,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))->h.cut.read.table.3
+h.cut.read.table.3$Frequency<-c("5.0%","2.0%","1.0%")
+h.cut.read.table.3$Sensitivity<-round(h.cut.read.table.3$Sensitivity,digits=3)
+h.cut.read.table.3$Specificity<-round(h.cut.read.table.3$Specificity,digits=3)
+
+tbl.3<-tableGrob(h.cut.read.table.3,rows=NULL,theme=ttheme_minimal()) # Make table
+
+ggdraw() +
+  draw_plot(roc.4, 0, .5, .5, .5) +
+  draw_plot(tbl.4, .5, .5, .5, .5) +
+  draw_plot(roc.3, 0, 0, 0.5, .5) +
+  draw_plot(tbl.3, 0.5, 0, 0.5, .5)+
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(1, 1, 0.5,0.5), size = 15)
+
 ```
 
-### 10^3
-```{r,roc_lower_inputs3}
-roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))
-
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.3.pdf")
-roc.3
-dev.off()
-
-knitr::kable(subset(h.cut.read.table,gc==3))
-```
-
-### Duplicates
+### Figure 6
 
@@
 
-dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1",name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))
+dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
 
-dups.roc.p
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/dups.roc.pdf")
-dups.roc.p
-dev.off()
+# dups.roc.p
+# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/dups.roc.pdf")
+# dups.roc.p
+# dev.off()
 
 dups.table<-hiseq.roc.table(dups,0.01,possible_tp,possible_vars)
-knitr::kable(subset(dups.table))
+#knitr::kable(subset(dups.table))
+
+rename(dups.table,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))->dups.table
+subset(dups.table,select=-c(gc))->dups.table
+dups.table$Frequency<-c("5.0%","2.0%","1.0%","0.5%")
+dups.table$Sensitivity<-round(dups.table$Sensitivity,digits=3)
+dups.table$Specificity<-round(dups.table$Specificity,digits=3)
+
+tbl<-tableGrob(dups.table,rows=NULL,theme=ttheme_minimal()) # Make table
+
+ggdraw() +
+  draw_plot(dups.roc.p, 0, .5, 0.5, .5) +
+  draw_plot(tbl, 0.5, 0.5, 0.5, .5) +
+  draw_plot_label(c("A", "B"), c(0, 0.5), c(1, 1), size = 15)
 
@@
 
-ggplot(subset(exp.meas.freq),aes(x=freq1,y=freq2,color=category))+geom_point(aes(size=3))+scale_y_log10()+scale_x_log10()+xlab("Frequency in sample 1")+ylab("Frequency in sample 2")+scale_color_hue(name="",labels=c("False positive","True positive"))+theme(legend.position=c(0.5,0.95))+geom_abline(intercept=0,slope=1,lty=2)->dup.freq
-dup.freq
-
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/dup.freq.pdf")
-dup.freq
-dev.off()
+# ggplot(subset(exp.meas.freq),aes(x=freq1,y=freq2,color=category))+geom_point(aes(size=3))+scale_y_log10()+scale_x_log10()+xlab("Frequency in sample 1")+ylab("Frequency in sample 2")+scale_color_hue(name="",labels=c("False positive","True positive"))+theme(legend.position=c(0.5,0.95))+geom_abline(intercept=0,slope=1,lty=2)->dup.freq
+# dup.freq
+# 
+# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/dup.freq.pdf")
+# dup.freq
+# dev.off()
 

### 2016-01-21 scripts/figures.Rmd
some work on axis

@@
 
-ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))+theme(legend.position="none")
+ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")->lofreq.roc.p
 
@@
 
-knitr::kable(subset(lofreq.table,gc==5,select=c(exp.freq,TP,FP)))
+#knitr::kable(subset(lofreq.table,gc==5,select=c(exp.freq,TP,FP)))
+
+tbl<-tableGrob(lofreq.table,rows=NULL,theme=ttheme_minimal()) # Make table
+
+ggdraw() +
+  draw_plot(lofreq.roc.p, 0, .5, 0.5, .5) +
+  draw_plot(tbl, 0.5, 0.5, 0.5, .5) +
+  draw_plot_label(c("A", "B"), c(0, 0.5), c(1, 1), size = 15)
 
@@
 
-roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005))+theme(legend.position="none")
+roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
 

### 2016-01-21 scripts/figures.Rmd
some table formatting

@@
 
+dd=4 # digits in tables
 ```
 
+> make sure the possible fp accounts for the fact that we aren't looking at the priming regions for both Miseq and Hiseq
 ## Figure 1
@@
 m.roc.table$Frequency<-c("5.0%","2.5%","1.25%","0.63%","0.016%")
-m.roc.table$Sensitivity<-round(m.roc.table$Sensitivity,digits=3)
-m.roc.table$Specificity<-round(m.roc.table$Specificity,digits=3)
+m.roc.table$Sensitivity<-round(m.roc.table$Sensitivity,digits=dd)
+m.roc.table$Specificity<-round(m.roc.table$Specificity,digits=dd)
 
 tbl<-tableGrob(m.roc.table,rows=NULL,theme=ttheme_minimal()) # Make table
+tbl<-gtable_add_grob(tbl,
+        grobs = segmentsGrob( # line across the bottom
+            x0 = unit(0,"npc"),
+            y0 = unit(0,"npc"),
+            x1 = unit(1,"npc"),
+            y1 = unit(0,"npc"),
+            gp = gpar(lwd = 2.0)),
+        t = 1, b = 1, l = 1, r = ncol(tbl))
+
+tbl<-gtable_add_grob(tbl,
+        grobs = segmentsGrob( # line across the bottom
+            x0 = unit(0,"npc"),
+            y0 = unit(0,"npc"),
+            x1 = unit(0,"npc"),
+            y1 = unit(0,"npc"),
+            gp = gpar(lwd = 2.0)),
+        t = 1, b = 1, l = 1, r = ncol(tbl))
 
@@
 ggdraw() +
-  draw_plot(miseq.setup, 0, .5, .5, .5) +
-  draw_plot(miseq.roc.p, .5, .5, .5, .5) +
+  draw_plot(miseq.setup, 0.01, .53, .45, .42) +
+  draw_plot(miseq.roc.p, .5, .5, .45, .45) +
   draw_plot(tbl, 0, 0, 1, .5) +
@@
 h.table.df$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
-h.table.df$Sensitivity<-round(h.table.df$Sensitivity,digits=3)
-h.table.df$Specificity<-round(h.table.df$Specificity,digits=3)
+h.table.df$Sensitivity<-round(h.table.df$Sensitivity,digits=dd)
+h.table.df$Specificity<-round(h.table.df$Specificity,digits=dd)
 
@@
 
-read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
+read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.0025),breaks=c(0,0.001,0.002),labels=c(0:2))+theme(legend.position="none")
 
@@
 h.cut.read.table.5$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
-h.cut.read.table.5$Sensitivity<-round(h.cut.read.table.5$Sensitivity,digits=3)
-h.cut.read.table.5$Specificity<-round(h.cut.read.table.5$Specificity,digits=3)
+h.cut.read.table.5$Sensitivity<-round(h.cut.read.table.5$Sensitivity,digits=dd)
+h.cut.read.table.5$Specificity<-round(h.cut.read.table.5$Specificity,digits=dd)
 
@@
 ```{r,roc_lower_inputs4}
-roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
+roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.0025),breaks=c(0,0.001,0.002),labels=c(0:2))+theme(legend.position="none")
 
@@
 
-roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
+roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.0025),breaks=c(0,0.001,0.002),labels=c(0:2))+theme(legend.position="none")
 
@@
 h.cut.read.table.4$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
-h.cut.read.table.4$Sensitivity<-round(h.cut.read.table.4$Sensitivity,digits=3)
-h.cut.read.table.4$Specificity<-round(h.cut.read.table.4$Specificity,digits=3)
+h.cut.read.table.4$Sensitivity<-round(h.cut.read.table.4$Sensitivity,digits=dd)
+h.cut.read.table.4$Specificity<-round(h.cut.read.table.4$Specificity,digits=dd)
 
@@
 h.cut.read.table.3$Frequency<-c("5.0%","2.0%","1.0%")
-h.cut.read.table.3$Sensitivity<-round(h.cut.read.table.3$Sensitivity,digits=3)
-h.cut.read.table.3$Specificity<-round(h.cut.read.table.3$Specificity,digits=3)
+h.cut.read.table.3$Sensitivity<-round(h.cut.read.table.3$Sensitivity,digits=dd)
+h.cut.read.table.3$Specificity<-round(h.cut.read.table.3$Specificity,digits=dd)
 
@@
 
-dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
+dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.001),breaks=c(0,0.0005,0.001),labels=c(0,0.5,1))+theme(legend.position="none")
 
@@
 dups.table$Frequency<-c("5.0%","2.0%","1.0%","0.5%")
-dups.table$Sensitivity<-round(dups.table$Sensitivity,digits=3)
-dups.table$Specificity<-round(dups.table$Specificity,digits=3)
+dups.table$Sensitivity<-round(dups.table$Sensitivity,digits=dd)
+dups.table$Specificity<-round(dups.table$Specificity,digits=dd)
 

### 2016-01-22 scripts/figures.Rmd
first 3 figs set

@@
 require(knitr)
-opts_chunk$set(fig.align = "center", warning = FALSE, echo = FALSE, fig.width = 10, 
-    fig.height = 7,message=F,cache=T)
+opts_chunk$set(fig.align = "center", warning = FALSE, echo = FALSE,message=F,cache=F)
 
@@
 ### ROC skematic
-```{r, roc_skematic}
+```{r, roc_skematic,fig.height=5,fig.width=3.4}
 set.seed(42) # make the figures reproducible
@@
   #plot the roc
-  roc.p<-ggplot(roc.df,aes(x=1-specificity,y=sensitivity))+geom_step(color="black",size=2)+xlab("1-Specificity")+ylab("Sensitivity")+theme(axis.title.y=element_text(vjust=1))
-  roc.p<-roc.p+geom_line(data=lines,aes(x=spec,y=sense,color=as.factor(group)),size=c(1.5),lty=c(2))+scale_color_brewer(palette="Set1")
+  roc.p<-ggplot(roc.df,aes(x=1-specificity,y=sensitivity))+geom_step(color="black",size=1.5)+xlab("1-Specificity")+ylab("Sensitivity")+theme(axis.title.y=element_text(vjust=1))
+  roc.p<-roc.p+geom_line(data=lines,aes(x=spec,y=sense,color=as.factor(group)),size=c(1),lty=c(2))+scale_color_brewer(palette="Set1")
 
@@
  #plot the points and cut offs
- points.p<-ggplot(distribution,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center")+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks = element_blank(), axis.text.x = element_blank(),axis.title.y = element_text(vjust=1.2))+xlab("")
+ points.p<-ggplot(distribution,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center",dotsize = 1.5)+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks = element_blank(), axis.text.x = element_blank(),axis.title.y = element_text(vjust=1.2),legend.position=c(0.8,0.6))+xlab("")
 
  points.p<-points.p+geom_hline(yintercept=output$threshold,
-                          size=1.5,lty=2,color=c("#e41a1c","#377eb8")) # these are the first 2 colors in set 1
+                          size=1,lty=2,color=c("#e41a1c","#377eb8")) # these are the first 2 colors in set 1
 
- plot_grid(points.p, roc.p, labels=c("A", "B"), ncol = 2, nrow = 1)
+ plot_grid(points.p, roc.p, labels=c("A", "B"), ncol = 1, nrow = 2)
 
@@
 }
+
+### tableGrob functions and settings #####
+tt1<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.75,fontfamily="serif")),core=list(fg_params=list(cex=0.75,fontfamily="serif")))
+make.table<-function(df,theme){
+  tbl<-tableGrob(df,rows=NULL,theme=theme) # Make table
+  tbl$widths <- unit(rep(1/ncol(tbl), ncol(tbl)), "npc")
+  tbl<-gtable_add_grob(tbl,
+        grobs = grobTree(
+          segmentsGrob( # line across the bottom
+            x0 = unit(0,"npc"),
+            y0 = unit(0,"npc"),
+            x1 = unit(1,"npc"),
+            y1 = unit(0,"npc"),
+            gp = gpar(lwd = 2.0)),
+        segmentsGrob( # line across the top
+            x0 = unit(0,"npc"),
+            y0 = unit(1,"npc"),
+            x1 = unit(1,"npc"),
+            y1 = unit(1,"npc"),
+            gp = gpar(lwd = 2.0))),
+        t = 1, b = 1, l = 1, r = ncol(tbl))
+
+  tbl<-gtable_add_grob(tbl,
+                     grobs=segmentsGrob(
+                       x0 = unit(0,"npc"),
+            y0 = unit(0,"npc"),
+            x1 = unit(1,"npc"),
+            y1 = unit(0,"npc"),
+            gp = gpar(lwd = 2.0)),
+        t = nrow(tbl), b = nrow(tbl), l = 1, r = ncol(tbl))
+}
 ```
@@
 ## Figure 2
-```{r,Miseq_out_of_the_box}
+```{r,Miseq_out_of_the_box,fig.width=7,fig.height=5}
 
@@
 
-miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position=c(0.8,.5))
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.5%","1.25%","0.63%","0.016%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position=c(1,.5))
 
@@
 
-tbl<-tableGrob(m.roc.table,rows=NULL,theme=ttheme_minimal()) # Make table
-tbl<-gtable_add_grob(tbl,
-        grobs = segmentsGrob( # line across the bottom
-            x0 = unit(0,"npc"),
-            y0 = unit(0,"npc"),
-            x1 = unit(1,"npc"),
-            y1 = unit(0,"npc"),
-            gp = gpar(lwd = 2.0)),
-        t = 1, b = 1, l = 1, r = ncol(tbl))
-
-tbl<-gtable_add_grob(tbl,
-        grobs = segmentsGrob( # line across the bottom
-            x0 = unit(0,"npc"),
-            y0 = unit(0,"npc"),
-            x1 = unit(0,"npc"),
-            y1 = unit(0,"npc"),
-            gp = gpar(lwd = 2.0)),
-        t = 1, b = 1, l = 1, r = ncol(tbl))
+tbl<-make.table(m.roc.table,tt1)
 
@@
 # Figure 2
+
 ggdraw() +
-  draw_plot(miseq.setup, 0.01, .53, .45, .42) +
-  draw_plot(miseq.roc.p, .5, .5, .45, .45) +
-  draw_plot(tbl, 0, 0, 1, .5) +
-  draw_plot_label(c("A", "B", "C"), c(0, 0.5, 0), c(1, 1, 0.5), size = 15)
+  draw_plot(miseq.setup, 0.01, .1, 0.4, 0.8) + #x y width height
+  draw_plot(miseq.roc.p, 0.43, 0.42, 0.5, 0.51) +
+  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
+  draw_plot_label(c("A", "B", "C"), c(0.01, 0.42, 0.42), c(0.78, 0.95, 0.46), size = 15)
+
+pdf("../results/figures/miseq1.pdf",width=7,height=5)
+ggdraw() +
+  draw_plot(miseq.setup, 0.01, .1, 0.4, 0.8) + #x y width height
+  draw_plot(miseq.roc.p, 0.43, 0.42, 0.5, 0.51) +
+  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
+  draw_plot_label(c("A", "B", "C"), c(0.01, 0.42, 0.42), c(0.78, 0.95, 0.46), size = 15)
+dev.off()
 
@@
 rm(list = setdiff(ls(), lsf.str())) # remove all variables except the functions
+tt1<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.75,fontfamily="serif")),core=list(fg_params=list(cex=0.75,fontfamily="serif")))
 
 ```
-
-### Hiseq expermintal setup
 
@@
 
-```{r,hiseq_roc}
+```{r,hiseq_roc,fig.width=7,fig.height=5}
 sum.df<-read.csv("../data/process/2015-6-23/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#") # reading in the variants from the pipeline output
@@
 
-Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
+Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")+theme(plot.margin=unit(c(0,0,0,0),"mm"))
 
@@
 
-tbl<-tableGrob(h.table.df,rows=NULL,theme=ttheme_minimal()) # Make table
+tbl<-make.table(h.table.df,tt1)    # Make table
 
@@
 ggdraw() +
-  draw_plot(hiseq.setup, 0, .5, .5, .5) +
-  draw_plot(Hiseq.roc, .5, .5, .5, .5) +
-  draw_plot(tbl, 0, 0, 1, .5) +
-  draw_plot_label(c("A", "B", "C"), c(0, 0.5, 0), c(1, 1, 0.5), size = 15)
+  draw_plot(hiseq.setup, 0, .1, 0.4, 0.8) + #x y width height
+  draw_plot(Hiseq.roc, 0.43, 0.43, 0.5, 0.5) +
+  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
+  draw_plot_label(c("A", "B", "C"), c(0.05, 0.4, 0.42), c(0.95, 0.95, 0.46), size = 15)
+
+pdf("../results/figures/hiseq1.pdf",width=7,height=5)
+ggdraw() +
+  draw_plot(hiseq.setup, 0, .1, 0.4, 0.8) + #x y width height
+  draw_plot(Hiseq.roc, 0.43, 0.43, 0.5, 0.5) +
+  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
+  draw_plot_label(c("A", "B", "C"), c(0.05, 0.4, 0.42), c(0.95, 0.95, 0.46), size = 15)
+dev.off()
 
@@
 
-tbl<-tableGrob(h.cut.read.table.5,rows=NULL,theme=ttheme_minimal()) # Make table
+tbl<-make.table(h.cut.read.table.5,tt1) # Make table
 
@@
 #knitr::kable(subset(lofreq.table,gc==5,select=c(exp.freq,TP,FP)))
+lofreq.table<-subset(lofreq.table,gc==5,select=c(exp.freq,TP,FP))
 
-tbl<-tableGrob(lofreq.table,rows=NULL,theme=ttheme_minimal()) # Make table
+tbl<-make.table(lofreq.table,tt1) # Make table
 
@@
 
-tbl.4<-tableGrob(h.cut.read.table.4,rows=NULL,theme=ttheme_minimal()) # Make table
+tbl.4<-make.table(h.cut.read.table.4,tt1) # Make table
 
@@
 
-tbl.3<-tableGrob(h.cut.read.table.3,rows=NULL,theme=ttheme_minimal()) # Make table
+tbl.3<-make.table(h.cut.read.table.3,tt1) # Make table
 
@@
 
-tbl<-tableGrob(dups.table,rows=NULL,theme=ttheme_minimal()) # Make table
+tbl<-make.table(dups.table,tt1) # Make table
 
@@
 This is so much better than before.
-### Concequences for diversity measurements

### 2016-01-22 scripts/figures.Rmd
deciding on legends

@@
 
-> make sure the possible fp accounts for the fact that we aren't looking at the priming regions for both Miseq and Hiseq
-## Figure 1
-
-The accuracy of variants callers is most appropiately described by the sensitivity and specificity of a the caller at a given qualtiy threshold.
-The sensitivity of a caller is the proportion of true, known variants identified by the algorith.
-The specificty is the the propotion of false variants corrected ignored.
-There is a trade off between sensitivity and specificity.
-One may be able to correctly identify all known variants from a data; however, if in doing the the algorithm identifies a large number of false positives then a sensitivity of 1 is of little use in practive.
-Conversly it is possible to increase the stringency of a caller to the point that it does not identify any erroneous variants however, if in doing so all the true variants are also removed, perfect specificity is of no use.
-A reciever operator curve (ROC) is a common way of visualizing the trade off between sensitivity and specificity.
-The DeepSNV algorithm returns a p value for each called variant.
-Figure 1 A depicts the relationship between various p value thresholds, and the resulting ROC using a hypothetical data set.
-
-### ROC skematic
+### Figure 1
 ```{r, roc_skematic,fig.height=5,fig.width=3.4}
@@
 
-### Experimental setup diagram
-
-The accuracy of DeepSNV algorithm has been extensively benchmarked using a region of the HIV genome (?kb); however, we wanted to determine its accuracy in calling variants accross the entire influenza A genome ( 8 segments totally roughly 13kb).
-To this end we diluted one lab strain WSN33 into another PR8 at decreasing frequencies.
-491 single nucleotide variants were confirmed between these strains using sanger sequencing of plaque isolated samples.
-
-### DeepSNV ROC 
 ```{r,roc_helper_runctions}
@@
 find_freq.v<-Vectorize(find_freq) # vectorized version of the helper function above
+
+find_freq.h<-function(Id){ # helper function to get the expected frequency from the sample name
+  if(Id==0.05){
+    x="5.0%"} else if ( Id==0.02){
+    x="2.0%"} else if ( Id==0.01){
+    x="1.0%"} else if ( Id==0.005){
+    x="0.5%"} else if ( Id==0.002){
+    x="0.2%"}
+}
+
+find_freq.v.h<-Vectorize(find_freq.h) # vectorized version of the helper function above
 
@@
   )-> roc.df.adj
-  #fills<-ddply(roc.df.adj,~samp,fill_in_plots) # take out these lines to not fill in the plots
-  #roc.df.adj<-rbind(fills,roc.df.adj) # here too
+  fills<-ddply(roc.df.adj,~samp,fill_in_plots) # take out these lines to not fill in the plots
+  roc.df.adj<-rbind(fills,roc.df.adj) # here too
   roc.df.adj$exp.freq<-as.character(roc.df.adj$exp.freq) # so that teh calculations below work since it is a factor here
@@
 
-The initial ROC of DeepSNV applied the viral dilutions above.
-Note for these analyses we only inlcude variants with p<0.05.
-I will extend this in the future.
-Doing so will extend the ROCs but will not change the results as we set a cut off of 0.01 at most in all the tables.
-
-> Uniform ROC axis?
-
-## Figure 2
+### Figure 2
 ```{r,Miseq_out_of_the_box,fig.width=7,fig.height=5}
@@
 
-To test our accuracy under conditions that more closely resemble those found in patient derived samples, 20 single point mutants were mixed and diluted into wild type decreasing frequencies.
-This mixtures were then diltued into viral media to yeild a spectrum of known variants at a range of input levels.
-
 ### Figure 3
-Out of the box here is how we do for the $10^5$ sample.
 
@@
 
-Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")+theme(plot.margin=unit(c(0,0,0,0),"mm"))
+x<-ddply(h.roc.df,~Id,fill_in_plots)
 
-# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/hiseq.roc.pdf")
-# Hiseq.roc
-# dev.off()
+x<-mutate(x,frequency=find_freq.v.h(exp.freq))
+
+Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1.1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")+geom_text(data=subset(x,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2,position=position_jitter(h=0.1,w=0))
+
+# This is so that the text is not cut off
+Hiseq.roc <- ggplot_gtable(ggplot_build(Hiseq.roc))
+Hiseq.roc$layout$clip[Hiseq.roc$layout$name == "panel"] <- "off"
 
@@
 
-## 3) Phred and MapQ filters
-
-It is common practice to account for base and mapping quality scores when calling variants.
-While DeepSNV is agnostic towards mapping qualty our initial analysis masked any base that had a phred score less than 30.
-Here we show that by focusing our analysis to varaints that have average phred >35 and average mapping quality >30 we can increase specificity without decrease sensitivity.
-
-> The first plot here is just to show you the distributions.
-p<0.01.
-We usually only look at this after the read cut off.
-But here you can see the cut offs to quite a bit, but do leave a bit to do.
-the cut offs of 30 and 35 can be adjusted as well.
-There must be a nice way to find the cutoffs algorithmically.
-
 ```{r, phred_mapq filter}
@@
 
-cut.df<-subset(sum.df,MapQ>30 & Phred>35)
+cut.df<-subset(sum.df,MapQ>30 & Phred>35) # This removes the variants that don't pass
+
+cut.df.all<-sum.df
+cut.df.all$p.val[cut.df.all$MapQ<30 | cut.df.all$Phred<35]<-1
 
@@
 
-## Figure 4
-
-### Read Position
-
-> Again the first plot is just to show the distribution is the significant variants that pass the map and prhed cutoff for all samples run.
+### Figure 4
 
@@
 
-cut.read.df<-subset(cut.df,Read_pos>32 & Read_pos<94)
+#cut.read.df<-subset(cut.df,Read_pos>32 & Read_pos<94) # This removes the variants that don't pass
 
-h.cut.read.df<-hiseq.roc(cut.read.df,possible_tp,possible_vars)
+cut.read.df.all<-cut.df.all # this set's the p.val of variants that don't pass the cut to 1
+cut.read.df.all$p.val[cut.read.df.all$Read_pos<32 | cut.read.df.all$Read_pos>94]<-1
 
-read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.0025),breaks=c(0,0.001,0.002),labels=c(0:2))+theme(legend.position="none")
+h.cut.read.df<-hiseq.roc(cut.read.df.all,possible_tp,possible_vars)
+
+read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
 
@@
 
-h.cut.read.table<-hiseq.roc.table(cut.read.df,0.01,possible_tp,possible_vars)
+h.cut.read.table<-hiseq.roc.table(cut.read.df.all,0.01,possible_tp,possible_vars)
 #knitr::kable(subset(h.cut.read.table,gc==5))
@@
 
-Same as above just with an added read cut off at 32 and 94 (middle 50%). p<0.01
+### Figure 5
 
-### Lofreq out of the box
-
-```{r,lofreq}
+```{r,lofreq,fig.height=5,fig.width=3.4}
 
@@
 
-ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")->lofreq.roc.p
+ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position=c(1,.5))->lofreq.roc.p
 
@@
 
-ggdraw() +
-  draw_plot(lofreq.roc.p, 0, .5, 0.5, .5) +
-  draw_plot(tbl, 0.5, 0.5, 0.5, .5) +
-  draw_plot_label(c("A", "B"), c(0, 0.5), c(1, 1), size = 15)
+# ggdraw() +
+#   draw_plot(lofreq.roc.p, 0, .5, 0.5, .5) +
+#   draw_plot(tbl, 0.5, 0.5, 0.5, .5) +
+#   draw_plot_label(c("A", "B"), c(0, 0.5), c(1, 1), size = 15)
+
+plot_grid(lofreq.roc.p, tbl, labels=c("A", "B"), ncol = 1, nrow = 2)
 
@@
 
-### Frequency accuracy
+### Figure 6
 
@@
 ## Frequency
-cut.read.sig.df<-subset(cut.read.df,p.val<0.01)
+cut.read.sig.df<-subset(cut.read.df.all,p.val<0.01)
 medians<-ddply(subset(cut.read.sig.df,category==T ),~exp.freq+gc, summarize, median=median(freq.var))
@@
 
-## Figure 5 Lower inputs
+### Figure 7
 
 ```{r,roc_lower_inputs4}
-roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.0025),breaks=c(0,0.001,0.002),labels=c(0:2))+theme(legend.position="none")
+roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
 
@@
 
-roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.0025),breaks=c(0,0.001,0.002),labels=c(0:2))+theme(legend.position="none")
+roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
 
@@
 
-### Figure 6
+### Figure 8
 
@@
 
-dups<-subset(dups,p.val!=1)
+#dups<-subset(dups,p.val!=1)
 
@@
 
-dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.001),breaks=c(0,0.0005,0.001),labels=c(0,0.5,1))+theme(legend.position="none")
+dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
 

### 2016-01-25 scripts/figures.Rmd
layouts

@@
 
-x<-ddply(h.roc.df,~Id,fill_in_plots)
+x<-ddply(h.roc.df,~samp,fill_in_plots)
 
@@
 
-Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1.1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")+geom_text(data=subset(x,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2,position=position_jitter(h=0.1,w=0))
+x<-subset(x,gc==5)
+
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.01
+#x$adj.sensitivity[x$frequency=="5.0%"]<-x$adj.sensitivity[x$frequency=="5.0%"]+0.0
+
+Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")+geom_text(data=subset(x,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2)
 
@@
   draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
-  draw_plot_label(c("A", "B", "C"), c(0.05, 0.4, 0.42), c(0.95, 0.95, 0.46), size = 15)
+  draw_plot_label(c("A", "B", "C"), c(0.05, 0.42, 0.42), c(0.95, 0.95, 0.46), size = 15)
 
@@
   draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
-  draw_plot_label(c("A", "B", "C"), c(0.05, 0.4, 0.42), c(0.95, 0.95, 0.46), size = 15)
+  draw_plot_label(c("A", "B", "C"), c(0.05, 0.42, 0.42), c(0.95, 0.95, 0.46), size = 15)
 dev.off()
@@
 
-mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)+geom_hline(yintercept=20,linetype=2)+geom_vline(xintercept=30,linetype=2)+scale_x_continuous(limits=c(30,40))+scale_color_hue(name="",labels=c("False posivite","True positive"))+ylab("Mean mapping quality")+xlab("Mean base qualtiy (Phred) ")
+mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)+geom_hline(yintercept=20,linetype=2)+geom_vline(xintercept=30,linetype=2)+scale_x_continuous(limits=c(30,40))+scale_color_hue(name="",labels=c("False posivite","True positive"))+ylab("Mean mapping quality")+xlab("Mean base qualtiy (Phred) ")+theme(legend.position="top")
 
@@
 
-read.pos<-ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')+scale_fill_hue(name="",labels=c("False positive","True positive"))+xlab("Average position on read")+ylab("Number of variants")+theme(legend.position=c(.5,.7))
+read.pos<-ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')+scale_fill_hue(name="",labels=c("False positive","True positive"))+xlab("Average position on read")+ylab("Number of variants")+theme(legend.position="top")
 # read.pos
@@
 
-read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
+x<-ddply(h.cut.read.df,~samp,fill_in_plots)
 
-# read.roc
-# 
-# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/read.roc.pdf")
-# read.roc
-# dev.off()
+x<-mutate(x,frequency=find_freq.v.h(exp.freq))
+
+x<-subset(x,gc==5)
+
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.05
+#x$adj.sensitivity[x$frequency=="5.0%"]<-x$adj.sensitivity[x$frequency=="5.0%"]+0.0
+
+read.roc<-ggplot(subset(h.cut.read.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")+geom_text(data=x,aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2)
+
+# This is so that the text is not cut off
+read.roc <- ggplot_gtable(ggplot_build(read.roc))
+read.roc$layout$clip[read.roc$layout$name == "panel"] <- "off"
 
@@
 
+pdf("../results/figures/figure4.pdf",width=7,height=6)
+ggdraw() +
+  draw_plot(mapq_phred.normal, 0, .5, .5, .48) + 
+  draw_plot(read.pos, .5, .5, .5, .48) +
+  draw_plot(read.roc, 0, 0, 0.42, .5) +
+  draw_plot(tbl, 0.47, 0, 0.52, 0.47) +
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(0.95, 0.95, 0.55,0.45), size = 15)
+dev.off()
 ```
@@
 plot_grid(lofreq.roc.p, tbl, labels=c("A", "B"), ncol = 1, nrow = 2)
+
+pdf("../results/figures/lofreq1.pdf",width=7,height=2.5)
+plot_grid(lofreq.roc.p, tbl, labels=c("A", "B"), ncol = 2, nrow = 1)
+dev.off()
 

### 2016-01-26 scripts/figures.Rmd
just need to move some lables around

@@
 
-ggdraw() +
-  draw_plot(miseq.setup, 0.01, .1, 0.4, 0.8) + #x y width height
-  draw_plot(miseq.roc.p, 0.43, 0.42, 0.5, 0.51) +
-  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
-  draw_plot_label(c("A", "B", "C"), c(0.01, 0.42, 0.42), c(0.78, 0.95, 0.46), size = 15)
-
 pdf("../results/figures/miseq1.pdf",width=7,height=5)
@@
 dev.off()
+
+ggdraw() +
+  draw_plot(miseq.setup, 0.01, .1, 0.4, 0.8) + #x y width height
+  draw_plot(miseq.roc.p, 0.43, 0.42, 0.5, 0.51) +
+  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
+  draw_plot_label(c("A", "B", "C"), c(0.01, 0.42, 0.42), c(0.78, 0.95, 0.46), size = 15)
 
@@
 
-rm(list = setdiff(ls(), lsf.str())) # remove all variables except the functions
-tt1<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.75,fontfamily="serif")),core=list(fg_params=list(cex=0.75,fontfamily="serif")))
-
 ```
@@
 ```{r,hiseq_roc,fig.width=7,fig.height=5}
+#rm(list = setdiff(ls(), lsf.str())) # remove all variables except the functions
+tt1<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.75,fontfamily="serif")),core=list(fg_params=list(cex=0.75,fontfamily="serif")))
+
 sum.df<-read.csv("../data/process/2015-6-23/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#") # reading in the variants from the pipeline output
@@
 
-ggdraw() +
-  draw_plot(hiseq.setup, 0, .1, 0.4, 0.8) + #x y width height
-  draw_plot(Hiseq.roc, 0.43, 0.43, 0.5, 0.5) +
-  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
-  draw_plot_label(c("A", "B", "C"), c(0.05, 0.42, 0.42), c(0.95, 0.95, 0.46), size = 15)
-
 pdf("../results/figures/hiseq1.pdf",width=7,height=5)
@@
 
+ggdraw() +
+  draw_plot(hiseq.setup, 0, .1, 0.4, 0.8) + #x y width height
+  draw_plot(Hiseq.roc, 0.43, 0.43, 0.5, 0.5) +
+  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
+  draw_plot_label(c("A", "B", "C"), c(0.05, 0.42, 0.42), c(0.95, 0.95, 0.46), size = 15)
+
 # 
@@
 
+### Figure 4 
 ```{r, phred_mapq filter}
@@
 
-mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)+geom_hline(yintercept=20,linetype=2)+geom_vline(xintercept=30,linetype=2)+scale_x_continuous(limits=c(30,40))+scale_color_hue(name="",labels=c("False posivite","True positive"))+ylab("Mean mapping quality")+xlab("Mean base qualtiy (Phred) ")+theme(legend.position="top")
+mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(size=1)+geom_hline(yintercept=20,linetype=2)+geom_vline(xintercept=30,linetype=2)+scale_x_continuous(limits=c(30,40))+scale_color_hue(name="",labels=c("False posivite","True positive"))+ylab("Mean mapping quality")+xlab("Mean base qualtiy (Phred) ")+theme(legend.position="top")
 
@@
 
-This is still for p<0.01 and 10^5 genomes/ul viral transport media.
-
-### Figure 4
-
-```{r,read_pos}
+```{r,read_pos,fig.width=7,fig.height=6}
 # ggplot(sig.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
@@
 
-ggdraw() +
-  draw_plot(mapq_phred.normal, 0, .5, .5, .5) + 
-  draw_plot(read.pos, .5, .5, .5, .5) +
-  draw_plot(read.roc, 0, 0, 0.5, .5) +
-  draw_plot(tbl, 0.5, 0, 0.5, .5)+
-  draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(1, 1, 0.5,0.5), size = 15)
-
 pdf("../results/figures/figure4.pdf",width=7,height=6)
@@
 dev.off()
+
+ggdraw() +
+  draw_plot(mapq_phred.normal, 0, .5, .5, .48) + 
+  draw_plot(read.pos, .5, .5, .5, .48) +
+  draw_plot(read.roc, 0, 0, 0.42, .5) +
+  draw_plot(tbl, 0.47, 0, 0.52, 0.47) +
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(0.95, 0.95, 0.55,0.45), size = 15)
 ```
@@
 
-ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position=c(1,.5))->lofreq.roc.p
+x<-ddply(lofreq.roc,~samp,fill_in_plots)
+
+x<-mutate(x,frequency=find_freq.v.h(exp.freq))
+
+x<-subset(x,gc==5)
+
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.1
+x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]-0.03
+
+ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position='none')+geom_text(data=x,aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2)->lofreq.roc.p
+# This is so that the text is not cut off
+lofreq.roc.p <- ggplot_gtable(ggplot_build(lofreq.roc.p))
+lofreq.roc.p$layout$clip[lofreq.roc.p$layout$name == "panel"] <- "off"
 
 #lofreq claims no further monkey bussiness is needed to we'll just make a table with all its output
-ddply(lofreq.df,~Id,summarize,gc=gc[1],exp.freq=exp.freq[1],TP=length(which(category==T)),FP=length(which(category==F)))->lofreq.table
+ddply(lofreq.df,~Id,summarize,gc=gc[1],exp.freq=exp.freq[1],TP=length(which(category==T)),Sensitivity=TP/20,FP=length(which(category==F)),Specificity=(possible_vars-possible_tp-FP)/(possible_vars-possible_tp))->lofreq.table
 lofreq.table<-lofreq.table[order(lofreq.table$exp.freq,decreasing = T),]
@@
 #knitr::kable(subset(lofreq.table,gc==5,select=c(exp.freq,TP,FP)))
-lofreq.table<-subset(lofreq.table,gc==5,select=c(exp.freq,TP,FP))
+lofreq.table<-head(subset(lofreq.table,gc==5,select=-c(gc,Id)),n=-1) # don't need the 0.1% sample
+rename(lofreq.table,c("exp.freq"="Frequency","TP"="True Positives","FP"="False Positives"))->lofreq.table
+lofreq.table$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
+lofreq.table$Sensitivity<-round(lofreq.table$Sensitivity,digits=dd)
+lofreq.table$Specificity<-round(lofreq.table$Specificity,digits=dd)
 
-tbl<-make.table(lofreq.table,tt1) # Make table
+tt2<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.65,fontfamily="serif")),core=list(fg_params=list(cex=0.65,fontfamily="serif")))
+
+tbl<-make.table(lofreq.table,tt2) # Make table
 
@@
 
-plot_grid(lofreq.roc.p, tbl, labels=c("A", "B"), ncol = 1, nrow = 2)
+pdf("../results/figures/lofreq1.pdf",width=3.4,height=5)
+ggdraw() +
+  draw_plot(lofreq.roc.p, 0, 0.5, 0.9, 0.45) + 
+  draw_plot(tbl, 0.025, 0, 0.95, .48) +
+  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
+dev.off()
 
-pdf("../results/figures/lofreq1.pdf",width=7,height=2.5)
-plot_grid(lofreq.roc.p, tbl, labels=c("A", "B"), ncol = 2, nrow = 1)
-dev.off()
+ggdraw() +
+  draw_plot(lofreq.roc.p, 0, 0.5, 0.9, 0.45) + 
+  draw_plot(tbl, 0.025, 0, 0.95, .48) +
+  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
 
@@
 For p<0.01 with phred, mapq and read position cut off.
-```{r, frequency_accuracy}
+```{r, frequency_accuracy,fig.width=3.4,fig.height=3.4}
 # Here we'll just be using
@@
 
-ac_freq<-ac_freq+geom_dotplot(binaxis = "y", stackdir = "center",dotsize=1)+scale_color_manual(values=rhg_cols)
+ac_freq<-ac_freq+geom_dotplot(binaxis = "y", stackdir = "center",dotsize=1)+scale_fill_manual(values=rhg_cols)
 
-ac_freq<-ac_freq+scale_y_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.05,0.1))#+theme(axis.text.y = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
-ac_freq<-ac_freq+scale_x_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.050,0.1))#+theme(axis.text.x = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
+ac_freq<-ac_freq+scale_y_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.05,0.1),label=c("0.1%","0.2%","0.5%","1.0%","2.0%","5.0%","10%"))#+theme(axis.text.y = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
+ac_freq<-ac_freq+scale_x_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.050,0.1),label=c("0.1%","0.2%","0.5%","1.0%","2.0%","5.0%","10%"))#+theme(axis.text.x = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
 ac_freq<-ac_freq+geom_abline(intercept = 0, slope = 1,linetype=2,size=1)
@@
 
-ac_freq<-ac_freq+geom_point(data=subset(medians,gc=="5"),mapping=aes(x=exp.freq,y=median),shape=95,size=30)
+ac_freq<-ac_freq+geom_point(data=subset(medians,gc=="5"),mapping=aes(x=exp.freq,y=median),shape=95,size=20)
 ac_freq
 
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/freq.pdf")
+pdf("../results/figures/freq.pdf",width=3.4,height=3.4)
 ac_freq
@@
 
-```{r,roc_lower_inputs4}
-roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
+```{r,roc_lower_inputs4,fig.width=7,fig.height=5}
+
+x<-ddply(h.cut.read.df,~samp,fill_in_plots)
+x<-subset(x,gc==4)
+
+x<-mutate(x,frequency=find_freq.v.h(exp.freq))
+
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.05
+x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]+0.05
+x$adj.sensitivity[x$frequency=="0.5%"]<-x$adj.sensitivity[x$frequency=="0.5%"]+0.025
+
+roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")+geom_text(data=x,aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2)
+
+# This is so that the text is not cut off
+roc.4 <- ggplot_gtable(ggplot_build(roc.4))
+roc.4$layout$clip[roc.4$layout$name == "panel"] <- "off"
 
@@
 
-roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
+x<-ddply(h.cut.read.df,~samp,fill_in_plots)
+x<-subset(x,gc==3)
+
+x<-mutate(x,frequency=find_freq.v.h(exp.freq))
+
+roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")+geom_text(data=x,aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2)
+
+# This is so that the text is not cut off
+roc.3 <- ggplot_gtable(ggplot_build(roc.3))
+roc.3$layout$clip[roc.3$layout$name == "panel"] <- "off"
 
@@
 
+pdf("../results/figures/lower.pdf",width=7,height=5)
 ggdraw() +
-  draw_plot(roc.4, 0, .5, .5, .5) +
-  draw_plot(tbl.4, .5, .5, .5, .5) +
-  draw_plot(roc.3, 0, 0, 0.5, .5) +
-  draw_plot(tbl.3, 0.5, 0, 0.5, .5)+
+  draw_plot(roc.4, 0, .48, .43, .5) +
+  draw_plot(tbl.4, 0.47, 0.55, 0.52, 0.47) +
+  draw_plot(roc.3, 0, 0, 0.43, .5) +
+  draw_plot(tbl.3, 0.47, 0, 0.52, 0.47)+
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(1, 1, 0.5,0.5), size = 15)
+dev.off()
+
+ggdraw() +
+  draw_plot(roc.4, 0, .48, .43, .5) +
+  draw_plot(tbl.4, 0.47, 0.55, 0.52, 0.47) +
+  draw_plot(roc.3, 0, 0, 0.43, .5) +
+  draw_plot(tbl.3, 0.47, 0, 0.52, 0.47)+
   draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(1, 1, 0.5,0.5), size = 15)
@@
 
-```{r,duplicates}
+```{r,duplicates,fig.height=5,fig.width=3.4}
 sum.df<-read.csv("../data/process/2015-11-14/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#")
@@
 
-dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
+x<-ddply(dups.roc,~samp,fill_in_plots)
+
+x<-mutate(x,frequency=find_freq.v.h(exp.freq))
+
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.15
+x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]-0.08
+
+dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")+geom_text(data=x,aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2)
+
+# This is so that the text is not cut off
+dups.roc.p <- ggplot_gtable(ggplot_build(dups.roc.p))
+dups.roc.p$layout$clip[dups.roc.p$layout$name == "panel"] <- "off"
 
@@
 
-tbl<-make.table(dups.table,tt1) # Make table
-
-ggdraw() +
-  draw_plot(dups.roc.p, 0, .5, 0.5, .5) +
-  draw_plot(tbl, 0.5, 0.5, 0.5, .5) +
-  draw_plot_label(c("A", "B"), c(0, 0.5), c(1, 1), size = 15)
+tbl<-make.table(dups.table,tt2) # Make table
 
@@
                                                                                       out=data.frame(freq1=freq1,freq2=freq2)})
+
+pdf("../results/figures/dups.roc.pdf",width=3.4,height=5)
+ggdraw() +
+  draw_plot(dups.roc.p, 0, 0.5, 0.9, 0.45) + 
+  draw_plot(tbl, 0.025, 0, 0.95, .48) +
+  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
+dev.off()
+
+ggdraw() +
+  draw_plot(dups.roc.p, 0, 0.5, 0.9, 0.45) + 
+  draw_plot(tbl, 0.025, 0, 0.95, .48) +
+  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
 

### 2016-01-26 scripts/figures.Rmd
labels adjusted started legends

@@
 
-Here are the figures (or figure panels) we are thinking of using in the Benchmarking paper.
-I will include my comments and decriptions on each and perhaps over time this will develop into the results section
 ```{r,settings_and_lib,echo=F,message=F,}
@@
 ```
+
+Figure 1.
+An example ROC.
+A) Hypothetical variants are stratified by p value.
+True variants are shown as filled cirlces, while those corresponding to RT, PCR, or sequencing errors are shown as open circles.
+Potential thresholds are indicated as colored lines.
+B) An ROC made from the data shown in A.
+The colored lines indicate the points made by the thresholds in A.
 
@@
 
+Figure 2.
+Initial DeepSNY accuracy.
+A) RT-PCR amplified WSN33 genomes were diluted into PR8 at decreasing frequencies and sequenced on an Illumina Miseq.
+B) An ROC showing our ability to identify WSN33 SNV relative to PR8 at the indicated frequencies.
+C) A table of the DeepSNV accuracy at a p value threshold of 0.01.
+
 ```{r,Miseq_read.cut,eval=F}
@@
 
-length(wt_mut)
+#length(wt_mut) How many wt muts do we find.
 sum.df$category[sum.df$mutation %in% wt_mut]<-"wt"
@@
 
-x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.01
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.03
 #x$adj.sensitivity[x$frequency=="5.0%"]<-x$adj.sensitivity[x$frequency=="5.0%"]+0.0
@@
 ```
+
+Figure 3. DeepSNV accuracy under condition that more accurately mimic those found in patient dervived samples. A) 20 known single point mutants were diluted into wt at decreasing frequencies.
 
@@
 
-x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.05
-#x$adj.sensitivity[x$frequency=="5.0%"]<-x$adj.sensitivity[x$frequency=="5.0%"]+0.0
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.01
+x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]+0.08
 
@@
 
-x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.1
-x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]-0.03
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.04
+x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]-0.12
 
@@
 
-x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.15
-x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]-0.08
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.18
+x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]-0.09
 
@@
 ```
-
-These are variants that pass mapq,phred, and read cut and p<0.01 in both samples.10^4
-
-This is so much better than before.

### 2016-01-26 scripts/figures.Rmd
rough,rough legends

@@
 A) Hypothetical variants are stratified by p value.
-True variants are shown as filled cirlces, while those corresponding to RT, PCR, or sequencing errors are shown as open circles.
+True variants are shown as filled circles, while those corresponding to RT, PCR, or sequencing errors are shown as open circles.
 Potential thresholds are indicated as colored lines.
@@
 
-Figure 3. DeepSNV accuracy under condition that more accurately mimic those found in patient dervived samples. A) 20 known single point mutants were diluted into wt at decreasing frequencies.
+Figure 3.
+DeepSNV accuracy under conditions that more accurately mimic those found in patient derived samples.
+A) 20 known single point mutants were diluted into wild type at decreasing frequencies.
+These were diluted further into viral media to match the genomes concentration found in patent samples (10$^5$-10$^3$ genomes/$\mu$l).
+B) An ROC indicating our ability to accurately identify the 20 known variants at 10 $^5$ genomes/$\mu$l.
+C) A table of the accuracy of our variant calling for the 10 $^5$ genomes/$\mu$l samples at a p value threshold of 0.01.
 
@@
 
+Figure 4.
+Investigating the quality of variants in the the 10 $^5$ genomes/$\mu$l samples.
+A) All called variants from the 10 $^5$ genomes/$\mu$l samples with p<0.01 stratified by the mean mapping quality of the reads containing the variants and the mean Phred scores of the variant bases.
+Dashed lines indicate the common cutoffs of 20 and 30 for mapping quality and Phred respectively.
+B) A histogram of all called variants from the 10 $^5$ genomes/$\mu$l samples with p<0.01 binned by the average position of the variant relative to the Hiseq reads on which it is found.
+C) An ROC of the accuracy of the variant caller when the following quality cut offs are applied : mean mapping quality of 30, mean Phred of 35, and an average read position between 32 and 94 (the middle 50% of the read).
+D) A table summarizing the ROC in C at a p value threshold of 0.01.
+
 ### Figure 5
@@
 
+Figure 5.
+The accuracy of the common variant caller Lofreq.
+The variant caller Lofreq was applied to the 10 $^5$ genomes/$\mu$l samples.
+A) An ROC of the accuracy of Lofreq using standard parameters.
+The specificity of the caller was scaled to account for the same number of tests as the DeepSNV algorithm.
+Lofreq provides qualtiy scores instead of p values.
+These were used to set cutoffs for the ROC.
+B) A table summarizing the accuracy of the ROC A.
+
 ### Figure 6
 
-For p<0.01 with phred, mapq and read position cut off.
 ```{r, frequency_accuracy,fig.width=3.4,fig.height=3.4}
@@
 ```
+
+Figure 6. The accuracy of the frequency measurements for the known variants in the 10 $^5$ genomes/$\mu$l samples. The median of each distribution is shown as a black bar.
 
@@
 ```
+
+Figure 7.
+The accuracy of DeepSNV at lower input levels.
+A) An ROC depicting the accuracy of our method when applied to the 10 $^4$ genomes/$\mu$l samples.
+B) A table summarizing the ROC in A at a p value threshold of 0.01.
+C) An ROC depicting the accuracy of our method when applied to the 10 $^3$ genomes/$\mu$l samples.
+B) A table summarizing the ROC in C at a p value threshold of 0.01.
 
@@
 ```
+
+Figure 8.
+The accucary of DeepSNV at lower input levels when run in duplicate.
+A subset of the 10 $^4$ genomes/$\mu$l samples were processed in duplicate to control for RT-PCR errors.
+A) The accuracy of our varant calling method at 10 $^4$ genomes/$\mu$l is greatly enhanced by requiring putative variants to be found in duplicate RT-PCR reactions.
+B) A table summarizing the ROC in C at a p value threshold of 0.01.

### 2016-01-26 scripts/figures.Rmd
page breaks

@@
 author: "JT McCrone"
-date: "December 30, 2015"
-output: html_document
+date: "January 25, 2015"
+output: pdf_document
 ---
@@
 The colored lines indicate the points made by the thresholds in A.
-
+\pagebreak
 ```{r,roc_helper_runctions}
@@
 Figure 2.
-Initial DeepSNY accuracy.
+Initial DeepSNV accuracy.
 A) RT-PCR amplified WSN33 genomes were diluted into PR8 at decreasing frequencies and sequenced on an Illumina Miseq.
@@
 C) A table of the DeepSNV accuracy at a p value threshold of 0.01.
+\pagebreak
 
@@
 
+\pagebreak
+
 ### Figure 4 
@@
 
+\pagebreak
+
 ### Figure 5
@@
 The specificity of the caller was scaled to account for the same number of tests as the DeepSNV algorithm.
-Lofreq provides qualtiy scores instead of p values.
+Lofreq provides quality scores instead of p values.
 These were used to set cutoffs for the ROC.
 B) A table summarizing the accuracy of the ROC A.
+
+\pagebreak
 
@@
 Figure 6. The accuracy of the frequency measurements for the known variants in the 10 $^5$ genomes/$\mu$l samples. The median of each distribution is shown as a black bar.
+
+\pagebreak
 
@@
 
+\pagebreak
+
 ### Figure 8
@@
 Figure 8.
-The accucary of DeepSNV at lower input levels when run in duplicate.
+The accuracy of DeepSNV at lower input levels when run in duplicate.
 A subset of the 10 $^4$ genomes/$\mu$l samples were processed in duplicate to control for RT-PCR errors.
-A) The accuracy of our varant calling method at 10 $^4$ genomes/$\mu$l is greatly enhanced by requiring putative variants to be found in duplicate RT-PCR reactions.
+A) The accuracy of our variant calling method at 10 $^4$ genomes/$\mu$l is greatly enhanced by requiring putative variants to be found in duplicate RT-PCR reactions.
 B) A table summarizing the ROC in C at a p value threshold of 0.01.

### 2016-01-26 doc/paper/results.Rmd
results started

---
title: "Results"
author: "JT McCrone"
date: "January 26, 2016"
output: pdf_document
---
```{r.message=F,echo=F}
xamp_roc<-"figure 1"
```

There are two major components to the accuracy of a variant caller. The first, sensitivity, describes how well a variant caller can identify variants that are know to be real. In other words what proportion of the true positives are found.  The second, specificity, measures how well a variant caller can correctly identify true negatives. For example, the influenza genome is roughly 13KB. There is potential for 3 variant nucleotides at each position leaving a variant caller with about 39,000 potential variants that can be called. The vast majority of these potential variants will not be present in the starting population and are potential false positives. A variant caller that correctly identifies all potential false positives as true negatives has a specificity of 1.

There is an obvious trade off between sensitivity and specificity. It may be the case that a variant caller has near perfect sensitivity, ie. it finds all the true positives in a sample. However, if in the process it has low specificity and identifies many false positives that will confound down stream analysis then the high sensitivity is all for naught. Conversely, a variant caller may have perfect specificity and identify no false positives; however, if in process it has low sensitivity and identifies only a few true positives again it will have limited applications.

The trade off between sensitivity and specificity can be represented as a reciter operator curve or ROC (`r xamp_roc`). A ROC plots the sensitivity of an assay along the y axis and 1-specificity along the x axis. Therefore a perfect assay with a sensitivity and specificity of 1 will lie in at the coordinates (0,1) in the top left hand corner of the plot.

### 2016-01-26 scripts/figures.Rmd
results started

@@
 A) RT-PCR amplified WSN33 genomes were diluted into PR8 at decreasing frequencies and sequenced on an Illumina Miseq.
-B) An ROC showing our ability to identify WSN33 SNV relative to PR8 at the indicated frequencies.
-C) A table of the DeepSNV accuracy at a p value threshold of 0.01.
+B) An ROC measuring the pipeline's ability to identify WSN33 SNVs relative to PR8 at the indicated frequencies.
+C) A summary of the DeepSNV accuracy at a p value threshold of 0.01.
 \pagebreak
@@
 Figure 3.
-DeepSNV accuracy under conditions that more accurately mimic those found in patient derived samples.
-A) 20 known single point mutants were diluted into wild type at decreasing frequencies.
-These were diluted further into viral media to match the genomes concentration found in patent samples (10$^5$-10$^3$ genomes/$\mu$l).
-B) An ROC indicating our ability to accurately identify the 20 known variants at 10 $^5$ genomes/$\mu$l.
+DeepSNV accuracy patient sample conditions A) 20 known single point mutants were diluted into wild type at decreasing frequencies.
+These were diluted further into viral media to match the genomes concentration found in patent samples (10$^5$-10$^3$ genomes/$\mu$l) and sequenced on an Illumina Hiseq (2 x 152bp reads.
+B) An ROC indicating the ability to accurately identify the 20 known variants in the 10 $^5$ genomes/$\mu$l samples.
 C) A table of the accuracy of our variant calling for the 10 $^5$ genomes/$\mu$l samples at a p value threshold of 0.01.
@@
 
-x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.01
-x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]+0.08
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.08
+x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]-0.01
 
@@
 Investigating the quality of variants in the the 10 $^5$ genomes/$\mu$l samples.
-A) All called variants from the 10 $^5$ genomes/$\mu$l samples with p<0.01 stratified by the mean mapping quality of the reads containing the variants and the mean Phred scores of the variant bases.
-Dashed lines indicate the common cutoffs of 20 and 30 for mapping quality and Phred respectively.
+A) All called variants from the 10 $^5$ genomes/$\mu$l samples with p<0.01 stratified by the mean mapping quality of the reads containing the variant and the mean Phred scores of the variant bases.
+Dashed lines indicate  common cutoffs of 20 and 30 for mapping quality and Phred respectively.
 B) A histogram of all called variants from the 10 $^5$ genomes/$\mu$l samples with p<0.01 binned by the average position of the variant relative to the Hiseq reads on which it is found.
@@
 Figure 5.
-The accuracy of the common variant caller Lofreq.
+The accuracy of Lofreq.
 The variant caller Lofreq was applied to the 10 $^5$ genomes/$\mu$l samples.
@@
 These were used to set cutoffs for the ROC.
-B) A table summarizing the accuracy of the ROC A.
+B) A table summarizing the accuracy of the ROC A using the standard cut offs applied by the algorithm.
 
@@
 A) An ROC depicting the accuracy of our method when applied to the 10 $^4$ genomes/$\mu$l samples.
-B) A table summarizing the ROC in A at a p value threshold of 0.01.
+B) A table summarizing the accuracy at 10 $^4$ genomes/$\mu$l with a  p value threshold of 0.01.
 C) An ROC depicting the accuracy of our method when applied to the 10 $^3$ genomes/$\mu$l samples.
-B) A table summarizing the ROC in C at a p value threshold of 0.01.
+B) A table summarizing the accuracy at 10 $^3$ genomes/$\mu$l with a  p value threshold of 0.01
 
@@
 A subset of the 10 $^4$ genomes/$\mu$l samples were processed in duplicate to control for RT-PCR errors.
-A) The accuracy of our variant calling method at 10 $^4$ genomes/$\mu$l is greatly enhanced by requiring putative variants to be found in duplicate RT-PCR reactions.
-B) A table summarizing the ROC in C at a p value threshold of 0.01.
+A) An ROC of  10 $^4$ genomes/$\mu$l samples processed in duplicate.
+A variant was required to be found in both duplicates to be considered.
+B) A table summarizing the accuracy when applied to duplicate samples with 10 $^4$ genomes/$\mu$l input, p<0.01.

### 2016-01-28 doc/paper/results.Rmd
a little more results

@@
 There are two major components to the accuracy of a variant caller.
-The first, sensitivity, describes how well a variant caller can identify variants that are know to be real.
+The first, sensitivity, describes how well a variant caller can identify variants that are known to be real.
 In other words what proportion of the true positives are found.
+A sensitivity of 1 indicates that all true positives are found.
 The second, specificity, measures how well a variant caller can correctly identify true negatives.
 For example, the influenza genome is roughly 13KB.
-There is potential for 3 variant nucleotides at each position leaving a variant caller with about 39,000 potential variants that can be called.
+There are 3  potential variant nucleotides at each position leaving a variant caller with about 39,000 potential variants that can be called.
 The vast majority of these potential variants will not be present in the starting population and are potential false positives.
 A variant caller that correctly identifies all potential false positives as true negatives has a specificity of 1.
+A perfect variant caller will have a sensitivity and specificity of 1.
+It will identify all true variants and only the true variants.
 
 There is an obvious trade off between sensitivity and specificity.
-It may be the case that a variant caller has near perfect sensitivity, ie.
-it finds all the true positives in a sample.
-However, if in the process it has low specificity and identifies many false positives that will confound down stream analysis then the high sensitivity is all for naught.
+It may be the case that a variant caller has near perfect sensitivity, and  can find all the true positives in a sample.
+However, if in the process it has low specificity and identifies many false positives down stream analysis will be confounded.
 Conversely, a variant caller may have perfect specificity and identify no false positives; however, if in process it has low sensitivity and identifies only a few true positives again it will have limited applications.
 
-The trade off between sensitivity and specificity can be represented as a reciter operator curve or ROC (`r xamp_roc`).
-A ROC plots the sensitivity of an assay along the y axis and 1-specificity along the x axis.
-Therefore a perfect assay with a sensitivity and specificity of 1 will lie in at the coordinates (0,1) in the top left hand corner of the plot.
+The trade off between sensitivity and specificity can be represented as a receiver operator curve or ROC (`r xamp_roc`).
+An ROC plots the sensitivity of an assay along the y axis and 1-specificity along the x axis.
+To make a ROC for a variant caller one must test an assay on known data.
+The outcomes can then be stratified according to some metric that represents our belief that a given variant is real.
+Often this is a p value or quality score.
+Cut offs can then be set (`r xamp_roc` A) and because the true outcomes are known one can calculate the sensitivity and specificity.
+The curve is then made up of points representing the sensitivity and specificity of the variant caller at various cut offs.
+A perfect ROC in which all the true positives can be separated from all potential false positives is a right angle that follows the upper left perimeter of the plot and passes through the point (0,1)  (`r xamp_roc` B).

### 2016-02-01 doc/paper/results.Rmd
more writing

@@
 author: "JT McCrone"
-date: "January 26, 2016"
+date: "Figure 1, 2016"
 output: pdf_document
@@
 xamp_roc<-"figure 1"
+miseq<-"figure 2"
+Miseq<-"Figure 2"
 ```
@@
 It may be the case that a variant caller has near perfect sensitivity, and  can find all the true positives in a sample.
-However, if in the process it has low specificity and identifies many false positives down stream analysis will be confounded.
+However, if in the process it has low specificity and identifies many false positives downstream analysis will be confounded.
 Conversely, a variant caller may have perfect specificity and identify no false positives; however, if in process it has low sensitivity and identifies only a few true positives again it will have limited applications.
@@
 Often this is a p value or quality score.
-Cut offs can then be set (`r xamp_roc` A) and because the true outcomes are known one can calculate the sensitivity and specificity.
-The curve is then made up of points representing the sensitivity and specificity of the variant caller at various cut offs.
-A perfect ROC in which all the true positives can be separated from all potential false positives is a right angle that follows the upper left perimeter of the plot and passes through the point (0,1)  (`r xamp_roc` B).
+Because the true are known, the senesitivity and specificity can be calculated at various cut offs (`r xamp_roc` A).
+The points are used to construct the curve (`r xamp_roc` B).
+A perfect ROC in which all the true positives can be separated from all potential false positives is a right angle that follows the upper left perimeter of the plot and passes through the point (0,1).
+
+DeepSNV has been previously benchmarked using small genomic regions involving minimal PCR.
+WSN33 and PR8, two lab strains that differ at roughly 500 positions, were mixed  (`r miseq` A) to ensure this algorith would be applicable to whole length influenza genomes that require an error prone RT-PCR reaction.
+Genomes were amplified separately in  segment specific RT-PCR reactions.
+This was done to avoid biases in amplification that may schew the down stream analysis.
+The products were mixed in a equimolar fashion to yield reconstituted DNA genomes.
+WSN33 was then serially diluted into PR8.
+These mixtures were sequenced using 2x250 reads on an Illumina Miseq and variants were called using DeepSNV.
+Because these are defined mixtures and the 491 true variants are known it is possible to measure the accuracy of our sequencing pipeline.
+We can see from `r miseq` B that as our sensitivity decreases with the the frequency of the vriants.
+`r Miseq` C summarize the accuracy of the pipeline at a p value cut off of 0.01.
+It should be noted that although the specificity is well above 99% for all samples seqeunced out of the box, we were finding 40-80 false positives per sample which is unacceptable.
+The fact that there are over 39,000 potential false positives inflates the specificity measurements.
+
+Encouraged by these results we endeavored to determined the accruacy of our sequencing under conditions that might more accurately represent those found in patient derived samples.

### 2016-02-02 doc/paper/results.Rmd
through qual

@@
 author: "JT McCrone"
-date: "Figure 1, 2016"
+date: "Febuary 2, 2016"
 output: pdf_document
@@
 Miseq<-"Figure 2"
+hiseq<-"figure 3"
+Hiseq<-"Figure 3"
+qual<-"figure 4"
+Qual<-"Figure 4"
 ```
@@
 Often this is a p value or quality score.
-Because the true are known, the senesitivity and specificity can be calculated at various cut offs (`r xamp_roc` A).
+Because the true are known, the sensitivity and specificity can be calculated at various cut offs (`r xamp_roc` A).
 The points are used to construct the curve (`r xamp_roc` B).
@@
 DeepSNV has been previously benchmarked using small genomic regions involving minimal PCR.
-WSN33 and PR8, two lab strains that differ at roughly 500 positions, were mixed  (`r miseq` A) to ensure this algorith would be applicable to whole length influenza genomes that require an error prone RT-PCR reaction.
-Genomes were amplified separately in  segment specific RT-PCR reactions.
-This was done to avoid biases in amplification that may schew the down stream analysis.
-The products were mixed in a equimolar fashion to yield reconstituted DNA genomes.
-WSN33 was then serially diluted into PR8.
+However, RNA viral samples require an RT-PCR step and the accuracy of this variant caller was previously unknown under these conditions.
+To characterize the accuracy of our sequencing pipeline, two lab strains, WSN33 and PR8, that differ at roughly 500 positions, were mixed  to create known viral populations (`r miseq` A).
+Briefly, genomes were amplified separately in segment specific RT-PCR reactions.
+This was done to avoid biases in amplification that may skew the down stream analysis.
+Reconstituted DNA genomes were made by mixing these products at equimolar ratios.
+The WSN33 mix was then serially diluted into PR8.
 These mixtures were sequenced using 2x250 reads on an Illumina Miseq and variants were called using DeepSNV.
+
 Because these are defined mixtures and the 491 true variants are known it is possible to measure the accuracy of our sequencing pipeline.
-We can see from `r miseq` B that as our sensitivity decreases with the the frequency of the vriants.
+Although sensitivity decreases slightly with variant frequency, we were able to detect variants at 0.5% with high accuracy.
 `r Miseq` C summarize the accuracy of the pipeline at a p value cut off of 0.01.
-It should be noted that although the specificity is well above 99% for all samples seqeunced out of the box, we were finding 40-80 false positives per sample which is unacceptable.
-The fact that there are over 39,000 potential false positives inflates the specificity measurements.
+_Something about the specificity_
 
-Encouraged by these results we endeavored to determined the accruacy of our sequencing under conditions that might more accurately represent those found in patient derived samples.
+Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under _adjective_ conditions.
+The virus populations found in patient derived samples are often less divergent than the two lab strains, and provide much less input than samples grown in cell culture.
+To mimic these conditions, 20 single point mutants made in the WSN33 background were diluted into wild type at decreasing frequencies (`r hiseq` A).
+This yielded 5 samples containing 20 known variants at frequencies ranging from 5% to 0.2% all at a concentration of $10^5$ genomes/$\mu$l.
+These samples were diluted serially into viral media to give a range of inputs matching that seen in patient samples.
+These samples were then sequenced on an Illumina Hiseq and variants called using DeepSNV.
+Focusing on the $10^5$ genomes/$\mu$l samples we find that again sensitivity is proportional to variant frequency (`r hiseq` B and C).
+The fact that the sensitivity drops more dramatically than in our previous experiment is not surprising given the fact the the RNA input derived from cell culture samples is often orders of magnitude higher than that provided in this experiment.
+During library prep, a disproportionate amount of the 2.0% sample was lost and this likely accounts for the decreased sensitivity in the 2.0% sample relative to the 1.0%.
+In addition to a more drastic drop in sensitivity we found a much larger number of false positives at a p value cut off of 0.01 (`r hiseq` C).
+It should be noted that in all cases the specificity was greater than 99%.
+However, given the large number of potential false positives even a specificity of 99% results in well over 200 false positives.
+Initially it is tempting to increase specificity by simply applying a more stringent p value cut off.
+This would result in marching backwards towards the y axis along the ROCs in `r hiseq` B.
+However it is apparent from the ROCs that any increase in specificity would result in a drastic decrease in sensitivity.
+In other words, we needed other means of separating true from false positives than simply the DeepSNV output.
+
+In deep sequencing studies of viral diversity mapping quality and base quality cutoffs are often used to ensure that only the highest caliber sequencing data is used to call variants.
+Mapping quality refers to our confidence that a given read is mapped to the correct position in the genome while base quality estimates the likelihood that the base read by the sequencer is correct.
+In our initial analysis we masked bases that had a base quality score less than 30.
+We did not apply any mapping quality cut offs.
+`r Qual` A shows the distribution of the mean quality scores of the variants (p <0.01) present in the $10^5$ genomes/ $\mu$l samples.
+In our case we find that while true and false positive variants can largely be separated by these metrics, the commonly used cut offs are not sufficient (dashed lines `r qual` A).
+Additionally, we found that in general false positives were biased near the ends of sequencing reads (`r qual` B).
+This is not surprising as it is well known that sequencing quality dips near the read end; however, recall that all bases with phred scores less than 30 were masked from our variant calling.
+In this case is seems that even high quality false positives are disproportionately called near the ends of reads, while true positives are called uniformly across the read resulting in an average read position near the middle of the read.
+_potential reasons for the bias?
+pcr errors in the lib step, locations in the genome..._
+
+In the light of these quality distributions we applied a number of cut offs and drastically increased our specificity without sacrificing sensitivity (`r qual` C and D).
+For a variant to be considered in our analysis we required a mean mapping quality of 30,a mean phred score of 35, and an  average read position within the middle 50% of the read length.

### 2016-02-02 doc/paper/results.Rmd
through low input

@@
 Qual<-"Figure 4"
+freq<-"figure 5"
+Freq<-"Figure 5"
+lower<-"figure 6"
+Lower<-"Figure 6"
+dups<-"figure 7"
+Dups<- "Figure 7"
+
 ```
@@
 However, if in the process it has low specificity and identifies many false positives downstream analysis will be confounded.
-Conversely, a variant caller may have perfect specificity and identify no false positives; however, if in process it has low sensitivity and identifies only a few true positives again it will have limited applications.
+Conversely, a variant caller may have perfect specificity and identify no false positives; however, if in process it has low sensitivity and identifies only a few true positives it will have limited applications.
 
@@
 Often this is a p value or quality score.
-Because the true are known, the sensitivity and specificity can be calculated at various cut offs (`r xamp_roc` A).
-The points are used to construct the curve (`r xamp_roc` B).
+Because the true variants are known, the sensitivity and specificity can be calculated at different cut offs (`r xamp_roc` A).
+These points are then used to construct the curve (`r xamp_roc` B).
 A perfect ROC in which all the true positives can be separated from all potential false positives is a right angle that follows the upper left perimeter of the plot and passes through the point (0,1).
@@
 DeepSNV has been previously benchmarked using small genomic regions involving minimal PCR.
-However, RNA viral samples require an RT-PCR step and the accuracy of this variant caller was previously unknown under these conditions.
+However, RNA viral samples require an RT-PCR step, and the accuracy of this variant caller was previously unknown under these conditions.
 To characterize the accuracy of our sequencing pipeline, two lab strains, WSN33 and PR8, that differ at roughly 500 positions, were mixed  to create known viral populations (`r miseq` A).
 Briefly, genomes were amplified separately in segment specific RT-PCR reactions.
-This was done to avoid biases in amplification that may skew the down stream analysis.
+This was done to avoid biases in amplification that may skew frequency measurements towards one strain.
 Reconstituted DNA genomes were made by mixing these products at equimolar ratios.
@@
 Although sensitivity decreases slightly with variant frequency, we were able to detect variants at 0.5% with high accuracy.
-`r Miseq` C summarize the accuracy of the pipeline at a p value cut off of 0.01.
+`r Miseq` C summarizes the accuracy of the pipeline at a p value cut off of 0.01.
 _Something about the specificity_
 
-Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under _adjective_ conditions.
-The virus populations found in patient derived samples are often less divergent than the two lab strains, and provide much less input than samples grown in cell culture.
+Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under laboratory conditions.
+The virus populations found in patient derived samples are often less divergent than the two strains used, and provide much less input than samples grown in cell culture.
 To mimic these conditions, 20 single point mutants made in the WSN33 background were diluted into wild type at decreasing frequencies (`r hiseq` A).
@@
 The fact that the sensitivity drops more dramatically than in our previous experiment is not surprising given the fact the the RNA input derived from cell culture samples is often orders of magnitude higher than that provided in this experiment.
-During library prep, a disproportionate amount of the 2.0% sample was lost and this likely accounts for the decreased sensitivity in the 2.0% sample relative to the 1.0%.
+During library prep, a disproportionate amount of the 2.0% sample was lost and this likely accounts for the decreased sensitivity in the 2.0% sample relative to the 1.0% sample.
 In addition to a more drastic drop in sensitivity we found a much larger number of false positives at a p value cut off of 0.01 (`r hiseq` C).
 It should be noted that in all cases the specificity was greater than 99%.
-However, given the large number of potential false positives even a specificity of 99% results in well over 200 false positives.
+However, given the large number of potential false positives even a specificity of 99% results in well over 200 false variant calls.
 Initially it is tempting to increase specificity by simply applying a more stringent p value cut off.
 This would result in marching backwards towards the y axis along the ROCs in `r hiseq` B.
-However it is apparent from the ROCs that any increase in specificity would result in a drastic decrease in sensitivity.
-In other words, we needed other means of separating true from false positives than simply the DeepSNV output.
+However it is apparent from the `r hiseq` B that any increase in specificity would result in a drastic decrease in sensitivity.
+In other words, we require other means of separating true from false positives than simply the DeepSNV p value.
+_perhaps more details about the setup mimicing real world.
+such as the mRT-PCR_
 
@@
 We did not apply any mapping quality cut offs.
-`r Qual` A shows the distribution of the mean quality scores of the variants (p <0.01) present in the $10^5$ genomes/ $\mu$l samples.
-In our case we find that while true and false positive variants can largely be separated by these metrics, the commonly used cut offs are not sufficient (dashed lines `r qual` A).
+`r Qual` A shows the distribution of the mean quality scores of the variants (p <0.01) present in the $10^5$ genomes/$\mu$l samples.
+In our case we find that while true and false positive variants can largely be separated by these metrics,  commonly used cut offs are not sufficient (dashed lines `r qual` A).
 Additionally, we found that in general false positives were biased near the ends of sequencing reads (`r qual` B).
 This is not surprising as it is well known that sequencing quality dips near the read end; however, recall that all bases with phred scores less than 30 were masked from our variant calling.
-In this case is seems that even high quality false positives are disproportionately called near the ends of reads, while true positives are called uniformly across the read resulting in an average read position near the middle of the read.
+In this case it seems that even high quality false positives are disproportionately called near the ends of reads, while true positives are called uniformly across the read resulting in an average read position near the middle of the read.
 _potential reasons for the bias?
@@
 In the light of these quality distributions we applied a number of cut offs and drastically increased our specificity without sacrificing sensitivity (`r qual` C and D).
-For a variant to be considered in our analysis we required a mean mapping quality of 30,a mean phred score of 35, and an  average read position within the middle 50% of the read length.
+For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean phred score of 35, and an average read position within the middle 50% of the read length.
+In addition to accurately identifying the presence of expected variants DeepSNV was able to fairly accurately measure the frequency of these variants (`r freq`).
+However, while the median frequencies are near the expected values, there is a fair spread about each median, something that should be kept in mind when basing down stream analysis on frequency measurements.
+
+Influenza patient samples vary in input levels by several orders of magnitude depending on a variety of factors ranging from biological such as day post infection to methodological such as how much mucus is collected on the swab.
+To ensure accuracy across a range of input levels we diluted our known mutant mixes serially into viral media.
+We found both decreases in sensitivity and specificity accompanied lower input samples (`r lower`).
+The decreased sensitivity is expected as the likelihood of detecting rare events decreases with sample size.
+The decreased specificity is most likely due to a greater dependence on RT-PCR to amplify our signal to detectable limits.
+Is it not surprising that an amplification of noise accompanies this dependence.
+However, our results do highlight the importance of controlling for input levels when comparing populations of RNA viruses.
+
+If RT-PCR errors are responsible for the decreased specificity found at lower input levels we would expect such errors to be controlled by processing the samples in duplicate.
+To test this hypothesis, a subset of the $10^4$ genomes/$\mu$l were run in duplicate.
+The 0.2% was not run as we expected a decrease in sensitivity would accompany this approach and had little hope of finding these rare variants.
+Two RT-PCR reactions were run from a new RNA prep.
+These samples were processed separately and sequenced on the same lane of an Illumina Hiseq.
+The aforementioned cut offs were applied and for a variant call to be considered it needed to be found in both duplicates.
+This drastically improved our specificity and brought it to a comparable level as the $10^5$ samples (`r dups`).
+Unexpectedly, an increase in sensitivity accompanied this sequencing run.
+This is most likely do to a cleaner library prep with less sample loss during size selection.

### 2016-02-08 doc/paper/results.Rmd
updated results text

@@
 author: "JT McCrone"
-date: "Febuary 2, 2016"
+date: "Febuary 8, 2016"
 output: pdf_document
@@
 Freq<-"Figure 5"
-lower<-"figure 6"
-Lower<-"Figure 6"
-dups<-"figure 7"
-Dups<- "Figure 7"
+lofreq<-"Figure 6"
+Lofreq<-"figure 6"
+lower<-"figure 7"
+Lower<-"Figure 7"
+dups<-"figure 8"
+Dups<- "Figure 8"
+Table<-"Table 1"
+table<-"table 1"
+```
+#### Accuracy
 
-```
-
-There are two major components to the accuracy of a variant caller.
-The first, sensitivity, describes how well a variant caller can identify variants that are known to be real.
-In other words what proportion of the true positives are found.
+There are two major components to the accuracy of a variant caller sensitivity and specificity.
+Sensitivity, describes how well a variant caller can identify variants that are known to be real and present in the viral population.
 A sensitivity of 1 indicates that all true positives are found.
-The second, specificity, measures how well a variant caller can correctly identify true negatives.
-For example, the influenza genome is roughly 13KB.
-There are 3  potential variant nucleotides at each position leaving a variant caller with about 39,000 potential variants that can be called.
-The vast majority of these potential variants will not be present in the starting population and are potential false positives.
+Specificity, on the other hand, measures how well a variant caller can correctly identify true negatives.
+For example, the influenza genome is slightly larger than 13KB.
+There are 3  potential variant nucleotides at each position leaving a variant caller with over 40,000 potential variants that can be called.
+The vast majority of these potential variants will not be present in the starting population.
+They will only be found as the result of a sequencing or RT-PCR error.
 A variant caller that correctly identifies all potential false positives as true negatives has a specificity of 1.
-A perfect variant caller will have a sensitivity and specificity of 1.
-It will identify all true variants and only the true variants.
+A perfect variant caller will have a sensitivity and specificity of 1 meaning it can correctly identify all true variants and only identifies the true variants.
 
@@
 It may be the case that a variant caller has near perfect sensitivity, and  can find all the true positives in a sample.
-However, if in the process it has low specificity and identifies many false positives downstream analysis will be confounded.
+However, if in the process it has low specificity and identifies many false positives downstream analysis will be confounded (`r table`).
 Conversely, a variant caller may have perfect specificity and identify no false positives; however, if in process it has low sensitivity and identifies only a few true positives it will have limited applications.
@@
 An ROC plots the sensitivity of an assay along the y axis and 1-specificity along the x axis.
-To make a ROC for a variant caller one must test an assay on known data.
+To make a ROC for a variant caller one must test the assay on known data.
 The outcomes can then be stratified according to some metric that represents our belief that a given variant is real.
@@
 
-DeepSNV has been previously benchmarked using small genomic regions involving minimal PCR.
-However, RNA viral samples require an RT-PCR step, and the accuracy of this variant caller was previously unknown under these conditions.
-To characterize the accuracy of our sequencing pipeline, two lab strains, WSN33 and PR8, that differ at roughly 500 positions, were mixed  to create known viral populations (`r miseq` A).
+#### Initial accuracy
+
+There is a seemingly endless supply of variant callers on the market today, and it is beyond the scope of this work to give a detailed comparison of the accuracy of all these options.
+Instead we focus our efforts towards robustly benchmarking one variant caller and highlighting a few means for improved accuracy.
+In the process we hope to stress the importance of robustly benchmarking variant callers under the experimental conditions of a give study.
+
+All variant callers attempt to distinguish true variants from errors by estimating accurate error rates across a given genome.
+These rates can take various metrics into account such as phred score, nucleotide position on a read, and the context of a given base call.
+The question then becomes is a putative variant found more often than would be expected given the applicable error rate.
+We have chosen DeepSNV as our variant caller of choice for this analysis.
+DeepSNV uses a clonal plasmid control to estimate an error rate for each possible nucleotide at each genomic position.
+The plasmid control contains is a clonal copy of the samples that are to be analyzed and is processed and sequenced in parallel.
+Thus the plasmid control allows DeepSNV to account for any context specific errors that may appear during the sequencing process.
+DeepSNV has been previously benchmarked using a cDNA mixture of plasmids containing a 1.5Kb regions of the HIV pol gene.
+This process involved minimal PCR and was devoid of a reverse transcription step.
+Patient derived RNA viral samples require an RT-PCR step, and the accuracy of this variant caller was previously unknown under these conditions.
+To characterize the accuracy of DeepSNV in the context of our sequencing pipeline, two lab strains, WSN33 and PR8, that differ at roughly 500 positions, were mixed at different frequencies to create known viral populations (`r miseq` A).
 Briefly, genomes were amplified separately in segment specific RT-PCR reactions.
 This was done to avoid biases in amplification that may skew frequency measurements towards one strain.
-Reconstituted DNA genomes were made by mixing these products at equimolar ratios.
+Reconstituted cDNA genomes were made by mixing these products at equimolar ratios.
 The WSN33 mix was then serially diluted into PR8.
@@
 
-Because these are defined mixtures and the 491 true variants are known it is possible to measure the accuracy of our sequencing pipeline.
-Although sensitivity decreases slightly with variant frequency, we were able to detect variants at 0.5% with high accuracy.
-`r Miseq` C summarizes the accuracy of the pipeline at a p value cut off of 0.01.
-_Something about the specificity_
+Because these are defined mixtures and the 491 true variants are known it is possible to measure the accuracy of our sequencing pipeline process.
+`r Miseq` summarizes the accuracy of DeepSNV when applied to our experimental conditions.
+In the past DeepSNV was found to have a a sensitivity of 86.0% and perfect specificity when applied to variants at 0.1%.
+Under our conditions  and with a slightly more stringent p value (p=0.01 versus p=0.05) we found a small decrease in both sensitivity and specificity when applied to variants at 0.63% (`r miseq` C).
+The  decrease in specificity is not surprising as our sequencing setup requires an error prone RT-PCR step.
+The drop in  sensitivity is likely due to the increased stringency expected when correcting for the number of test required to study a 13Kb region compared to the 1.5Kb region that was tested in the past.
+(_this can be tested_) It should also be noted that although the drop in specificity is small, 1 compared to 0.9983, the change in the number of false positives is fairly large (0 compared to 66).
 
-Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under laboratory conditions.
+#### Realistic conditions
+
+Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions.
 The virus populations found in patient derived samples are often less divergent than the two strains used, and provide much less input than samples grown in cell culture.
 To mimic these conditions, 20 single point mutants made in the WSN33 background were diluted into wild type at decreasing frequencies (`r hiseq` A).
-This yielded 5 samples containing 20 known variants at frequencies ranging from 5% to 0.2% all at a concentration of $10^5$ genomes/$\mu$l.
-These samples were diluted serially into viral media to give a range of inputs matching that seen in patient samples.
-These samples were then sequenced on an Illumina Hiseq and variants called using DeepSNV.
-Focusing on the $10^5$ genomes/$\mu$l samples we find that again sensitivity is proportional to variant frequency (`r hiseq` B and C).
+This yielded 5 samples containing 20 known variants at frequencies ranging from 5% to 0.2% at a concentration of $10^5$ genomes/$\mu$l.
+These samples were diluted serially into viral media down to $10^3$ genomes/$\mu$l.
+This matches the range of inputs found in patient samples.
+The samples were then sequenced on an Illumina Hiseq and variants called using DeepSNV.
+Focusing on the $10^5$ genomes/$\mu$l samples we find that again sensitivity is correlated with variant frequency (`r hiseq` B and C).
 The fact that the sensitivity drops more dramatically than in our previous experiment is not surprising given the fact the the RNA input derived from cell culture samples is often orders of magnitude higher than that provided in this experiment.
-During library prep, a disproportionate amount of the 2.0% sample was lost and this likely accounts for the decreased sensitivity in the 2.0% sample relative to the 1.0% sample.
+During library prep, a disproportionate amount of the 2.0% sample was lost compared to the others, and this loss likely accounts for the decreased sensitivity in the 2.0% sample relative to the 1.0% sample.
 In addition to a more drastic drop in sensitivity we found a much larger number of false positives at a p value cut off of 0.01 (`r hiseq` C).
 It should be noted that in all cases the specificity was greater than 99%.
-However, given the large number of potential false positives even a specificity of 99% results in well over 200 false variant calls.
-Initially it is tempting to increase specificity by simply applying a more stringent p value cut off.
+However, given the large number of potential false positives in an influenza genome (>40,000) even a specificity of 99% results in well over 200 erroneous variant calls.
+The drop in specificity is likely due to the decease in input levels.
+In our experience samples grown in cell culture have at least $10^7$ genomes/$\mu$l.
+As the genomic input decreases, we are more and more reliant on RT-PCR to amplify the true positive "signal" to a detectable level.
+It is not surprising that is error prone method also amplifies the noise in the process.Initially it is tempting to increase specificity by simply applying a more stringent p value cut off.
 This would result in marching backwards towards the y axis along the ROCs in `r hiseq` B.
-However it is apparent from the `r hiseq` B that any increase in specificity would result in a drastic decrease in sensitivity.
-In other words, we require other means of separating true from false positives than simply the DeepSNV p value.
-_perhaps more details about the setup mimicing real world.
-such as the mRT-PCR_
+However it is apparent from  `r hiseq` B that any increase in specificity would result in a drastic decrease in sensitivity.
+It is clear an increase in specificity requires other means of separating true from false positives than the output of DeepSNV provides.
 
@@
 We did not apply any mapping quality cut offs.
-`r Qual` A shows the distribution of the mean quality scores of the variants (p <0.01) present in the $10^5$ genomes/$\mu$l samples.
+`r Qual` A shows the distribution of the mean quality scores of the variants (p <0.01) present in the 5% $10^5$ genomes/$\mu$l sample.
 In our case we find that while true and false positive variants can largely be separated by these metrics,  commonly used cut offs are not sufficient (dashed lines `r qual` A).
@@
 In this case it seems that even high quality false positives are disproportionately called near the ends of reads, while true positives are called uniformly across the read resulting in an average read position near the middle of the read.
-_potential reasons for the bias?
-pcr errors in the lib step, locations in the genome..._
+(_do I need potential reasons for the bias?
+locations in the genome...)_
 
@@
 For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean phred score of 35, and an average read position within the middle 50% of the read length.
-In addition to accurately identifying the presence of expected variants DeepSNV was able to fairly accurately measure the frequency of these variants (`r freq`).
+_These cut offs were choosen as they ..._ By applying our these quality controls we were able to greatly decrease the number of false positives from over 200 in many cases to an average of 9 with out sacrificing a single true positive.
+We recognize that by applying cut offs after variants have been called we are treating potential variant bases more stringently than those matching the consensus sequence.
+However,because we are applying cut offs to mean quality scores and there are many more consensus bases than variant bases at each position, it is unlikely these biases result in any biases in downstream analysis.
+In addition to accurately identifying the presence of expected variants DeepSNV was able to  accurately measure the frequency of these variants (`r freq`).
 However, while the median frequencies are near the expected values, there is a fair spread about each median, something that should be kept in mind when basing down stream analysis on frequency measurements.
+
+#### Lofreq
+
+In addition to DeepSNV, we analyzed our $10^5$ genomes/$\mu$l influenza populations with the variant caller Lofreq.
+Lofreq had marginally decreased sensitivity compared DeepSNV when applied to the variant frequencies $\geq$ 1.0% but marginally increased sensitivity when applied to variant frequencies $\l$ 1.0% (`r lofreq`).
+Additionally, Lofreq had specificity near that seen in our cell culture benchmarking experiment, and much better than DeepSNV prior to our quality controls.
+This is not completely surprising as the Lofreq algorithm takes MapQ and Phred scores into account when calling variants, something DeepSNV does not do.
+Additionally, Lofreq was not prone to calling false positives near the ends of reads _(verify whether or not it uses read position)_.
+_Show distributions of Lofreq variants?_ The "out of the box" accuracy of Lofreq was better than that of DeepSNV.
+However, our modified DeepSNV protocol greatly surpasses that of Lofreq, and as was the case with DeepSNV, the Lofreq ROCs (`r lofreq` A) suggests that more stringent cut offs will greatly sacrifice sensitivity for specificity.
+
+##### Buyer beware
+
+#### Lower populations
 

### 2016-02-08 doc/paper/results.Rmd
updated results text

@@
 _Show distributions of Lofreq variants?_ The "out of the box" accuracy of Lofreq was better than that of DeepSNV.
-However, our modified DeepSNV protocol greatly surpasses that of Lofreq, and as was the case with DeepSNV, the Lofreq ROCs (`r lofreq` A) suggests that more stringent cut offs will greatly sacrifice sensitivity for specificity.
+However, our modified DeepSNV protocol greatly surpasses that of Lofreq.
+`r Lofreq` A suggests that more stringent cut offs will greatly sacrifice sensitivity for specificity.
 
-##### Buyer beware
+Lofreq has been reported to have perfect specificity and a sensitivity of 96% for variants present at 0.2%.
+However, this accuracy reflects a best case scenario using simulated sequencing reads.
+As was the case with DeepSNV, the accuracy of Lofreq was much lower when applied under conditions that often face virologist using patient derived samples.
+Our data suggests that is inappropriate to apply Lofreq or DeepSNV under diverse experimental conditions and expect the performance to be unaffected.
 
-#### Lower populations
+It may be possible to further filter the putative variants called by Lofreq and achieve a similar accuracy to what was achieved using DeepSNV.
+However, because DeepSNV was much less computationally intensive, more user friendly _not sure if I can or should say that_, and capable of high accuracy, we continued to benchmark DeepSNV using the less concentrated samples from our dilution series.
 
-Influenza patient samples vary in input levels by several orders of magnitude depending on a variety of factors ranging from biological such as day post infection to methodological such as how much mucus is collected on the swab.
+_Include Lofreq in supplimentary?_
+
+#### Lower input levels
+
+Influenza patient samples vary in input concentration by several orders of magnitude depending on a variety of factors ranging from biological such as day post infection to methodological such as how much mucus is collected on the swab.
 To ensure accuracy across a range of input levels we diluted our known mutant mixes serially into viral media.
@@
 To test this hypothesis, a subset of the $10^4$ genomes/$\mu$l were run in duplicate.
-The 0.2% was not run as we expected a decrease in sensitivity would accompany this approach and had little hope of finding these rare variants.
+The 0.2% was not run as we expected a decrease in sensitivity would accompany this approach and had little hope of finding these rare variants based on the sensitivity we saw in `r lower` B.
 Two RT-PCR reactions were run from a new RNA prep.
@@
 Unexpectedly, an increase in sensitivity accompanied this sequencing run.
-This is most likely do to a cleaner library prep with less sample loss during size selection.
+This is most likely do to a superior library prep with less sample loss during size selection.

### 2016-02-08 doc/paper/results.Rmd
updated results text

@@
 
-There are two major components to the accuracy of a variant caller sensitivity and specificity.
+There are two major components that define the accuracy of a variant caller sensitivity and specificity.
 Sensitivity, describes how well a variant caller can identify variants that are known to be real and present in the viral population.
@@
 Specificity, on the other hand, measures how well a variant caller can correctly identify true negatives.
-For example, the influenza genome is slightly larger than 13KB.
+For example, the influenza genome is slightly larger than 13Kb.
 There are 3  potential variant nucleotides at each position leaving a variant caller with over 40,000 potential variants that can be called.
@@
 A variant caller that correctly identifies all potential false positives as true negatives has a specificity of 1.
-A perfect variant caller will have a sensitivity and specificity of 1 meaning it can correctly identify all true variants and only identifies the true variants.
+A perfect variant caller will have a sensitivity and specificity of 1 meaning it identifies all of the true variants an nothing else.
 
@@
 Instead we focus our efforts towards robustly benchmarking one variant caller and highlighting a few means for improved accuracy.
-In the process we hope to stress the importance of robustly benchmarking variant callers under the experimental conditions of a give study.
+In the process we hope to stress the importance of robustly benchmarking variant callers under the experimental conditions to which it is being applied.
 
@@
 DeepSNV uses a clonal plasmid control to estimate an error rate for each possible nucleotide at each genomic position.
-The plasmid control contains is a clonal copy of the samples that are to be analyzed and is processed and sequenced in parallel.
+The plasmid control contains is a clonal copy of the genome that is to be analyzed and is processed and sequenced in parallel with the test samples.
 Thus the plasmid control allows DeepSNV to account for any context specific errors that may appear during the sequencing process.
-DeepSNV has been previously benchmarked using a cDNA mixture of plasmids containing a 1.5Kb regions of the HIV pol gene.
+DeepSNV has previously been benchmarked using a cDNA mixture of plasmids containing a 1.5Kb regions of the HIV pol gene.
 This process involved minimal PCR and was devoid of a reverse transcription step.

### 2016-02-09 doc/paper/results.Rmd
updated results text

@@
 
-There are two major components that define the accuracy of a variant caller sensitivity and specificity.
+There are two major components to the accuracy of a variant caller, sensitivity and specificity.
 Sensitivity, describes how well a variant caller can identify variants that are known to be real and present in the viral population.
@@
 For example, the influenza genome is slightly larger than 13Kb.
-There are 3  potential variant nucleotides at each position leaving a variant caller with over 40,000 potential variants that can be called.
+There are 3  potential variant nucleotides at each genomic position leaving a variant caller with over 40,000 potential variants that can be called.
 The vast majority of these potential variants will not be present in the starting population.
@@
 A variant caller that correctly identifies all potential false positives as true negatives has a specificity of 1.
-A perfect variant caller will have a sensitivity and specificity of 1 meaning it identifies all of the true variants an nothing else.
+A perfect variant caller will have a sensitivity and specificity of 1 meaning it identifies all of the true variants and nothing else.
 
@@
 It may be the case that a variant caller has near perfect sensitivity, and  can find all the true positives in a sample.
-However, if in the process it has low specificity and identifies many false positives downstream analysis will be confounded (`r table`).
+However, if in the process it has low specificity and identifies many false positives, downstream analysis will be confounded (`r table`).
 Conversely, a variant caller may have perfect specificity and identify no false positives; however, if in process it has low sensitivity and identifies only a few true positives it will have limited applications.
@@
 
-There is a seemingly endless supply of variant callers on the market today, and it is beyond the scope of this work to give a detailed comparison of the accuracy of all these options.
-Instead we focus our efforts towards robustly benchmarking one variant caller and highlighting a few means for improved accuracy.
-In the process we hope to stress the importance of robustly benchmarking variant callers under the experimental conditions to which it is being applied.
+There is a seemingly endless supply of variant callers on the market today, and it is beyond the scope of this work to give a comprehensive comparison of their accuracy.
+Instead we focus our efforts towards robustly benchmarking one variant caller and highlighting a few means for improving accuracy.
+In the process we hope to stress the importance of robustly benchmarking any variant caller under the experimental conditions to which it is being applied.
 
-All variant callers attempt to distinguish true variants from errors by estimating accurate error rates across a given genome.
+All variant callers attempt to distinguish true variants from errors by estimating  error rates across a given genome.
 These rates can take various metrics into account such as phred score, nucleotide position on a read, and the context of a given base call.
@@
 Thus the plasmid control allows DeepSNV to account for any context specific errors that may appear during the sequencing process.
-DeepSNV has previously been benchmarked using a cDNA mixture of plasmids containing a 1.5Kb regions of the HIV pol gene.
+DeepSNV has previously been benchmarked using a mixture of plasmids containing a 1.5Kb regions of the HIV pol gene.
 This process involved minimal PCR and was devoid of a reverse transcription step.
 Patient derived RNA viral samples require an RT-PCR step, and the accuracy of this variant caller was previously unknown under these conditions.
-To characterize the accuracy of DeepSNV in the context of our sequencing pipeline, two lab strains, WSN33 and PR8, that differ at roughly 500 positions, were mixed at different frequencies to create known viral populations (`r miseq` A).
+To characterize the accuracy of DeepSNV in the context of our sequencing pipeline, two lab strains, WSN33 and PR8, were mixed at different frequencies to create known viral populations (`r miseq` A).
 Briefly, genomes were amplified separately in segment specific RT-PCR reactions.
@@
 
-Because these are defined mixtures and the 491 true variants are known it is possible to measure the accuracy of our sequencing pipeline process.
+Because these are defined mixtures and the 491 true variants are known it is possible to measure the accuracy of our sequencing pipeline.
 `r Miseq` summarizes the accuracy of DeepSNV when applied to our experimental conditions.
-In the past DeepSNV was found to have a a sensitivity of 86.0% and perfect specificity when applied to variants at 0.1%.
+In the past DeepSNV was found to have a sensitivity of 86.0% and perfect specificity when applied to variants at 0.1%.
 Under our conditions  and with a slightly more stringent p value (p=0.01 versus p=0.05) we found a small decrease in both sensitivity and specificity when applied to variants at 0.63% (`r miseq` C).
@@
 Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions.
-The virus populations found in patient derived samples are often less divergent than the two strains used, and provide much less input than samples grown in cell culture.
-To mimic these conditions, 20 single point mutants made in the WSN33 background were diluted into wild type at decreasing frequencies (`r hiseq` A).
+The virus populations found in patient derived samples are expected to be less divergent than the two strains used, and provide much less input than samples grown in cell culture.
+To mimic these conditions, 20 single point mutants were made in the WSN33 background and were diluted into wild type at decreasing frequencies (`r hiseq` A).
 This yielded 5 samples containing 20 known variants at frequencies ranging from 5% to 0.2% at a concentration of $10^5$ genomes/$\mu$l.
@@
 As the genomic input decreases, we are more and more reliant on RT-PCR to amplify the true positive "signal" to a detectable level.
-It is not surprising that is error prone method also amplifies the noise in the process.Initially it is tempting to increase specificity by simply applying a more stringent p value cut off.
+It is not surprising that is error prone method also amplifies the noise in the process.
+Initially it is tempting to increase specificity by simply applying a more stringent p value cut off.
 This would result in marching backwards towards the y axis along the ROCs in `r hiseq` B.
-However it is apparent from  `r hiseq` B that any increase in specificity would result in a drastic decrease in sensitivity.
-It is clear an increase in specificity requires other means of separating true from false positives than the output of DeepSNV provides.
+However, it is apparent from  `r hiseq` B that any increase in specificity would result in a drastic decrease in sensitivity.
+It is clear from `r hiseq` that under experimental conditions the DeepSNV output alone is insufficient to accurately separate true positives from false positives.
 
@@
 Additionally, we found that in general false positives were biased near the ends of sequencing reads (`r qual` B).
-This is not surprising as it is well known that sequencing quality dips near the read end; however, recall that all bases with phred scores less than 30 were masked from our variant calling.
+This is not surprising as it is well known that sequencing quality dips near the end of a read; however, recall that all bases with phred scores less than 30 were masked from our variant calling.
 In this case it seems that even high quality false positives are disproportionately called near the ends of reads, while true positives are called uniformly across the read resulting in an average read position near the middle of the read.
@@
 For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean phred score of 35, and an average read position within the middle 50% of the read length.
-_These cut offs were choosen as they ..._ By applying our these quality controls we were able to greatly decrease the number of false positives from over 200 in many cases to an average of 9 with out sacrificing a single true positive.
+_These cut offs were choosen as they ..._ By applying these quality controls we were able to greatly decrease the number of false positives from over 200 in many cases to an average of 9 without sacrificing a single true positive.
 We recognize that by applying cut offs after variants have been called we are treating potential variant bases more stringently than those matching the consensus sequence.
-However,because we are applying cut offs to mean quality scores and there are many more consensus bases than variant bases at each position, it is unlikely these biases result in any biases in downstream analysis.
+However,because we are applying cut offs to mean quality scores and there are many more consensus bases than variant bases at each position, it is unlikely these biases have any noticable affect on our analysis.
 In addition to accurately identifying the presence of expected variants DeepSNV was able to  accurately measure the frequency of these variants (`r freq`).
@@
 In addition to DeepSNV, we analyzed our $10^5$ genomes/$\mu$l influenza populations with the variant caller Lofreq.
+This was done to ensure that the decreased accuracy that accompanied our realistic experimental conditions was not simply based on the DeepSNV variant calling method.
 Lofreq had marginally decreased sensitivity compared DeepSNV when applied to the variant frequencies $\geq$ 1.0% but marginally increased sensitivity when applied to variant frequencies $\l$ 1.0% (`r lofreq`).
@@
 This is not completely surprising as the Lofreq algorithm takes MapQ and Phred scores into account when calling variants, something DeepSNV does not do.
-Additionally, Lofreq was not prone to calling false positives near the ends of reads _(verify whether or not it uses read position)_.
-_Show distributions of Lofreq variants?_ The "out of the box" accuracy of Lofreq was better than that of DeepSNV.
-However, our modified DeepSNV protocol greatly surpasses that of Lofreq.
-`r Lofreq` A suggests that more stringent cut offs will greatly sacrifice sensitivity for specificity.
+Additionally, Lofreq was not prone to calling false positives near the ends of reads, _(verify whether or not it uses read position Show distributions of Lofreq variants?)_, but   `r lofreq` A suggests that sacrificing sensitivity is the only means of increasing specificity.
 
@@
 
-Influenza patient samples vary in input concentration by several orders of magnitude depending on a variety of factors ranging from biological such as day post infection to methodological such as how much mucus is collected on the swab.
+Influenza patient samples vary in input concentration by several orders of magnitude and depend on a variety of factors often outside the control of a researcher (day of collection, volume of mucus collected ect.).
 To ensure accuracy across a range of input levels we diluted our known mutant mixes serially into viral media.
-We found both decreases in sensitivity and specificity accompanied lower input samples (`r lower`).
-The decreased sensitivity is expected as the likelihood of detecting rare events decreases with sample size.
-The decreased specificity is most likely due to a greater dependence on RT-PCR to amplify our signal to detectable limits.
+We found both decreases in sensitivity and specificity accompanied lower input concentrations (`r lower`).
+The decreased sensitivity is expected as the likelihood of detecting rare events decreases with sample sizem while the decreased specificity is most likely due to a greater dependence on RT-PCR to amplify the signal to detectable limits.
 Is it not surprising that an amplification of noise accompanies this dependence.
@@
 
-If RT-PCR errors are responsible for the decreased specificity found at lower input levels we would expect such errors to be controlled by processing the samples in duplicate.
+If RT-PCR errors are responsible for the decreased specificity found at lower input levels, and RT_PCR are randomly distributed across the genome, we would expect such errors to be eliminated by processing the samples in duplicate.
 To test this hypothesis, a subset of the $10^4$ genomes/$\mu$l were run in duplicate.
@@
 These samples were processed separately and sequenced on the same lane of an Illumina Hiseq.
-The aforementioned cut offs were applied and for a variant call to be considered it needed to be found in both duplicates.
-This drastically improved our specificity and brought it to a comparable level as the $10^5$ samples (`r dups`).
+The previously mentioned quality cut offs were applied and for a variant call to be considered it was required to be found in both duplicates.
+This drastically improved our specificity and resulted an accuracy that was comparable to that found in the $10^5$ samples (`r dups`).
 Unexpectedly, an increase in sensitivity accompanied this sequencing run.

### 2016-02-09 scripts/figures.Rmd
updated draft of figs and results

@@
 ### Figure 1
-```{r, roc_skematic,fig.height=5,fig.width=3.4}
+```{r, roc_skematic,fig.height=4,fig.width=3.4}
 set.seed(42) # make the figures reproducible
@@
  #plot the points and cut offs
- points.p<-ggplot(distribution,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center",dotsize = 1.5)+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks = element_blank(), axis.text.x = element_blank(),axis.title.y = element_text(vjust=1.2),legend.position=c(0.8,0.6))+xlab("")
+ points.p<-ggplot(distribution,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center",dotsize = 1.5)+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks = element_blank(), axis.text.x = element_blank(),axis.title.y = element_text(vjust=1.2),legend.position=c(0.9,0.65),legend.background = element_rect(fill = alpha("white",0)))+xlab("")
 
  points.p<-points.p+geom_hline(yintercept=output$threshold,
-                          size=1,lty=2,color=c("#e41a1c","#377eb8")) # these are the first 2 colors in set 1
+                          size=1,lty=2,color=c("#08519cff","#787878ff")) # these are the first 2 colors in set 1
 
- plot_grid(points.p, roc.p, labels=c("A", "B"), ncol = 1, nrow = 2)
+ggdraw() +
+  draw_plot(points.p, 0, .45, .8, 0.48) + #x y width height
+  draw_plot(roc.p, 0, 0, .8, 0.48) +
+  draw_plot_label(c("A", "B"), c(0.05, 0.05), c(1, 0.55), size = 15)
 
@@
 plot.ex(0.35,0.95,roc.df,example)
+
+pdf("../results/figures/roc.example.pdf",width=3.4,height=4)
+plot.ex(0.35,0.95,roc.df,example)
+dev.off()
 # pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.ex.pdf",width = 12,height = 8)
@@
 ### tableGrob functions and settings #####
-tt1<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.75,fontfamily="serif")),core=list(fg_params=list(cex=0.75,fontfamily="serif")))
+tt1<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.75)),core=list(fg_params=list(cex=0.75)))#,fontfamily="serif")))
 make.table<-function(df,theme){
@@
 ### Figure 2
-```{r,Miseq_out_of_the_box,fig.width=7,fig.height=5}
+```{r,Miseq_out_of_the_box,fig.width=3.4,fig.height=7}
 
@@
 
-miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.5%","1.25%","0.63%","0.016%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position=c(1,.5))
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.5%","1.25%","0.63%","0.016%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
 
@@
 #knitr::kable(m.roc.table)
-m.roc.table<-rename(m.roc.table,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))
+m.roc.table<-rename(m.roc.table,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True\nPositives","adj.specificity"="Specificity","FP"="False\nPositives"))
 
@@
 
-pdf("../results/figures/miseq1.pdf",width=7,height=5)
+pdf("../results/figures/miseq1.pdf",width=3.4,height=7)
 ggdraw() +
-  draw_plot(miseq.setup, 0.01, .1, 0.4, 0.8) + #x y width height
-  draw_plot(miseq.roc.p, 0.43, 0.42, 0.5, 0.51) +
-  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
-  draw_plot_label(c("A", "B", "C"), c(0.01, 0.42, 0.42), c(0.78, 0.95, 0.46), size = 15)
+  draw_plot(miseq.setup, 0.2, .66, .6, 0.33) + #x y width height
+  draw_plot(miseq.roc.p, 0.05, 0.33, .9, 0.33) +
+  draw_plot(tbl, 0.05, 0, .9, 0.33) +
+  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(0.95, 0.7, 0.35), size = 15)
 dev.off()
@@
 ggdraw() +
-  draw_plot(miseq.setup, 0.01, .1, 0.4, 0.8) + #x y width height
-  draw_plot(miseq.roc.p, 0.43, 0.42, 0.5, 0.51) +
-  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
-  draw_plot_label(c("A", "B", "C"), c(0.01, 0.42, 0.42), c(0.78, 0.95, 0.46), size = 15)
+  draw_plot(miseq.setup, 0.2, .66, .6, 0.33) + #x y width height
+  draw_plot(miseq.roc.p, 0.05, 0.33, .9, 0.33) +
+  draw_plot(tbl, 0.05, 0, .9, 0.33) +
+  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(0.95, 0.7, 0.35), size = 15)
 
@@
 
-```{r,hiseq_roc,fig.width=7,fig.height=5}
+```{r,hiseq_roc,fig.width=3.4,fig.height=7}
 #rm(list = setdiff(ls(), lsf.str())) # remove all variables except the functions
-tt1<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.75,fontfamily="serif")),core=list(fg_params=list(cex=0.75,fontfamily="serif")))
+#tt1<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.75,fontfamily="serif")),core=list(fg_params=list(cex=0.75,fontfamily="serif")))
 
@@
 
-h.table.df<-rename(h.table.df,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))
+h.table.df<-rename(h.table.df,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True\nPositives","adj.specificity"="Specificity","FP"="False\nPositives"))
 
@@
 
-pdf("../results/figures/hiseq1.pdf",width=7,height=5)
+pdf("../results/figures/hiseq1.pdf",width=3.4,height=7)
 ggdraw() +
-  draw_plot(hiseq.setup, 0, .1, 0.4, 0.8) + #x y width height
-  draw_plot(Hiseq.roc, 0.43, 0.43, 0.5, 0.5) +
-  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
-  draw_plot_label(c("A", "B", "C"), c(0.05, 0.42, 0.42), c(0.95, 0.95, 0.46), size = 15)
+  draw_plot(hiseq.setup, 0.1, .57, 0.8, 0.43) + #x y width height
+  draw_plot(Hiseq.roc, 0.05, 0.33, .8, 0.33) +
+  draw_plot(tbl, 0.05, 0, .9, 0.33) +
+  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(1, 0.7, 0.35), size = 15)
 dev.off()
@@
 ggdraw() +
-  draw_plot(hiseq.setup, 0, .1, 0.4, 0.8) + #x y width height
-  draw_plot(Hiseq.roc, 0.43, 0.43, 0.5, 0.5) +
-  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
-  draw_plot_label(c("A", "B", "C"), c(0.05, 0.42, 0.42), c(0.95, 0.95, 0.46), size = 15)
-
+  draw_plot(hiseq.setup, 0.1, .57, 0.8, 0.43) + #x y width height
+  draw_plot(Hiseq.roc, 0.05, 0.33, .8, 0.33) +
+  draw_plot(tbl, 0.05, 0, .9, 0.33) +
+  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(1, 0.7, 0.35), size = 15)
 # 
@@
 
-mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(size=1)+geom_hline(yintercept=20,linetype=2)+geom_vline(xintercept=30,linetype=2)+scale_x_continuous(limits=c(30,40))+scale_color_hue(name="",labels=c("False posivite","True positive"))+ylab("Mean mapping quality")+xlab("Mean base qualtiy (Phred) ")+theme(legend.position="top")
+mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(size=1)+geom_hline(yintercept=20,linetype=2)+geom_vline(xintercept=30,linetype=2)+scale_x_continuous(limits=c(30,40))+scale_color_manual(name="",labels=c("False posivite","True positive"),values=c("#787878ff","#08519cff"))+ylab("Mean mapping quality")+xlab("Mean base qualtiy (Phred) ")+theme(legend.position="top")
 
@@
 
-```{r,read_pos,fig.width=7,fig.height=6}
+```{r,read_pos,fig.width=3.4,fig.height=9}
 # ggplot(sig.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
@@
 
-read.pos<-ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')+scale_fill_hue(name="",labels=c("False positive","True positive"))+xlab("Average position on read")+ylab("Number of variants")+theme(legend.position="top")
+read.pos<-ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')+scale_fill_manual(name="",labels=c("False positive","True positive"),values=c("#787878ff","#08519cff"))+xlab("Average position on read")+ylab("Number of variants")+theme(legend.position="top")
 # read.pos
@@
 h.cut.read.table.5<-subset(h.cut.read.table,gc==5,select=-c(gc))
-rename(h.cut.read.table.5,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))->h.cut.read.table.5
+rename(h.cut.read.table.5,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True\nPositives","adj.specificity"="Specificity","FP"="False\nPositives"))->h.cut.read.table.5
 h.cut.read.table.5$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
@@
 
+h.cut.read.table.5$Specificity[h.cut.read.table.5$Specificity>0.9999]<-">0.9999"
+
 tbl<-make.table(h.cut.read.table.5,tt1) # Make table
 
-pdf("../results/figures/figure4.pdf",width=7,height=6)
+pdf("../results/figures/figure4.pdf",width=3.4,height=9)
 ggdraw() +
-  draw_plot(mapq_phred.normal, 0, .5, .5, .48) + 
-  draw_plot(read.pos, .5, .5, .5, .48) +
-  draw_plot(read.roc, 0, 0, 0.42, .5) +
-  draw_plot(tbl, 0.47, 0, 0.52, 0.47) +
-  draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(0.95, 0.95, 0.55,0.45), size = 15)
+  draw_plot(mapq_phred.normal,0.03, 0.77, .85, 0.23) + 
+  draw_plot(read.pos, 0.03, .53, .85, .25) +
+  draw_plot(read.roc, 0.03, 0.23, 0.85, .3) +
+  draw_plot(tbl,   0.025, 0, 0.95, .25) +
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(1, 0.78, 0.55,0.25), size = 15)
 dev.off()
@@
 ggdraw() +
-  draw_plot(mapq_phred.normal, 0, .5, .5, .48) + 
-  draw_plot(read.pos, .5, .5, .5, .48) +
-  draw_plot(read.roc, 0, 0, 0.42, .5) +
-  draw_plot(tbl, 0.47, 0, 0.52, 0.47) +
-  draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(0.95, 0.95, 0.55,0.45), size = 15)
+  draw_plot(mapq_phred.normal,0.03, 0.77, .85, 0.23) + 
+  draw_plot(read.pos, 0.03, .53, .85, .25) +
+  draw_plot(read.roc, 0.03, 0.23, 0.85, .3) +
+  draw_plot(tbl,   0.025, 0, 0.95, .25) +
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(1, 0.78, 0.55,0.25), size = 15)
 ```
@@
 lofreq.table<-head(subset(lofreq.table,gc==5,select=-c(gc,Id)),n=-1) # don't need the 0.1% sample
-rename(lofreq.table,c("exp.freq"="Frequency","TP"="True Positives","FP"="False Positives"))->lofreq.table
+rename(lofreq.table,c("exp.freq"="Frequency","TP"="True\nPositives","FP"="False\nPositives"))->lofreq.table
 lofreq.table$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
@@
 lofreq.table$Specificity<-round(lofreq.table$Specificity,digits=dd)
-
+lofreq.table<-subset(lofreq.table,select=c("Frequency","Sensitivity","True\nPositives","Specificity","False\nPositives"))
 tt2<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.65,fontfamily="serif")),core=list(fg_params=list(cex=0.65,fontfamily="serif")))
 
-tbl<-make.table(lofreq.table,tt2) # Make table
+tbl<-make.table(lofreq.table,tt1) # Make table
 
@@
 
-ac_freq<-ac_freq+scale_y_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.05,0.1),label=c("0.1%","0.2%","0.5%","1.0%","2.0%","5.0%","10%"))#+theme(axis.text.y = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
-ac_freq<-ac_freq+scale_x_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.050,0.1),label=c("0.1%","0.2%","0.5%","1.0%","2.0%","5.0%","10%"))#+theme(axis.text.x = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
+ac_freq<-ac_freq+scale_y_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.05,0.1))#+theme(axis.text.y = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
+ac_freq<-ac_freq+scale_x_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.050,0.1))#+theme(axis.text.x = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
 ac_freq<-ac_freq+geom_abline(intercept = 0, slope = 1,linetype=2,size=1)
-ac_freq<-ac_freq+xlab("Expected Frequency")+ylab("Measured Frequency")+theme(legend.position="none")
+ac_freq<-ac_freq+xlab("Expected Frequency")+ylab("Measured Frequency")+theme(legend.position="none",axis.text.x = element_text(angle = 45, hjust = 1))
 
@@
 
-```{r,roc_lower_inputs4,fig.width=7,fig.height=5}
+```{r,roc_lower_inputs4,fig.width=3.4,fig.height=9}
 
@@
 h.cut.read.table.4<-subset(h.cut.read.table,gc==4,select=-c(gc))
-rename(h.cut.read.table.4,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))->h.cut.read.table.4
+rename(h.cut.read.table.4,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True\nPositives","adj.specificity"="Specificity","FP"="False\nPositives"))->h.cut.read.table.4
 h.cut.read.table.4$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
@@
 h.cut.read.table.3<-subset(h.cut.read.table,gc==3,select=-c(gc))
-rename(h.cut.read.table.3,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))->h.cut.read.table.3
+rename(h.cut.read.table.3,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True\nPositives","adj.specificity"="Specificity","FP"="False\nPositives"))->h.cut.read.table.3
 h.cut.read.table.3$Frequency<-c("5.0%","2.0%","1.0%")
@@
 
-pdf("../results/figures/lower.pdf",width=7,height=5)
+pdf("../results/figures/lower.pdf",width=3.4,height=9)
 ggdraw() +
-  draw_plot(roc.4, 0, .48, .43, .5) +
-  draw_plot(tbl.4, 0.47, 0.55, 0.52, 0.47) +
-  draw_plot(roc.3, 0, 0, 0.43, .5) +
-  draw_plot(tbl.3, 0.47, 0, 0.52, 0.47)+
-  draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(1, 1, 0.5,0.5), size = 15)
+  draw_plot(roc.4,0.03, 0.69, .85, 0.25) + 
+  draw_plot(tbl.4, 0.025, .47,  0.95, .25) +
+  draw_plot(roc.3, 0.03, 0.22, 0.85, .25) +
+  draw_plot(tbl.3,   0.025, 0, 0.95, .25) +
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(0.97, 0.73, 0.5,0.23), size = 15)
 dev.off()
@@
 ggdraw() +
-  draw_plot(roc.4, 0, .48, .43, .5) +
-  draw_plot(tbl.4, 0.47, 0.55, 0.52, 0.47) +
-  draw_plot(roc.3, 0, 0, 0.43, .5) +
-  draw_plot(tbl.3, 0.47, 0, 0.52, 0.47)+
-  draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(1, 1, 0.5,0.5), size = 15)
-
+  draw_plot(roc.4,0.03, 0.69, .85, 0.25) + 
+  draw_plot(tbl.4, 0.025, .47,  0.95, .25) +
+  draw_plot(roc.3, 0.03, 0.22, 0.85, .25) +
+  draw_plot(tbl.3,   0.025, 0, 0.95, .25) +
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(0.97, 0.73, 0.5,0.23), size = 15)
 ```

### 2016-02-09 doc/paper/results.Rmd
updated draft of figs and results
drafts given to Adam at weekly meeting

@@
 author: "JT McCrone"
-date: "Febuary 8, 2016"
-output: pdf_document
+date: "Febuary 9, 2016"
+output: word_document
 ---
@@
 We did not apply any mapping quality cut offs.
-`r Qual` A shows the distribution of the mean quality scores of the variants (p <0.01) present in the 5% $10^5$ genomes/$\mu$l sample.
+`r Qual` A shows the distribution of the mean quality scores of the variants (p <0.01) present in the $10^5$ genomes/$\mu$l samples.
 In our case we find that while true and false positive variants can largely be separated by these metrics,  commonly used cut offs are not sufficient (dashed lines `r qual` A).

### 2016-02-16 doc/paper/results.Rmd
starting 2nd draft

@@
 author: "JT McCrone"
-date: "Febuary 9, 2016"
+date: "Febuary 15, 2016"
 output: word_document
@@
 ```
-#### Accuracy
 
-There are two major components to the accuracy of a variant caller, sensitivity and specificity.
-Sensitivity, describes how well a variant caller can identify variants that are known to be real and present in the viral population.
-A sensitivity of 1 indicates that all true positives are found.
-Specificity, on the other hand, measures how well a variant caller can correctly identify true negatives.
-For example, the influenza genome is slightly larger than 13Kb.
-There are 3  potential variant nucleotides at each genomic position leaving a variant caller with over 40,000 potential variants that can be called.
-The vast majority of these potential variants will not be present in the starting population.
-They will only be found as the result of a sequencing or RT-PCR error.
-A variant caller that correctly identifies all potential false positives as true negatives has a specificity of 1.
-A perfect variant caller will have a sensitivity and specificity of 1 meaning it identifies all of the true variants and nothing else.
+Accurately identifying rare single nucleotide variants depends on two components, sensitivity and specificity.
+Sensitivity, also known as the true positive rate, is the proportion of true variants that are found.
+Specificity, on the other hand, is the proportion of true negatives that are correctly identified.
+$1-\text{specificity}$ is the false positive rate, the ratio of false positive variants to true negatives.
+Both a sensitivity and specificity of 1 describe perfect accurarcy in which all the true variants are found and nothing else.
 
 There is an obvious trade off between sensitivity and specificity.
-It may be the case that a variant caller has near perfect sensitivity, and  can find all the true positives in a sample.
-However, if in the process it has low specificity and identifies many false positives, downstream analysis will be confounded (`r table`).
-Conversely, a variant caller may have perfect specificity and identify no false positives; however, if in process it has low sensitivity and identifies only a few true positives it will have limited applications.
-
-The trade off between sensitivity and specificity can be represented as a receiver operator curve or ROC (`r xamp_roc`).
+Improved sensitivity often requires less strigent criteria in variant calling which reduces specificity.
+Conversly, increased stringency can improve specificity but reduce sensitivity.
+This relationship can be represented as a receiver operating characteristic (ROC) curve (`r xamp_roc`).
 An ROC plots the sensitivity of an assay along the y axis and 1-specificity along the x axis.
-To make a ROC for a variant caller one must test the assay on known data.
-The outcomes can then be stratified according to some metric that represents our belief that a given variant is real.
-Often this is a p value or quality score.
-Because the true variants are known, the sensitivity and specificity can be calculated at different cut offs (`r xamp_roc` A).
+To make an ROC curve for a variant callling algorithm one must test the assay on known data.
+The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real; often this is a p value or quality score.
+In a controled benchmakring experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut offs (`r xamp_roc` A).
 These points are then used to construct the curve (`r xamp_roc` B).
-A perfect ROC in which all the true positives can be separated from all potential false positives is a right angle that follows the upper left perimeter of the plot and passes through the point (0,1).
+A perfect ROC curve in which all the true positives can be separated from all false positives is a right angle that follows the upper left perimeter of the plot.
 
@@
 
-There is a seemingly endless supply of variant callers on the market today, and it is beyond the scope of this work to give a comprehensive comparison of their accuracy.
-Instead we focus our efforts towards robustly benchmarking one variant caller and highlighting a few means for improving accuracy.
-In the process we hope to stress the importance of robustly benchmarking any variant caller under the experimental conditions to which it is being applied.
+It is beyond the scope of this work to comprehensively compare all variant calling methods virologists use.
+I Instead we focus our efforts towards robustly benchmarking one variant caller and highlighting a few means for improving accuracy.
+In the process we demonstrate the importance of robustly benchmarking any variant calling pipeline under the experimental conditions to which it is being applied.
 
-All variant callers attempt to distinguish true variants from errors by estimating  error rates across a given genome.
-These rates can take various metrics into account such as phred score, nucleotide position on a read, and the context of a given base call.
-The question then becomes is a putative variant found more often than would be expected given the applicable error rate.
  We have chosen DeepSNV as our variant caller of choice for this analysis.
-DeepSNV uses a clonal plasmid control to estimate an error rate for each possible nucleotide at each genomic position.
-The plasmid control contains is a clonal copy of the genome that is to be analyzed and is processed and sequenced in parallel with the test samples.
-Thus the plasmid control allows DeepSNV to account for any context specific errors that may appear during the sequencing process.
-DeepSNV has previously been benchmarked using a mixture of plasmids containing a 1.5Kb regions of the HIV pol gene.
-This process involved minimal PCR and was devoid of a reverse transcription step.
-Patient derived RNA viral samples require an RT-PCR step, and the accuracy of this variant caller was previously unknown under these conditions.
-To characterize the accuracy of DeepSNV in the context of our sequencing pipeline, two lab strains, WSN33 and PR8, were mixed at different frequencies to create known viral populations (`r miseq` A).
-Briefly, genomes were amplified separately in segment specific RT-PCR reactions.
-This was done to avoid biases in amplification that may skew frequency measurements towards one strain.
-Reconstituted cDNA genomes were made by mixing these products at equimolar ratios.
-The WSN33 mix was then serially diluted into PR8.
-These mixtures were sequenced using 2x250 reads on an Illumina Miseq and variants were called using DeepSNV.
+DeepSNV uses a clonal, plasmid-derived, control to estimate local error rates controlling for sequence context accross the genome.
+Because the control is processed together with the experimental samples it has identical noise characteristics.
+DeepSNV applies a hierarchiacal binomial model at each genomic position and identifies true positive variants with frequencies greater than expected based on the control.
+Like many variant callers the accuracy of DeepSNV was initially determined using samples that required mimal PCR amplication.
+To characterize the accuracy of DeepSNV when applied to RT-amplified RNA virus we creadted difed mixutres of two winfluenza strains, WSN33 and PR8 (`r miseq` A).
 
-Because these are defined mixtures and the 491 true variants are known it is possible to measure the accuracy of our sequencing pipeline.
-`r Miseq` summarizes the accuracy of DeepSNV when applied to our experimental conditions.
-In the past DeepSNV was found to have a sensitivity of 86.0% and perfect specificity when applied to variants at 0.1%.
-Under our conditions  and with a slightly more stringent p value (p=0.01 versus p=0.05) we found a small decrease in both sensitivity and specificity when applied to variants at 0.63% (`r miseq` C).
-The  decrease in specificity is not surprising as our sequencing setup requires an error prone RT-PCR step.
-The drop in  sensitivity is likely due to the increased stringency expected when correcting for the number of test required to study a 13Kb region compared to the 1.5Kb region that was tested in the past.
-(_this can be tested_) It should also be noted that although the drop in specificity is small, 1 compared to 0.9983, the change in the number of false positives is fairly large (0 compared to 66).
+Because these viruses differ at 491 positions, there are 491 true positives at each dilution.
+On plasmids subserted to limited amplification, DeepSNV ad a sensitivity 0.860 and a specificity of 1 for SNV at 0.1% frequency.
+Under our realistic expermental conditions and with a slightly more stringent p value cutoff ( p=0.01 vs.
+p=0.05) we found a moderate reducttion in sensitivity  (0.830 for variants at 0.63% and 0.14 for variants at 0.16%).
+Our specificity was above 0.998 in all dilution.
+While this reduction in specificity appears minor, a false positive rate of 0.002 applied to the xx,xxx potential variants in the 13.xkb influenza genome results in 78 false positives.
 
-#### Realistic conditions
+#### An experimental intrahost population
 
 Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions.
-The virus populations found in patient derived samples are expected to be less divergent than the two strains used, and provide much less input than samples grown in cell culture.
-To mimic these conditions, 20 single point mutants were made in the WSN33 background and were diluted into wild type at decreasing frequencies (`r hiseq` A).
-This yielded 5 samples containing 20 known variants at frequencies ranging from 5% to 0.2% at a concentration of $10^5$ genomes/$\mu$l.
-These samples were diluted serially into viral media down to $10^3$ genomes/$\mu$l.
-This matches the range of inputs found in patient samples.
-The samples were then sequenced on an Illumina Hiseq and variants called using DeepSNV.
-Focusing on the $10^5$ genomes/$\mu$l samples we find that again sensitivity is correlated with variant frequency (`r hiseq` B and C).
-The fact that the sensitivity drops more dramatically than in our previous experiment is not surprising given the fact the the RNA input derived from cell culture samples is often orders of magnitude higher than that provided in this experiment.
-During library prep, a disproportionate amount of the 2.0% sample was lost compared to the others, and this loss likely accounts for the decreased sensitivity in the 2.0% sample relative to the 1.0% sample.
+The virus populations found in patient-derived samples are expected to be less divergent than the two lab strains.
+Also the nucleic acid concentration in patient samples is much lower than that found samples grown in cell culture.
+To mimic these conditions, we generated 20 viral colones, each with a single point mutation in the WSN33 background.
+We sequenced these mutants on the Illumina Miseq platform and cofirmed that no additional mutations >1% frequency had arisen between transfection and the passage 1 stock.
+We also quantified the genome copy number of each stock using quantitative PCR.
+We then mixed equal genome equivalents of these 20 virsuses, generating an artificial population of $10^5$ copies per microliter in which each mutation was present at 5% frequency.
+This population was serially diluted into a stock of WSN33, generating artificial populations witha ll 20 mutations present a 2,1,0.5,0.2, and 0.1% frequency (`r hiseq` A).
+We then serially diluted each of these $10^5$ copy number populations into basal media to obtain populations with lower input.
+The range ($10^3-10^5$ copies per microliter) matches the inpus typically found in pateint-derived samples (data not shown).
+We sequenced the samples on the Illumina Hiseq platform, and variants were called using DeepSNV.
+We also processed and sequenced the WSN33 stock in duplicate.
+
+We considered the 20 mutations present in our initial viral mixture as the only true positives in our analysis.
+Any variant that was present in the wild type stock was masked and neither considered a true positive nor a true negative.
+This was done to avoid benchmarking DeepSNV with variants identified only be DeepSNV.
+
+The samples with the highest input concentration had the best sensitivity (`r hiseq` B and C).
+Sensitivity dropped in the 0.5% frequency samples.
+In general, our sensitivity was lower than that seen previously in the cell culture samples where the input concentration was at least $10^7$ copies per microliter.
+The drop in sensitivity is likely due to the 100 fold decrease in variant genomes present in the $10^5$ samples.
+During library prep, a disproportionate amount of the 2.0% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0% sample.
+
 In addition to a more drastic drop in sensitivity we found a much larger number of false positives at a p value cut off of 0.01 (`r hiseq` C).

### 2016-02-23 doc/paper/results.Rmd

@@
 ---
-title: "Results"
-author: "JT McCrone"
-date: "Febuary 15, 2016"
-output: word_document
+output: pdf_document
 ---
-```{r.message=F,echo=F}
+
+```{r,message=F,echo=FALSE}
 xamp_roc<-"figure 1"
@@
 table<-"table 1"
+
+potential_var<-"XX"
 ```
 
-Accurately identifying rare single nucleotide variants depends on two components, sensitivity and specificity.
-Sensitivity, also known as the true positive rate, is the proportion of true variants that are found.
-Specificity, on the other hand, is the proportion of true negatives that are correctly identified.
+The ability to reliably distinguish true from false single nucleotide variants (SNV) in next generation sequencing data is integral to accurate studies of viral diversity.
+Sensitivity is the proportion of true variants identified; while, specificity is the proportion of true negatives identified.
 $1-\text{specificity}$ is the false positive rate, the ratio of false positive variants to true negatives.
-Both a sensitivity and specificity of 1 describe perfect accurarcy in which all the true variants are found and nothing else.
+Both a sensitivity and specificity of 1 describe perfect accuracy in which all the true variants are found and nothing else.
 
 There is an obvious trade off between sensitivity and specificity.
-Improved sensitivity often requires less strigent criteria in variant calling which reduces specificity.
-Conversly, increased stringency can improve specificity but reduce sensitivity.
+Improved sensitivity often requires less stringent criteria in variant calling which reduces specificity.
+Conversely, increased stringency can improve specificity but reduce sensitivity.
 This relationship can be represented as a receiver operating characteristic (ROC) curve (`r xamp_roc`).
 An ROC plots the sensitivity of an assay along the y axis and 1-specificity along the x axis.
-To make an ROC curve for a variant callling algorithm one must test the assay on known data.
+To make an ROC curve for a variant calling algorithm one must test the assay on known data.
 The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real; often this is a p value or quality score.
-In a controled benchmakring experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut offs (`r xamp_roc` A).
+In a controlled benchmakring experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut offs (`r xamp_roc` A).
 These points are then used to construct the curve (`r xamp_roc` B).
@@
 
-It is beyond the scope of this work to comprehensively compare all variant calling methods virologists use.
-I Instead we focus our efforts towards robustly benchmarking one variant caller and highlighting a few means for improving accuracy.
+It is beyond the scope of this work to comprehensively compare all variant calling methods employed by virologists.
+Instead we focus our efforts towards robustly benchmarking one variant caller and highlighting a few means for improving accuracy.
 In the process we demonstrate the importance of robustly benchmarking any variant calling pipeline under the experimental conditions to which it is being applied.
@@
  We have chosen DeepSNV as our variant caller of choice for this analysis.
-DeepSNV uses a clonal, plasmid-derived, control to estimate local error rates controlling for sequence context accross the genome.
-Because the control is processed together with the experimental samples it has identical noise characteristics.
-DeepSNV applies a hierarchiacal binomial model at each genomic position and identifies true positive variants with frequencies greater than expected based on the control.
-Like many variant callers the accuracy of DeepSNV was initially determined using samples that required mimal PCR amplication.
-To characterize the accuracy of DeepSNV when applied to RT-amplified RNA virus we creadted difed mixutres of two winfluenza strains, WSN33 and PR8 (`r miseq` A).
+DeepSNV uses a clonal, plasmid-derived, control to estimate local error rates across the genome.
+Because it is clonal, the sequence of the control is known with a high degree of confidence and so any nonconsensus base is considered noise.
+The control is processed together with experimental samples, and so it has identical noise characteristics.
+DeepSNV applies a hierarchical binomial model at each genomic position and identifies true variants as those with frequencies greater than expected based on the noise found in the plasmid control.
+Like many variant callers the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification.
+We created defined mixtures of two influenza strains, WSN33 and PR8 to characterize the accuracy of DeepSNV when applied to RT-amplified RNA virus populations (`r Miseq` A).
 
-Because these viruses differ at 491 positions, there are 491 true positives at each dilution.
-On plasmids subserted to limited amplification, DeepSNV ad a sensitivity 0.860 and a specificity of 1 for SNV at 0.1% frequency.
-Under our realistic expermental conditions and with a slightly more stringent p value cutoff ( p=0.01 vs.
-p=0.05) we found a moderate reducttion in sensitivity  (0.830 for variants at 0.63% and 0.14 for variants at 0.16%).
-Our specificity was above 0.998 in all dilution.
-While this reduction in specificity appears minor, a false positive rate of 0.002 applied to the xx,xxx potential variants in the 13.xkb influenza genome results in 78 false positives.
+WSN33 and PR8 genomes were mixed serially such that WSN33 was present at relative frequencies of 5, 2.5, 1.25, 0.63 and 0.16%.
+Because these viruses differ at 491 positions,there are 491 true positives in each dilution.
+On plasmids subjected to limited amplification, DeepSNV has a sensitivity of 0.860 and perfect specificity for SNV at 0.1% frequency.
+Under our experimental conditions and with a slightly more stringent p value cutoff ( p=0.01 vs.
+p=0.05) we found a moderate reduction in sensitivity  (0.830 for variants at 0.63% and 0.14 for variants at 0.16%).
+The specificity was above 0.998 in all dilutions.
+While this drop in specificity appears minor, it is important to keep in mind that even a false positive rate of 0.002 results in 78 false positives when applied across the xx,xxx potential variants in the 13.xkb influenza genome.
 
@@
 Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions.
-The virus populations found in patient-derived samples are expected to be less divergent than the two lab strains.
+The virus populations found in patient-derived samples are expected to be less divergent than WSN33 and PR8.
 Also the nucleic acid concentration in patient samples is much lower than that found samples grown in cell culture.
 To mimic these conditions, we generated 20 viral colones, each with a single point mutation in the WSN33 background.
-We sequenced these mutants on the Illumina Miseq platform and cofirmed that no additional mutations >1% frequency had arisen between transfection and the passage 1 stock.
+We sequenced these mutants on the Illumina Miseq platform and confirmed that no additional mutations >1% frequency had arisen between transfection and the passage 1 stock.
 We also quantified the genome copy number of each stock using quantitative PCR.
-We then mixed equal genome equivalents of these 20 virsuses, generating an artificial population of $10^5$ copies per microliter in which each mutation was present at 5% frequency.
-This population was serially diluted into a stock of WSN33, generating artificial populations witha ll 20 mutations present a 2,1,0.5,0.2, and 0.1% frequency (`r hiseq` A).
-We then serially diluted each of these $10^5$ copy number populations into basal media to obtain populations with lower input.
-The range ($10^3-10^5$ copies per microliter) matches the inpus typically found in pateint-derived samples (data not shown).
+We then mixed equal genome equivalents of these 20 viruses, to generate an artificial population with $10^5$ copies per microliter and each mutation  present at 5% frequency.
+This population was serially diluted into a stock of WSN33, generating artificial populations with all 20 mutations present a 2,1,0.5,0.2, and 0.1% frequency (`r Hiseq` A).
+We then serially diluted each of these populations into basal media to obtain populations with lower input.
+The range ($10^3-10^5$ copies per microliter) matches the inputs typically found in patient-derived samples (data not shown).
 We sequenced the samples on the Illumina Hiseq platform, and variants were called using DeepSNV.
-We also processed and sequenced the WSN33 stock in duplicate.
+We also processed and sequenced the WSN33 stock in duplicate to control for any background mutations.
 
 We considered the 20 mutations present in our initial viral mixture as the only true positives in our analysis.
-Any variant that was present in the wild type stock was masked and neither considered a true positive nor a true negative.
-This was done to avoid benchmarking DeepSNV with variants identified only be DeepSNV.
+Any variant that was present in both duplicates of the wild type stock was masked and  considered neither a true positive nor a true negative.
+This was done to avoid benchmarking our pipeline with variants identified by our pipeline.
 
-The samples with the highest input concentration had the best sensitivity (`r hiseq` B and C).
-Sensitivity dropped in the 0.5% frequency samples.
-In general, our sensitivity was lower than that seen previously in the cell culture samples where the input concentration was at least $10^7$ copies per microliter.
-The drop in sensitivity is likely due to the 100 fold decrease in variant genomes present in the $10^5$ samples.
+The samples with the greatest input concentration and variant frequencies of 1% and greater had the highest sensitivity (`r hiseq` B and C).
+Sensitivity dropped at the 0.5% frequency, and in general was lower than that seen previously in the cell culture samples.
+Our cell culture samples had an  input concentration of at least $10^8$ copies per microliter; whereas, these samples had $10^5$ copies per microliter.
+The reduced sensitivity is likely due to the fact that there were only 7,000 of each variant genome in the 0.5% frequency samples compared to the equivalent of 62,500,000 in the 0.63% sample from the cell culture experiment.
+
+> This is a little tricky since we amplified the genomes first and then mixed them.
+
 During library prep, a disproportionate amount of the 2.0% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0% sample.
 
-In addition to a more drastic drop in sensitivity we found a much larger number of false positives at a p value cut off of 0.01 (`r hiseq` C).
-It should be noted that in all cases the specificity was greater than 99%.
-However, given the large number of potential false positives in an influenza genome (>40,000) even a specificity of 99% results in well over 200 erroneous variant calls.
-The drop in specificity is likely due to the decease in input levels.
-In our experience samples grown in cell culture have at least $10^7$ genomes/$\mu$l.
-As the genomic input decreases, we are more and more reliant on RT-PCR to amplify the true positive "signal" to a detectable level.
-It is not surprising that is error prone method also amplifies the noise in the process.
-Initially it is tempting to increase specificity by simply applying a more stringent p value cut off.
-This would result in marching backwards towards the y axis along the ROCs in `r hiseq` B.
-However, it is apparent from  `r hiseq` B that any increase in specificity would result in a drastic decrease in sensitivity.
-It is clear from `r hiseq` that under experimental conditions the DeepSNV output alone is insufficient to accurately separate true positives from false positives.
+The specificity was greater than 0.99 in all the samples with 10^5^ genomes per microliter.
+The decreased input concentration and the accompanying increase in RT-PCR cycles is likely the cause of reduced specificity compared with that seen in the cell culture experiment.
+As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r potential_var` potential variants in the influenza genome.
+Increased specificity can be achieved by applying a more stringent p-value cutoff.
+However, as shown by the ROC curves in `r hiseq`, this move towards the y axis would markedly reduce sensitivity.
+Our data demonstrate that with moderate concentrations of input nucleic acid, even statistically significant p-values from a robust variant caller are not sufficient to accurately separate true from false positive variants.
 
-In deep sequencing studies of viral diversity mapping quality and base quality cutoffs are often used to ensure that only the highest caliber sequencing data is used to call variants.
-Mapping quality refers to our confidence that a given read is mapped to the correct position in the genome while base quality estimates the likelihood that the base read by the sequencer is correct.
-In our initial analysis we masked bases that had a base quality score less than 30.
-We did not apply any mapping quality cut offs.
-`r Qual` A shows the distribution of the mean quality scores of the variants (p <0.01) present in the $10^5$ genomes/$\mu$l samples.
-In our case we find that while true and false positive variants can largely be separated by these metrics,  commonly used cut offs are not sufficient (dashed lines `r qual` A).
-Additionally, we found that in general false positives were biased near the ends of sequencing reads (`r qual` B).
-This is not surprising as it is well known that sequencing quality dips near the end of a read; however, recall that all bases with phred scores less than 30 were masked from our variant calling.
-In this case it seems that even high quality false positives are disproportionately called near the ends of reads, while true positives are called uniformly across the read resulting in an average read position near the middle of the read.
-(_do I need potential reasons for the bias?
-locations in the genome...)_
+#### Additional filtering criterea
 
-In the light of these quality distributions we applied a number of cut offs and drastically increased our specificity without sacrificing sensitivity (`r qual` C and D).
-For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean phred score of 35, and an average read position within the middle 50% of the read length.
-_These cut offs were choosen as they ..._ By applying these quality controls we were able to greatly decrease the number of false positives from over 200 in many cases to an average of 9 without sacrificing a single true positive.
-We recognize that by applying cut offs after variants have been called we are treating potential variant bases more stringently than those matching the consensus sequence.
-However,because we are applying cut offs to mean quality scores and there are many more consensus bases than variant bases at each position, it is unlikely these biases have any noticable affect on our analysis.
-In addition to accurately identifying the presence of expected variants DeepSNV was able to  accurately measure the frequency of these variants (`r freq`).
-However, while the median frequencies are near the expected values, there is a fair spread about each median, something that should be kept in mind when basing down stream analysis on frequency measurements.
+Many next generation sequencing studies utilize mapping (MapQ) and base (Phred) quality cut-offs to ensure that only the highest caliber sequencing data is used to call variants.
+Mapping quality measures the probability that a given read is mapped to the correct position in the genome while base quality estimates the likelihood that the base call by the sequencer is correct.
+In our initial analysis we masked bases that had a base quality score less than 30 and did not apply any mapping quality cut offs.
+We therefore sought to determine if we could improve our specificity by imposing more stringent MapQ and Phred cut-offs.
+Initially,frequently employed cut-offs such as a MapQ of 20 and a Phred of 30 but were unable to distinguish true from false positives in our 10^5^ samples (`r Qual` A).
+Many false positives occur on well mapped reads with high quality base calls.
 
-#### Lofreq
+We further parsed our false variant calls by locating them within individual sequencing reads.
+It is well known that sequence quality drops near the end of a read, and we found that our false positives clustered to these regions (`r Qual` B).
+The average Phred score of these false positives was xx, and no base calls with a Phred less than 30 was included in the analysis, further demonstrating that filtering on quality score alone is insufficient.
+In contrast, true positives were uniformly distributed across the reads resulting in an average read position near the middle of the read.
 
-In addition to DeepSNV, we analyzed our $10^5$ genomes/$\mu$l influenza populations with the variant caller Lofreq.
-This was done to ensure that the decreased accuracy that accompanied our realistic experimental conditions was not simply based on the DeepSNV variant calling method.
-Lofreq had marginally decreased sensitivity compared DeepSNV when applied to the variant frequencies $\geq$ 1.0% but marginally increased sensitivity when applied to variant frequencies $\l$ 1.0% (`r lofreq`).
-Additionally, Lofreq had specificity near that seen in our cell culture benchmarking experiment, and much better than DeepSNV prior to our quality controls.
-This is not completely surprising as the Lofreq algorithm takes MapQ and Phred scores into account when calling variants, something DeepSNV does not do.
-Additionally, Lofreq was not prone to calling false positives near the ends of reads, _(verify whether or not it uses read position Show distributions of Lofreq variants?)_, but   `r lofreq` A suggests that sacrificing sensitivity is the only means of increasing specificity.
+Based on these data we applied a number of empirically determined cut-offs and drastically increased our specificity without sacrificing sensitivity (`r Qual` C and D).
+For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean phred score of 35, and an average read position within the middle 50% of the read.
+In addition to these quality cut-offs we applied a number of other criteria in an attempt to increase accuracy such as correcting p-values using the Benjamin Hochman p-value correction, utilizing more stringent p values or frequency cut-offs, retaining PCR duplicate reads, trimming the ends of the influenza genome, and employing various distributions to estimate the error rate in the control sample.
+All of these results have be summarized in an interactive shiny application available for download at https://github.com/lauringlab/benchmarking_shiny.git and we invite the reader to explore the affects of all these criteria on our data set.
 
-Lofreq has been reported to have perfect specificity and a sensitivity of 96% for variants present at 0.2%.
-However, this accuracy reflects a best case scenario using simulated sequencing reads.
-As was the case with DeepSNV, the accuracy of Lofreq was much lower when applied under conditions that often face virologist using patient derived samples.
-Our data suggests that is inappropriate to apply Lofreq or DeepSNV under diverse experimental conditions and expect the performance to be unaffected.
+DeepSNV was able to accurately measure the frequency of the true positive variants (`r Freq`).
+However, while the median frequencies are near the expected values, there is a fair spread in each sample (mean standard deviation of xx).
+This error should  be kept in mind when employing down stream analysis that depend on frequency measurements such as diversity metrics and comparisons of population structure.
 
-It may be possible to further filter the putative variants called by Lofreq and achieve a similar accuracy to what was achieved using DeepSNV.
-However, because DeepSNV was much less computationally intensive, more user friendly _not sure if I can or should say that_, and capable of high accuracy, we continued to benchmark DeepSNV using the less concentrated samples from our dilution series.
+#### Alternative variant callers
 
-_Include Lofreq in supplimentary?_
+DeepSNV is one of many variant callers that use a combination of empiric and statistical approaches to model error rates and separate true from false variants.
+We asked whether the decreased accuracy observed in our data set was due simply topeculiarities specific to the DeepSNV algorithm.
+We therefore analyzed our $10^5$ input populations using Lofreq, another alogorthim that is commonly used in viral next generation sequencing studies.
+Lofreq has been reported to have perfect specificity in the past.
+Under our conditions Lofreq had marginally reduced sensitivity compared DeepSNV when applied to the variant frequencies \geq 1.0% but marginally increased sensitivity when applied to variant frequencies \l 1.0% (`r Lofreq`).
+The specificity of LoFreq was comparable to what we observed with DeepSNV in our high-input cell culture-derived populations (`r Miseq`), and better than DeepSNV in our initial implementation (`r Hiseq` prior to Phred, MapQ, and read position filtering).
+This increased specificity is most likely due to the fast that  the LoFreq algorithm takes MapQ and Phred scores into account when calling variants and has a stringent strand bias filter that removes many of the variants fount only at one end of the reads.
+However even with these additional characteristics, the specificity of LoFreq was lower than our improved DeepSNV pipeline (`r Qual`) with over 40 false positives per sample.
+It appears that higher than expected false positive rates are not specific to DeepSNV and most likely plague all variant callers applied to patient-derived viral samples.
 
-#### Lower input levels
+\subsection{Lower input levels}
 
-Influenza patient samples vary in input concentration by several orders of magnitude and depend on a variety of factors often outside the control of a researcher (day of collection, volume of mucus collected ect.).
-To ensure accuracy across a range of input levels we diluted our known mutant mixes serially into viral media.
-We found both decreases in sensitivity and specificity accompanied lower input concentrations (`r lower`).
-The decreased sensitivity is expected as the likelihood of detecting rare events decreases with sample sizem while the decreased specificity is most likely due to a greater dependence on RT-PCR to amplify the signal to detectable limits.
-Is it not surprising that an amplification of noise accompanies this dependence.
-However, our results do highlight the importance of controlling for input levels when comparing populations of RNA viruses.
+Host-derived viral populations vary in copy number and titer by several orders of magnitude.
+This variability can be attributed to a variety of factors including but not limited to: collection site, ease of of nucleic acid isolation, library preparation, and host and viral variability.
+To ensure accuracy across a range of input levels, we diluted our experimental populations serially into basal media and applied our emperically determined quality cut-offs to our list of putative variants (`r Lower`).
+As expected, the sensitivity for minority variants was lower in populations with lower genome copy number.
+A variant with a 0.5% frequency in a 10^4^ genomes per microliter sample is only present on 700 genomes and any number of these could be lost in the amplification and library preparation process.
+We also found that specificity was reduced in samples with lower starting copy numbers.
+This is presumably due to a greater dependence on RT-PCR amplification.
+These data highlight the importance of controlling for input levels when comparing diversity across experimental samples.
 
-If RT-PCR errors are responsible for the decreased specificity found at lower input levels, and RT_PCR are randomly distributed across the genome, we would expect such errors to be eliminated by processing the samples in duplicate.
-To test this hypothesis, a subset of the $10^4$ genomes/$\mu$l were run in duplicate.
-The 0.2% was not run as we expected a decrease in sensitivity would accompany this approach and had little hope of finding these rare variants based on the sensitivity we saw in `r lower` B.
-Two RT-PCR reactions were run from a new RNA prep.
-These samples were processed separately and sequenced on the same lane of an Illumina Hiseq.
-The previously mentioned quality cut offs were applied and for a variant call to be considered it was required to be found in both duplicates.
-This drastically improved our specificity and resulted an accuracy that was comparable to that found in the $10^5$ samples (`r dups`).
-Unexpectedly, an increase in sensitivity accompanied this sequencing run.
-This is most likely do to a superior library prep with less sample loss during size selection.
+In most cases, RT-PCRs error should be randomly distributed across the amplified region.
+If RT-PCR errors are responsible for the reduced specificity found at lower input levels, and RT-PCR are randomly distributed across the genome, we would expect such error to be eliminated by processing the samples in duplicate.
+To test this hypothesis, we performed duplicate RT-PCR reactions on for the 5,2,1, and 0.5% samples with 10^4^ genomes per microliter.
+This was done using a separate RNA prep to that shown in `r lower` A.
+The duplicates were processed separately, but sequenced on the same lane of an Illumina Hiseq.
+We applied the previously mentioned quality cut-offs and required that a given variant be found in both duplicates.
+This drastically improved our specificity and resulted in an accuracy that was comparable to that found in the 10^5^ samples (`r Dups`).

### 2016-02-23 scripts/figures.Rmd

@@
 author: "JT McCrone"
-date: "January 25, 2015"
+date: "February 11, 2015"
 output: pdf_document
@@
   roc.p<-ggplot(roc.df,aes(x=1-specificity,y=sensitivity))+geom_step(color="black",size=1.5)+xlab("1-Specificity")+ylab("Sensitivity")+theme(axis.title.y=element_text(vjust=1))
-  roc.p<-roc.p+geom_line(data=lines,aes(x=spec,y=sense,color=as.factor(group)),size=c(1),lty=c(2))+scale_color_brewer(palette="Set1")
+  roc.p<-roc.p+geom_line(data=lines,aes(x=spec,y=sense,group=as.factor(group),color=rep(c("#08519cff","#787878ff"),times=3)),size=c(1),lty=c(2))+scale_color_manual(values=c("#08519cff","#787878ff"))
 
@@
 ```{r,duplicates,fig.height=5,fig.width=3.4}
-sum.df<-read.csv("../data/process/2015-11-14/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#")
+dup.df<-read.csv("../data/process/2015-11-14/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#")
 #sum.df<-read.csv("~/Box Sync/Lab_work/NGS/Benchmarking/data/2015-11-14/all.sum.csv")
 pat<-"([0-9]+)_([0-9]+)"
-mutate(sum.df, exp.freq = sub(pat,"\\1",Id),
+mutate(dup.df, exp.freq = sub(pat,"\\1",Id),
        dup = as.numeric(sub(pat,"\\2",Id)),
-)-> sum.df
+)-> dup.df
 
-sum.df<-subset(sum.df,Id!="PC") # remove the plasmid control that was run
-mutate(sum.df,Id=paste(4,exp.freq,sep="_"),exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->sum.df
+dup.df<-subset(dup.df,Id!="PC") # remove the plasmid control that was run
+mutate(dup.df,Id=paste(4,exp.freq,sep="_"),exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->dup.df
 
-sum.df<-mutate(sum.df,category=mutation %in% true_snv$mutant) # add Column for True and false variants
-sum.df$category[sum.df$mutation %in% wt_mut]<-"wt"
+dup.df<-mutate(dup.df,category=mutation %in% true_snv$mutant) # add Column for True and false variants
+dup.df$category[dup.df$mutation %in% wt_mut]<-"wt"
 
-cut.df<-subset(sum.df,Phred>35 & MapQ>30 & Read_pos<94 & Read_pos>32)
+cut.df<-subset(dup.df,Phred>35 & MapQ>30 & Read_pos<94 & Read_pos>32)
 
@@
 B) A table summarizing the accuracy when applied to duplicate samples with 10 $^4$ genomes/$\mu$l input, p<0.01.
+
+\pagebreak
+
+### Table 1
+
+Richness
+Inverse Simpsons D ( without replacement)
+Shannon Entropy
+For these I'm assuming all varaints are independent
+
+```{r}
+perfect<-data.frame(mutation=rep(true_snv$mutant,times=4),freq.var=rep(c(0.05,0.02,0.01,0.005),each=20))
+dups<-subset(dups,p.val<0.01)
+dups<-mutate(dups,freq.var=pmax(freq1,freq2))
+
+rich<-function(x){
+  dim(x)[1]  
+}
+
+H<-function(x){
+  p=x$freq.var*log(x$freq.var)
+  -sum(p)
+}
+
+D<-function(x){
+  p=(x$freq.var)^2
+  1/sum(p)
+}
+
+euc<-function(x,ref){
+  x$Id="test"
+  ref$Id="ref"
+  all<-rbind(subset(x,select=c(mutation,freq.var,Id)),ref)
+  tab<-as.matrix(subset(dcast(all,mutation ~ Id, value.var = "freq.var"),select=-c(mutation)))
+  tab[is.na(tab)]<-0
+  as.dist(dist(tab))
+}
+get_data<-function(x,setting){
+  if(length(unique(x$freq.var))==1){
+    exp.freq=unique(x$freq.var)
+  }else{
+  exp.freq=unique(x$exp.freq)
+  }
+  out<-data.frame(Richness=rich(x), Entropy=H(x),Simpson=D(x))#,dist=euc(x,subset(perfect,freq.var==exp.freq))[1])
+  rownames(out)<-setting
+  return(t(out))
+}
+tabulate<-function(input,freq){
+get_data(subset(perfect,freq.var==freq),"Perfect")->perf
+get_data(subset(sum.df,gc==input & exp.freq==freq),"DeepSNV")->deepSNV
+get_data(subset(lofreq.df,gc==input & exp.freq==freq),"Lofreq")->lo
+get_data(subset(cut.read.sig.df,gc==input & exp.freq==freq),"DeepSNV Final")->deepSNV.f
+knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f),digits = 2)
+}
+tabulate.4<-function(input,freq){
+get_data(subset(perfect,freq.var==freq),"Perfect")->perf
+get_data(subset(sum.df,gc==input & exp.freq==freq),"DeepSNV")->deepSNV
+get_data(subset(lofreq.df,gc==input & exp.freq==freq),"Lofreq")->lo
+get_data(subset(cut.read.sig.df,gc==input & exp.freq==freq),"DeepSNV Final")->deepSNV.f
+get_data(subset(dups,exp.freq==freq),"DeepSNV duplicates")->deepSNV.d
+
+knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f,deepSNV.d),digits = 2)
+}
+
+```
+
+$10^5$ 5%
+
+```{r}
+tabulate(5,0.05)
+```
+
+$10^5$ 1%
+
+```{r}
+tabulate(5,0.01)
+```
+
+$10^5$ 0.5%
+
+```{r}
+tabulate(5,0.005)
+```
+
+$10^4$ 5%
+
+```{r}
+tabulate.4(4,0.05)
+```
+
+$10^4$ 1%
+
+```{r}
+tabulate.4(4,0.01)
+```
+
+$10^4$ 0.5%
+
+```{r}
+tabulate.4(4,0.005)
+```

### 2016-02-23 doc/paper/draft.Rmd
intro and results drafts

---
output: 
  word_document:
    reference_docx: style.docx
bibliography: bib.bib
csl: journal-of-virology.csl

---

```{r message=F,echo=FALSE,cache=T,results='hide'}
xamp_roc<-"figure 1"
miseq<-"figure 2"
Miseq<-"Figure 2"
hiseq<-"figure 3"
Hiseq<-"Figure 3"
qual<-"figure 4"
Qual<-"Figure 4"
freq<-"figure 5"
Freq<-"Figure 5"
lofreq<-"Figure 6"
Lofreq<-"figure 6"
lower<-"figure 7"
Lower<-"Figure 7"
dups<-"figure 8"
Dups<- "Figure 8"
Table<-"Table 1"
table<-"table 1"

require("Biostrings")
require(plyr)

knitr::knit("../../scripts/figures.Rmd",tangle = T)

```

```{r,message=F,echo=FALSE,cache=T}

miseq_vars<-format(miseq_vars,scientific=F)
miseq_kb<-round(miseq_kb,digits=2)

hiseq_vars<-format(hiseq_vars,scientific=F)
hiseq_kb<-round(hiseq_kb,digits=2)

st_error="XXX"
```

##Introduction

The popularization of next generation sequencing (NGS) is revolutionizing the study of viral diversity. Until recently, our understanding of viral evolution was limited by high-effort, low-throughput sequencing methods. Thus much of what is known regarding \emph{in vivo} viral evolution is derived from epidemiological studies (possibly cite examples, flu foot and mouth ect). However, with the advent of NGS it is now feasible to detect rare single nucleotide variants present in hundreds of patient derived samples (cite). To many, next generation sequencing promises a high resolution glimpse of intrahost viral dynamics, and there has been an explosion of papers that apply these highly sensitive and cost effective  techniques to viral populations from a myriad of sample sources (cite). Although NGS is highly sensitive, its specificity in the context of intrahost viral populations has not been robustly measured. Each sequencing platform has inherent error profiles, and sample collection, target amplification, and library preparation are additional processes where by errors can be introduce and propagated. The next generation sequencing of viral populations, in particular those of RNA viruses, is error prone, and these errors must be accounted for in order to ensure accurate analysis.

Many sample preparation pipelines have been developed to control for errors in NGS based viral studies; hover, each approach has its own caveats and ultimately limitted application. For example, Cirseq is an ingenious approach in which  template RNA is sheared and circularized prior to reverse transcription  [@Acevedo:2014ej]. Rolling circle cDNA synthesis then produces on average three tandem reads of the same RNA fragment. Consensus sequences are assigned to each RNA fragment controlling for reverse transcription, PCR, and sequencing errors. However, the sensitivity of the method has not been robustly benchmarked against known rare variants. Additionally, Cirseq requires 1 microgram of starting RNA and provides no mechanism for targeting viral RNA [@Acevedo:2014df]. Thus the method has limited application to patient-derived samples  that  yield low nucleic acid concentrations and are highly enriched for host RNA.

Alternatively, bar-coded primers can be used to construct consensus sequences for each cDNA template and control for PCR and sequencing erros [@Jabara:2011eo]. While this approach requires less input and targets sequencing to the viral genome it is unable to control for RT errors present in the cDNA. Additionally, primer Id based methods require that each bar code be physically attached to a PCR product, and so the region of the genome that can be bar-coded and analyzed is limited by the read length of the sequencing platform (300-500 bp for Illumina).  Therefore primer ID is useful in interrogating specific regions of a genome, but has limited application in whole genome studies. An adapted form of Sequence independent single primer amplification (SISPA) is an alternative approach that allows for the whole genome sequencing and controls for error propogated by library preparation [@Djikeng:2008cd]. RT-PCR products are sheared and tagged with bar-coded random primers in a klenow reaction, prior to library preparation. This method controls for any errors that may arise in the PCR step of the library prep and controls for PCR duplicates and frequency biases. However, the bar-coding reaction used in SISPA is biased by the bar codes resulting in uneven coverage and thus sensitivity across the genome [@Rosseel:2013dn].

In addition to sample processing pipelines, a number of statistical approaches have been developed to distinguish true variants from sequencing errors using only sequence data [@Gerstung:1js; @Isakov:2015gq; @Wilm:2012br; @Macalalad:2012fx; @Koboldt:2009dk]. In general these algorithms distinguish true variants from errors by calibrating base specific error rates across the genome. The rates are determined by various metrics including but not limited to mapping quality, phred score, nucleotide position on a read, and the sequence context of a given base. True variants are identified as those with frequencies that exceed the expected error rate according to some predetermined statistical test. Despite being employed in many NGS based studies of viral diversity (cite?) few variant calling algorithms have been benchmarked using viral populations, and to our knowledge none have been tested under patient-derived sample conditions. The accuracy of such algorithms in the context of NGS studies of patient-derived samples is not known.

Here, we use defined populations of influenza A to robustly benchmark the accuracy of a variant calling pipeline under patient derived conditions. In our analysis, samples are enriched for influenza genomes through a multiplex RT-PCR reaction prior to sequencing on the Illumina platform. PCR products are sheared using Covaris sonication which gives unbiased coverage across the genome and, variants are identified using the DeepSNV algorithm. In this work, we highlight the challenges that accompany NGS based studies of viral diversity and  make some suggestions for improving the accuracy of our approach. It is our hope that this work exemplifies the quality controls that should be run prior to any NGS based viral study and provides a comprehensive data set for benchmarking future variant calling approaches.

##Results

The ability to reliably distinguish true from false single nucleotide variants (SNV) in next generation sequencing data is integral to accurate studies of viral diversity. Sensitivity is the proportion of true variants identified; while, specificity is the proportion of true negatives identified. $1-\text{specificity}$ is the false positive rate, the ratio of false positive variants to true negatives. Both a sensitivity and specificity of 1 describe perfect accuracy in which all the true variants are found and nothing else.

There is an obvious trade off between sensitivity and specificity. Improved sensitivity often requires less stringent criteria in variant calling which reduces specificity. Conversely, increased stringency can improve specificity but reduce sensitivity. This relationship can be represented as a receiver operating characteristic (ROC) curve (`r xamp_roc`). An ROC plots the sensitivity of an assay along the y axis and 1-specificity along the x axis. To make an ROC curve for a variant calling algorithm one must test the assay on known data. The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real; often this is a p value or quality score. In a controlled benchmakring experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut offs (`r xamp_roc` A). These points are then used to construct the curve (`r xamp_roc` B). A perfect ROC curve in which all the true positives can be separated from all false positives is a right angle that follows the upper left perimeter of the plot.

###Initial accuracy

A comprehensive comparison of SNV calling approaches  is beyond the scope of this work. Here, we robustly benchmark one variant caller and highlight few means for improving accuracy. In the process we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is being applied.

DeepSNV is a variant calling algorithm that uses a clonal, plasmid-derived, control to estimate local error rates across the genome. Because it is clonal, the sequence of the control is known with a high degree of confidence and so any nonconsensus base is indicative of a sequencing error. Additionally, the control and experimental samples are processed together and have identical noise characteristics. DeepSNV then applies a hierarchical binomial model at each genomic position and identifies true variants as those with frequencies significantly above the noise found in the plasmid control. Like many variant calling method, the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification. We created defined mixtures of two influenza strains, WSN33 and PR8, to characterize the accuracy of DeepSNV when applied to RT-amplified RNA virus populations.

WSN33 and PR8 genomes were mixed serially such that WSN33 was present at  frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq` A). Because these viruses differ at 491 positions, there are 491 true positives in each dilution. On plasmids subjected to limited PCR, DeepSNV identified known variants at 0.1\% frequency with a sensitivity of 0.860 and  specificity of 1.0. Under our experimental conditions and with a slightly more stringent p value  ( p=0.01 vs. p=0.05) we found a moderate reduction in sensitivity  (0.830 for variants at 0.63\% and 0.14 for variants at 0.16\%). The specificity was above 0.998 in all dilutions. While this drop in specificity appears minor, it is important to keep in mind that even a false positive rate of 0.002 results in 78 false positives when applied across the `r miseq_vars` potential variants in the `r miseq_kb`kb of the influenza genome we analyzed.

###An experimental intrahost population

Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions. The virus populations found in patient-derived samples are  less divergent than WSN33 and PR8. Also the nucleic acid concentration in patient samples is much lower than that found samples grown in cell culture. To mimic these conditions, we generated 20 viral clones, each with a single point mutation in the WSN33 background. We sequenced these mutants on the Illumina Miseq platform and confirmed that no additional mutations >1\% frequency had arisen between transfection and the passage 1 stock. We also quantified the genome copy number of each stock using quantitative PCR. We then mixed equal genome equivalents of these 20 viruses, to generate an artificial population with $10^5$ copies per microliter with each mutation  present at 5\% frequency. This population was serially diluted into a stock of WSN33, generating artificial populations with all 20 mutations present a 2, 1, 0.5, 0.2, and 0.1\% frequency (`r Hiseq` A). We then serially diluted these populations into basal media to obtain populations with lower input. The range ($10^3-10^5$ copies per microliter) matches the inputs typically found in patient-derived samples (data not shown). We sequenced the samples on the Illumina Hiseq platform, and variants were called using DeepSNV. We also processed and sequenced the WSN33 stock in duplicate to control for any background mutations.

The 20 mutations present in our initial viral mixture were the  only true positives considered in our analysis. Any variant that was present in both duplicates of the wild type stock was masked and  considered neither a true positive nor a true negative. This was done to avoid benchmarking our pipeline with variants identified by our pipeline.

The samples with the greatest input concentration and variant frequencies of 1\% and greater had the highest sensitivity (`r hiseq` B and C). Sensitivity dropped at the 0.5\% frequency, and in general was lower than that seen previously in the cell culture samples. Typically our samples grown in cell culture have  at least $10^8$ genomes per microliter; whereas, the nucleic acid concentration in these samples was 1,000 fold less . The reduced sensitivity is likely due to the fact that we expect only 7,000 of each variant genome in the 0.5\% frequency samples compared to the equivalent of 62,500,000 variant genomes in the 0.63\% sample from the cell culture experiment above.

%This is a little tricky since we amplified the genomes first and then mixed them.

During library prep, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.

The specificity was greater than 0.99 in all the samples with $10^5$ genomes per microliter. The decreased input concentration and the accompanying increase in RT-PCR cycles is likely the cause of reduced specificity compared with that seen in the cell culture experiment. As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome. Increased specificity can be achieved by applying a more stringent p-value cutoff. However, as shown by the ROC curves in `r hiseq`, this move towards the y axis would markedly reduce sensitivity. Our data demonstrate that with moderate concentrations of input nucleic acid, even statistically significant p-values from a robust variant caller are not sufficient to accurately separate true from false positive variants.

###Additional filtering criteria

Many next generation sequencing studies utilize mapping (MapQ) and base (Phred) quality cut-offs to ensure that only the highest caliber sequencing data is used to call variants. Mapping quality measures the probability that a given read is mapped to the correct position in the genome while base quality estimates the likelihood that the base call by the sequencer is correct. In our initial analysis we masked bases that had a base  score less than 30 (0.001 probability of being incorrect) and did not apply any mapping quality cut offs. We  sought to determine if we could improve our specificity by imposing more stringent MapQ and Phred cut-offs. Initially, we applied frequently employed cut-offs such as a MapQ of 20 and a Phred of 30 but were unable to distinguish true from false positives in our $10^5$ samples (`r Qual` A). Many false positives occur on well mapped reads with high quality base calls.

We further parsed our false variant calls by locating them within individual sequencing reads. It is well known that sequence quality drops near the end of a read, and we found that our false positives clustered to these regions (`r Qual` B). The average Phred score of these false positives was xx, further demonstrating that filtering on quality score alone is insufficient. In contrast, true positives were uniformly distributed across the reads resulting in an average read position near the middle of the read.

Based on these data we applied a number of empirically determined cut-offs and drastically increased our specificity without sacrificing sensitivity (`r Qual` C and D). For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean phred score of 35, and an average read position within the middle 50\% of the read. In addition to these quality cut-offs we applied a number of other criteria in an attempt to increase accuracy. We tried using the Benjamin Hochman p-value correction, utilizing more stringent p values or frequency cut-offs, retaining PCR duplicate reads, trimming the ends of the influenza genome, and employing various distributions to estimate the error rate in the control sample. All of these results have be summarized in an interactive shiny application available for download at $https://github.com/lauringlab/benchmarking_shiny.git$  We invite the reader to explore the affects of all these criteria on our data set.

In addition to distinguishing true from false SNV, DeepSNV was able to estimate the frequency of the true positive variants with reasonable accuracy (`r Freq`). However, while the median frequencies are near the expected values, there is a fair spread in each sample (mean standard error of `r st_error`). This error should be kept in mind when employing down stream analysis that depend on frequency measurements such as diversity metrics like Shannon's Entropy and Simpson's D, as well as inter-sample comparison of population structure .

###Alternative variant callers

DeepSNV is one of many variant callers that use a combination of empiric and statistical approaches to model error rates and separate true from false variants. We asked whether the decreased accuracy observed in our data set was due simply to peculiarities specific to the DeepSNV algorithm. We therefore analyzed our $10^5$ input populations using Lofreq, another algorithm that is commonly used in viral next generation sequencing studies. Lofreq has been reported to have perfect specificity in the past. Under our conditions Lofreq had marginally reduced sensitivity compared to DeepSNV when applied to the variant frequencies $\geq$ 1.0\% but marginally increased sensitivity when applied to variant frequencies $<$ 1.0\% (`r Lofreq`). The specificity of LoFreq was comparable to what we observed with DeepSNV in our high-input cell culture-derived populations (`r Miseq`), and better than DeepSNV in our initial implementation (`r Hiseq` prior to Phred, MapQ, and read position filtering). This increased specificity is most likely due to the fast that  the LoFreq algorithm takes MapQ and Phred scores into account when calling variants and has a stringent strand bias filter that removes many of the variants fount only at one end of the reads. However even with these additional characteristics, the specificity of LoFreq was lower than our improved DeepSNV pipeline (`r Qual`) with over 40 false positives per sample. It appears that higher than expected false positive rates are not specific to DeepSNV and most likely plague all variant callers applied to patient-derived viral samples.

%%%% if the Errors are PCR related why does Lofreq remove them.

###Lower input levels

Host-derived viral populations vary in copy number and titer by several orders of magnitude. This variability can be attributed to a variety of factors including but not limited to: collection site, ease of of nucleic acid isolation, library preparation pipeline, and host and viral variability. To ensure accuracy across a range of input levels, we diluted our experimental populations serially into basal media and and called variants using our modified DeepSNV analysis pipeline (`r Lower`). As expected, the sensitivity was lower in populations with fewer genomes. A variant with a 0.5\% frequency in a $10^4$ genomes per microliter sample is expected to present on only 700 genomes in the initial RT-PCR and any number of these genomes could be lost due to bottlenecks in the amplification and library preparation process. We also found that specificity was reduced in samples with lower starting input. This is presumably due to a greater dependence on RT-PCR amplification. These data highlight the importance of controlling for input levels when comparing diversity across experimental samples.

In most cases, RT-PCRs error should be randomly distributed across the amplified region. If RT-PCR errors are responsible for the reduced specificity found at lower input levels, and RT-PCR are randomly distributed across the genome, we would expect such error to be eliminated by processing the samples in duplicate. To test this hypothesis, we performed duplicate RT-PCR reactions on the 5, 2, 1, and 0.5\% variant frequency samples that had a starting concentration of $10^4$ genomes per microliter. This was done using a separate RNA prep than that shown in `r lower` A. The duplicates were processed separately, but sequenced on the same lane of an Illumina Hiseq. We applied the previously mentioned quality cut-offs and required that a given variant be found in both duplicates. This drastically improved our specificity and resulted in an accuracy that was comparable to that found in the $10^5$ samples (`r Dups`).

#References

### 2016-02-24 doc/draft.Rmd
sent intro to Adam

--- 2016-02-23 doc/paper/draft.Rmd
@@
 
-```{r message=F,echo=FALSE,cache=T,results='hide'}
+```{r message=F,echo=FALSE,cache=T,results='hide',warning=FALSE}
 xamp_roc<-"figure 1"
+Xamp_roc<-"Figure 1"
 miseq<-"figure 2"
@@
 
-knitr::knit("../../scripts/figures.Rmd",tangle = T)
-
+knitr::knit("../scripts/figures.Rmd",tangle = T)
+source("./figures.R")
+#this is so the cache resets
 ```
@@
 st_error="XXX"
+false_p_phred<-round(false_p_phred,digits=1)
+#csl: journal-of-virology.csl
 ```
@@
 
-The popularization of next generation sequencing (NGS) is revolutionizing the study of viral diversity.
-Until recently, our understanding of viral evolution was limited by high-effort, low-throughput sequencing methods.
-Thus much of what is known regarding \emph{in vivo} viral evolution is derived from epidemiological studies (possibly cite examples, flu foot and mouth ect).
-However, with the advent of NGS it is now feasible to detect rare single nucleotide variants present in hundreds of patient derived samples (cite).
-To many, next generation sequencing promises a high resolution glimpse of intrahost viral dynamics, and there has been an explosion of papers that apply these highly sensitive and cost effective  techniques to viral populations from a myriad of sample sources (cite).
+Until recently, our understanding of intrahost viral evolution was limited by high-effort, low-throughput sequencing methods.
+Much of what is known regarding \emph{in vivo} viral evolution has been derived from epidemiological studies (possibly cite examples, flu foot and mouth ect).
+However, with the advent of next generation sequencing (NGS) it is now feasible to detect rare single nucleotide variants present in hundreds of patient-derived samples (cite examples?).
+Next generation sequencing offers a high resolution glimpse of intrahost viral dynamics, and there has been an explosion of papers that apply these highly sensitive and cost effective  techniques to viral populations from a myriad of sample sources (cite examples?).
 Although NGS is highly sensitive, its specificity in the context of intrahost viral populations has not been robustly measured.
-Each sequencing platform has inherent error profiles, and sample collection, target amplification, and library preparation are additional processes where by errors can be introduce and propagated.
-The next generation sequencing of viral populations, in particular those of RNA viruses, is error prone, and these errors must be accounted for in order to ensure accurate analysis.
+Each sequencing platform has inherent error profiles, and sample collection, target amplification, and library preparation are additional processes whereby errors can be introduce and propagated.
+The next generation sequencing of viral populations, in particular those of RNA viruses, is error prone, and these errors must be accounted for to ensure accurate secondary analysis.
 
-Many sample preparation pipelines have been developed to control for errors in NGS based viral studies; hover, each approach has its own caveats and ultimately limitted application.
+Many sample preparation pipelines have been developed to control for errors in NGS based studies of RNA virus populations; however, each approach has its own caveats that ultimately limit its application.
 For example, Cirseq is an ingenious approach in which  template RNA is sheared and circularized prior to reverse transcription  [@Acevedo:2014ej].
-Rolling circle cDNA synthesis then produces on average three tandem reads of the same RNA fragment.
+Rolling circle cDNA synthesis then produces tandem reads of the same RNA fragment.
 Consensus sequences are assigned to each RNA fragment controlling for reverse transcription, PCR, and sequencing errors.
 However, the sensitivity of the method has not been robustly benchmarked against known rare variants.
-Additionally, Cirseq requires 1 microgram of starting RNA and provides no mechanism for targeting viral RNA [@Acevedo:2014df].
-Thus the method has limited application to patient-derived samples  that  yield low nucleic acid concentrations and are highly enriched for host RNA.
+Additionally, Cirseq requires 1 microgram of starting RNA and provides no mechanism for targeting sequencing to viral RNA [@Acevedo:2014df].
+As such, the method has limited application to patient-derived samples that are often characterized by low nucleic acid concentrations and are enriched for host RNA.
 
-Alternatively, bar-coded primers can be used to construct consensus sequences for each cDNA template and control for PCR and sequencing erros [@Jabara:2011eo].
-While this approach requires less input and targets sequencing to the viral genome it is unable to control for RT errors present in the cDNA.
-Additionally, primer Id based methods require that each bar code be physically attached to a PCR product, and so the region of the genome that can be bar-coded and analyzed is limited by the read length of the sequencing platform (300-500 bp for Illumina).
-Therefore primer ID is useful in interrogating specific regions of a genome, but has limited application in whole genome studies.
-An adapted form of Sequence independent single primer amplification (SISPA) is an alternative approach that allows for the whole genome sequencing and controls for error propogated by library preparation [@Djikeng:2008cd].
+Alternatively, bar-coded primers can be used to construct consensus sequences for each cDNA template and control for PCR and sequencing errors [@Jabara:2011eo].
+This approach requires less input and targets sequencing to the viral genome, but unlike Cirseq, it is unable to account for RT errors present in the cDNA.
+Additionally, primer Id based methods require each bar code be physically attached to a PCR product, and so the region of the genome that can be targeted is limited by the read length of the sequencing platform (250-500 bp for Illumina).
+Primer ID based methods are useful in interrogating specific regions of a genome (cite examples?), but have limited application in whole genome sequencing.
+An adapted form of Sequence independent single primer amplification (SISPA) is an alternative approach that allows for whole genome sequencing and controls for errors propagated during library preparation [@Djikeng:2008cd].
 RT-PCR products are sheared and tagged with bar-coded random primers in a klenow reaction, prior to library preparation.
-This method controls for any errors that may arise in the PCR step of the library prep and controls for PCR duplicates and frequency biases.
-However, the bar-coding reaction used in SISPA is biased by the bar codes resulting in uneven coverage and thus sensitivity across the genome [@Rosseel:2013dn].
+SISPA controls for any errors that may arise during the library amplification and controls for PCR duplicate frequency biases.
+However, the bar-coding reaction used in SISPA is biased and results in uneven coverage and sensitivity across the genome [@Rosseel:2013dn].
 
-In addition to sample processing pipelines, a number of statistical approaches have been developed to distinguish true variants from sequencing errors using only sequence data [@Gerstung:1js; @Isakov:2015gq; @Wilm:2012br; @Macalalad:2012fx; @Koboldt:2009dk].
-In general these algorithms distinguish true variants from errors by calibrating base specific error rates across the genome.
-The rates are determined by various metrics including but not limited to mapping quality, phred score, nucleotide position on a read, and the sequence context of a given base.
-True variants are identified as those with frequencies that exceed the expected error rate according to some predetermined statistical test.
-Despite being employed in many NGS based studies of viral diversity (cite?) few variant calling algorithms have been benchmarked using viral populations, and to our knowledge none have been tested under patient-derived sample conditions.
-The accuracy of such algorithms in the context of NGS studies of patient-derived samples is not known.
+Given the limitations described above, a number of statistical approaches have been developed to distinguish true variants from sequencing errors using only sequencing data [@Gerstung:1js; @Isakov:2015gq; @Wilm:2012br; @Macalalad:2012fx; @Koboldt:2009dk].
+In general these algorithms calibrate base specific error rates across the genome.
+The rates are determined by various metrics including but not limited to mapping quality, phred score, nucleotide position on a read, and sequence context.
+True variants are identified as those with frequencies exceeding the expected error rate according to some predetermined statistical test.
+Despite being employed in many NGS based studies of viral diversity (cite?) few variant calling algorithms have been benchmarked using defined viral populations, and to our knowledge none have been tested under patient-derived sample conditions.
+The accuracy of such algorithms in the context of NGS studies of patient-derived RNA viral populations is not known.
 
-Here, we use defined populations of influenza A to robustly benchmark the accuracy of a variant calling pipeline under patient derived conditions.
-In our analysis, samples are enriched for influenza genomes through a multiplex RT-PCR reaction prior to sequencing on the Illumina platform.
-PCR products are sheared using Covaris sonication which gives unbiased coverage across the genome and, variants are identified using the DeepSNV algorithm.
+We are interested in using NGS to study the intrahost dynamics of RNA viruses.
+Because our experimental design relies on whole genome sequencing of patient-derived samples, we are unable to use any of the above error-controlling sample preparation pipelines.
+We, are left with only statistical algorithms to control for errors.
+However, the accuracy of these algorithms under patient-derived sample conditions is unknown.
+Here, we use defined populations of Influenza A to robustly benchmark the accuracy of a variant calling pipeline under patient-derived conditions.
+In our analysis, samples are enriched for influenza genomes through a multiplex RT-PCR reaction, sheared using Covaris sonication, sequenced on the Illumina platform, and analyzed using the DeepSNV algorithm and in house scripts (all available at \emph{https://github.com/lauringlab/variant_pipeline})[@Gerstung:1js].
 In this work, we highlight the challenges that accompany NGS based studies of viral diversity and  make some suggestions for improving the accuracy of our approach.
-It is our hope that this work exemplifies the quality controls that should be run prior to any NGS based viral study and provides a comprehensive data set for benchmarking future variant calling approaches.
+This work exemplifies the quality controls that should be run prior to any NGS based study of RNA virus populations and provides a comprehensive data set for benchmarking future variant calling approaches.
 
@@
 
-The ability to reliably distinguish true from false single nucleotide variants (SNV) in next generation sequencing data is integral to accurate studies of viral diversity.
-Sensitivity is the proportion of true variants identified; while, specificity is the proportion of true negatives identified.
-$1-\text{specificity}$ is the false positive rate, the ratio of false positive variants to true negatives.
-Both a sensitivity and specificity of 1 describe perfect accuracy in which all the true variants are found and nothing else.
+The ability to reliably distinguish true from false single nucleotide variants (SNV) is integral to accurate NGS based studies of viral diversity.
+The accuracy of any SNV calling pipeline can be described in terms of sensitivity and specificity.
+Sensitivity is the proportion of true variants properly identified; while, specificity is the proportion of true negatives properly identified.
+Specificity is often reported as 1-specificity or the false positive rate.
+The false positive rate gives the ratio of false positive variants to true negatives.
+Both a sensitivity and specificity of 1 describe perfect accuracy in which all the true variants are found and only true variants are found.
 
@@
 Improved sensitivity often requires less stringent criteria in variant calling which reduces specificity.
-Conversely, increased stringency can improve specificity but reduce sensitivity.
-This relationship can be represented as a receiver operating characteristic (ROC) curve (`r xamp_roc`).
-An ROC plots the sensitivity of an assay along the y axis and 1-specificity along the x axis.
-To make an ROC curve for a variant calling algorithm one must test the assay on known data.
+Conversely, increased stringency can improve specificity but often reduces sensitivity.
+This relationship can be represented as a receiver operating characteristic (ROC) curve (`r Xamp_roc`).
+An ROC curve plots the sensitivity of an assay along the y axis and 1-specificity along the x axis.
+To make an ROC curve for a variant calling pipeline one must test the assay on known data.
 The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real; often this is a p value or quality score.
-In a controlled benchmakring experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut offs (`r xamp_roc` A).
-These points are then used to construct the curve (`r xamp_roc` B).
+In a controlled benchmakring experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut-offs (`r Xamp_roc` A).
+These points are then used to construct the curve (`r Xamp_roc` B).
 A perfect ROC curve in which all the true positives can be separated from all false positives is a right angle that follows the upper left perimeter of the plot.
@@
 A comprehensive comparison of SNV calling approaches is beyond the scope of this work.
-Here, we robustly benchmark one variant caller and highlight few means for improving accuracy.
+Here, we robustly benchmark one variant caller, DeepSNV, and highlight approaches for improving accuracy.
 In the process we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is being applied.
@@
 DeepSNV is a variant calling algorithm that uses a clonal, plasmid-derived, control to estimate local error rates across the genome.
-Because it is clonal, the sequence of the control is known with a high degree of confidence and so any nonconsensus base is indicative of a sequencing error.
-Additionally, the control and experimental samples are processed together and have identical noise characteristics.
+Because it is clonal, the sequence of the control is known with a high degree of confidence and so any nonconsensus base is indicative of an error.
+Additionally, the control and experimental samples are processed together and are assumed to have identical noise characteristics.
 DeepSNV then applies a hierarchical binomial model at each genomic position and identifies true variants as those with frequencies significantly above the noise found in the plasmid control.
-Like many variant calling method, the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification.
-We created defined mixtures of two influenza strains, WSN33 and PR8, to characterize the accuracy of DeepSNV when applied to RT-amplified RNA virus populations.
+Like many variant calling methods, the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification.
+We created defined mixtures of two influenza strains, WSN33 and PR8, to determine the accuracy of DeepSNV when applied to RT-PCR amplified RNA virus populations.
 
-WSN33 and PR8 genomes were mixed serially such that WSN33 was present at  frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq` A).
-Because these viruses differ at 491 positions, there are 491 true positives in each dilution.
+WSN33 and PR8 were plaque purified and sanger sequenced.
+cDNA genomes from both viruses were mixed serially such that WSN33 was present at frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq` A).
+These viruses differ at 491 positions (primer sites were excluded from analysis), providing 491 true positives in each dilution.
 On plasmids subjected to limited PCR, DeepSNV identified known variants at 0.1\% frequency with a sensitivity of 0.860 and  specificity of 1.0.
@@
 The specificity was above 0.998 in all dilutions.
-While this drop in specificity appears minor, it is important to keep in mind that even a false positive rate of 0.002 results in 78 false positives when applied across the `r miseq_vars` potential variants in the `r miseq_kb`kb of the influenza genome we analyzed.
+While this drop in specificity appears minor, it is important to keep in mind that even a false positive rate of 0.002 results in 78 false positives when applied across the `r miseq_vars` potential variants present in the `r miseq_kb`kb of the influenza genome we analyzed (3 variants per genomic position).
 
@@
 Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions.
-The virus populations found in patient-derived samples are  less divergent than WSN33 and PR8.
-Also the nucleic acid concentration in patient samples is much lower than that found samples grown in cell culture.
+The virus populations found in patient-derived samples are  less divergent than WSN33 and PR8 (cite).
+Also the nucleic acid concentration in patient samples is much lower than that found in samples grown in cell culture (data not shown).
 To mimic these conditions, we generated 20 viral clones, each with a single point mutation in the WSN33 background.
 We sequenced these mutants on the Illumina Miseq platform and confirmed that no additional mutations >1\% frequency had arisen between transfection and the passage 1 stock.
+Accross all 20 clones 4 additional mutations were found (frequencies 1.2%-3.7%).
 We also quantified the genome copy number of each stock using quantitative PCR.
@@
 This population was serially diluted into a stock of WSN33, generating artificial populations with all 20 mutations present a 2, 1, 0.5, 0.2, and 0.1\% frequency (`r Hiseq` A).
-We then serially diluted these populations into basal media to obtain populations with lower input.
+We then serially diluted these populations into basal media to obtain mixtures with lower nucleic acid input.
 The range ($10^3-10^5$ copies per microliter) matches the inputs typically found in patient-derived samples (data not shown).
@@
 The 20 mutations present in our initial viral mixture were the  only true positives considered in our analysis.
-Any variant that was present in both duplicates of the wild type stock was masked and  considered neither a true positive nor a true negative.
+Any variant that was present in both duplicates of the wild type stock or in one of the viral clones (>1.0%) was masked and  considered neither a true positive nor a true negative.
 This was done to avoid benchmarking our pipeline with variants identified by our pipeline.
@@
 Sensitivity dropped at the 0.5\% frequency, and in general was lower than that seen previously in the cell culture samples.
-Typically our samples grown in cell culture have  at least $10^8$ genomes per microliter; whereas, the nucleic acid concentration in these samples was 1,000 fold less .
-The reduced sensitivity is likely due to the fact that we expect only 7,000 of each variant genome in the 0.5\% frequency samples compared to the equivalent of 62,500,000 variant genomes in the 0.63\% sample from the cell culture experiment above.
-
-%This is a little tricky since we amplified the genomes first and then mixed them.
-
+Typically our samples grown in cell culture have  at least $10^8$ genomes per microliter; whereas, the nucleic acid concentration in these samples was lower by 1,000 fold.
+The reduced sensitivity is likely due to the fact that we expect only 7,000 copies of each variant genome in the 0.5\% frequency samples compared to the equivalent of 62,500,000 variant genomes in the 0.63\% sample from the cell culture experiment above.
 During library prep, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.
@@
 The specificity was greater than 0.99 in all the samples with $10^5$ genomes per microliter.
-The decreased input concentration and the accompanying increase in RT-PCR cycles is likely the cause of reduced specificity compared with that seen in the cell culture experiment.
-As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome.
+As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome (again primer sites were excluded).
 Increased specificity can be achieved by applying a more stringent p-value cutoff.
@@
 Mapping quality measures the probability that a given read is mapped to the correct position in the genome while base quality estimates the likelihood that the base call by the sequencer is correct.
-In our initial analysis we masked bases that had a base  score less than 30 (0.001 probability of being incorrect) and did not apply any mapping quality cut offs.
+In our initial analysis we masked bases that had a Phred  score less than 30 (0.001 probability of being incorrect) and did not apply any mapping quality cut-offs.
 We  sought to determine if we could improve our specificity by imposing more stringent MapQ and Phred cut-offs.
@@
 We further parsed our false variant calls by locating them within individual sequencing reads.
-It is well known that sequence quality drops near the end of a read, and we found that our false positives clustered to these regions (`r Qual` B).
-The average Phred score of these false positives was xx, further demonstrating that filtering on quality score alone is insufficient.
+It is well known that sequence quality drops near the end of a read (cite?), and we found that our false positives clustered to these regions (`r Qual` B).
+The average Phred score of these false positives was `r false_p_phred`, further demonstrating that filtering on quality score alone is insufficient.
 In contrast, true positives were uniformly distributed across the reads resulting in an average read position near the middle of the read.
@@
 We tried using the Benjamin Hochman p-value correction, utilizing more stringent p values or frequency cut-offs, retaining PCR duplicate reads, trimming the ends of the influenza genome, and employing various distributions to estimate the error rate in the control sample.
-All of these results have be summarized in an interactive shiny application available for download at $https://github.com/lauringlab/benchmarking_shiny.git$  We invite the reader to explore the affects of all these criteria on our data set.
+All of these results have be summarized in an interactive shiny application available for download at \emph{https://github.com/lauringlab/benchmarking_shiny.git}  We invite the reader to explore the affects of all these criteria on our data set.
 

### 2016-02-24 scripts/verifying_tp.R

@@ -1,0 +1,32 @@
+require(plyr)
+sum.1<-read.csv("~/Desktop/adams_data/SeptA.sum.csv",stringsAsFactors = F)
+sum.1$Id[sum.1$Id=="NA_36"]<-"NA_26" #corrects a typo in the file naming scheme
+sum.2<-read.csv("~/Desktop/adams_data/SeptB.sum.csv",stringsAsFactors = F)
+sum.2<-mutate(sum.2, Id=sub("-","_",Id))
+sum.3<-read.csv("~/Desktop/adams_data/JuneA.sum.csv",stringsAsFactors = F)
+sum.4<-read.csv("~/Desktop/adams_data/Shawn1.sum.csv",stringsAsFactors = F)
+sum.4<-mutate(sum.4, Id=sub("-","_",Id))
+tp<-read.csv("~/Desktop/Benchmarking_paper/data/reference/mutant_id.csv")
+
+sum.1$run="SeptA"
+sum.2$run="SeptB"
+sum.3$run="JuneA"
+sum.4$run="Shawn1"
+
+sum.df<-rbind(sum.1,sum.2)
+sum.df<-rbind(sum.df,sum.3)
+sum.df<-rbind(sum.df,sum.4)
+
+sum(unique(sum.df$Id) %in% tp$Id) # checks that we have all 20
+
+## now let's subset
+
+cut.df<-subset(sum.df, Id %in% tp$Id & p.val<0.01 & MapQ>30 & Phred>35 & Read_pos>62 & Read_pos<188 )
+cut.df<-mutate(cut.df,category=mutation %in% tp$mutant)
+
+
+subset(cut.df, category==F )
+
+#There are some overlaps so I'll remove those now.
+
+

### 2016-02-24 doc/draft.Rmd
added correlation coeff

@@
 Table<-"Table 1"
-table<-"table 1"
 
@@
 false_p_phred<-round(false_p_phred,digits=1)
+corr_coef<-round(corr_coef,digits=4)
+mean_fold<-round(mean_fold,digits=2)*100
+#print(interest)
 #csl: journal-of-virology.csl
+```
+
+```{r,echo=F}
+#Is it clear that I'm only talking about the 10^5 samples until we get to the lower input section
 ```
@@
 The ability to reliably distinguish true from false single nucleotide variants (SNV) is integral to accurate NGS based studies of viral diversity.
-The accuracy of any SNV calling pipeline can be described in terms of sensitivity and specificity.
-Sensitivity is the proportion of true variants properly identified; while, specificity is the proportion of true negatives properly identified.
+The accuracy of any SNV calling pipeline can be described in terms of its sensitivity and specificity.
+Sensitivity is the proportion of true variants that are properly identified; while, specificity is the proportion of true negatives that properly identified.
 Specificity is often reported as 1-specificity or the false positive rate.
 The false positive rate gives the ratio of false positive variants to true negatives.
-Both a sensitivity and specificity of 1 describe perfect accuracy in which all the true variants are found and only true variants are found.
+A sensitivity and specificity of 1 describe perfect accuracy in which all the true variants are found and only true variants are found.
 
@@
 An ROC curve plots the sensitivity of an assay along the y axis and 1-specificity along the x axis.
-To make an ROC curve for a variant calling pipeline one must test the assay on known data.
+A variant calling pipeline must be tested against known data in order to construct an ROC curve.
 The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real; often this is a p value or quality score.
@@
 A comprehensive comparison of SNV calling approaches is beyond the scope of this work.
-Here, we robustly benchmark one variant caller, DeepSNV, and highlight approaches for improving accuracy.
+Instead, we robustly benchmark one variant caller, DeepSNV, and highlight approaches for improving accuracy.
 In the process we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is being applied.
@@
 Like many variant calling methods, the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification.
-We created defined mixtures of two influenza strains, WSN33 and PR8, to determine the accuracy of DeepSNV when applied to RT-PCR amplified RNA virus populations.
+However, the accuracy had not been tested when applied to whole genome sequencing of RT-PCR amplified viral population.
+We created defined mixtures of two influenza strains, WSN33 and PR8, to determine the accuracy of DeepSNV under these conditions.
 
-WSN33 and PR8 were plaque purified and sanger sequenced.
+WSN33 and PR8 samples were plaque purified and sanger sequenced.
 cDNA genomes from both viruses were mixed serially such that WSN33 was present at frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq` A).
@@
 To mimic these conditions, we generated 20 viral clones, each with a single point mutation in the WSN33 background.
-We sequenced these mutants on the Illumina Miseq platform and confirmed that no additional mutations >1\% frequency had arisen between transfection and the passage 1 stock.
-Accross all 20 clones 4 additional mutations were found (frequencies 1.2%-3.7%).
+We sequenced these mutants on the Illumina Miseq platform to account for any additional mutations that might have arisen between transfection and the passage 1 stock.
+Four additional mutations were found above 1% frequency and were masked from analysis (frequencies 1.2%-3.7%).
 We also quantified the genome copy number of each stock using quantitative PCR.
@@
 
-The samples with the greatest input concentration and variant frequencies of 1\% and greater had the highest sensitivity (`r hiseq` B and C).
-Sensitivity dropped at the 0.5\% frequency, and in general was lower than that seen previously in the cell culture samples.
+The samples with the greatest input concentration and variant frequencies of 1\% and greater sensitivities of 0.850 or greater (`r Hiseq` B and C).
+Unlike what was seen in cell culture samples abover, sensitivity dropped drastically at 0.5\% frequency.
 Typically our samples grown in cell culture have  at least $10^8$ genomes per microliter; whereas, the nucleic acid concentration in these samples was lower by 1,000 fold.
 The reduced sensitivity is likely due to the fact that we expect only 7,000 copies of each variant genome in the 0.5\% frequency samples compared to the equivalent of 62,500,000 variant genomes in the 0.63\% sample from the cell culture experiment above.
-During library prep, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.
+During library preperation, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.
 
@@
 Increased specificity can be achieved by applying a more stringent p-value cutoff.
-However, as shown by the ROC curves in `r hiseq`, this move towards the y axis would markedly reduce sensitivity.
+However, as shown by the ROC curves in `r Hiseq`, this move towards the y axis would markedly reduce sensitivity.
 Our data demonstrate that with moderate concentrations of input nucleic acid, even statistically significant p-values from a robust variant caller are not sufficient to accurately separate true from false positive variants.
@@
 
-In addition to distinguishing true from false SNV, DeepSNV was able to estimate the frequency of the true positive variants with reasonable accuracy (`r Freq`).
-However, while the median frequencies are near the expected values, there is a fair spread in each sample (mean standard error of `r st_error`).
+In addition to distinguishing true from false SNV, DeepSNV was able to estimate the frequency of the true positive variants (`r Freq`).
+Although the median of the measured frequencies match the expetect values we found substantial spread about the median and the overall the fit was quite poor ( R^2^=`r corr_coef`).
+On average the percent difference between the measured frequency an the expected was `r mean_fold`\%.
 This error should be kept in mind when employing down stream analysis that depend on frequency measurements such as diversity metrics like Shannon's Entropy and Simpson's D, as well as inter-sample comparison of population structure .

### 2016-02-25 doc/draft.Rmd
updated tables and lofreq

@@
 #Is it clear that I'm only talking about the 10^5 samples until we get to the lower input section
+# verfiy # of Tp in 0.5% 10^5 sample
 ```
@@
 
-WSN33 and PR8 samples were plaque purified and sanger sequenced.
+WSN33 and PR8 samples were plaque purified and Sanger sequenced.
 cDNA genomes from both viruses were mixed serially such that WSN33 was present at frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq` A).
@@
 The samples with the greatest input concentration and variant frequencies of 1\% and greater sensitivities of 0.850 or greater (`r Hiseq` B and C).
-Unlike what was seen in cell culture samples abover, sensitivity dropped drastically at 0.5\% frequency.
+Unlike what was seen in cell culture samples above, sensitivity dropped drastically at 0.5\% frequency.
 Typically our samples grown in cell culture have  at least $10^8$ genomes per microliter; whereas, the nucleic acid concentration in these samples was lower by 1,000 fold.
 The reduced sensitivity is likely due to the fact that we expect only 7,000 copies of each variant genome in the 0.5\% frequency samples compared to the equivalent of 62,500,000 variant genomes in the 0.63\% sample from the cell culture experiment above.
-During library preperation, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.
+During library preparation, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.
 
@@
 In addition to distinguishing true from false SNV, DeepSNV was able to estimate the frequency of the true positive variants (`r Freq`).
-Although the median of the measured frequencies match the expetect values we found substantial spread about the median and the overall the fit was quite poor ( R^2^=`r corr_coef`).
-On average the percent difference between the measured frequency an the expected was `r mean_fold`\%.
+Although the median of the measured frequencies match the expect values, we found substantial spread in each sample and the overall the fit was quite poor ( R^2^=`r corr_coef`).
+The mean percent difference between the measured frequency an the expected was `r mean_fold`\%.
 This error should be kept in mind when employing down stream analysis that depend on frequency measurements such as diversity metrics like Shannon's Entropy and Simpson's D, as well as inter-sample comparison of population structure .
@@
 DeepSNV is one of many variant callers that use a combination of empiric and statistical approaches to model error rates and separate true from false variants.
-We asked whether the decreased accuracy observed in our data set was due simply to peculiarities specific to the DeepSNV algorithm.
+We asked whether the decreased accuracy observed in our data set was due simply to peculiarities specific to  DeepSNV.
 We therefore analyzed our $10^5$ input populations using Lofreq, another algorithm that is commonly used in viral next generation sequencing studies.
@@
 
-%%%% if the Errors are PCR related why does Lofreq remove them.
-
 ###Lower input levels
@@
 Host-derived viral populations vary in copy number and titer by several orders of magnitude.
-This variability can be attributed to a variety of factors including but not limited to: collection site, ease of of nucleic acid isolation, library preparation pipeline, and host and viral variability.
-To ensure accuracy across a range of input levels, we diluted our experimental populations serially into basal media and and called variants using our modified DeepSNV analysis pipeline (`r Lower`).
+This variability can be attributed to a variety of factors including but not limited to: collection site, ease of of nucleic acid isolation, library preparation, and host and viral variability.
+To ensure accuracy across a range of input levels, we diluted our experimental populations serially into basal media and and identified variants using our modified DeepSNV analysis pipeline (`r Lower`).
 As expected, the sensitivity was lower in populations with fewer genomes.
-A variant with a 0.5\% frequency in a $10^4$ genomes per microliter sample is expected to present on only 700 genomes in the initial RT-PCR and any number of these genomes could be lost due to bottlenecks in the amplification and library preparation process.
+A variant with a 0.5\% frequency in a $10^4$ genomes per microliter sample is expected to present on only 700 genomes in the initial RT-PCR and it is expected that many of these genomes will be lost due to bottlenecks in the amplification and library preparation process.
 We also found that specificity was reduced in samples with lower starting input.
@@
 This drastically improved our specificity and resulted in an accuracy that was comparable to that found in the $10^5$ samples (`r Dups`).
+Thus accurate analysis of low input RNA virus samples can be achieved through duplicate RT-PCR and careful benchmarking experiments.
+
+####The consequences of inaccurate SNV identification
+
+`r Table` stresses the importance of accurate SNV indentification in studies of viral diversity.
+We have apply three commonly used diversity metrics, (Richness, Shannon's Entropy, the inverse of Simpson's D), to our $10^5$ genomes per microliter populations.
 

### 2016-02-25 scripts/figures.Rmd
updated tables and lofreq

@@
 
-pdf("../results/figures/roc.example.pdf",width=3.4,height=4)
-plot.ex(0.35,0.95,roc.df,example)
-dev.off()
+# pdf("../results/figures/roc.example.pdf",width=3.4,height=4)
+# plot.ex(0.35,0.95,roc.df,example)
+# dev.off()
 # pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.ex.pdf",width = 12,height = 8)
@@
 }
-
-sum_roc<-function(x){
+sum_roc<-function(x,direction){
   sum.df<-subset(x,category %in% c(TRUE,FALSE),select=c(category,p.val))# get the TRUE and false variant calls and p.vals
@@
     }  
-    roc(sum.df$category~sum.df$p.val,plot=F,CI=T,direction=">")
+    roc(sum.df$category~sum.df$p.val,plot=F,CI=T,direction=direction)
   }
@@
 ######### Miseq roc functions ##########
-miseq.roc<-function(sum.df,possible_tp,possible_vars){ # A function to run the roc calculations and adjustments
-  roc.ls<-dlply(sum.df,~Id,sum_roc)
+miseq.roc<-function(sum.df,possible_tp,possible_vars,direction){ # A function to run the roc calculations and adjustments
+  roc.ls<-dlply(sum.df,~Id,sum_roc,direction)
 
@@
 }
-miseq.roc.table<-function(sum.df,cut.off,possible_tp,possible_vars){
-  roc.ls<-dlply(sum.df,~Id,sum_roc)
+miseq.roc.table<-function(sum.df,cut.off,possible_tp,possible_vars,direction){
+  roc.ls<-dlply(sum.df,~Id,sum_roc,direction)
   roc.df<-roc_df.one(roc.ls,cut.off)
@@
 ######### Hiseq roc functions ##########
-hiseq.roc<-function(sum.df,possible_tp,possible_vars){ # A function to run the roc calculations and adjustments
-  roc.ls<-dlply(sum.df,~Id,sum_roc)
+hiseq.roc<-function(sum.df,possible_tp,possible_vars,direction){ # A function to run the roc calculations and adjustments
+  roc.ls<-dlply(sum.df,~Id,sum_roc,direction)
 
@@
 
-hiseq.roc.table<-function(sum.df,cut.off,possible_tp,possible_vars){
-  roc.ls<-dlply(sum.df,~Id,sum_roc)
+hiseq.roc.table<-function(sum.df,cut.off,possible_tp,possible_vars,direction){
+  roc.ls<-dlply(sum.df,~Id,sum_roc,direction)
   roc.df<-roc_df.one(roc.ls,cut.off)
@@
 possible_vars<-sum(regions.bed$length)*3 # three possible variants/ position
+miseq_vars<-sum(regions.bed$length)*3
+miseq_kb<-sum(regions.bed$length)/1000
 
@@
 
-m.roc.df<-miseq.roc(sum.df,possible_tp,possible_vars)
+m.roc.df<-miseq.roc(sum.df,possible_tp,possible_vars,">")
 
@@
 
-m.roc.table<-miseq.roc.table(sum.df,0.01,possible_tp,possible_vars)
+m.roc.table<-miseq.roc.table(sum.df,0.01,possible_tp,possible_vars,">")
 
@@
 
-pdf("../results/figures/miseq1.pdf",width=3.4,height=7)
-ggdraw() +
-  draw_plot(miseq.setup, 0.2, .66, .6, 0.33) + #x y width height
-  draw_plot(miseq.roc.p, 0.05, 0.33, .9, 0.33) +
-  draw_plot(tbl, 0.05, 0, .9, 0.33) +
-  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(0.95, 0.7, 0.35), size = 15)
-dev.off()
+# pdf("../results/figures/miseq1.pdf",width=3.4,height=7)
+# ggdraw() +
+#   draw_plot(miseq.setup, 0.2, .66, .6, 0.33) + #x y width height
+#   draw_plot(miseq.roc.p, 0.05, 0.33, .9, 0.33) +
+#   draw_plot(tbl, 0.05, 0, .9, 0.33) +
+#   draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(0.95, 0.7, 0.35), size = 15)
+# dev.off()
 
@@
 read.df<-subset(sum.df,Read_pos>63 & Read_pos<188)
-m.roc.df<-miseq.roc(read.df,possible_tp,possible_vars)
+m.roc.df<-miseq.roc(read.df,possible_tp,possible_vars,">")
 
@@
 
-m.roc.table<-miseq.roc.table(read.df,0.01,possible_tp,possible_vars)
+m.roc.table<-miseq.roc.table(read.df,0.01,possible_tp,possible_vars,">")
 knitr::kable(m.roc.table)
@@
 
-h.roc.df<-hiseq.roc(sum.df,possible_tp,possible_vars)
+hiseq_vars<-possible_vars
+hiseq_kb<-sum(regions.bed$length)/1000
+
+h.roc.df<-hiseq.roc(sum.df,possible_tp,possible_vars,">")
 
@@
 
-h.table.df<-hiseq.roc.table(sum.df,0.01,possible_tp,possible_vars)
+h.table.df<-hiseq.roc.table(sum.df,0.01,possible_tp,possible_vars,">")
 
@@
 
-pdf("../results/figures/hiseq1.pdf",width=3.4,height=7)
+# pdf("../results/figures/hiseq1.pdf",width=3.4,height=7)
+# ggdraw() +
+#   draw_plot(hiseq.setup, 0.1, .57, 0.8, 0.43) + #x y width height
+#   draw_plot(Hiseq.roc, 0.05, 0.33, .8, 0.33) +
+#   draw_plot(tbl, 0.05, 0, .9, 0.33) +
+#   draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(1, 0.7, 0.35), size = 15)
+# dev.off()
+
 ggdraw() +
@@
   draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(1, 0.7, 0.35), size = 15)
-dev.off()
-
-ggdraw() +
-  draw_plot(hiseq.setup, 0.1, .57, 0.8, 0.43) + #x y width height
-  draw_plot(Hiseq.roc, 0.05, 0.33, .8, 0.33) +
-  draw_plot(tbl, 0.05, 0, .9, 0.33) +
-  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(1, 0.7, 0.35), size = 15)
 # 
@@
 
-cut.df<-subset(sum.df,MapQ>30 & Phred>35) # This removes the variants that don't pass
+#cut.df<-subset(sum.df,MapQ>30 & Phred>35) # This removes the variants that don't pass
 
@@
 
-sig.cut.df<-subset(cut.df,p.val<0.01)
+sig.cut.df<-subset(cut.df.all,p.val<0.01)
 
@@
 
+fp_ends<-subset(sig.cut.df, category==F & Read_pos<32 | Read_pos>94)
+false_p_phred<-mean(fp_ends$Phred)
+
 #cut.read.df<-subset(cut.df,Read_pos>32 & Read_pos<94) # This removes the variants that don't pass
@@
 
-h.cut.read.df<-hiseq.roc(cut.read.df.all,possible_tp,possible_vars)
+h.cut.read.df<-hiseq.roc(cut.read.df.all,possible_tp,possible_vars,">")
 
@@
 
-h.cut.read.table<-hiseq.roc.table(cut.read.df.all,0.01,possible_tp,possible_vars)
+h.cut.read.table<-hiseq.roc.table(cut.read.df.all,0.01,possible_tp,possible_vars,">")
 #knitr::kable(subset(h.cut.read.table,gc==5))
@@
 
-pdf("../results/figures/figure4.pdf",width=3.4,height=9)
-ggdraw() +
-  draw_plot(mapq_phred.normal,0.03, 0.77, .85, 0.23) + 
-  draw_plot(read.pos, 0.03, .53, .85, .25) +
-  draw_plot(read.roc, 0.03, 0.23, 0.85, .3) +
-  draw_plot(tbl,   0.025, 0, 0.95, .25) +
-  draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(1, 0.78, 0.55,0.25), size = 15)
-dev.off()
+# pdf("../results/figures/figure4.pdf",width=3.4,height=9)
+# ggdraw() +
+#   draw_plot(mapq_phred.normal,0.03, 0.77, .85, 0.23) + 
+#   draw_plot(read.pos, 0.03, .53, .85, .25) +
+#   draw_plot(read.roc, 0.03, 0.23, 0.85, .3) +
+#   draw_plot(tbl,   0.025, 0, 0.95, .25) +
+#   draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(1, 0.78, 0.55,0.25), size = 15)
+# dev.off()
 
@@
 
-lofreq.df<-read.csv("../data/process/2015-6-23/lofreq/all.removed.mapq_vcf.vcf_csv.csv")
+lofreq.df<-read.csv("../data/process/2015-6-23/lofreq/all.removed.mapq_vcf.vcf_csv.csv",stringsAsFactors = F)
 
@@
 
-lofreq.roc<-hiseq.roc(cut.df,possible_tp,possible_vars)
+direction<-">"
+lofreq.roc<-hiseq.roc(lofreq.df,possible_tp,possible_vars,"<")
 
@@
 
-x<-mutate(x,frequency=find_freq.v.h(exp.freq))
+x<-mutate(x,frequency=as.character(find_freq.v.h(exp.freq)))
 
@@
 
-x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.04
-x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]-0.12
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]+0.02
+x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]-0.02
 
@@
 
-pdf("../results/figures/lofreq1.pdf",width=3.4,height=5)
+# pdf("../results/figures/lofreq1.pdf",width=3.4,height=5)
+# ggdraw() +
+#   draw_plot(lofreq.roc.p, 0, 0.5, 0.9, 0.45) + 
+#   draw_plot(tbl, 0.025, 0, 0.95, .48) +
+#   draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
+# dev.off()
+
 ggdraw() +
@@
   draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
-dev.off()
-
-ggdraw() +
-  draw_plot(lofreq.roc.p, 0, 0.5, 0.9, 0.45) + 
-  draw_plot(tbl, 0.025, 0, 0.95, .48) +
-  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
-
+direction<-"<"
 ```
@@
 
+### here weget the R2 of the sample
+
+model<-lm(formula = freq.var~exp.freq,data = subset(cut.read.sig.df,category==T & gc=="5"))
+
+corr_coef<-summary(model)$adj.r.squared
+
+interest<-subset(cut.read.sig.df,category==T & gc=="5")
+
+mean_fold<-mean(abs(interest$freq.var-interest$exp.freq)/interest$exp.freq)
+
+print(mean_fold)
+
 ac_freq<-ggplot(subset(cut.read.sig.df,category==T & gc=="5"),aes(x=exp.freq,y=freq.var,fill=factor(exp.freq,levels=rev(levels(as.factor(exp.freq))))))
@@
 
-pdf("../results/figures/freq.pdf",width=3.4,height=3.4)
-ac_freq
-dev.off()
+# pdf("../results/figures/freq.pdf",width=3.4,height=3.4)
+# ac_freq
+# dev.off()
 
@@
 
-pdf("../results/figures/lower.pdf",width=3.4,height=9)
-ggdraw() +
-  draw_plot(roc.4,0.03, 0.69, .85, 0.25) + 
-  draw_plot(tbl.4, 0.025, .47,  0.95, .25) +
-  draw_plot(roc.3, 0.03, 0.22, 0.85, .25) +
-  draw_plot(tbl.3,   0.025, 0, 0.95, .25) +
-  draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(0.97, 0.73, 0.5,0.23), size = 15)
-dev.off()
+# pdf("../results/figures/lower.pdf",width=3.4,height=9)
+# ggdraw() +
+#   draw_plot(roc.4,0.03, 0.69, .85, 0.25) + 
+#   draw_plot(tbl.4, 0.025, .47,  0.95, .25) +
+#   draw_plot(roc.3, 0.03, 0.22, 0.85, .25) +
+#   draw_plot(tbl.3,   0.025, 0, 0.95, .25) +
+#   draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(0.97, 0.73, 0.5,0.23), size = 15)
+# dev.off()
 
@@
 
-dups.roc<-hiseq.roc(dups,possible_tp,possible_vars)
+dups.roc<-hiseq.roc(dups,possible_tp,possible_vars,">")
 
@@
 
-dups.table<-hiseq.roc.table(dups,0.01,possible_tp,possible_vars)
+dups.table<-hiseq.roc.table(dups,0.01,possible_tp,possible_vars,">")
 #knitr::kable(subset(dups.table))
@@
 
-pdf("../results/figures/dups.roc.pdf",width=3.4,height=5)
-ggdraw() +
-  draw_plot(dups.roc.p, 0, 0.5, 0.9, 0.45) + 
-  draw_plot(tbl, 0.025, 0, 0.95, .48) +
-  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
-dev.off()
+# pdf("../results/figures/dups.roc.pdf",width=3.4,height=5)
+# ggdraw() +
+#   draw_plot(dups.roc.p, 0, 0.5, 0.9, 0.45) + 
+#   draw_plot(tbl, 0.025, 0, 0.95, .48) +
+#   draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
+# dev.off()
 
@@
 ```{r}
+mut_pat="(.*)_[A,T,C,G](\\d+)[A,T,C,G]"
 perfect<-data.frame(mutation=rep(true_snv$mutant,times=4),freq.var=rep(c(0.05,0.02,0.01,0.005),each=20))
+perfect=mutate(perfect,chr=gsub(pattern = mut_pat,replacement = "\\1",x=mutation),pos=gsub(pattern = mut_pat,replacement = "\\2",x=mutation))
 dups<-subset(dups,p.val<0.01)
@@
 H<-function(x){
-  p=x$freq.var*log(x$freq.var)
-  -sum(p)
+  H_pos<-ddply(x,~chr+pos,summarize,wt=1-sum(freq.var),H=-sum(c(freq.var*log(freq.var),wt*log(wt))))
+sum(H_pos$H)/(possible_vars/3)
 }
@@
 D<-function(x){
-  p=(x$freq.var)^2
-  1/sum(p)
+  D_pos<-ddply(x,~chr+pos,summarize,wt=1-sum(freq.var),D=1/sum(c(freq.var^2,wt^2)))
+  others=(possible_vars/3)-dim(D_pos)[1]
+  sum(D_pos$D,others)/(possible_vars/3)
 }
@@
   ref$Id="ref"
-  all<-rbind(subset(x,select=c(mutation,freq.var,Id)),ref)
+  all<-rbind(subset(x,select=c(mutation,freq.var,Id)),subset(ref,select=c(mutation,freq.var,Id)))
   tab<-as.matrix(subset(dcast(all,mutation ~ Id, value.var = "freq.var"),select=-c(mutation)))
   tab[is.na(tab)]<-0
-  as.dist(dist(tab))
+  dist(t(tab),"manhattan")[1]
 }
@@
   if(length(unique(x$freq.var))==1){
-    exp.freq=unique(x$freq.var)
+    exp.freq=unique(x$freq.var) # this is the perfect condition where there is only 1 frequency and it is all the same
   }else{
@@
   }
-  out<-data.frame(Richness=rich(x), Entropy=H(x),Simpson=D(x))#,dist=euc(x,subset(perfect,freq.var==exp.freq))[1])
+  out<-data.frame(Richness=round(rich(x),digits = 0), Entropy=format(H(x),scientific = T,digits = 3),"1/D"=D(x),Manhattan=round(euc(x,subset(perfect,freq.var==exp.freq)),digits=3))
   rownames(out)<-setting
@@
 get_data(subset(perfect,freq.var==freq),"Perfect")->perf
-get_data(subset(sum.df,gc==input & exp.freq==freq),"DeepSNV")->deepSNV
+get_data(subset(sum.df,gc==input & exp.freq==freq & p.val<0.01),"DeepSNV")->deepSNV
 get_data(subset(lofreq.df,gc==input & exp.freq==freq),"Lofreq")->lo
-get_data(subset(cut.read.sig.df,gc==input & exp.freq==freq),"DeepSNV Final")->deepSNV.f
-knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f),digits = 2)
-}
-tabulate.4<-function(input,freq){
-get_data(subset(perfect,freq.var==freq),"Perfect")->perf
-get_data(subset(sum.df,gc==input & exp.freq==freq),"DeepSNV")->deepSNV
-get_data(subset(lofreq.df,gc==input & exp.freq==freq),"Lofreq")->lo
-get_data(subset(cut.read.sig.df,gc==input & exp.freq==freq),"DeepSNV Final")->deepSNV.f
-get_data(subset(dups,exp.freq==freq),"DeepSNV duplicates")->deepSNV.d
-
-knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f,deepSNV.d),digits = 2)
-}
+get_data(subset(cut.read.df.all,gc==input & exp.freq==freq & p.val<0.01),"DeepSNV Final")->deepSNV.f
+knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f))
+}
+# tabulate.4<-function(input,freq){
+# get_data(subset(perfect,freq.var==freq),"Perfect")->perf
+# get_data(subset(sum.df,gc==input & exp.freq==freq),"DeepSNV")->deepSNV
+# get_data(subset(lofreq.df,gc==input & exp.freq==freq),"Lofreq")->lo
+# get_data(subset(cut.read.sig.df,gc==input & exp.freq==freq),"DeepSNV Final")->deepSNV.f
+# get_data(subset(dups,exp.freq==freq),"DeepSNV duplicates")->deepSNV.d
+# 
+# knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f,deepSNV.d),digits = 2)
+# }
 
@@
 ```
-
-$10^4$ 5%
-
-```{r}
-tabulate.4(4,0.05)
-```
-
-$10^4$ 1%
-
-```{r}
-tabulate.4(4,0.01)
-```
-
-$10^4$ 0.5%
-
-```{r}
-tabulate.4(4,0.005)
-```

### 2016-03-16 doc/draft.Rmd
starting methods

@@
     reference_docx: style.docx
-bibliography: bib.bib
-csl: journal-of-virology.csl
 
@@
 #this is so the cache resets
+# 
+# bibliography: bib.bib
+# csl: journal-of-virology.csl
+
 ```
@@
 miseq_vars<-format(miseq_vars,scientific=F)
-miseq_kb<-round(miseq_kb,digits=2)
+miseq_kb<-format(miseq_kb*1000,scientific = F)
 
 hiseq_vars<-format(hiseq_vars,scientific=F)
-hiseq_kb<-round(hiseq_kb,digits=2)
+hiseq_kb<-format(hiseq_kb*1000,scientific = F)
 
@@
 mean_fold<-round(mean_fold,digits=2)*100
+read_depth<-"a lot"
 #print(interest)
@@
 ```{r,echo=F}
-#Is it clear that I'm only talking about the 10^5 samples until we get to the lower input section
-# verfiy # of Tp in 0.5% 10^5 sample
+ ####################### To do List #######################
+# Is it clear that I'm only talking about the 10^5 samples until we get to the lower input section?
+
+# verfy # of Tp in 0.5% 10^5 sample - done
+
+# verity that the possible variants in each case accounts for the wt and carry over variants - done
+
+#Mapq exploration - concordant vs nonconcordant pairs
+
+# add mfe raw data to repo. currently pulls data from desktop directory.
 ```
@@
 
-Until recently, our understanding of intrahost viral evolution was limited by high-effort, low-throughput sequencing methods.
-Much of what is known regarding \emph{in vivo} viral evolution has been derived from epidemiological studies (possibly cite examples, flu foot and mouth ect).
-However, with the advent of next generation sequencing (NGS) it is now feasible to detect rare single nucleotide variants present in hundreds of patient-derived samples (cite examples?).
-Next generation sequencing offers a high resolution glimpse of intrahost viral dynamics, and there has been an explosion of papers that apply these highly sensitive and cost effective  techniques to viral populations from a myriad of sample sources (cite examples?).
-Although NGS is highly sensitive, its specificity in the context of intrahost viral populations has not been robustly measured.
-Each sequencing platform has inherent error profiles, and sample collection, target amplification, and library preparation are additional processes whereby errors can be introduce and propagated.
+The advent of next generation sequencing (NGS) is revolutionizing the study of viral evolution and molecular epidemiology through high-throughput measures of intrahost diversity.
+Until recently investigations of viral diversity were limited by labor intensive sequencing approaches; however, it is now feasible to sequence patient-derived samples at sufficient read depth to detect rare single nucleotide variants (SNV) (cite examples?).
+The high-throughput, cost-effective nature of NGS has lead to an explosion of studies that quantify viral diversity within and between hosts (cite examples?).
+Given the increasing contribution of NGS to the field of virology it is important to reflect on the accuracy of these methods and the controls that are needed to ensure robust conclusions.
+
+Because next generation sequencing of viral populations often produces tens of thousands of reads per genomic position, the sensitivity of these approaches has been assumed to be quite robust (cite).
+However, patient derived samples often differ in input titer by orders of magnitude, and this variability can have drastic impacts on the sensitivity of detecting rare events.
+Furthermore, some library preparations result in variable sequencing coverage both within and between samples.
+These sources of ascertainment bias are often  overlooked in NGS based studies of viral diversity.
+
+The specificity of NGS is another factor to consider when designing an experiment using patient-derived samples.
+Each sequencing platform has inherent error profiles (cite), and sample collection, target amplification, and library preparation are additional processes whereby errors can be introduced and propagated.
 The next generation sequencing of viral populations, in particular those of RNA viruses, is error prone, and these errors must be accounted for to ensure accurate secondary analysis.
@@
 For example, Cirseq is an ingenious approach in which  template RNA is sheared and circularized prior to reverse transcription  [@Acevedo:2014ej].
-Rolling circle cDNA synthesis then produces tandem reads of the same RNA fragment.
-Consensus sequences are assigned to each RNA fragment controlling for reverse transcription, PCR, and sequencing errors.
-However, the sensitivity of the method has not been robustly benchmarked against known rare variants.
-Additionally, Cirseq requires 1 microgram of starting RNA and provides no mechanism for targeting sequencing to viral RNA [@Acevedo:2014df].
-As such, the method has limited application to patient-derived samples that are often characterized by low nucleic acid concentrations and are enriched for host RNA.
+Rolling circle cDNA synthesis produces tandem reads, generating a consensus sequence for each RNA fragment.
+While this method is likely to be highly sensitive for rare variant detection and can control for reverse transcription, PCR, and sequencing errors, the requirement for a large and relatively pure population of viral RNA limits its applicability to patient derived samples [@Acevedo:2014df].
 
-Alternatively, bar-coded primers can be used to construct consensus sequences for each cDNA template and control for PCR and sequencing errors [@Jabara:2011eo].
-This approach requires less input and targets sequencing to the viral genome, but unlike Cirseq, it is unable to account for RT errors present in the cDNA.
-Additionally, primer Id based methods require each bar code be physically attached to a PCR product, and so the region of the genome that can be targeted is limited by the read length of the sequencing platform (250-500 bp for Illumina).
-Primer ID based methods are useful in interrogating specific regions of a genome (cite examples?), but have limited application in whole genome sequencing.
-An adapted form of Sequence independent single primer amplification (SISPA) is an alternative approach that allows for whole genome sequencing and controls for errors propagated during library preparation [@Djikeng:2008cd].
-RT-PCR products are sheared and tagged with bar-coded random primers in a klenow reaction, prior to library preparation.
-SISPA controls for any errors that may arise during the library amplification and controls for PCR duplicate frequency biases.
+"Primer Id" based methods, on the other hand, require less input and target sequencing to the viral genome.
+These methods use bar-coded primers to construct consensus sequences for each cDNA template and can control for PCR and sequencing errors  [@Jabara:2011eo].
+Primer Id based methods require each bar code be physically attached to a PCR product and are most easily applied to a small targeted regions of the genome.
+As such, they have limited application in whole genome sequencing.
+
+An adapted form of sequence independent single primer amplification (SISPA) is an alternative approach that allows for whole genome sequencing and controls for errors propagated during library preparation [@Djikeng:2008cd].
+In this method RT-PCR products are sheared and tagged with bar-coded random primers in a klenow reaction, prior to library preparation.
+SISPA controls for any errors that may arise during the library amplification and controls for PCR biases.
+This method has been used in conjunction with statistical algorithms to control for accuracy in studies of intrahost influenza diversity (cite).
 However, the bar-coding reaction used in SISPA is biased and results in uneven coverage and sensitivity across the genome  [@Rosseel:2013dn].
 
-Given the limitations described above, a number of statistical approaches have been developed to distinguish true variants from sequencing errors using only sequencing data [@Gerstung:1js; @Isakov:2015gq; @Wilm:2012br; @Macalalad:2012fx; @Koboldt:2009dk].
-In general these algorithms calibrate base specific error rates across the genome.
-The rates are determined by various metrics including but not limited to mapping quality, phred score, nucleotide position on a read, and sequence context.
+Given the above limitations, many statistical approaches have been developed to distinguish true variants from sequencing errors [@Gerstung:1js; @Isakov:2015gq; @Wilm:2012br; @Macalalad:2012fx; @Koboldt:2009dk].
+In general, these algorithms calculate base specific error rates from various metrics including but not limited to: mapping quality (MapQ), base quality (Phred), strand bias, and sequence context.
 True variants are identified as those with frequencies exceeding the expected error rate according to some predetermined statistical test.
-Despite being employed in many NGS based studies of viral diversity (cite?) few variant calling algorithms have been benchmarked using defined viral populations, and to our knowledge none have been tested under patient-derived sample conditions.
-The accuracy of such algorithms in the context of NGS studies of patient-derived RNA viral populations is not known.
+Despite being employed in many NGS based studies of viral diversity (cite?), few of these algorithms have been benchmarked using defined viral populations, and to our knowledge none have been tested under conditions that mimic those found in patient-derived samples.
+The accuracy of such algorithms in the context of NGS studies of patient-derived viral populations is not known.
 
-We are interested in using NGS to study the intrahost dynamics of RNA viruses.
-Because our experimental design relies on whole genome sequencing of patient-derived samples, we are unable to use any of the above error-controlling sample preparation pipelines.
-We, are left with only statistical algorithms to control for errors.
-However, the accuracy of these algorithms under patient-derived sample conditions is unknown.
-Here, we use defined populations of Influenza A to robustly benchmark the accuracy of a variant calling pipeline under patient-derived conditions.
-In our analysis, samples are enriched for influenza genomes through a multiplex RT-PCR reaction, sheared using Covaris sonication, sequenced on the Illumina platform, and analyzed using the DeepSNV algorithm and in house scripts (all available at \emph{https://github.com/lauringlab/variant_pipeline})[@Gerstung:1js].
-In this work, we highlight the challenges that accompany NGS based studies of viral diversity and  make some suggestions for improving the accuracy of our approach.
-This work exemplifies the quality controls that should be run prior to any NGS based study of RNA virus populations and provides a comprehensive data set for benchmarking future variant calling approaches.
+Here, we use defined populations of Influenza A and variable input titers to indirectly determine the accuracy of rare variant detection in patient-derived samples.
+We highlight the challenges that accompany NGS based studies of viral diversity and include a few means for improved the accuracy.
+This work exemplifies the controls that should be run prior to any NGS based study of viral populations and provides a comprehensive data set for benchmarking other pipelines.
 
@@
 The accuracy of any SNV calling pipeline can be described in terms of its sensitivity and specificity.
-Sensitivity is the proportion of true variants that are properly identified; while, specificity is the proportion of true negatives that properly identified.
-Specificity is often reported as 1-specificity or the false positive rate.
-The false positive rate gives the ratio of false positive variants to true negatives.
+Sensitivity is the proportion of true variants that are properly identified; while, specificity is the proportion of true negatives that are properly identified.
+A closely related concept is the false positive rate, which is 1-specificity, and gives the ratio of false positive variants to true negatives.
 A sensitivity and specificity of 1 describe perfect accuracy in which all the true variants are found and only true variants are found.
@@
 Conversely, increased stringency can improve specificity but often reduces sensitivity.
-This relationship can be represented as a receiver operating characteristic (ROC) curve (`r Xamp_roc`).
+This relationship can be visualized using a receiver operating characteristic (ROC) curve (`r Xamp_roc`).
 An ROC curve plots the sensitivity of an assay along the y axis and 1-specificity along the x axis.
 A variant calling pipeline must be tested against known data in order to construct an ROC curve.
-The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real; often this is a p value or quality score.
-In a controlled benchmakring experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut-offs (`r Xamp_roc` A).
-These points are then used to construct the curve (`r Xamp_roc` B).
+The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real, often a p value or quality score.
+In a controlled benchmarking experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut-offs (`r Xamp_roc`A).
+These points are then used to construct the curve (`r Xamp_roc`B).
 A perfect ROC curve in which all the true positives can be separated from all false positives is a right angle that follows the upper left perimeter of the plot.
@@
 A comprehensive comparison of SNV calling approaches is beyond the scope of this work.
-Instead, we robustly benchmark one variant caller, DeepSNV, and highlight approaches for improving accuracy.
-In the process we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is being applied.
+Instead, we robustly benchmark one variant caller, DeepSNV, and highlight approaches for improving its accuracy.
+In doing so we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is being applied.
 
-DeepSNV is a variant calling algorithm that uses a clonal, plasmid-derived, control to estimate local error rates across the genome.
-Because it is clonal, the sequence of the control is known with a high degree of confidence and so any nonconsensus base is indicative of an error.
+DeepSNV is a variant calling algorithm that uses a clonal, plasmid-derived, control to estimate local error rates across the genome [@Gerstung:1js].
+Because it is clonal, the sequence of the control is known with a high degree of confidence, and any nonconsensus base is indicative of an error.
 Additionally, the control and experimental samples are processed together and are assumed to have identical noise characteristics.
@@
 Like many variant calling methods, the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification.
-However, the accuracy had not been tested when applied to whole genome sequencing of RT-PCR amplified viral population.
-We created defined mixtures of two influenza strains, WSN33 and PR8, to determine the accuracy of DeepSNV under these conditions.
+However, the accuracy has not been tested when applied to whole genome sequencing of a viral population amplified by reverse transcription PCR (RT-PCR).
 
-WSN33 and PR8 samples were plaque purified and Sanger sequenced.
-cDNA genomes from both viruses were mixed serially such that WSN33 was present at frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq` A).
-These viruses differ at 491 positions (primer sites were excluded from analysis), providing 491 true positives in each dilution.
-On plasmids subjected to limited PCR, DeepSNV identified known variants at 0.1\% frequency with a sensitivity of 0.860 and  specificity of 1.0.
-Under our experimental conditions and with a slightly more stringent p value  ( p=0.01 vs.
+In our first benchmarking data set we created defined mixtures of two influenza strains,WSN33 and PR8.
+We plaque purified and expanded WSN33 and PR8 populations.
+Complementary DNA from both viruses were mixed serially such that WSN33 cDNA was present at frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq`A).
+These viruses differ at 491 positions (primer sites used in RT-PCR were excluded from analysis), providing 491 true positives in each dilution.
+On plasmids subjected to limited PCR, DeepSNV identified known variants at 0.1\% frequency with a sensitivity of 0.860 and  specificity of 1.0 [@Gerstung:1js].
+Under our experimental conditions and with a slightly more stringent p value (p=0.01 vs.
 p=0.05) we found a moderate reduction in sensitivity  (0.830 for variants at 0.63\% and 0.14 for variants at 0.16\%).
 The specificity was above 0.998 in all dilutions.
-While this drop in specificity appears minor, it is important to keep in mind that even a false positive rate of 0.002 results in 78 false positives when applied across the `r miseq_vars` potential variants present in the `r miseq_kb`kb of the influenza genome we analyzed (3 variants per genomic position).
+While the drop in specificity appears small, a false positive rate of 0.002 corresponds to 78 false positives when applied across the `r miseq_vars` potential variants present in the `r miseq_kb` bp of the influenza genome we analyzed.
 
@@
 Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions.
-The virus populations found in patient-derived samples are  less divergent than WSN33 and PR8 (cite).
-Also the nucleic acid concentration in patient samples is much lower than that found in samples grown in cell culture (data not shown).
+Patient-derived populations are typically less divergent than WSN33 and PR8 (cite?), and the number of viral genomes in patient samples is much lower than in cell-culture derived populations (data not shown).
 To mimic these conditions, we generated 20 viral clones, each with a single point mutation in the WSN33 background.
 We sequenced these mutants on the Illumina Miseq platform to account for any additional mutations that might have arisen between transfection and the passage 1 stock.
-Four additional mutations were found above 1% frequency and were masked from analysis (frequencies 1.2%-3.7%).
-We also quantified the genome copy number of each stock using quantitative PCR.
+Four additional mutations were found above 1% frequency (frequencies 1.2%-3.7%).
+We also determined the genome copy number of each stock using quantitative RT-PCR.
 We then mixed equal genome equivalents of these 20 viruses, to generate an artificial population with $10^5$ copies per microliter with each mutation  present at 5\% frequency.
@@
 The range ($10^3-10^5$ copies per microliter) matches the inputs typically found in patient-derived samples (data not shown).
-We sequenced the samples on the Illumina Hiseq platform, and variants were called using DeepSNV.
-We also processed and sequenced the WSN33 stock in duplicate to control for any background mutations.
+We sequenced the samples on the Illumina Hiseq platform, and called variants using DeepSNV.
+We also processed and sequenced the WSN33 stock in duplicate to control for any mutations in the viral dilutent.
 
 The 20 mutations present in our initial viral mixture were the only true positives considered in our analysis.
-Any variant that was present in both duplicates of the wild type stock or in one of the viral clones (>1.0%) was masked and  considered neither a true positive nor a true negative.
-This was done to avoid benchmarking our pipeline with variants identified by our pipeline.
+Any SNV that was present in both duplicates of the wild type stock or in one of the viral clones (>1.0%) was masked and  considered neither a true positive nor a true negative.
 
-The samples with the greatest input concentration and variant frequencies of 1\% and greater sensitivities of 0.850 or greater (`r Hiseq` B and C).
-Unlike what was seen in cell culture samples above, sensitivity dropped drastically at 0.5\% frequency.
-Typically our samples grown in cell culture have  at least $10^8$ genomes per microliter; whereas, the nucleic acid concentration in these samples was lower by 1,000 fold.
-The reduced sensitivity is likely due to the fact that we expect only 7,000 copies of each variant genome in the 0.5\% frequency samples compared to the equivalent of 62,500,000 variant genomes in the 0.63\% sample from the cell culture experiment above.
+In these populations with lower diversity and input titer, we maintained greater than 0.85 sensitivity for SNV at 1% frequency or higher.
+Our sensitivity was considerably lower for variants at or below 0.5% frequency (`r Hiseq`B and C).
+We observed an average read depth of over 10,000 reads/bp in these samples.
+Therefore, the drop in sensitivity is most likely due to the 1,000-fold decrease in nucleaic acid concentration compared to the first data set and the fact that library preparation requires a number of sampling steps that may limit detection.
 During library preparation, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.
@@
 The specificity was greater than 0.99 in all the samples with $10^5$ genomes per microliter.
-As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome (again primer sites were excluded).
-Increased specificity can be achieved by applying a more stringent p-value cutoff.
-However, as shown by the ROC curves in `r Hiseq`, this move towards the y axis would markedly reduce sensitivity.
+As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome.
+The false positives outnumber the true positives by 100-fold in this population with realistic diversity and input.
+We were able to increased our specificity  by applying a more stringent p-value cutoff.
+However, as shown by the ROC curves in `r Hiseq`, this move towards the y axis markedly reduces sensitivity.
 Our data demonstrate that with moderate concentrations of input nucleic acid, even statistically significant p-values from a robust variant caller are not sufficient to accurately separate true from false positive variants.
@@
 
-Many next generation sequencing studies utilize mapping (MapQ) and base (Phred) quality cut-offs to ensure that only the highest caliber sequencing data is used to call variants.
+Many next generation sequencing studies utilize mapping quality (MapQ) and/or base quality (Phred) cut-offs to ensure that only the highest caliber sequencing data is used to call variants.
 Mapping quality measures the probability that a given read is mapped to the correct position in the genome while base quality estimates the likelihood that the base call by the sequencer is correct.
-In our initial analysis we masked bases that had a Phred  score less than 30 (0.001 probability of being incorrect) and did not apply any mapping quality cut-offs.
-We  sought to determine if we could improve our specificity by imposing more stringent MapQ and Phred cut-offs.
-Initially, we applied frequently employed cut-offs such as a MapQ of 20 and a Phred of 30 but were unable to distinguish true from false positives in our $10^5$ samples (`r Qual` A).
-Many false positives occur on well mapped reads with high quality base calls.
+In the above analysis we masked bases that had a Phred  score less than 30 (0.001 probability of being incorrect) and did not apply any MapQ cut-offs.
+In an initial analysis, we applied frequently employed cut-offs such as a MapQ of 20 and a Phred of 30 to our data.
+These criteria were unable to distinguish true from false positives in our $10^5$ samples (`r Qual`A) and indicate that many false positives occur on well mapped reads with high quality base calls.
 
 We further parsed our false variant calls by locating them within individual sequencing reads.
-It is well known that sequence quality drops near the end of a read (cite?), and we found that our false positives clustered to these regions (`r Qual` B).
+It is well known that sequence quality drops near the end of a read (cite?), and we found that our false positives clustered at the termini of our paired end reads (`r Qual`B).
 The average Phred score of these false positives was `r false_p_phred`, further demonstrating that filtering on quality score alone is insufficient.
@@
 
-Based on these data we applied a number of empirically determined cut-offs and drastically increased our specificity without sacrificing sensitivity (`r Qual` C and D).
+Based on these data we applied a number of empirically determined cut-offs and markedly improved our specificity to >0.999 without sacrificing sensitivity (`r Qual` C and D).
 For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean phred score of 35, and an average read position within the middle 50\% of the read.
-In addition to these quality cut-offs we applied a number of other criteria in an attempt to increase accuracy.
-We tried using the Benjamin Hochman p-value correction, utilizing more stringent p values or frequency cut-offs, retaining PCR duplicate reads, trimming the ends of the influenza genome, and employing various distributions to estimate the error rate in the control sample.
-All of these results have be summarized in an interactive shiny application available for download at \emph{https://github.com/lauringlab/benchmarking_shiny.git}  We invite the reader to explore the affects of all these criteria on our data set.
+Under these conditions we found 10 or fewer false positives in 4 of the 5 samples.
+Given this success we applied a number of strategies to further increase our accuracy, including but not limited to: Benjamin Hochman p-value correction,  more stringent p values (<0.01) or frequency cut-offs (>0.2%), retaining PCR duplicate reads, trimming the ends of the influenza genome, and employing various distributions to estimate the error rate in the control sample.
+The impact of various filtering criteria on our data can be visualized in an interactive shiny application available for download at  \emph{https://github.com/lauringlab/benchmarking_shiny.git}.
 
-In addition to distinguishing true from false SNV, DeepSNV was able to estimate the frequency of the true positive variants (`r Freq`).
-Although the median of the measured frequencies match the expect values, we found substantial spread in each sample and the overall the fit was quite poor ( R^2^=`r corr_coef`).
-The mean percent difference between the measured frequency an the expected was `r mean_fold`\%.
-This error should be kept in mind when employing down stream analysis that depend on frequency measurements such as diversity metrics like Shannon's Entropy and Simpson's D, as well as inter-sample comparison of population structure .
+We also benchmarked the accuracy of our DeepSNV pipeline in estimating the frequency of the true positive variants (`r Freq`).
+Although the median of the measured frequencies match the expected values, we found substantial spread in each sample and the overall the fit was quite poor ( R^2^=`r corr_coef`).
+The mean percent difference between the measured frequency and the expected was `r mean_fold`\%.
+This error is likely due to stochastic PCR biases (cite?) in the initial RT-PCR reaction and should be kept in mind when employing down stream analysis that depend on frequency measurements such as variant fitness and diversity metrics that incorporate species abundance (e.g.
+Shannon entropy).
 
@@
 We asked whether the decreased accuracy observed in our data set was due simply to peculiarities specific to  DeepSNV.
-We therefore analyzed our $10^5$ input populations using Lofreq, another algorithm that is commonly used in viral next generation sequencing studies.
-Lofreq has been reported to have perfect specificity in the past.
-Under our conditions Lofreq had marginally reduced sensitivity compared to DeepSNV when applied to the variant frequencies $\geq$ 1.0\% but marginally increased sensitivity when applied to variant frequencies $<$ 1.0\% (`r Lofreq`).
+We therefore analyzed our $10^5$ input populations using Lofreq, another variant caller that is commonly used in next generation sequencing studies, and has been reported to have perfect specificity [@Wilm:2012br].
+Under our experimental conditions, Lofreq had marginally reduced sensitivity compared to DeepSNV when applied to the variant frequencies $\geq$ 1.0\% but marginally increased sensitivity when applied to variant frequencies $<$ 1.0\% (`r Lofreq`).
 The specificity of LoFreq was comparable to what we observed with DeepSNV in our high-input cell culture-derived populations (`r Miseq`), and better than DeepSNV in our initial implementation (`r Hiseq` prior to Phred, MapQ, and read position filtering).
-This increased specificity is most likely due to the fast that  the LoFreq algorithm takes MapQ and Phred scores into account when calling variants and has a stringent strand bias filter that removes many of the variants fount only at one end of the reads.
-However even with these additional characteristics, the specificity of LoFreq was lower than our improved DeepSNV pipeline (`r Qual`) with over 40 false positives per sample.
-It appears that higher than expected false positive rates are not specific to DeepSNV and most likely plague all variant callers applied to patient-derived viral samples.
+This increased specificity is most likely due to the fact that the LoFreq algorithm already takes MapQ and Phred scores into account when calling variants and has a stringent strand bias filter that removes many of the variants found only at one end of a paired-end read.
+However even with these additional characteristics, the specificity of LoFreq was lower than our improved DeepSNV pipeline (`r Qual`), with over 40 false positives per sample.
+It appears that higher than expected false positive rates are not specific to DeepSNV and most likely plague many variant callers applied to patient-derived viral samples.
 
@@
 Host-derived viral populations vary in copy number and titer by several orders of magnitude.
-This variability can be attributed to a variety of factors including but not limited to: collection site, ease of of nucleic acid isolation, library preparation, and host and viral variability.
-To ensure accuracy across a range of input levels, we diluted our experimental populations serially into basal media and and identified variants using our modified DeepSNV analysis pipeline (`r Lower`).
-As expected, the sensitivity was lower in populations with fewer genomes.
-A variant with a 0.5\% frequency in a $10^4$ genomes per microliter sample is expected to present on only 700 genomes in the initial RT-PCR and it is expected that many of these genomes will be lost due to bottlenecks in the amplification and library preparation process.
-We also found that specificity was reduced in samples with lower starting input.
-This is presumably due to a greater dependence on RT-PCR amplification.
+This variability can be attributed to a variety of factors including, but not limited to: collection site, ease of of nucleic acid isolation, the presence of host nucleic acid, efficiency of library preparation, and host and viral factors.
+To ensure accuracy across a range of input levels, we diluted our experimental populations serially into basal media (`r Hiseq`A) and identified variants using our modified DeepSNV analysis pipeline (`r Lower`).
+As expected, our sensitivity was lower in populations with fewer genomes.
+For example, a variant with a 0.5\% frequency in a $10^4$ genomes per microliter sample is expected to present on only 700 genomes in the initial RT-PCR, many of these will be lost due to bottlenecks in the amplification and library preparation process.
+In addition to reduced sensitivity we found reduced specificity in samples with lower starting input.
+The increase in false positives is presumably due to a greater dependence on RT-PCR amplification and consequent propagation of errors.
 These data highlight the importance of controlling for input levels when comparing diversity across experimental samples.
 
-In most cases, RT-PCRs error should be randomly distributed across the amplified region.
-If RT-PCR errors are responsible for the reduced specificity found at lower input levels, and RT-PCR are randomly distributed across the genome, we would expect such error to be eliminated by processing the samples in duplicate.
-To test this hypothesis, we performed duplicate RT-PCR reactions on the 5, 2, 1, and 0.5\% variant frequency samples that had a starting concentration of $10^4$ genomes per microliter.
-This was done using a separate RNA prep than that shown in `r lower` A.
+In most cases, RT-PCR errors should be sporadic and randomly distributed across the amplified region.
+If RT-PCR errors are responsible for the reduced specificity found at lower input levels, they should be easily identified as variants present in only one of two RT-PCR reactions performed on the same RNA.
+To test this hypothesis, we performed RT-PCR reactions and sequenced duplicates of the  5, 2, 1, and 0.5\% variant frequency samples from our $10^4$ genomes per microliter collection.
 The duplicates were processed separately, but sequenced on the same lane of an Illumina Hiseq.
 We applied the previously mentioned quality cut-offs and required that a given variant be found in both duplicates.
-This drastically improved our specificity and resulted in an accuracy that was comparable to that found in the $10^5$ samples (`r Dups`).
-Thus accurate analysis of low input RNA virus samples can be achieved through duplicate RT-PCR and careful benchmarking experiments.
+By running samples in duplicates we reduced the number of false positives in each sample to 10 or fewer resulting in and a specificity of 0.9998 or greater (`r Dups`).
+This increased specificity was not accompanied by a decreased sensitivity.
+In fact, we found a slight increase in sensitivity most likely due to variability in library preparation.
+Thus, accurate analysis of low input samples can be achieved through duplicate RT-PCR and careful benchmarking experiments.
 
-####The consequences of inaccurate SNV identification
+###The consequences of sub-optimal SNV identification
 
-`r Table` stresses the importance of accurate SNV indentification in studies of viral diversity.
-We have apply three commonly used diversity metrics, (Richness, Shannon's Entropy, the inverse of Simpson's D), to our $10^5$ genomes per microliter populations.
+NGS of intrahost populations is commonly used determine the interaction between host or environmental factors and viral diversity.
+However calculations of viral diversity rely entirely on SNV identified in NGS data, and are very sensitive to the accuracy of these variant calls.
+To illustrate this problem, we calculated the diversity of our 10^5^ genomes per microliter samples at each step of our benchmarking process using three complementary metrics (`r Table`).
+Richness is the count of nonconsensus variants present in a population (often referred to as intrahost SNV).
+Shannon's Entropy is a diversity metric that accounts for  both the number of variants present (richness) and their frequencies (evenness).
+Because our data are unphased (i.e.
+no haplotypes), we have reported the average entropy per nucleotide position (cite).
+The last metric, L1-norm, is a distance measurement that describes how similar two populations are to one another.
+Identical populations will have an L1-norm of 0.
+To mimic experimental conditions, we included all variants identified in each analysis regardless of whether or not subsequent benchmarking distinguished them as true or false positives.
+We found that the accuracy of the SNV calling method has a profound effect on measurements of diversity.
+It is clear from the richness measurements in `r Table` that the number of false SNV largely determines the accuracy of the other diversity metrics.
+Thus, our adapted DeepSNV protocol, which was able to distinguish between true and false SNV with the highest accuracy, gave the most accurate measures of diversity, followed by Lofreq and the default version of DeepSNV.
+
+## Methods
+
+### Viruses and cells
+
+We passaged plaque purified WSN33 and PR8 lab strains 3 times in
+
+### Viral populations
+
+We extracted viral RNA from cleared supernatants using  QIAamp viral RNA minikits (Qiagen 5204) according to the manufactures' instructions, and generated cDNA using Superscript III one-step with HiFi platinum Taq (Invitrogen 12574).
+For the WSN33, PR8 data set we used a segment specific primers PB2-Forward-JT	(GCAGGTCAATTATATTCAATATGGAAA),PB2-Reverse-JT	(CAAGGTCGTTTTTAAACTATTCGACAC), PB1-Forward-JT (GCAGGCAAACCATTTGAATGG), PB1-Reverse-JT	(CAAGGCATTTTTTCATGAAGGACAAG), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG), NA-Forward-JT	(GCAGGAGTTTAAATGAATCCAAACC), NA-Reverse-JT	(CAAGGAGTTTTTTGAACAAACTACTTG), HA-Forward-JT	(GCAGGGGAAAATAAAAACAACCAAAAT), HA-Reverse-JT	(CAAGGGTGTTTTTCCTTATATTTCTGAA), NP-Forward-JT	(GCAGGGTAGATAATCACTCACAG), NP-Reverse-JT	(CAAGGGTATTTTTCTTTAATTGTCGTACT), M-Forward-JT	(GCAGGTAGATATTGAAAGATGAGTC), M-Reverse-JT	(CAAGGTAGTTTTTTACTCCAGCTCT), NS-Forward-JT	(GCAGGGTGACAAAGACATAATG), NS-Reverse-JT	(CAAAGGGTGTTTTTTATTATTAAATAAGCTG), which with the reaction conditions were adapted from [@Anonymous:A25Ka2tY].
+Each 25 $\mu$ l reaction included 1 $\mu$ l of each primer (5 $\mu$ M) and 2 $\mu$ l of RNA template.
+Reaction conditions were 50$^\circ$ C (60 min), 94$^\circ$ C (2 min), and then 30 cycles of 94$^\circ$ C (30 sec), 54$^\circ$ C (30 sec), and 68$^\circ$ C (3 min).
+
+### Mutagenesis
+
+### Viral populations
+
+### Plasmid controls
+
+### Library preperation
+
+### Sequence analysis
 

### 2016-03-16 doc/paper/results.Rmd
starting methods

@@
 Lofreq has been reported to have perfect specificity in the past.
-Under our conditions Lofreq had marginally reduced sensitivity compared DeepSNV when applied to the variant frequencies \geq 1.0% but marginally increased sensitivity when applied to variant frequencies \l 1.0% (`r Lofreq`).
+Under our conditions Lofreq had marginally reduced sensitivity compared DeepSNV when applied to the variant frequencies $\geq$ 1.0% but marginally increased sensitivity when applied to variant frequencies $\l$ 1.0% (`r Lofreq`).
 The specificity of LoFreq was comparable to what we observed with DeepSNV in our high-input cell culture-derived populations (`r Miseq`), and better than DeepSNV in our initial implementation (`r Hiseq` prior to Phred, MapQ, and read position filtering).

### 2016-03-16 scripts/figures.Rmd
starting methods

@@
 ### With read position cut off 63 and 188
-read.df<-subset(sum.df,Read_pos>63 & Read_pos<188)
+read.df<-subset(sum.df,Read_pos>62 & Read_pos<188)
 m.roc.df<-miseq.roc(read.df,possible_tp,possible_vars,">")
@@
 sum.df<-read.csv("../data/process/2015-6-23/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#") # reading in the variants from the pipeline output
-
+mfe.df<-read.csv("../data/reference/mfe.csv") 
 # Getting genome lenght and segment information from the wsn33 fasta file
@@
 sum.df$category[sum.df$mutation %in% wt_mut]<-"wt"
-sum.df<-subset(sum.df,!(grepl("WT",Id))) # remove the WT samples
+sum.df<-subset(sum.df,!(grepl("WT",Id))) # remove the WT sample
+
+sum.df$category[sum.df$mutation %in% mfe.df$mutation]<-"mfe"
 
@@
 
-possible_vars<-sum(regions.bed$length)*3 # three possible variants/ position
-
+total_possible_vars<-sum(regions.bed$length)*3 # three possible variants/ position
+
+possible_vars<-total_possible_vars-length(wt_mut)-dim(mfe.df)[1] # removing the mfe and wt_mut sites
 hiseq_vars<-possible_vars
@@
 
+lofreq.df$category[lofreq.df$mutation %in% wt_mut]<-"wt"
+lofreq.df$categroy[lofreq.df$mutation %in% mfe.df$mutation]<-"mfe"
+
 lofreq.df$p.val<-lofreq.df$QUAL # to make the roc based on the quality scores given by lofreq
 
-direction<-">"
 lofreq.roc<-hiseq.roc(lofreq.df,possible_tp,possible_vars,"<")
@@
   draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
-direction<-"<"
 ```
@@
   }
-  out<-data.frame(Richness=round(rich(x),digits = 0), Entropy=format(H(x),scientific = T,digits = 3),"1/D"=D(x),Manhattan=round(euc(x,subset(perfect,freq.var==exp.freq)),digits=3))
+  out<-data.frame(Richness=round(rich(x),digits = 0), Entropy=format(H(x),scientific = T,digits = 3),Manhattan=round(euc(x,subset(perfect,freq.var==exp.freq)),digits=3)) #"1/D"=D(x),
   rownames(out)<-setting
@@
 tabulate<-function(input,freq){
-get_data(subset(perfect,freq.var==freq),"Perfect")->perf
-get_data(subset(sum.df,gc==input & exp.freq==freq & p.val<0.01),"DeepSNV")->deepSNV
-get_data(subset(lofreq.df,gc==input & exp.freq==freq),"Lofreq")->lo
-get_data(subset(cut.read.df.all,gc==input & exp.freq==freq & p.val<0.01),"DeepSNV Final")->deepSNV.f
+get_data(subset(perfect,freq.var==freq),"Expected")->perf
+get_data(subset(sum.df,gc==input & exp.freq==freq & p.val<0.01 & category %in% c(T,F)),"DeepSNV")->deepSNV
+get_data(subset(lofreq.df,gc==input & exp.freq==freq & category %in% c(T,F)),"Lofreq")->lo
+get_data(subset(cut.read.df.all,gc==input & exp.freq==freq & p.val<0.01 & category %in% c(T,F) ),"DeepSNV Final")->deepSNV.f
 knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f))

### 2016-03-16 scripts/verifying_tp.R

@@ -25,7 +25,12 @@ cut.df<-subset(sum.df, Id %in% tp$Id & p.val<0.01 & MapQ>30 & Phred>35 & Read_po
 cut.df<-mutate(cut.df,category=mutation %in% tp$mutant)
 
 
-subset(cut.df, category==F )
+subset(cut.df, category==F & freq.var>0.01)->carry_over;carry_over
+write.csv(carry_over,file = "../data/reference/mfe.csv")
+
+bench.df<-read.csv("../data/process/2015-6-23/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#") # reading in the variants from the pipeline output
+
+subset(bench.df,mutation %in% carry_over$mutation)
 
 #There are some overlaps so I'll remove those now.
 

### 2016-03-16 scripts/figures.Rmd
dummy commit

--- 2016-02-23 scripts/figures.Rmd
@@
 author: "JT McCrone"
-date: "February 11, 2015"
-output: pdf_document
+date: "December 30, 2015"
+output: html_document
 ---
 
+Here are the figures (or figure panels) we are thinking of using in the Benchmarking paper.
+I will include my comments and decriptions on each and perhaps over time this will develop into the results section
 ```{r,settings_and_lib,echo=F,message=F,}
@@
 ### Figure 1
-```{r, roc_skematic,fig.height=4,fig.width=3.4}
+```{r, roc_skematic,fig.height=5,fig.width=3.4}
 set.seed(42) # make the figures reproducible
@@
   roc.p<-ggplot(roc.df,aes(x=1-specificity,y=sensitivity))+geom_step(color="black",size=1.5)+xlab("1-Specificity")+ylab("Sensitivity")+theme(axis.title.y=element_text(vjust=1))
-  roc.p<-roc.p+geom_line(data=lines,aes(x=spec,y=sense,group=as.factor(group),color=rep(c("#08519cff","#787878ff"),times=3)),size=c(1),lty=c(2))+scale_color_manual(values=c("#08519cff","#787878ff"))
+  roc.p<-roc.p+geom_line(data=lines,aes(x=spec,y=sense,color=as.factor(group)),size=c(1),lty=c(2))+scale_color_brewer(palette="Set1")
 
@@
  #plot the points and cut offs
- points.p<-ggplot(distribution,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center",dotsize = 1.5)+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks = element_blank(), axis.text.x = element_blank(),axis.title.y = element_text(vjust=1.2),legend.position=c(0.9,0.65),legend.background = element_rect(fill = alpha("white",0)))+xlab("")
+ points.p<-ggplot(distribution,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center",dotsize = 1.5)+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks = element_blank(), axis.text.x = element_blank(),axis.title.y = element_text(vjust=1.2),legend.position=c(0.8,0.6))+xlab("")
 
  points.p<-points.p+geom_hline(yintercept=output$threshold,
-                          size=1,lty=2,color=c("#08519cff","#787878ff")) # these are the first 2 colors in set 1
+                          size=1,lty=2,color=c("#e41a1c","#377eb8")) # these are the first 2 colors in set 1
 
-ggdraw() +
-  draw_plot(points.p, 0, .45, .8, 0.48) + #x y width height
-  draw_plot(roc.p, 0, 0, .8, 0.48) +
-  draw_plot_label(c("A", "B"), c(0.05, 0.05), c(1, 0.55), size = 15)
+ plot_grid(points.p, roc.p, labels=c("A", "B"), ncol = 1, nrow = 2)
 
@@
 plot.ex(0.35,0.95,roc.df,example)
-
-pdf("../results/figures/roc.example.pdf",width=3.4,height=4)
-plot.ex(0.35,0.95,roc.df,example)
-dev.off()
 # pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.ex.pdf",width = 12,height = 8)
@@
 
-Figure 1.
-An example ROC.
-A) Hypothetical variants are stratified by p value.
-True variants are shown as filled circles, while those corresponding to RT, PCR, or sequencing errors are shown as open circles.
-Potential thresholds are indicated as colored lines.
-B) An ROC made from the data shown in A.
-The colored lines indicate the points made by the thresholds in A.
-\pagebreak
 ```{r,roc_helper_runctions}
@@
 ### tableGrob functions and settings #####
-tt1<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.75)),core=list(fg_params=list(cex=0.75)))#,fontfamily="serif")))
+tt1<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.75,fontfamily="serif")),core=list(fg_params=list(cex=0.75,fontfamily="serif")))
 make.table<-function(df,theme){
@@
 ### Figure 2
-```{r,Miseq_out_of_the_box,fig.width=3.4,fig.height=7}
+```{r,Miseq_out_of_the_box,fig.width=7,fig.height=5}
 
@@
 
-miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.5%","1.25%","0.63%","0.016%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
+miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.5%","1.25%","0.63%","0.016%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position=c(1,.5))
 
@@
 #knitr::kable(m.roc.table)
-m.roc.table<-rename(m.roc.table,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True\nPositives","adj.specificity"="Specificity","FP"="False\nPositives"))
+m.roc.table<-rename(m.roc.table,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))
 
@@
 
-pdf("../results/figures/miseq1.pdf",width=3.4,height=7)
 ggdraw() +
-  draw_plot(miseq.setup, 0.2, .66, .6, 0.33) + #x y width height
-  draw_plot(miseq.roc.p, 0.05, 0.33, .9, 0.33) +
-  draw_plot(tbl, 0.05, 0, .9, 0.33) +
-  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(0.95, 0.7, 0.35), size = 15)
+  draw_plot(miseq.setup, 0.01, .1, 0.4, 0.8) + #x y width height
+  draw_plot(miseq.roc.p, 0.43, 0.42, 0.5, 0.51) +
+  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
+  draw_plot_label(c("A", "B", "C"), c(0.01, 0.42, 0.42), c(0.78, 0.95, 0.46), size = 15)
+
+pdf("../results/figures/miseq1.pdf",width=7,height=5)
+ggdraw() +
+  draw_plot(miseq.setup, 0.01, .1, 0.4, 0.8) + #x y width height
+  draw_plot(miseq.roc.p, 0.43, 0.42, 0.5, 0.51) +
+  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
+  draw_plot_label(c("A", "B", "C"), c(0.01, 0.42, 0.42), c(0.78, 0.95, 0.46), size = 15)
 dev.off()
-
-ggdraw() +
-  draw_plot(miseq.setup, 0.2, .66, .6, 0.33) + #x y width height
-  draw_plot(miseq.roc.p, 0.05, 0.33, .9, 0.33) +
-  draw_plot(tbl, 0.05, 0, .9, 0.33) +
-  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(0.95, 0.7, 0.35), size = 15)
 
@@
 ```
-
-Figure 2.
-Initial DeepSNV accuracy.
-A) RT-PCR amplified WSN33 genomes were diluted into PR8 at decreasing frequencies and sequenced on an Illumina Miseq.
-B) An ROC measuring the pipeline's ability to identify WSN33 SNVs relative to PR8 at the indicated frequencies.
-C) A summary of the DeepSNV accuracy at a p value threshold of 0.01.
-\pagebreak
 
@@
 
+rm(list = setdiff(ls(), lsf.str())) # remove all variables except the functions
+tt1<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.75,fontfamily="serif")),core=list(fg_params=list(cex=0.75,fontfamily="serif")))
+
 ```
@@
 
-```{r,hiseq_roc,fig.width=3.4,fig.height=7}
-#rm(list = setdiff(ls(), lsf.str())) # remove all variables except the functions
-#tt1<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.75,fontfamily="serif")),core=list(fg_params=list(cex=0.75,fontfamily="serif")))
-
+```{r,hiseq_roc,fig.width=7,fig.height=5}
 sum.df<-read.csv("../data/process/2015-6-23/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#") # reading in the variants from the pipeline output
@@
 
-#length(wt_mut) How many wt muts do we find.
+length(wt_mut)
 sum.df$category[sum.df$mutation %in% wt_mut]<-"wt"
@@
 
-x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.03
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.01
 #x$adj.sensitivity[x$frequency=="5.0%"]<-x$adj.sensitivity[x$frequency=="5.0%"]+0.0
@@
 
-h.table.df<-rename(h.table.df,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True\nPositives","adj.specificity"="Specificity","FP"="False\nPositives"))
+h.table.df<-rename(h.table.df,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))
 
@@
 
-pdf("../results/figures/hiseq1.pdf",width=3.4,height=7)
 ggdraw() +
-  draw_plot(hiseq.setup, 0.1, .57, 0.8, 0.43) + #x y width height
-  draw_plot(Hiseq.roc, 0.05, 0.33, .8, 0.33) +
-  draw_plot(tbl, 0.05, 0, .9, 0.33) +
-  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(1, 0.7, 0.35), size = 15)
+  draw_plot(hiseq.setup, 0, .1, 0.4, 0.8) + #x y width height
+  draw_plot(Hiseq.roc, 0.43, 0.43, 0.5, 0.5) +
+  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
+  draw_plot_label(c("A", "B", "C"), c(0.05, 0.42, 0.42), c(0.95, 0.95, 0.46), size = 15)
+
+pdf("../results/figures/hiseq1.pdf",width=7,height=5)
+ggdraw() +
+  draw_plot(hiseq.setup, 0, .1, 0.4, 0.8) + #x y width height
+  draw_plot(Hiseq.roc, 0.43, 0.43, 0.5, 0.5) +
+  draw_plot(tbl, 0.45, 0, 0.53, 0.47) +
+  draw_plot_label(c("A", "B", "C"), c(0.05, 0.42, 0.42), c(0.95, 0.95, 0.46), size = 15)
 dev.off()
 
-ggdraw() +
-  draw_plot(hiseq.setup, 0.1, .57, 0.8, 0.43) + #x y width height
-  draw_plot(Hiseq.roc, 0.05, 0.33, .8, 0.33) +
-  draw_plot(tbl, 0.05, 0, .9, 0.33) +
-  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(1, 0.7, 0.35), size = 15)
 # 
@@
 
-Figure 3.
-DeepSNV accuracy patient sample conditions A) 20 known single point mutants were diluted into wild type at decreasing frequencies.
-These were diluted further into viral media to match the genomes concentration found in patent samples (10$^5$-10$^3$ genomes/$\mu$l) and sequenced on an Illumina Hiseq (2 x 152bp reads.
-B) An ROC indicating the ability to accurately identify the 20 known variants in the 10 $^5$ genomes/$\mu$l samples.
-C) A table of the accuracy of our variant calling for the 10 $^5$ genomes/$\mu$l samples at a p value threshold of 0.01.
+### Figure 4
 
-\pagebreak
-
-### Figure 4 
 ```{r, phred_mapq filter}
@@
 
-mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(size=1)+geom_hline(yintercept=20,linetype=2)+geom_vline(xintercept=30,linetype=2)+scale_x_continuous(limits=c(30,40))+scale_color_manual(name="",labels=c("False posivite","True positive"),values=c("#787878ff","#08519cff"))+ylab("Mean mapping quality")+xlab("Mean base qualtiy (Phred) ")+theme(legend.position="top")
+mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(size=1)+geom_hline(yintercept=20,linetype=2)+geom_vline(xintercept=30,linetype=2)+scale_x_continuous(limits=c(30,40))+scale_color_hue(name="",labels=c("False posivite","True positive"))+ylab("Mean mapping quality")+xlab("Mean base qualtiy (Phred) ")+theme(legend.position="top")
 
@@
 
-```{r,read_pos,fig.width=3.4,fig.height=9}
+```{r,read_pos,fig.width=7,fig.height=6}
 # ggplot(sig.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
@@
 
-read.pos<-ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')+scale_fill_manual(name="",labels=c("False positive","True positive"),values=c("#787878ff","#08519cff"))+xlab("Average position on read")+ylab("Number of variants")+theme(legend.position="top")
+read.pos<-ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')+scale_fill_hue(name="",labels=c("False positive","True positive"))+xlab("Average position on read")+ylab("Number of variants")+theme(legend.position="top")
 # read.pos
@@
 
-x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.08
-x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]-0.01
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.05
+#x$adj.sensitivity[x$frequency=="5.0%"]<-x$adj.sensitivity[x$frequency=="5.0%"]+0.0
 
@@
 h.cut.read.table.5<-subset(h.cut.read.table,gc==5,select=-c(gc))
-rename(h.cut.read.table.5,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True\nPositives","adj.specificity"="Specificity","FP"="False\nPositives"))->h.cut.read.table.5
+rename(h.cut.read.table.5,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))->h.cut.read.table.5
 h.cut.read.table.5$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
@@
 
-h.cut.read.table.5$Specificity[h.cut.read.table.5$Specificity>0.9999]<-">0.9999"
-
 tbl<-make.table(h.cut.read.table.5,tt1) # Make table
 
-pdf("../results/figures/figure4.pdf",width=3.4,height=9)
 ggdraw() +
-  draw_plot(mapq_phred.normal,0.03, 0.77, .85, 0.23) + 
-  draw_plot(read.pos, 0.03, .53, .85, .25) +
-  draw_plot(read.roc, 0.03, 0.23, 0.85, .3) +
-  draw_plot(tbl,   0.025, 0, 0.95, .25) +
-  draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(1, 0.78, 0.55,0.25), size = 15)
+  draw_plot(mapq_phred.normal, 0, .5, .5, .5) + 
+  draw_plot(read.pos, .5, .5, .5, .5) +
+  draw_plot(read.roc, 0, 0, 0.5, .5) +
+  draw_plot(tbl, 0.5, 0, 0.5, .5)+
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(1, 1, 0.5,0.5), size = 15)
+
+pdf("../results/figures/figure4.pdf",width=7,height=6)
+ggdraw() +
+  draw_plot(mapq_phred.normal, 0, .5, .5, .48) + 
+  draw_plot(read.pos, .5, .5, .5, .48) +
+  draw_plot(read.roc, 0, 0, 0.42, .5) +
+  draw_plot(tbl, 0.47, 0, 0.52, 0.47) +
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(0.95, 0.95, 0.55,0.45), size = 15)
 dev.off()
-
-ggdraw() +
-  draw_plot(mapq_phred.normal,0.03, 0.77, .85, 0.23) + 
-  draw_plot(read.pos, 0.03, .53, .85, .25) +
-  draw_plot(read.roc, 0.03, 0.23, 0.85, .3) +
-  draw_plot(tbl,   0.025, 0, 0.95, .25) +
-  draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(1, 0.78, 0.55,0.25), size = 15)
 ```
-
-Figure 4.
-Investigating the quality of variants in the the 10 $^5$ genomes/$\mu$l samples.
-A) All called variants from the 10 $^5$ genomes/$\mu$l samples with p<0.01 stratified by the mean mapping quality of the reads containing the variant and the mean Phred scores of the variant bases.
-Dashed lines indicate  common cutoffs of 20 and 30 for mapping quality and Phred respectively.
-B) A histogram of all called variants from the 10 $^5$ genomes/$\mu$l samples with p<0.01 binned by the average position of the variant relative to the Hiseq reads on which it is found.
-C) An ROC of the accuracy of the variant caller when the following quality cut offs are applied : mean mapping quality of 30, mean Phred of 35, and an average read position between 32 and 94 (the middle 50% of the read).
-D) A table summarizing the ROC in C at a p value threshold of 0.01.
-
-\pagebreak
 
@@
 
-x<-ddply(lofreq.roc,~samp,fill_in_plots)
-
-x<-mutate(x,frequency=find_freq.v.h(exp.freq))
-
-x<-subset(x,gc==5)
-
-x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.04
-x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]-0.12
-
-ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position='none')+geom_text(data=x,aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2)->lofreq.roc.p
-# This is so that the text is not cut off
-lofreq.roc.p <- ggplot_gtable(ggplot_build(lofreq.roc.p))
-lofreq.roc.p$layout$clip[lofreq.roc.p$layout$name == "panel"] <- "off"
+ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position=c(1,.5))->lofreq.roc.p
 
 #lofreq claims no further monkey bussiness is needed to we'll just make a table with all its output
-ddply(lofreq.df,~Id,summarize,gc=gc[1],exp.freq=exp.freq[1],TP=length(which(category==T)),Sensitivity=TP/20,FP=length(which(category==F)),Specificity=(possible_vars-possible_tp-FP)/(possible_vars-possible_tp))->lofreq.table
+ddply(lofreq.df,~Id,summarize,gc=gc[1],exp.freq=exp.freq[1],TP=length(which(category==T)),FP=length(which(category==F)))->lofreq.table
 lofreq.table<-lofreq.table[order(lofreq.table$exp.freq,decreasing = T),]
@@
 #knitr::kable(subset(lofreq.table,gc==5,select=c(exp.freq,TP,FP)))
-lofreq.table<-head(subset(lofreq.table,gc==5,select=-c(gc,Id)),n=-1) # don't need the 0.1% sample
-rename(lofreq.table,c("exp.freq"="Frequency","TP"="True\nPositives","FP"="False\nPositives"))->lofreq.table
-lofreq.table$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
-lofreq.table$Sensitivity<-round(lofreq.table$Sensitivity,digits=dd)
-lofreq.table$Specificity<-round(lofreq.table$Specificity,digits=dd)
-lofreq.table<-subset(lofreq.table,select=c("Frequency","Sensitivity","True\nPositives","Specificity","False\nPositives"))
-tt2<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.65,fontfamily="serif")),core=list(fg_params=list(cex=0.65,fontfamily="serif")))
+lofreq.table<-subset(lofreq.table,gc==5,select=c(exp.freq,TP,FP))
 
@@
 
-pdf("../results/figures/lofreq1.pdf",width=3.4,height=5)
-ggdraw() +
-  draw_plot(lofreq.roc.p, 0, 0.5, 0.9, 0.45) + 
-  draw_plot(tbl, 0.025, 0, 0.95, .48) +
-  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
+plot_grid(lofreq.roc.p, tbl, labels=c("A", "B"), ncol = 1, nrow = 2)
+
+pdf("../results/figures/lofreq1.pdf",width=7,height=2.5)
+plot_grid(lofreq.roc.p, tbl, labels=c("A", "B"), ncol = 2, nrow = 1)
 dev.off()
-
-ggdraw() +
-  draw_plot(lofreq.roc.p, 0, 0.5, 0.9, 0.45) + 
-  draw_plot(tbl, 0.025, 0, 0.95, .48) +
-  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
 
@@
 
-Figure 5.
-The accuracy of Lofreq.
-The variant caller Lofreq was applied to the 10 $^5$ genomes/$\mu$l samples.
-A) An ROC of the accuracy of Lofreq using standard parameters.
-The specificity of the caller was scaled to account for the same number of tests as the DeepSNV algorithm.
-Lofreq provides quality scores instead of p values.
-These were used to set cutoffs for the ROC.
-B) A table summarizing the accuracy of the ROC A using the standard cut offs applied by the algorithm.
-
-\pagebreak
-
 ### Figure 6
 
-```{r, frequency_accuracy,fig.width=3.4,fig.height=3.4}
+For p<0.01 with phred, mapq and read position cut off.
+```{r, frequency_accuracy}
 # Here we'll just be using
@@
 
-ac_freq<-ac_freq+geom_dotplot(binaxis = "y", stackdir = "center",dotsize=1)+scale_fill_manual(values=rhg_cols)
+ac_freq<-ac_freq+geom_dotplot(binaxis = "y", stackdir = "center",dotsize=1)+scale_color_manual(values=rhg_cols)
 
@@
 ac_freq<-ac_freq+geom_abline(intercept = 0, slope = 1,linetype=2,size=1)
-ac_freq<-ac_freq+xlab("Expected Frequency")+ylab("Measured Frequency")+theme(legend.position="none",axis.text.x = element_text(angle = 45, hjust = 1))
+ac_freq<-ac_freq+xlab("Expected Frequency")+ylab("Measured Frequency")+theme(legend.position="none")
 
-ac_freq<-ac_freq+geom_point(data=subset(medians,gc=="5"),mapping=aes(x=exp.freq,y=median),shape=95,size=20)
+ac_freq<-ac_freq+geom_point(data=subset(medians,gc=="5"),mapping=aes(x=exp.freq,y=median),shape=95,size=30)
 ac_freq
 
-pdf("../results/figures/freq.pdf",width=3.4,height=3.4)
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/freq.pdf")
 ac_freq
@@
 
-Figure 6. The accuracy of the frequency measurements for the known variants in the 10 $^5$ genomes/$\mu$l samples. The median of each distribution is shown as a black bar.
-
-\pagebreak
-
 ### Figure 7
 
-```{r,roc_lower_inputs4,fig.width=3.4,fig.height=9}
-
-x<-ddply(h.cut.read.df,~samp,fill_in_plots)
-x<-subset(x,gc==4)
-
-x<-mutate(x,frequency=find_freq.v.h(exp.freq))
-
-x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.05
-x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]+0.05
-x$adj.sensitivity[x$frequency=="0.5%"]<-x$adj.sensitivity[x$frequency=="0.5%"]+0.025
-
-roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")+geom_text(data=x,aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2)
-
-# This is so that the text is not cut off
-roc.4 <- ggplot_gtable(ggplot_build(roc.4))
-roc.4$layout$clip[roc.4$layout$name == "panel"] <- "off"
+```{r,roc_lower_inputs4}
+roc.4<-ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
 
@@
 
-x<-ddply(h.cut.read.df,~samp,fill_in_plots)
-x<-subset(x,gc==3)
-
-x<-mutate(x,frequency=find_freq.v.h(exp.freq))
-
-roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")+geom_text(data=x,aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2)
-
-# This is so that the text is not cut off
-roc.3 <- ggplot_gtable(ggplot_build(roc.3))
-roc.3$layout$clip[roc.3$layout$name == "panel"] <- "off"
+roc.3<-ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
 
@@
 h.cut.read.table.4<-subset(h.cut.read.table,gc==4,select=-c(gc))
-rename(h.cut.read.table.4,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True\nPositives","adj.specificity"="Specificity","FP"="False\nPositives"))->h.cut.read.table.4
+rename(h.cut.read.table.4,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))->h.cut.read.table.4
 h.cut.read.table.4$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
@@
 h.cut.read.table.3<-subset(h.cut.read.table,gc==3,select=-c(gc))
-rename(h.cut.read.table.3,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True\nPositives","adj.specificity"="Specificity","FP"="False\nPositives"))->h.cut.read.table.3
+rename(h.cut.read.table.3,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))->h.cut.read.table.3
 h.cut.read.table.3$Frequency<-c("5.0%","2.0%","1.0%")
@@
 
-pdf("../results/figures/lower.pdf",width=3.4,height=9)
 ggdraw() +
-  draw_plot(roc.4,0.03, 0.69, .85, 0.25) + 
-  draw_plot(tbl.4, 0.025, .47,  0.95, .25) +
-  draw_plot(roc.3, 0.03, 0.22, 0.85, .25) +
-  draw_plot(tbl.3,   0.025, 0, 0.95, .25) +
-  draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(0.97, 0.73, 0.5,0.23), size = 15)
-dev.off()
+  draw_plot(roc.4, 0, .5, .5, .5) +
+  draw_plot(tbl.4, .5, .5, .5, .5) +
+  draw_plot(roc.3, 0, 0, 0.5, .5) +
+  draw_plot(tbl.3, 0.5, 0, 0.5, .5)+
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0.5, 0,0.5), c(1, 1, 0.5,0.5), size = 15)
 
-ggdraw() +
-  draw_plot(roc.4,0.03, 0.69, .85, 0.25) + 
-  draw_plot(tbl.4, 0.025, .47,  0.95, .25) +
-  draw_plot(roc.3, 0.03, 0.22, 0.85, .25) +
-  draw_plot(tbl.3,   0.025, 0, 0.95, .25) +
-  draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(0.97, 0.73, 0.5,0.23), size = 15)
 ```
-
-Figure 7.
-The accuracy of DeepSNV at lower input levels.
-A) An ROC depicting the accuracy of our method when applied to the 10 $^4$ genomes/$\mu$l samples.
-B) A table summarizing the accuracy at 10 $^4$ genomes/$\mu$l with a  p value threshold of 0.01.
-C) An ROC depicting the accuracy of our method when applied to the 10 $^3$ genomes/$\mu$l samples.
-B) A table summarizing the accuracy at 10 $^3$ genomes/$\mu$l with a  p value threshold of 0.01
-
-\pagebreak
 
@@
 
-```{r,duplicates,fig.height=5,fig.width=3.4}
-dup.df<-read.csv("../data/process/2015-11-14/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#")
+```{r,duplicates}
+sum.df<-read.csv("../data/process/2015-11-14/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#")
 #sum.df<-read.csv("~/Box Sync/Lab_work/NGS/Benchmarking/data/2015-11-14/all.sum.csv")
 pat<-"([0-9]+)_([0-9]+)"
-mutate(dup.df, exp.freq = sub(pat,"\\1",Id),
+mutate(sum.df, exp.freq = sub(pat,"\\1",Id),
        dup = as.numeric(sub(pat,"\\2",Id)),
-)-> dup.df
+)-> sum.df
 
-dup.df<-subset(dup.df,Id!="PC") # remove the plasmid control that was run
-mutate(dup.df,Id=paste(4,exp.freq,sep="_"),exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->dup.df
+sum.df<-subset(sum.df,Id!="PC") # remove the plasmid control that was run
+mutate(sum.df,Id=paste(4,exp.freq,sep="_"),exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->sum.df
 
-dup.df<-mutate(dup.df,category=mutation %in% true_snv$mutant) # add Column for True and false variants
-dup.df$category[dup.df$mutation %in% wt_mut]<-"wt"
+sum.df<-mutate(sum.df,category=mutation %in% true_snv$mutant) # add Column for True and false variants
+sum.df$category[sum.df$mutation %in% wt_mut]<-"wt"
 
-cut.df<-subset(dup.df,Phred>35 & MapQ>30 & Read_pos<94 & Read_pos>32)
+cut.df<-subset(sum.df,Phred>35 & MapQ>30 & Read_pos<94 & Read_pos>32)
 
@@
 
-x<-ddply(dups.roc,~samp,fill_in_plots)
-
-x<-mutate(x,frequency=find_freq.v.h(exp.freq))
-
-x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.18
-x$adj.sensitivity[x$frequency=="2.0%"]<-x$adj.sensitivity[x$frequency=="2.0%"]-0.09
-
-dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")+geom_text(data=x,aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2)
-
-# This is so that the text is not cut off
-dups.roc.p <- ggplot_gtable(ggplot_build(dups.roc.p))
-dups.roc.p$layout$clip[dups.roc.p$layout$name == "panel"] <- "off"
+dups.roc.p<-ggplot(dups.roc,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")
 
@@
 
-tbl<-make.table(dups.table,tt2) # Make table
+tbl<-make.table(dups.table,tt1) # Make table
+
+ggdraw() +
+  draw_plot(dups.roc.p, 0, .5, 0.5, .5) +
+  draw_plot(tbl, 0.5, 0.5, 0.5, .5) +
+  draw_plot_label(c("A", "B"), c(0, 0.5), c(1, 1), size = 15)
 
@@
                                                                                       out=data.frame(freq1=freq1,freq2=freq2)})
-
-pdf("../results/figures/dups.roc.pdf",width=3.4,height=5)
-ggdraw() +
-  draw_plot(dups.roc.p, 0, 0.5, 0.9, 0.45) + 
-  draw_plot(tbl, 0.025, 0, 0.95, .48) +
-  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
-dev.off()
-
-ggdraw() +
-  draw_plot(dups.roc.p, 0, 0.5, 0.9, 0.45) + 
-  draw_plot(tbl, 0.025, 0, 0.95, .48) +
-  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
 
@@
 
-Figure 8.
-The accuracy of DeepSNV at lower input levels when run in duplicate.
-A subset of the 10 $^4$ genomes/$\mu$l samples were processed in duplicate to control for RT-PCR errors.
-A) An ROC of  10 $^4$ genomes/$\mu$l samples processed in duplicate.
-A variant was required to be found in both duplicates to be considered.
-B) A table summarizing the accuracy when applied to duplicate samples with 10 $^4$ genomes/$\mu$l input, p<0.01.
+These are variants that pass mapq,phred, and read cut and p<0.01 in both samples.10^4
 
-\pagebreak
-
-### Table 1
-
-Richness
-Inverse Simpsons D ( without replacement)
-Shannon Entropy
-For these I'm assuming all varaints are independent
-
-```{r}
-perfect<-data.frame(mutation=rep(true_snv$mutant,times=4),freq.var=rep(c(0.05,0.02,0.01,0.005),each=20))
-dups<-subset(dups,p.val<0.01)
-dups<-mutate(dups,freq.var=pmax(freq1,freq2))
-
-rich<-function(x){
-  dim(x)[1]  
-}
-
-H<-function(x){
-  p=x$freq.var*log(x$freq.var)
-  -sum(p)
-}
-
-D<-function(x){
-  p=(x$freq.var)^2
-  1/sum(p)
-}
-
-euc<-function(x,ref){
-  x$Id="test"
-  ref$Id="ref"
-  all<-rbind(subset(x,select=c(mutation,freq.var,Id)),ref)
-  tab<-as.matrix(subset(dcast(all,mutation ~ Id, value.var = "freq.var"),select=-c(mutation)))
-  tab[is.na(tab)]<-0
-  as.dist(dist(tab))
-}
-get_data<-function(x,setting){
-  if(length(unique(x$freq.var))==1){
-    exp.freq=unique(x$freq.var)
-  }else{
-  exp.freq=unique(x$exp.freq)
-  }
-  out<-data.frame(Richness=rich(x), Entropy=H(x),Simpson=D(x))#,dist=euc(x,subset(perfect,freq.var==exp.freq))[1])
-  rownames(out)<-setting
-  return(t(out))
-}
-tabulate<-function(input,freq){
-get_data(subset(perfect,freq.var==freq),"Perfect")->perf
-get_data(subset(sum.df,gc==input & exp.freq==freq),"DeepSNV")->deepSNV
-get_data(subset(lofreq.df,gc==input & exp.freq==freq),"Lofreq")->lo
-get_data(subset(cut.read.sig.df,gc==input & exp.freq==freq),"DeepSNV Final")->deepSNV.f
-knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f),digits = 2)
-}
-tabulate.4<-function(input,freq){
-get_data(subset(perfect,freq.var==freq),"Perfect")->perf
-get_data(subset(sum.df,gc==input & exp.freq==freq),"DeepSNV")->deepSNV
-get_data(subset(lofreq.df,gc==input & exp.freq==freq),"Lofreq")->lo
-get_data(subset(cut.read.sig.df,gc==input & exp.freq==freq),"DeepSNV Final")->deepSNV.f
-get_data(subset(dups,exp.freq==freq),"DeepSNV duplicates")->deepSNV.d
-
-knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f,deepSNV.d),digits = 2)
-}
-
-```
-
-$10^5$ 5%
-
-```{r}
-tabulate(5,0.05)
-```
-
-$10^5$ 1%
-
-```{r}
-tabulate(5,0.01)
-```
-
-$10^5$ 0.5%
-
-```{r}
-tabulate(5,0.005)
-```
-
-$10^4$ 5%
-
-```{r}
-tabulate.4(4,0.05)
-```
-
-$10^4$ 1%
-
-```{r}
-tabulate.4(4,0.01)
-```
-
-$10^4$ 0.5%
-
-```{r}
-tabulate.4(4,0.005)
-```
+This is so much better than before.

### 2016-03-17 doc/draft.Rmd
methods draft

@@
 
-### Viral populations
+### Mutagenesis
 
-We extracted viral RNA from cleared supernatants using  QIAamp viral RNA minikits (Qiagen 5204) according to the manufactures' instructions, and generated cDNA using Superscript III one-step with HiFi platinum Taq (Invitrogen 12574).
-For the WSN33, PR8 data set we used a segment specific primers PB2-Forward-JT	(GCAGGTCAATTATATTCAATATGGAAA),PB2-Reverse-JT	(CAAGGTCGTTTTTAAACTATTCGACAC), PB1-Forward-JT (GCAGGCAAACCATTTGAATGG), PB1-Reverse-JT	(CAAGGCATTTTTTCATGAAGGACAAG), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG), NA-Forward-JT	(GCAGGAGTTTAAATGAATCCAAACC), NA-Reverse-JT	(CAAGGAGTTTTTTGAACAAACTACTTG), HA-Forward-JT	(GCAGGGGAAAATAAAAACAACCAAAAT), HA-Reverse-JT	(CAAGGGTGTTTTTCCTTATATTTCTGAA), NP-Forward-JT	(GCAGGGTAGATAATCACTCACAG), NP-Reverse-JT	(CAAGGGTATTTTTCTTTAATTGTCGTACT), M-Forward-JT	(GCAGGTAGATATTGAAAGATGAGTC), M-Reverse-JT	(CAAGGTAGTTTTTTACTCCAGCTCT), NS-Forward-JT	(GCAGGGTGACAAAGACATAATG), NS-Reverse-JT	(CAAAGGGTGTTTTTTATTATTAAATAAGCTG), which with the reaction conditions were adapted from [@Anonymous:A25Ka2tY].
-Each 25 $\mu$ l reaction included 1 $\mu$ l of each primer (5 $\mu$ M) and 2 $\mu$ l of RNA template.
-Reaction conditions were 50$^\circ$ C (60 min), 94$^\circ$ C (2 min), and then 30 cycles of 94$^\circ$ C (30 sec), 54$^\circ$ C (30 sec), and 68$^\circ$ C (3 min).
+Point mutants were made in the WSN33 background using the pHW2000 reverse genetics system (cite)  as previously described.
+Briefly, primers ..........
+were used in overlap PCR to generate mutations at HA\_T1583G,HA\_G1006T,HA\_G542T,M\_T861G,M\_A541C,NA\_G1168T,NA\_C454T,NP\_A454C,NP\_A1160T,NS\_G227T,NS\_A809G,PA\_T964G,PA\_T237A,PA\_A1358T,PB1\_G599A,PB1\_G1764T,PB1\_T1288A,PB2\_A1854G,PB2\_A440T,PB2\_A1167T.
 
-### Mutagenesis
+### Library preperation
+
+Seven hundred and fifty nanograms of purified PCR products were sheared to an average size of 300-400 bp using a Covaris S220 focused ultrasonicator at the following settings : Intensity 4, Duty cycle 10%, Burst/second 200, time 80 seconds.
+Sequencing libraries were prepared from these fragmented products using the NEBNext Ultra DNA library prep kit (NEB E7370L) and NEBNext multiplex oligonucleotides for Illumina (NEB E7600S).
+We performed size selection during library preparation using Agencourt AMPure XP beads (Beckman Coulter A63881) with a target insert size of ......
+according to the manufactures' instructions.
 
@@
 
+We extracted viral RNA from infected supernatants using  QIAamp viral RNA minikits (Qiagen 5204) according to the manufactures' instructions, and generated cDNA using Superscript III one-step with HiFi platinum Taq (Invitrogen 12574).
+PCR products were purified using the Thermo Scientific PCR clean-up according to the manufactures' instructions.
+
+For the WSN33, PR8 data set we used segment specific reactions with the following primers : PB2-Forward-JT	(GCAGGTCAATTATATTCAATATGGAAA), PB2-Reverse-JT	(CAAGGTCGTTTTTAAACTATTCGACAC), PB1-Forward-JT (GCAGGCAAACCATTTGAATGG), PB1-Reverse-JT	(CAAGGCATTTTTTCATGAAGGACAAG), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG), NA-Forward-JT	(GCAGGAGTTTAAATGAATCCAAACC), NA-Reverse-JT	(CAAGGAGTTTTTTGAACAAACTACTTG), HA-Forward-JT	(GCAGGGGAAAATAAAAACAACCAAAAT), HA-Reverse-JT	(CAAGGGTGTTTTTCCTTATATTTCTGAA), NP-Forward-JT	(GCAGGGTAGATAATCACTCACAG), NP-Reverse-JT	(CAAGGGTATTTTTCTTTAATTGTCGTACT), M-Forward-JT	(GCAGGTAGATATTGAAAGATGAGTC), M-Reverse-JT	(CAAGGTAGTTTTTTACTCCAGCTCT), NS-Forward-JT	(GCAGGGTGACAAAGACATAATG), NS-Reverse-JT	(CAAAGGGTGTTTTTTATTATTAAATAAGCTG), which with the reaction conditions were adapted from [@Anonymous:A25Ka2tY].
+Each 25 $\mu$ l reaction included 1 $\mu$ l of each primer (5 $\mu$ M) and 2 $\mu$ l of RNA template.
+Reaction conditions were 50$^\circ$ C (60 min), 94$^\circ$ C (2 min), followed by 30 cycles of 94$^\circ$ C (30 sec), 54$^\circ$ C (30 sec), and 68$^\circ$ C (3 min).
+PCR products were pooled to create equimolar mixes of WSN33 and PR8 segments.
+The WNS33 sample was then diluted into PR8 serially to yield mixtures in which WSN33 made up 5,2.5,1.25,and 0.63% of the population.
+We processed the samples for sequencing as below, and sequenced pooled libraries on an Illumina Miseq machine using 2X250 paired end reads.
+
+To more closely mimic patient-derived viral populations we quantified the genome copy number of each of the 20 WSN33 point mutants (created above) and WSN33 wt using ...
+and the primers ...
+as described in ...
+.
+The mutants were mixed at equal concentrations and diluted to generate a mixture with each mutant present at 5% frequency and a total concentration of 10^5^ genomes per microliter.
+We diluted this population into the wt sample to create populations in which each mutant was present at 2, 1, 0.5,and 0.2% frequency all with a total concentration of 10^5^ genomes per micoliter.
+These 5 populations were diluted serially into basal media to generate samples with the same mutant frequencies as before but at nucleic acid concentrations of 10^4^ and 10^3^ genomes per microliter.
+Viral RNA was extracted from these samples as above and cDNA was generated in a multiplex RT-PCR as in [@Anonymous:A25Ka2tY] using the primers primers Uni12/Inf1 (GGGGGGAGCAAAAGCAGG), Uni12/Inf3 (GGGGGAGCGAAAGCAGG), and Uni13/Inf1 (5CGGGTTATTAGTAGAAACAAGG).
+Two multiplex RT-PCR reactions were run on the 10^5^ genomes per micoliter wt sample.
+We prepared libraries as before and used picogreen (.......) to quantify the concentration of each indexed library.
+_____ ng of each library was pooled together, gel isolated using __________, and sequenced using 2X125 paired end reads on an Illumina Hiseq machine.
+
 ### Plasmid controls
-
-### Library preperation
 
@@
 
+Reads were aligned to either to PR8 (PR8 WSN33 data set) or WSN33 (pateint derived conditions) using Bowtie2 (cite).
+Aligments were sorted and processed using Picard (......),variants were called using either DeepSNV or Lofreq and filtered using in house python and R scripts.
+We connected these steps into an analysis pipeline using bpipe.
+All commands needed to complete the analysis presented in this work are publically available and free for download at .
+
 #References

### 2016-03-17 doc/draft.Rmd
methods work

@@
 We observed an average read depth of over 10,000 reads/bp in these samples.
-Therefore, the drop in sensitivity is most likely due to the 1,000-fold decrease in nucleaic acid concentration compared to the first data set and the fact that library preparation requires a number of sampling steps that may limit detection.
+Therefore, the drop in sensitivity is most likely due to the 1,000-fold decrease in nucleic acid concentration compared to the first data set and the fact that library preparation requires a number of sampling steps that may limit detection.
 During library preparation, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.
@@
 
-We passaged plaque purified WSN33 and PR8 lab strains 3 times in
+Madin-Darby canine kidney cells were provided by Arnold S.
+Monto (University of Michigan School of Public Health) and were maintained in Dulbecco's Modified Eagle Medium (DMEM, Invitrogen) with 10% fetal bovine serum (Gibco and HyClone), 25mM HEPES (Invitogen), and 0.1875% bovine serum albumin (Life Technologies) as in (MFE submitted).
+WSN33 was rescued from transfections using the 8 plasmid reverse genetic system containing the genomic segments for influenza A/WSN33 (pHW181-188) was a kind gift from Rober Webster (St.
+Jude's Children's Research Hospital) [@Hoffmann:2002ih].
+A biological clone of influenza A/Puerto Rico/8/1934 was obtained from ATCC (VR-1469) and was cloned into the pHW2000 reverse genetic system as before [@Pauly:2015eq]
 
@@
 
-Point mutants were made in the WSN33 background using the pHW2000 reverse genetics system (cite)  as previously described.
-Briefly, primers ..........
-were used in overlap PCR to generate mutations at HA\_T1583G,HA\_G1006T,HA\_G542T,M\_T861G,M\_A541C,NA\_G1168T,NA\_C454T,NP\_A454C,NP\_A1160T,NS\_G227T,NS\_A809G,PA\_T964G,PA\_T237A,PA\_A1358T,PB1\_G599A,PB1\_G1764T,PB1\_T1288A,PB2\_A1854G,PB2\_A440T,PB2\_A1167T.
+Point mutants were made in the WSN33 background using the pHW2000 reverse genetics system (cite)  as previously described (MFE citation) .
+Briefly, we used  PCR mutagenesis to generate mutations at HA\_T1583G, HA\_G1006T, HA\_G542T, M\_T861G, M\_A541C, NA\_G1168T, NA\_C454T, NP\_A454C, NP\_A1160T, NS\_G227T, NS\_A809G, PA\_T964G, PA\_T237A, PA\_A1358T, PB1\_G599A, PB1\_G1764T, PB1\_T1288A, PB2\_A1854G, PB2\_A440T, PB2\_A1167T.
+Mutants were transfected and passaged once in MDCK cells prior to sequencing.
 
@@
 Sequencing libraries were prepared from these fragmented products using the NEBNext Ultra DNA library prep kit (NEB E7370L) and NEBNext multiplex oligonucleotides for Illumina (NEB E7600S).
-We performed size selection during library preparation using Agencourt AMPure XP beads (Beckman Coulter A63881) with a target insert size of ......
-according to the manufactures' instructions.
+We performed size selection during library preparation using Agencourt AMPure XP beads (Beckman Coulter A63881) with a target insert size of 300-400bp or 250 bp for sequencing on Illumina Miseq and Hiseq respectively.
 
@@
 
-We extracted viral RNA from infected supernatants using  QIAamp viral RNA minikits (Qiagen 5204) according to the manufactures' instructions, and generated cDNA using Superscript III one-step with HiFi platinum Taq (Invitrogen 12574).
+We extracted viral RNA from infected supernaturals using  QIAamp viral RNA minikits (Qiagen 5204) according to the manufactures' instructions, and generated cDNA using Superscript III one-step with HiFi platinum Taq (Invitrogen 12574).
 PCR products were purified using the Thermo Scientific PCR clean-up according to the manufactures' instructions.
 
-For the WSN33, PR8 data set we used segment specific reactions with the following primers : PB2-Forward-JT	(GCAGGTCAATTATATTCAATATGGAAA), PB2-Reverse-JT	(CAAGGTCGTTTTTAAACTATTCGACAC), PB1-Forward-JT (GCAGGCAAACCATTTGAATGG), PB1-Reverse-JT	(CAAGGCATTTTTTCATGAAGGACAAG), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG), NA-Forward-JT	(GCAGGAGTTTAAATGAATCCAAACC), NA-Reverse-JT	(CAAGGAGTTTTTTGAACAAACTACTTG), HA-Forward-JT	(GCAGGGGAAAATAAAAACAACCAAAAT), HA-Reverse-JT	(CAAGGGTGTTTTTCCTTATATTTCTGAA), NP-Forward-JT	(GCAGGGTAGATAATCACTCACAG), NP-Reverse-JT	(CAAGGGTATTTTTCTTTAATTGTCGTACT), M-Forward-JT	(GCAGGTAGATATTGAAAGATGAGTC), M-Reverse-JT	(CAAGGTAGTTTTTTACTCCAGCTCT), NS-Forward-JT	(GCAGGGTGACAAAGACATAATG), NS-Reverse-JT	(CAAAGGGTGTTTTTTATTATTAAATAAGCTG), which with the reaction conditions were adapted from [@Anonymous:A25Ka2tY].
+#### PR8 WSN33 population 
+WSN33 and PR8 isolates were plaque purified and passaged three times in MCDK cells.
+We varified the sequences of the viruses using Sanger sequencing.
+cDNA was generated in segment specific reactions using : PB2-Forward-JT	(GCAGGTCAATTATATTCAATATGGAAA), PB2-Reverse-JT	(CAAGGTCGTTTTTAAACTATTCGACAC), PB1-Forward-JT (GCAGGCAAACCATTTGAATGG), PB1-Reverse-JT	(CAAGGCATTTTTTCATGAAGGACAAG), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG), NA-Forward-JT	(GCAGGAGTTTAAATGAATCCAAACC), NA-Reverse-JT	(CAAGGAGTTTTTTGAACAAACTACTTG), HA-Forward-JT	(GCAGGGGAAAATAAAAACAACCAAAAT), HA-Reverse-JT	(CAAGGGTGTTTTTCCTTATATTTCTGAA), NP-Forward-JT	(GCAGGGTAGATAATCACTCACAG), NP-Reverse-JT	(CAAGGGTATTTTTCTTTAATTGTCGTACT), M-Forward-JT	(GCAGGTAGATATTGAAAGATGAGTC), M-Reverse-JT	(CAAGGTAGTTTTTTACTCCAGCTCT), NS-Forward-JT	(GCAGGGTGACAAAGACATAATG), NS-Reverse-JT	(CAAAGGGTGTTTTTTATTATTAAATAAGCTG), which with the reaction conditions were adapted from [@Anonymous:A25Ka2tY].
 Each 25 $\mu$ l reaction included 1 $\mu$ l of each primer (5 $\mu$ M) and 2 $\mu$ l of RNA template.
@@
 
-Reads were aligned to either to PR8 (PR8 WSN33 data set) or WSN33 (pateint derived conditions) using Bowtie2 (cite).
-Aligments were sorted and processed using Picard (......),variants were called using either DeepSNV or Lofreq and filtered using in house python and R scripts.
+Reads were aligned to either to PR8 (PR8 WSN33 data set) or WSN33 (patient derived conditions) using Bowtie2 (cite).
+Alignments were sorted and processed using Picard (......),variants were called using either DeepSNV or Lofreq and filtered using in house python and R scripts.
 We connected these steps into an analysis pipeline using bpipe.
-All commands needed to complete the analysis presented in this work are publically available and free for download at .
+All commands needed to complete the analysis presented in this work are public ally available and free for download at .
 

### 2016-03-18 doc/draft.Rmd
methods draft

@@
 Monto (University of Michigan School of Public Health) and were maintained in Dulbecco's Modified Eagle Medium (DMEM, Invitrogen) with 10% fetal bovine serum (Gibco and HyClone), 25mM HEPES (Invitogen), and 0.1875% bovine serum albumin (Life Technologies) as in (MFE submitted).
-WSN33 was rescued from transfections using the 8 plasmid reverse genetic system containing the genomic segments for influenza A/WSN33 (pHW181-188) was a kind gift from Rober Webster (St.
+WSN33 was rescued from transfections using the 8 plasmid reverse genetic system containing the genomic segments for influenza A/WSN33 (pHW181-188), a kind gift from Rober Webster (St.
 Jude's Children's Research Hospital) [@Hoffmann:2002ih].
-A biological clone of influenza A/Puerto Rico/8/1934 was obtained from ATCC (VR-1469) and was cloned into the pHW2000 reverse genetic system as before [@Pauly:2015eq]
+A biological clone of influenza A/Puerto Rico/8/1934 was obtained from ATCC (VR-1469) and was cloned into the pHW2000 reverse genetic system to be used as a sequencing control as before [@Pauly:2015eq].
 
@@
 
-Point mutants were made in the WSN33 background using the pHW2000 reverse genetics system (cite)  as previously described (MFE citation) .
-Briefly, we used  PCR mutagenesis to generate mutations at HA\_T1583G, HA\_G1006T, HA\_G542T, M\_T861G, M\_A541C, NA\_G1168T, NA\_C454T, NP\_A454C, NP\_A1160T, NS\_G227T, NS\_A809G, PA\_T964G, PA\_T237A, PA\_A1358T, PB1\_G599A, PB1\_G1764T, PB1\_T1288A, PB2\_A1854G, PB2\_A440T, PB2\_A1167T.
-Mutants were transfected and passaged once in MDCK cells prior to sequencing.
+Point mutants were made in the WSN33 background using the pHW2000 reverse genetics system [@Hoffmann:2002ih] as previously described (MFE citation) .
+Briefly, we used  overlap PCR mutagenesis to generate mutations at HA\_T1583G, HA\_G1006T, HA\_G542T, M\_T861G, M\_A541C, NA\_G1168T, NA\_C454T, NP\_A454C, NP\_A1160T, NS\_G227T, NS\_A809G, PA\_T964G, PA\_T237A, PA\_A1358T, PB1\_G599A, PB1\_G1764T, PB1\_T1288A, PB2\_A1854G, PB2\_A440T, PB2\_A1167T.
 
@@
 
-Seven hundred and fifty nanograms of purified PCR products were sheared to an average size of 300-400 bp using a Covaris S220 focused ultrasonicator at the following settings : Intensity 4, Duty cycle 10%, Burst/second 200, time 80 seconds.
+Seven hundred and fifty nanograms of purified PCR products were sheared to an average size of 300-400 bp using a Covaris S220 focused ultrasonicator using the following settings : Intensity:4, Duty cycle:10%, Burst/second: 200, Duration: 80 seconds.
 Sequencing libraries were prepared from these fragmented products using the NEBNext Ultra DNA library prep kit (NEB E7370L) and NEBNext multiplex oligonucleotides for Illumina (NEB E7600S).
@@
 
-We extracted viral RNA from infected supernaturals using  QIAamp viral RNA minikits (Qiagen 5204) according to the manufactures' instructions, and generated cDNA using Superscript III one-step with HiFi platinum Taq (Invitrogen 12574).
-PCR products were purified using the Thermo Scientific PCR clean-up according to the manufactures' instructions.
+We extracted viral RNA from infected supernaturants using  QIAamp viral RNA minikits (Qiagen 5204) according to the manufactures' instructions, and generated cDNA using Superscript III one-step with HiFi platinum Taq (Invitrogen 12574).
+PCR products were purified using the GeneJet PCR Purification Kit (ThermoFisher Scientific K0701) according to the manufactures' instructions.
 
 #### PR8 WSN33 population
+
 WSN33 and PR8 isolates were plaque purified and passaged three times in MCDK cells.
-We varified the sequences of the viruses using Sanger sequencing.
+We then varified the sequences of the these viruses using Sanger sequencing.
 cDNA was generated in segment specific reactions using : PB2-Forward-JT	(GCAGGTCAATTATATTCAATATGGAAA), PB2-Reverse-JT	(CAAGGTCGTTTTTAAACTATTCGACAC), PB1-Forward-JT (GCAGGCAAACCATTTGAATGG), PB1-Reverse-JT	(CAAGGCATTTTTTCATGAAGGACAAG), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG), NA-Forward-JT	(GCAGGAGTTTAAATGAATCCAAACC), NA-Reverse-JT	(CAAGGAGTTTTTTGAACAAACTACTTG), HA-Forward-JT	(GCAGGGGAAAATAAAAACAACCAAAAT), HA-Reverse-JT	(CAAGGGTGTTTTTCCTTATATTTCTGAA), NP-Forward-JT	(GCAGGGTAGATAATCACTCACAG), NP-Reverse-JT	(CAAGGGTATTTTTCTTTAATTGTCGTACT), M-Forward-JT	(GCAGGTAGATATTGAAAGATGAGTC), M-Reverse-JT	(CAAGGTAGTTTTTTACTCCAGCTCT), NS-Forward-JT	(GCAGGGTGACAAAGACATAATG), NS-Reverse-JT	(CAAAGGGTGTTTTTTATTATTAAATAAGCTG), which with the reaction conditions were adapted from [@Anonymous:A25Ka2tY].
 Each 25 $\mu$ l reaction included 1 $\mu$ l of each primer (5 $\mu$ M) and 2 $\mu$ l of RNA template.
-Reaction conditions were 50$^\circ$ C (60 min), 94$^\circ$ C (2 min), followed by 30 cycles of 94$^\circ$ C (30 sec), 54$^\circ$ C (30 sec), and 68$^\circ$ C (3 min).
-PCR products were pooled to create equimolar mixes of WSN33 and PR8 segments.
-The WNS33 sample was then diluted into PR8 serially to yield mixtures in which WSN33 made up 5,2.5,1.25,and 0.63% of the population.
-We processed the samples for sequencing as below, and sequenced pooled libraries on an Illumina Miseq machine using 2X250 paired end reads.
+Reaction conditions were 50$^\circ\text{C}$ (60 min), 94$^\circ\text{C}$ (2 min), followed by 30 cycles of 94$^\circ\text{C}$ (30 sec), 54$^\circ\text{C}$ (30 sec), and 68$^\circ\text{C}$ (3 min).
+PCR products were pooled equimolarly to reconstituded cDNA genomes of WSN33 and PR8.
+WSN33 was then diluted into PR8 serially to yield mixtures in which WSN33 made up 5,2.5,1.25,and 0.63% of the population.
+We processed the samples for sequencing as above, pooled equal volumes of the indexed libraries, sequenced the pooled libraries on an Illumina Miseq machine using 2X250 paired end reads.
+A clonal plasmid control library was prepared from 8 plasmids containing PR8 genomic segments.
+These plasmids were mixed equimolarly and cDNA was generated using a multiplex RT-PCR with the primers Uni12/Inf1 (GGGGGGAGCAAAAGCAGG), Uni12/Inf3 (GGGGGAGCGAAAGCAGG), and Uni13/Inf1 (5CGGGTTATTAGTAGAAACAAGG) as in [@Anonymous:A25Ka2tY].
+This library was then prepared in identical fashion to the experimental populations and was sequenced on the same Miseq lane.
 
-To more closely mimic patient-derived viral populations we quantified the genome copy number of each of the 20 WSN33 point mutants (created above) and WSN33 wt using ...
-and the primers ...
-as described in ...
-.
+#### Experimental intrahost population
+
+We quantified the genome copy number of each of the 20 WSN33 point mutants (created above) and WSN33 wt using SuperScript III Platinum One-Step qRT-PCR (Invitrogen 11732).
+Primers and probes were designed by the by the CDC Influenza Branch for the universal detection of influenza A and B.
 The mutants were mixed at equal concentrations and diluted to generate a mixture with each mutant present at 5% frequency and a total concentration of 10^5^ genomes per microliter.
-We diluted this population into the wt sample to create populations in which each mutant was present at 2, 1, 0.5,and 0.2% frequency all with a total concentration of 10^5^ genomes per micoliter.
+We diluted this mixture into the wt sample to create populations in which each mutant was present at 2, 1, 0.5,and 0.2% frequency all with a total concentration of 10^5^ genomes per micoliter.
 These 5 populations were diluted serially into basal media to generate samples with the same mutant frequencies as before but at nucleic acid concentrations of 10^4^ and 10^3^ genomes per microliter.
-Viral RNA was extracted from these samples as above and cDNA was generated in a multiplex RT-PCR as in [@Anonymous:A25Ka2tY] using the primers primers Uni12/Inf1 (GGGGGGAGCAAAAGCAGG), Uni12/Inf3 (GGGGGAGCGAAAGCAGG), and Uni13/Inf1 (5CGGGTTATTAGTAGAAACAAGG).
+Viral RNA was extracted from these samples as above and cDNA was generated in a multiplex RT-PCR as in [@Anonymous:A25Ka2tY] using the primers Uni12/Inf1 (GGGGGGAGCAAAAGCAGG), Uni12/Inf3 (GGGGGAGCGAAAGCAGG), and Uni13/Inf1 (5CGGGTTATTAGTAGAAACAAGG).
 Two multiplex RT-PCR reactions were run on the 10^5^ genomes per micoliter wt sample.
-We prepared libraries as before and used picogreen (.......) to quantify the concentration of each indexed library.
-_____ ng of each library was pooled together, gel isolated using __________, and sequenced using 2X125 paired end reads on an Illumina Hiseq machine.
-
-### Plasmid controls
+We prepared libraries as before and used  Quanti PicoGreen dsDNA quantification (ThermoFisher Scientific P7589) to quantify the concentration of each indexed library.
+Forty six ng of each indexed library was pooled together, gel isolated using the GeneJET Gel Extraction Kit (ThermoFisher Scientific K0691), and sequenced using 2X125 paired end reads on an Illumina Hiseq 2000 machine.
+A clonal control library was processed in an identical fashion starting from an equimolar mix of 8 plasmids containing the WSN33 genomic segments.
 
@@
 
-Reads were aligned to either to PR8 (PR8 WSN33 data set) or WSN33 (patient derived conditions) using Bowtie2 (cite).
-Alignments were sorted and processed using Picard (......),variants were called using either DeepSNV or Lofreq and filtered using in house python and R scripts.
-We connected these steps into an analysis pipeline using bpipe.
-All commands needed to complete the analysis presented in this work are public ally available and free for download at .
+Reads were aligned to either to PR8 (PR8 WSN33 data set) or WSN33 (patient derived conditions) using Bowtie2 [@Langmead:2012jh].
+Alignments were sorted and processed using Picard (http://broadinstitute.github.io/picard/), variants were called using either DeepSNV [@Gerstung:1js] or Lofreq [@Wilm:2012br] and filtered using the Pysam module in python (https://github.com/pysam-developers/pysam) and R scripts.
+Bases with a Phred<30 were masked in the DeepSNV analysis.
+We connected these steps into an analysis pipeline using bpipe [@em:2012wd] which is available for download at .https://github.com/lauringlab/variant\_pipeline.
+Additionally, all commands needed to complete the analysis presented in this work are publically available and free for download at https://github.com/lauringlab/Benchmarking\_paper .
+Finaly and interactive shiny app of our  benchmarking work can be
+downloaded at https://github.com/lauringlab/benchmarking\_shiny.
 

### 2016-03-18 doc/draft.Rmd
methods work

--- 2016-03-17 doc/draft.Rmd
@@
 
-##Introduction
+**Benchmarking and quality controls are required in sequence based studies of intrahost diversity.**
+
+\n
+
+\n
+
+\n
+
+John T. McCrone^1^ and Adam S. Lauring^1,2^ *
+
+\n
+
+\n
+
+1 Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109
+
+2 Division of Infectious Diseases, Department of Internal Medicine, University of Michigan, Ann Arbor, MI 48109
+
+\n
+
+\n
+
+\n
+
+\n
+
+\n
+
+\n
+
+\n
+* Corresponding author
+
+\n
+
+Adam S. Lauring
+
+1150 W. Medical Center Dr.
+
+MSRB1 Room 5510B
+
+Ann Arbor, MI 48109-5680
+
+alauring@med.umich.edu
+
+#Introduction
 
 The advent of next generation sequencing (NGS) is revolutionizing the study of viral evolution and molecular epidemiology through high-throughput measures of intrahost diversity.
-Until recently investigations of viral diversity were limited by labor intensive sequencing approaches; however, it is now feasible to sequence patient-derived samples at sufficient read depth to detect rare single nucleotide variants (SNV) (cite examples?).
-The high-throughput, cost-effective nature of NGS has lead to an explosion of studies that quantify viral diversity within and between hosts (cite examples?).
+Until recently investigations of viral diversity were limited by labor intensive sequencing approaches; however, it is now feasible to sequence patient-derived samples at sufficient read depth to detect rare single nucleotide variants (SNV), and the high-throughput, cost-effective nature of NGS has lead to an explosion of studies that quantify viral diversity within and between hosts (e.g.[@Andersen:2015jj], [@Grubaugh:2015by], [@Rogers:2015ij],[@Poon:2016gf]).
 Given the increasing contribution of NGS to the field of virology it is important to reflect on the accuracy of these methods and the controls that are needed to ensure robust conclusions.
@@
 Because next generation sequencing of viral populations often produces tens of thousands of reads per genomic position, the sensitivity of these approaches has been assumed to be quite robust (cite).
-However, patient derived samples often differ in input titer by orders of magnitude, and this variability can have drastic impacts on the sensitivity of detecting rare events.
+However, patient derived samples can differ in input titer by orders of magnitude (e.g.
+[@Lau:2013ima],[@Teunis:2015iw],[@Takeyama:2015jg]), and this variability can have drastic impacts on the sensitivity of detecting rare events.
 Furthermore, some library preparations result in variable sequencing coverage both within and between samples.
@@
 This work exemplifies the controls that should be run prior to any NGS based study of viral populations and provides a comprehensive data set for benchmarking other pipelines.
+
+## Methods
+
+### Viruses and cells
+
+Madin-Darby canine kidney cells were provided by Arnold S.
+Monto (University of Michigan School of Public Health) and were maintained in Dulbecco's Modified Eagle Medium (DMEM, Invitrogen) with 10% fetal bovine serum (Gibco and HyClone), 25mM HEPES (Invitogen), and 0.1875% bovine serum albumin (Life Technologies).
+Influenza A/WSN/33(H1N1) virus was rescued from transfected cells using the 8 plasmid reverse genetic system containing the genomic segments (pHW181-188), a kind gift from Robert Webster (St.
+Jude's Children's Research Hospital) [@Hoffmann:2002ih].
+A biological clone of influenza A/Puerto Rico/8/1934 was obtained from ATCC (VR-1469) and the genomic segments were cloned into the pHW2000 reverse genetic system [@Pauly:2015eq].
+
+### Viral populations
+
+We extracted viral RNA from infected supernaturants using  QIAamp Viral RNA kits (Qiagen 5204) and generated cDNA using Superscript III one-step with HiFi platinum Taq (Invitrogen 12574).
+PCR products were purified using the GeneJet PCR Purification Kit (ThermoFisher Scientific K0701) according to the manufactures' instructions.
+
+#### PR8 WSN33 population
+
+For the experiment in `r Miseq`, WSN33 and PR8 viruses were plaque purified and passaged three times in MDCK cells.
+We then verified the sequences of the these viruses by Sanger sequencing.
+Two microliters of RNA template was used  to generate cDNA in 25 $\mu\text{l}$ segment specific one step RT-PCR with $0.2\mu\text{M}$ of the following primers :   PB2-Forward-JT	(5'-GCAGGTCAATTATATTCAATATGGAAA-3'), PB2-Reverse-JT	(5'-CAAGGTCGTTTTTAAACTATTCGACAC-3'), PB1-Forward-JT (5'-GCAGGCAAACCATTTGAATGG-3'), PB1-Reverse-JT	(5'-CAAGGCATTTTTTCATGAAGGACAAG-3'), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG-3'), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG-3'), NA-Forward-JT	5'-(GCAGGAGTTTAAATGAATCCAAACC-3'), NA-Reverse-JT	(5'-CAAGGAGTTTTTTGAACAAACTACTTG-3'), HA-Forward-JT	(5'-GCAGGGGAAAATAAAAACAACCAAAAT-3'), HA-Reverse-JT	(5'-CAAGGGTGTTTTTCCTTATATTTCTGAA-3'), NP-Forward-JT	(5'-GCAGGGTAGATAATCACTCACAG-3'), NP-Reverse-JT	(5'-CAAGGGTATTTTTCTTTAATTGTCGTACT-3'), M-Forward-JT	(5'-GCAGGTAGATATTGAAAGATGAGTC-3'), M-Reverse-JT	(5'-CAAGGTAGTTTTTTACTCCAGCTCT-3'), NS-Forward-JT	(5'-GCAGGGTGACAAAGACATAATG-3'), NS-Reverse-JT	(5'-CAAAGGGTGTTTTTTATTATTAAATAAGCTG-3').
+Reaction conditions were $50^\circ\text{C}$ (60 min), $94^\circ\text{C}$ (2 min), followed by 30 cycles of $94^\circ\text{C}$ (30 sec), $54^\circ\text{C}$ (30 sec), and $68^{\circ}\text{C}$ (3 min).
+Molar equivalents of each PCR product were pooled to generate reconstituted cDNA genomes of WSN33 and PR8.
+The WSN33 cDNA pool was then serially diluted into the PR8 cDNA pool to yield mixtures in which WSN33 made up 5, 2.5, 1.25,and 0.63% of the population.
+Seven hundred and fifty nanograms of purified PCR products were sheared to an average size of 300-400 bp using a Covaris S220 focused ultrasonicator with the following settings : Intensity: 4, Duty cycle: 10%, Burst/second: 200, Duration: 80 seconds.
+Sequencing libraries were prepared from these fragmented products using the NEBNext Ultra DNA library prep kit (NEB E7370L), Agencourt AMPure XP beads (Beckman Coulter A63881), and NEBNext multiplex oligonucleotides for Illumina (NEB E7600S).
+Pooled libraries were sequenced on an Illumina Miseq machine using 2X250 paired end reads.
+A clonal plasmid control library was prepared from 8 plasmids containing PR8 genomic segments.
+These plasmids were mixed equimolarly and cDNA was generated using a multiplex one step RT-PCR with the primers Uni12/Inf1 (5'-GGGGGGAGCAAAAGCAGG-3'), Uni12/Inf3 (5'-GGGGGAGCGAAAGCAGG-3'), and Uni13/Inf1 (5'-5CGGGTTATTAGTAGAAACAAGG-3') as in [@Anonymous:A25Ka2tY].
+This library was then prepared in identical fashion to the experimental populations and was sequenced in the same Miseq lane.
+
+#### Experimental intrahost population
+
+Twenty point mutants were generated in the WSN33 background using the pHW2000 reverse genetics system [@Hoffmann:2002ih] as previously described (Visher et al., manuscript in preparation).
+In short, we used  overlap PCR mutagenesis to introduce the following mutations: HA\_T1583G, HA\_G1006T, HA\_G542T, M\_T861G, M\_A541C, NA\_G1168T, NA\_C454T, NP\_A454C, NP\_A1160T, NS\_G227T, NS\_A809G, PA\_T964G, PA\_T237A, PA\_A1358T, PB1\_G599A, PB1\_G1764T, PB1\_T1288A, PB2\_A1854G, PB2\_A440T, PB2\_A1167T.
+
+We quantified the genome copy number of each of the 20 WSN33 point mutants and the WSN33 WT using a SuperScript III Platinum One-Step qRT-PCR (Invitrogen 11732) and univeral influenza A/B primer and probe sets [@Anonymous:tc].
+Infected supernatants were mixed at equal concentrations and diluted to generate a population with each mutant present at 5% frequency and a total concentration of 10^5^ genomes per microliter.
+We diluted this mixture into WT WSN33 supernatant to create populations in which each mutant was present at 2, 1, 0.5,and 0.2% frequency all with a total concentration of 10^5^ genomes per micoliter.
+These 5 populations were diluted serially into basal media to generate samples with total nucleic acid concentrations of 10^4^ and 10^3^ genomes per microliter.
+Viral RNA was extracted from these samples and cDNA was generated in a one step multiplex RT-PCR as above.
+The WT WSN33 sample (10^5^ genomes per micoliter) was processed and sequenced in duplicate.
+We prepared libraries as before and used Quanti PicoGreen dsDNA quantification (ThermoFisher Scientific P7589) to quantify the concentration of each indexed library.
+We pooled equal quantifies (by nanogram) of each indexed library, and removed adaptor dimers using gel isolation with the GeneJET Gel Extraction Kit (ThermoFisher Scientific K0691) prior to sequencing using 2 x 125 paired end reads on an Illumina Hiseq 2500.
+A clonal control library was processed in an identical fashion starting from an equimolar mix of 8 plasmids containing the WSN33 genomic segments.
+
+#### Duplicate populations
+
+Fresh RNA was isolated from the 5%, 2%, 1%, and 0.5% 10^4^ genomes per microliter samples and processed in duplicate as above, and these duplicates were sequenced with a plasmid control using 2 x 125 paired end reads on an Illumina Hiseq 2500 machine.
+
+### Sequence analysis
+
+ Reads were aligned to either a PR8 or a WSN33 reference sequence using Bowtie2 [@Langmead:2012jh].
+Alignments were sorted and PCR duplicateds removed using Picard (http://broadinstitute.github.io/picard/).
+Variants were called using either DeepSNV [@Gerstung:1js] or Lofreq [@Wilm:2012br] and filtered using the Pysam module in python (https://github.com/pysam-developers/pysam) and custom R scripts.
+Bases with a Phred<30 were masked in the DeepSNV analysis.
+We connected all of these steps into an analysis pipeline using bpipe [@em:2012wd] which is available for download at https://github.com/lauringlab/variant\_pipeline.
+Additionally, this work is completely reproducible.
+All raw fastq files were submitted to the Sequence Read Archive (SRA) under accession number ___________, and all commands required generate the figures in the manuscript are available for anonymous download at https://github.com/lauringlab/Benchmarking\_paper .
+Finally an interactive shiny app of our  benchmarking work can be
+downloaded at https://github.com/lauringlab/benchmarking\_shiny.
 
@@
 
-## Methods
+**Acknowledgements**
 
-### Viruses and cells
-
-We passaged plaque purified WSN33 and PR8 lab strains 3 times in
-
-### Mutagenesis
-
-Point mutants were made in the WSN33 background using the pHW2000 reverse genetics system (cite)  as previously described.
-Briefly, primers ..........
-were used in overlap PCR to generate mutations at HA\_T1583G,HA\_G1006T,HA\_G542T,M\_T861G,M\_A541C,NA\_G1168T,NA\_C454T,NP\_A454C,NP\_A1160T,NS\_G227T,NS\_A809G,PA\_T964G,PA\_T237A,PA\_A1358T,PB1\_G599A,PB1\_G1764T,PB1\_T1288A,PB2\_A1854G,PB2\_A440T,PB2\_A1167T.
-
-### Library preperation
-
-Seven hundred and fifty nanograms of purified PCR products were sheared to an average size of 300-400 bp using a Covaris S220 focused ultrasonicator at the following settings : Intensity 4, Duty cycle 10%, Burst/second 200, time 80 seconds.
-Sequencing libraries were prepared from these fragmented products using the NEBNext Ultra DNA library prep kit (NEB E7370L) and NEBNext multiplex oligonucleotides for Illumina (NEB E7600S).
-We performed size selection during library preparation using Agencourt AMPure XP beads (Beckman Coulter A63881) with a target insert size of ......
-according to the manufactures' instructions.
-
-### Viral populations
-
-We extracted viral RNA from infected supernatants using  QIAamp viral RNA minikits (Qiagen 5204) according to the manufactures' instructions, and generated cDNA using Superscript III one-step with HiFi platinum Taq (Invitrogen 12574).
-PCR products were purified using the Thermo Scientific PCR clean-up according to the manufactures' instructions.
-
-For the WSN33, PR8 data set we used segment specific reactions with the following primers : PB2-Forward-JT	(GCAGGTCAATTATATTCAATATGGAAA), PB2-Reverse-JT	(CAAGGTCGTTTTTAAACTATTCGACAC), PB1-Forward-JT (GCAGGCAAACCATTTGAATGG), PB1-Reverse-JT	(CAAGGCATTTTTTCATGAAGGACAAG), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG), NA-Forward-JT	(GCAGGAGTTTAAATGAATCCAAACC), NA-Reverse-JT	(CAAGGAGTTTTTTGAACAAACTACTTG), HA-Forward-JT	(GCAGGGGAAAATAAAAACAACCAAAAT), HA-Reverse-JT	(CAAGGGTGTTTTTCCTTATATTTCTGAA), NP-Forward-JT	(GCAGGGTAGATAATCACTCACAG), NP-Reverse-JT	(CAAGGGTATTTTTCTTTAATTGTCGTACT), M-Forward-JT	(GCAGGTAGATATTGAAAGATGAGTC), M-Reverse-JT	(CAAGGTAGTTTTTTACTCCAGCTCT), NS-Forward-JT	(GCAGGGTGACAAAGACATAATG), NS-Reverse-JT	(CAAAGGGTGTTTTTTATTATTAAATAAGCTG), which with the reaction conditions were adapted from [@Anonymous:A25Ka2tY].
-Each 25 $\mu$ l reaction included 1 $\mu$ l of each primer (5 $\mu$ M) and 2 $\mu$ l of RNA template.
-Reaction conditions were 50$^\circ$ C (60 min), 94$^\circ$ C (2 min), followed by 30 cycles of 94$^\circ$ C (30 sec), 54$^\circ$ C (30 sec), and 68$^\circ$ C (3 min).
-PCR products were pooled to create equimolar mixes of WSN33 and PR8 segments.
-The WNS33 sample was then diluted into PR8 serially to yield mixtures in which WSN33 made up 5,2.5,1.25,and 0.63% of the population.
-We processed the samples for sequencing as below, and sequenced pooled libraries on an Illumina Miseq machine using 2X250 paired end reads.
-
-To more closely mimic patient-derived viral populations we quantified the genome copy number of each of the 20 WSN33 point mutants (created above) and WSN33 wt using ...
-and the primers ...
-as described in ...
-.
-The mutants were mixed at equal concentrations and diluted to generate a mixture with each mutant present at 5% frequency and a total concentration of 10^5^ genomes per microliter.
-We diluted this population into the wt sample to create populations in which each mutant was present at 2, 1, 0.5,and 0.2% frequency all with a total concentration of 10^5^ genomes per micoliter.
-These 5 populations were diluted serially into basal media to generate samples with the same mutant frequencies as before but at nucleic acid concentrations of 10^4^ and 10^3^ genomes per microliter.
-Viral RNA was extracted from these samples as above and cDNA was generated in a multiplex RT-PCR as in [@Anonymous:A25Ka2tY] using the primers primers Uni12/Inf1 (GGGGGGAGCAAAAGCAGG), Uni12/Inf3 (GGGGGAGCGAAAGCAGG), and Uni13/Inf1 (5CGGGTTATTAGTAGAAACAAGG).
-Two multiplex RT-PCR reactions were run on the 10^5^ genomes per micoliter wt sample.
-We prepared libraries as before and used picogreen (.......) to quantify the concentration of each indexed library.
-_____ ng of each library was pooled together, gel isolated using __________, and sequenced using 2X125 paired end reads on an Illumina Hiseq machine.
-
-### Plasmid controls
-
-### Sequence analysis
-
-Reads were aligned to either to PR8 (PR8 WSN33 data set) or WSN33 (pateint derived conditions) using Bowtie2 (cite).
-Aligments were sorted and processed using Picard (......),variants were called using either DeepSNV or Lofreq and filtered using in house python and R scripts.
-We connected these steps into an analysis pipeline using bpipe.
-All commands needed to complete the analysis presented in this work are publically available and free for download at .
+We thank Pat Schloss, Mike Imperiale, and Robert Woods for helpful discussion and a critical reading of the manuscript.
+This work was supported by a Clinician Scientist Development Award from the Doris Duke Charitable Foundation and R01 AI118886, both to ASL.
+JTM was supported by the Michigan Predoctoral Training Program in Genetics (T32GM007544).
 

### 2016-03-20 doc/draft.Rmd
discussion-crap

@@
 
-# verfy # of Tp in 0.5% 10^5 sample - done
-
-# verity that the possible variants in each case accounts for the wt and carry over variants - done
-
 #Mapq exploration - concordant vs nonconcordant pairs
@@
 # add mfe raw data to repo. currently pulls data from desktop directory.
+
+# add WSN33 and PR8 consensus sequences to repo
+
+# plot of read start postion vs. genome
+# do we not find the same variants
 ```
@@
 Furthermore, some library preparations result in variable sequencing coverage both within and between samples.
-These sources of ascertainment bias are often  overlooked in NGS based studies of viral diversity.
+These sources of ascertainment bias can be easily  overlooked in NGS based studies of viral diversity.
 
@@
 Each sequencing platform has inherent error profiles (cite), and sample collection, target amplification, and library preparation are additional processes whereby errors can be introduced and propagated.
-The next generation sequencing of viral populations, in particular those of RNA viruses, is error prone, and these errors must be accounted for to ensure accurate secondary analysis.
+As we will show, the next generation sequencing of viral populations, in particular those of RNA viruses, is error prone, and these errors must be accounted for to ensure accurate secondary analysis.
 
@@
 
-"Primer Id" based methods, on the other hand, require less input and target sequencing to the viral genome.
+"Primer Id" based methods require less input and target sequencing to the viral genome.
 These methods use bar-coded primers to construct consensus sequences for each cDNA template and can control for PCR and sequencing errors  [@Jabara:2011eo].
@@
 In this method RT-PCR products are sheared and tagged with bar-coded random primers in a klenow reaction, prior to library preparation.
-SISPA controls for any errors that may arise during the library amplification and controls for PCR biases.
-This method has been used in conjunction with statistical algorithms to control for accuracy in studies of intrahost influenza diversity (cite).
+SISPA controls for any errors that may arise during the library amplification such as PCR biases.
+This method has been used in conjunction with statistical algorithms to control for accuracy in studies of intrahost influenza diversity (Ghedin group citations).
 However, the bar-coding reaction used in SISPA is biased and results in uneven coverage and sensitivity across the genome  [@Rosseel:2013dn].
@@
 True variants are identified as those with frequencies exceeding the expected error rate according to some predetermined statistical test.
-Despite being employed in many NGS based studies of viral diversity (cite?), few of these algorithms have been benchmarked using defined viral populations, and to our knowledge none have been tested under conditions that mimic those found in patient-derived samples.
+Despite being employed in many NGS based studies of viral diversity (e.g.
+), few of these algorithms have been benchmarked using defined viral populations, and to our knowledge none have been tested under conditions that mimic those found in patient-derived samples.
 The accuracy of such algorithms in the context of NGS studies of patient-derived viral populations is not known.
@@
 Jude's Children's Research Hospital) [@Hoffmann:2002ih].
-A biological clone of influenza A/Puerto Rico/8/1934 was obtained from ATCC (VR-1469) and the genomic segments were cloned into the pHW2000 reverse genetic system [@Pauly:2015eq].
+A biological clone of influenza A/Puerto Rico/8/1934(H1N1) was obtained from ATCC (VR-1469) and the genomic segments were cloned into the pHW2000 reverse genetic system [@Pauly:2015eq].
 
@@
 Molar equivalents of each PCR product were pooled to generate reconstituted cDNA genomes of WSN33 and PR8.
-The WSN33 cDNA pool was then serially diluted into the PR8 cDNA pool to yield mixtures in which WSN33 made up 5, 2.5, 1.25,and 0.63% of the population.
+The WSN33 cDNA pool was then serially diluted into the PR8 cDNA pool to yield mixtures in which WSN33 made up 5, 2.5, 1.25, 0.63, and 0.16\% of the population.
 Seven hundred and fifty nanograms of purified PCR products were sheared to an average size of 300-400 bp using a Covaris S220 focused ultrasonicator with the following settings : Intensity: 4, Duty cycle: 10%, Burst/second: 200, Duration: 80 seconds.
 Sequencing libraries were prepared from these fragmented products using the NEBNext Ultra DNA library prep kit (NEB E7370L), Agencourt AMPure XP beads (Beckman Coulter A63881), and NEBNext multiplex oligonucleotides for Illumina (NEB E7600S).
-Pooled libraries were sequenced on an Illumina Miseq machine using 2X250 paired end reads.
+Pooled libraries were sequenced on an Illumina Miseq machine using 2 x 250 paired end reads.
 A clonal plasmid control library was prepared from 8 plasmids containing PR8 genomic segments.
 These plasmids were mixed equimolarly and cDNA was generated using a multiplex one step RT-PCR with the primers Uni12/Inf1 (5'-GGGGGGAGCAAAAGCAGG-3'), Uni12/Inf3 (5'-GGGGGAGCGAAAGCAGG-3'), and Uni13/Inf1 (5'-5CGGGTTATTAGTAGAAACAAGG-3') as in [@Anonymous:A25Ka2tY].
-This library was then prepared in identical fashion to the experimental populations and was sequenced in the same Miseq lane.
+This library was prepared in identical fashion to the experimental populations and was sequenced in the same Miseq lane.
 
@@
 
-We quantified the genome copy number of each of the 20 WSN33 point mutants and the WSN33 WT using a SuperScript III Platinum One-Step qRT-PCR (Invitrogen 11732) and univeral influenza A/B primer and probe sets [@Anonymous:tc].
+We passaged the 20 WSN33 point mutants and the WSN33 WT once in MDCK cells and varified the identity of the mutants by sequencing on a Illumina Miseq as above.
+We quantified the genome copy number of each supernant using a SuperScript III Platinum One-Step qRT-PCR (Invitrogen 11732) and univeral influenza A/B primer and probe sets [@Anonymous:tc].
 Infected supernatants were mixed at equal concentrations and diluted to generate a population with each mutant present at 5% frequency and a total concentration of 10^5^ genomes per microliter.
@@
 We prepared libraries as before and used Quanti PicoGreen dsDNA quantification (ThermoFisher Scientific P7589) to quantify the concentration of each indexed library.
-We pooled equal quantifies (by nanogram) of each indexed library, and removed adaptor dimers using gel isolation with the GeneJET Gel Extraction Kit (ThermoFisher Scientific K0691) prior to sequencing using 2 x 125 paired end reads on an Illumina Hiseq 2500.
+We pooled equal quantities (by nanogram) of each indexed library, and removed adaptor dimers using gel isolation with the GeneJET Gel Extraction Kit (ThermoFisher Scientific K0691) prior to sequencing an on Illumina Hiseq 2500 with 2 x 125 paired end reads.
 A clonal control library was processed in an identical fashion starting from an equimolar mix of 8 plasmids containing the WSN33 genomic segments.
@@
 We connected all of these steps into an analysis pipeline using bpipe [@em:2012wd] which is available for download at https://github.com/lauringlab/variant\_pipeline.
-Additionally, this work is completely reproducible.
+To save memory during SNV processing only variants with p<0.9 were included in our ROC curve analysis, as the vast majority of true negatives are trivial to identify and have a p=1.
+For ease of viewing, and to account for this analysis artifact, we extended the ROC curves horizontally from the last observed change in sensitivity.
 All raw fastq files were submitted to the Sequence Read Archive (SRA) under accession number ___________, and all commands required generate the figures in the manuscript are available for anonymous download at https://github.com/lauringlab/Benchmarking\_paper .
@@
 Under our experimental conditions and with a slightly more stringent p value (p=0.01 vs.
-p=0.05) we found a moderate reduction in sensitivity  (0.830 for variants at 0.63\% and 0.14 for variants at 0.16\%).
+p=0.05) we found a moderate reduction in sensitivity  (`r miseq.sense.63` for variants at 0.63\% and `r miseq.sense.16` for variants at 0.16\%).
 The specificity was above 0.998 in all dilutions.
@@
 As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome.
-The false positives outnumber the true positives by 100-fold in this population with realistic diversity and input.
+The false positives outnumber the true positives by 100-fold in these populations with realistic diversity and input.
 We were able to increased our specificity  by applying a more stringent p-value cutoff.
@@
 Under these conditions we found 10 or fewer false positives in 4 of the 5 samples.
-Given this success we applied a number of strategies to further increase our accuracy, including but not limited to: Benjamin Hochman p-value correction,  more stringent p values (<0.01) or frequency cut-offs (>0.2%), retaining PCR duplicate reads, trimming the ends of the influenza genome, and employing various distributions to estimate the error rate in the control sample.
+Given this success we applied a number of strategies to further increase our accuracy, including but not limited to: Benjamin Hochman p-value correction,  more stringent p values (<0.01) or frequency cut-offs (>0.2%), retaining PCR duplicate reads, trimming the ends of the influenza genome, and employing multiple distributions to estimate the error rate in the control sample.
 The impact of various filtering criteria on our data can be visualized in an interactive shiny application available for download at  \emph{https://github.com/lauringlab/benchmarking_shiny.git}.
@@
 We found that the accuracy of the SNV calling method has a profound effect on measurements of diversity.
-It is clear from the richness measurements in `r Table` that the number of false SNV largely determines the accuracy of the other diversity metrics.
+It is clear from the richness measurements in `r Table` that the number of false SNV (i.e.
+the specificity) largely determines the accuracy of the down stream anlaysis.
 Thus, our adapted DeepSNV protocol, which was able to distinguish between true and false SNV with the highest accuracy, gave the most accurate measures of diversity, followed by Lofreq and the default version of DeepSNV.
+
+##Discussion
+
+We have shown that benchmarking is vital to accurate NGS based studies of viral diversity.
+In our opinion DeepSNV represents the most rigorously benchmarked SNV indentifier.
+We show that even under the best conditions the accuracy of DeepSNV drops when applied to whole genome sequencing of an RNA virus.
+The accuracy of SNV identification under more realistic conditions was even more sobering.
+We found input nucleic acid concentrations, which can vary greatly in patient-derived samples, had large effects on both sensitivity and specificity, and thatat moderate levels of nucleaic acid input we could greatly improve accuracy by filtering putative SNV based on quality metrics and read positions.
+We further improved our accuracy at low input levels by controlling for RT-PCR errors by processing samples in duplicates.
+While our quality cut offs may not be needed by all variant callers, our work with Lofreq indicates that experimental design is important in acurate SNV detection and few variant callers have been benchmarked under patient derived conditions .
+Finally, we show that these inaccuracies in SNV calling drastically impact down stream analysis, and that accurate conclusions require quanitifying this error.
+
+ To our knowledge, DeepSNV is the only variant calling method that has been benchmarked on real sequencing reads in which all true positive variants and their frequencies were known *a priori* and independent of NGS.
+Given this robust benchmarking we thought it ideal for identifying rare variants in populations of influenza virus.
+However, we found decreased sensitivity and specificity when applied to cell culture derived mixtures of WSN33 and PR8.
+The most stiking drop in sensitivity came for variants found at 0.16%.
+Under these conditions we found true variants at a rate of `r miseq.sense.16` compared to the reported 0.860 for variants at 0.1%.
+This decrease is most likely due to the fact that we are looking for variants across `r miseq_kb`bp of the influenza genome where previous benchmarking focused on just 1,500 bp of the HIV pol gene.
+Therefore, our sequencing efforts are diluted and we expect fewer reads per variant which limits sensitivity.
+Additionally, in our analysis we aligned all reads to a PR8 reference genome which biases our detection of WSN33 variants and this would play a larger role for rare variants that have fewer reads to begin with.
+Dispite the use of bonferronni corrected p values, the relatively modest decrease in specificity is mostlikely due to multiple testing errors caused by applying the algorithm across `r miseq_kb`bp (cite).
+
+Because, our WSN3-PR8 populations were constructed from reconstituded cDNA genomes they were devoid of any biases that might accompany PCR amplification.
+Gerstung *et.
+al.* have shown that PCR amplification biases lead to overdispered data.
+We found that the dispersion parameter estimated by DeepSNV was relatively low (<5) indicating that the variability in the data was relatively close to what would be expected in a binomial distribution.
+On the other hand, the dispersions estimated for our 20 mutant populations were much higher (>100) indicating more variability in the data.
+This variability is mostlikely due to the RT-PCR based biases as has been observed in the past, and likely accounts for the drastic drop in specificity that accompanied our 20 mutant data set.
+We were able to increase specificity through a number of empirically determined quality cut offs.
+We have sequenced over 300 influenza libraries in on more that 5 Hiseq lanes and the mean MapQ and mean Phred distributions for subconsensus variants is consistant.
+We trust then that our cut offs are valid when applied to influenza sequencing.
+However, these distributions are likely influenced by the articheture of the influenza geneome which contains conserved ends of every fragment.
+We expect poorly mapping reads to map to these conserved areas and so by removing this reads we are indirectly removing variants that are being called in the poorly covered regions of the genome.
+
+One of the take home points of this manuscript is that accuracy is a function of the conditions under which a test is applied.
+Therefore to determine the accuracy of DeepSNV in indentifying SNV from patient derived influenza populations we
 

### 2016-03-20 doc/draft.Rmd
discussion outline

@@
 # do we not find the same variants
+# L1-norm of 4_5 and 5_5
 ```
@@
 The accuracy of SNV identification under more realistic conditions was even more sobering.
-We found input nucleic acid concentrations, which can vary greatly in patient-derived samples, had large effects on both sensitivity and specificity, and thatat moderate levels of nucleaic acid input we could greatly improve accuracy by filtering putative SNV based on quality metrics and read positions.
+We found input nucleic acid concentrations, which can vary greatly in patient-derived samples, had large effects on both sensitivity and specificity, and that at moderate levels of nucleaic acid input we could greatly improve accuracy by filtering putative SNV based on quality metrics and read positions.
 We further improved our accuracy at low input levels by controlling for RT-PCR errors by processing samples in duplicates.
 While our quality cut offs may not be needed by all variant callers, our work with Lofreq indicates that experimental design is important in acurate SNV detection and few variant callers have been benchmarked under patient derived conditions .
-Finally, we show that these inaccuracies in SNV calling drastically impact down stream analysis, and that accurate conclusions require quanitifying this error.
+Finally, we show that these inaccuracies in SNV calling drastically impact down stream analysis, and that accurate conclusions require quanitifying and limitting this error.
 
@@
 The most stiking drop in sensitivity came for variants found at 0.16%.
-Under these conditions we found true variants at a rate of `r miseq.sense.16` compared to the reported 0.860 for variants at 0.1%.
+Under these conditions we found true variants at a rate of `r miseq.sense.16` compared to the reported sensitivity of  0.860 for variants at 0.1%.
 This decrease is most likely due to the fact that we are looking for variants across `r miseq_kb`bp of the influenza genome where previous benchmarking focused on just 1,500 bp of the HIV pol gene.
 Therefore, our sequencing efforts are diluted and we expect fewer reads per variant which limits sensitivity.
-Additionally, in our analysis we aligned all reads to a PR8 reference genome which biases our detection of WSN33 variants and this would play a larger role for rare variants that have fewer reads to begin with.
-Dispite the use of bonferronni corrected p values, the relatively modest decrease in specificity is mostlikely due to multiple testing errors caused by applying the algorithm across `r miseq_kb`bp (cite).
+Additionally, in our analysis we aligned all reads to a PR8 reference genome which biases our detection of WSN33 variants.
+Any bias against WSN33 would decrease our detection of rare variants, which already have few reads.
+A relatively modest decrease in specificity accompanied the reduced sensitivity.Dispite the use of bonferronni corrected p values, the drop in specificity is mostlikely due to multiple testing errors caused by applying the algorithm across `r miseq_kb`bp (cite).
 
-Because, our WSN3-PR8 populations were constructed from reconstituded cDNA genomes they were devoid of any biases that might accompany PCR amplification.
-Gerstung *et.
-al.* have shown that PCR amplification biases lead to overdispered data.
-We found that the dispersion parameter estimated by DeepSNV was relatively low (<5) indicating that the variability in the data was relatively close to what would be expected in a binomial distribution.
-On the other hand, the dispersions estimated for our 20 mutant populations were much higher (>100) indicating more variability in the data.
-This variability is mostlikely due to the RT-PCR based biases as has been observed in the past, and likely accounts for the drastic drop in specificity that accompanied our 20 mutant data set.
-We were able to increase specificity through a number of empirically determined quality cut offs.
-We have sequenced over 300 influenza libraries in on more that 5 Hiseq lanes and the mean MapQ and mean Phred distributions for subconsensus variants is consistant.
-We trust then that our cut offs are valid when applied to influenza sequencing.
-However, these distributions are likely influenced by the articheture of the influenza geneome which contains conserved ends of every fragment.
-We expect poorly mapping reads to map to these conserved areas and so by removing this reads we are indirectly removing variants that are being called in the poorly covered regions of the genome.
+ Despite this reduced accuracy we applied DeepSNV to experimental populations that more closely mimic the diversity and nucleaic acid input levels found in patient derived samples.
+At a modest input level of 10^5^ genomes per microliter we found roughly a four-fold increase in false positives.
+This decrease is likely dues to how the populations were constructed and stresses the importance of benchmarking SNV identifying pipelines under conditions to which they will be applied.
+III. Experimental populations
+  A. The large increase in false positives compared to PR8-WSN33
+    i) Our WSN3-PR8 populations were constructed from reconstituded cDNA genomes and were devoid of any biases that might accompany PCR amplification. 
+    ii) Not the case with the experimental samples
+    iii) Gerstung *et.
+al.* have shown that PCR amplification biases lead to increased variability in the subconsensus nucleotide counts used to call SNV.
+This variability in erreoneous base calls in test samples compared to the control increases the likelihood \ that the frequency of a false positive SNV will exceed the error rate estimated by the plasmid control.
+    iv) DeepSNV can attempt to control for such variablity inpart by estimating the dispersion of a sample.
+Indeed, our experimental samples were more dispersed than those in cell culture.
+However, the variability estimates are limitted  and this likely accounts for the decreased specificity.
+  B. Quality cut offs
+    We hypothesis that the RT-PCR biases introduced in our experimental populations reduced the efficacy of teh DeepSNV algorithm. We,therefore, used alternative means of increasing accuracy.
+    i) The average MapQ and Phred scores of putative SNV were bimodal with true SNV lying in the higher of the two distributions. 
+    ii) We plotted the cdf of these distributions and applied cuttoffs at the streepest increase in the true positive distributions.
+    iii) We have sequenced over 300 influenza libraries on over 5 Hiseq lanes and observe the consistant qualty distributions for DeepSNV identified putative SNV.
+    iv) We have also sequenced many polio libraries and see the same bimodal trends; with slightly shift means.
+    v) Mostlikely due to influenza genomic structures.
+    vi) This is different than filtering reads and masking phred, as we are filtering SNV call after they have been identified.
+Treats False positives and True negatives differently, and perhaps inflates coverage estimates.
+    vi) MapQ, Phred, and Read position were not correlated with eachother and so remove different false positive variants
+    vii) We found many false SNV were identified only at the termi of paired end reads.
+We removed these by filtering putatuive SNV calls based on their position on a paired end read; however, there are many ways to account for these errors.
+    viii) Strand bias, trimming
+    ix) In particular these were found almost exclusively in regions of the genome that were enriched for read start sites.
+This enrichment is mostlikely a consequence  of sequence context, the fragmentation process, and our size selection protocol; however, it is unclear why our plasmid control would not contain similar biases.
+The cause may have a biological foundation in DI particals which have been identified to conatian truncated genomic segements and would only be present in infected supernantants and not a plasmid control.
+    x) These ammendments were able to greatly increase the accuracy of our DeepSNV based SNV pipeline.
 
-One of the take home points of this manuscript is that accuracy is a function of the conditions under which a test is applied.
-Therefore to determine the accuracy of DeepSNV in indentifying SNV from patient derived influenza populations we
+  C. Discussion of frequency of variants
+    i) due to variability in frequency measurments cut offs limited sensitivity we did not apply frequency cut offs as is commonly done
+    ii) Before our cut off no variants were found below XXX%. after our cut offs non were found below X% showing we indirectly applied a frequency threshold.
+    iii) discussion of coverage
+    iv) Lofreq and frequency cut offs.
+    v) reiterate importance of decreases sensitivity when comparing rare SNV between samples.
+
+  D. Lower inputs
+  i) RT-PCR errors resulted in decreased specificity at lower levels of nucleic acid concentration input. 
+  ii) As these errors result from base changes, they appear as TP to algorithms and have high quality scores.
+  iii) Where contorled for by processing samples in duplicate.
+  iv) could duplicates have been pooled and sequenced as one sample?
+
+IV. Comparision to Lofreq  and Consequences of errors
+  i) our benchmarking shows that DeepSNVs means of distinguishing true from false variants is partitially breaks down under conditions that mimic patient derived samples.
+  ii) This can be overcome with the addition of specific quality filters.
+  iii) We would not have known about these inaccuracies had we not run our benchmarking experiment.
+  iv) Do other variant callers have similar inaccuracies.
+  v) Yes Lofreq, although more accurate than our unmodified DeepSNV pipeline, had decreased specificity than had previously been reported.
+  vi) The initial Lofreq sensitivity and specificity metrics were based largely on a *in silico* data sets.
+  vii) Lofreq did not display this high accuracy in under our experimental conditions.
+  viii) While we did not robustly attempt to improve the Lofreq accuracy we did find that frequency threshold could improve its specificity.
+  ix) Experimental conditions matter, regardless of variant calling method and so accurate study requires controls to be run.
+
+V. Consequences of errors
+  A. 
+  i) Accuracy in SNV indetification has consequences on down stream analysis.
+  ii) Diversity measurements and populations comparisions are affected.
+  iii) Differences in titer must be accounted for in population comparisions.
+
+VI. Take home message
+i) Benchmarking controls are needed to ensure accurate NGS based studies of viral populations.
+i) As influenza virologist we benefit from an easily accessible reverse genetics system that made for a straight forward way to benchmark our variant calling.
+ii) Hopefully we have convince the reader that such controls are necessisary.
+iii) While we do not recomemd a varaints calling pipeline we have provided our raw data and computational approach.
+iv) It is provides the most robust data set to data and avaialable for anyone to test their methods against.
 

### 2016-03-21 doc/draft.Rmd
rough draft

@@
 source("./figures.R")
-#this is so the cache resets
+
+#This is to reset the cache
 # 
@@
 It is well known that sequence quality drops near the end of a read (cite?), and we found that our false positives clustered at the termini of our paired end reads (`r Qual`B).
-The average Phred score of these false positives was `r false_p_phred`, further demonstrating that filtering on quality score alone is insufficient.
+The average Phred score of these false positives was `r false_p_Phred`, further demonstrating that filtering on quality score alone is insufficient.
 In contrast, true positives were uniformly distributed across the reads resulting in an average read position near the middle of the read.
@@
 Based on these data we applied a number of empirically determined cut-offs and markedly improved our specificity to >0.999 without sacrificing sensitivity (`r Qual` C and D).
-For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean phred score of 35, and an average read position within the middle 50\% of the read.
+For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean Phred score of 35, and an average read position within the middle 50\% of the read.
 Under these conditions we found 10 or fewer false positives in 4 of the 5 samples.
@@
 This decrease is likely dues to how the populations were constructed and stresses the importance of benchmarking SNV identifying pipelines under conditions to which they will be applied.
-III. Experimental populations
-  A. The large increase in false positives compared to PR8-WSN33
-    i) Our WSN3-PR8 populations were constructed from reconstituded cDNA genomes and were devoid of any biases that might accompany PCR amplification. 
-    ii) Not the case with the experimental samples
-    iii) Gerstung *et.
-al.* have shown that PCR amplification biases lead to increased variability in the subconsensus nucleotide counts used to call SNV.
-This variability in erreoneous base calls in test samples compared to the control increases the likelihood \ that the frequency of a false positive SNV will exceed the error rate estimated by the plasmid control.
-    iv) DeepSNV can attempt to control for such variablity inpart by estimating the dispersion of a sample.
-Indeed, our experimental samples were more dispersed than those in cell culture.
+Our WSN3-PR8 populations were constructed from reconstituded cDNA genomes and were devoid of any biases that might accompany PCR amplification.
+However, this was not the case in our 20 mutant populations.
+To more accurately mimic the method we use to process patient samples we mixed all 20 point mutants and WSN33 WT supernants and amplified influenza genomes from isolate RNA using a multiplex RT-PCR reaction.
+This method undoubitbly introduced unavaidable PCR biases into our populations.
+Gerstung *et.
+al.* have shown that PCR amplification biases lead to increased variability in the subconsensus nucleotide counts used to call SNV (cite DeepSNV documentation).
+Variability in the test samples compared to the control increases the likelihood that an erroneous SNV will exceed the error rate estimated in the plasmid control and thus the false positive rate.
+DeepSNV attempts to control for such variablity inpart by estimating the dispersion of a sample, which incorporates greater variance into the null model.
+In fact, our 20 mutant libraries were more dispersed than those in cell culture, indicating a need for greater variance.
 However, the variability estimates are limitted and this likely accounts for the decreased specificity.
-  B. Quality cut offs
-    We hypothesis that the RT-PCR biases introduced in our experimental populations reduced the efficacy of teh DeepSNV algorithm. We,therefore, used alternative means of increasing accuracy.
-    i) The average MapQ and Phred scores of putative SNV were bimodal with true SNV lying in the higher of the two distributions. 
-    ii) We plotted the cdf of these distributions and applied cuttoffs at the streepest increase in the true positive distributions.
-    iii) We have sequenced over 300 influenza libraries on over 5 Hiseq lanes and observe the consistant qualty distributions for DeepSNV identified putative SNV.
-    iv) We have also sequenced many polio libraries and see the same bimodal trends; with slightly shift means.
-    v) Mostlikely due to influenza genomic structures.
-    vi) This is different than filtering reads and masking phred, as we are filtering SNV call after they have been identified.
-Treats False positives and True negatives differently, and perhaps inflates coverage estimates.
-    vi) MapQ, Phred, and Read position were not correlated with eachother and so remove different false positive variants
-    vii) We found many false SNV were identified only at the termi of paired end reads.
-We removed these by filtering putatuive SNV calls based on their position on a paired end read; however, there are many ways to account for these errors.
-    viii) Strand bias, trimming
-    ix) In particular these were found almost exclusively in regions of the genome that were enriched for read start sites.
+
+Because RT-PCR biases reduced the efficacy of the DeepSNV algorithm, we search for alternative means of increasing accuracy.
+Out-of-the box DeepSNV is rather agnostic towards the mapping quality and base quality of a given variant, and only masks bases calls with Phred scores below a user designated threshold.
+We were able to use the qualtiy metrics to filter putative variants and greatly increase the accuracy of SNV identifivation.
+However, commonly used thresholds (MapQ>20, Phred>30) did little to increase the accuracy of our method.
+Instead,we realied on our own empirically determined thresholds.
+We found the average MapQ and Phred scores of putative SNV were bimodal with true SNV gound in the higher of the two distributions.
+In both the MapQ and Phred distributions there was a clear distinction between the two populations, and we applied qualtiy thresholds to elimnate any putative SNV found in the lower of the two distributions.
+We found no correlation betwee MapQ and Phred (p>x), suggesting these metrics are independent and thus both thresholds eliminate distinct false positives.
+We have sequenced over 300 influenza libraries on over 5 Hiseq lanes and observe the consistant qualty distributions for DeepSNV identified putative SNV.
+We have sequenced many polio virus libraries and see the same bimodal trends.
+We do see shifts in the means of our MapQ distibutions when comparing polio and influenza libraries.
+This is most likely due to differences in the genomic structures of the viruses.
+The tremial regions of all 8 influenza genomic segments are conserved and this conservation decreases the MapQ of reads that map to these areas.
+Therefore, while our MapQ and Phred thresholds are robust in our system, they may need to be adjusted for use in other systems.
+Even in the face of stringent MapQ and Phred cut-offs, we found many high qualtiy false SNV were identified only at the termi of paired end reads.
+We removed these by filtering putatuive SNV calls based on their position on a paired end read.
+In particular these were found almost exclusively in regions of the genome that were enriched for read start sites.
 This enrichment is mostlikely a consequence  of sequence context, the fragmentation process, and our size selection protocol; however, it is unclear why our plasmid control would not contain similar biases.
 The cause may have a biological foundation in DI particals which have been identified to conatian truncated genomic segements and would only be present in infected supernantants and not a plasmid control.
-    x) These ammendments were able to greatly increase the accuracy of our DeepSNV based SNV pipeline.
+Despite finding these false positives in across a number of libraries, because our variants were made by overlap PCR mutatgenesis we can be confident they are an artifact of the experimental design and/or sample preparation, and have robustly controlled for such errors.
 
-  C. Discussion of frequency of variants
-    i) due to variability in frequency measurments cut offs limited sensitivity we did not apply frequency cut offs as is commonly done
-    ii) Before our cut off no variants were found below XXX%. after our cut offs non were found below X% showing we indirectly applied a frequency threshold.
-    iii) discussion of coverage
-    iv) Lofreq and frequency cut offs.
-    v) reiterate importance of decreases sensitivity when comparing rare SNV between samples.
+It is common to filter and trim sequencing reads prior to SNV identification.
+However, we have taken a slightly different approach in our anlaysis.
+In the initial SNV identification step we have masked bases with Phred<30 but made no additional restrinctions on the raw data.
+Only after putative SNV (that exceed the frequency expected given a plasmid control) have been identified do we inpose quality restrictions.
+Our approach treats variant nucleotides more stringently than consensus base calls, but this is of little concern for two reasons.
+Firstly, we have identified specificity as the problem in accrate SNV identification, and so stringenlty filtering potential false positives seems appropriate.
+Additional, since there are orders of magnitude more sequenced reads for a concensus base than there are for variant bases, and our thresholds are based on mean quality scores it is unlikely this affects the accruacy of our method.
 
-  D. Lower inputs
-  i) RT-PCR errors resulted in decreased specificity at lower levels of nucleic acid concentration input. 
-  ii) As these errors result from base changes, they appear as TP to algorithms and have high quality scores.
-  iii) Where contorled for by processing samples in duplicate.
+Frequency thresholds are one quality filter that is often applied to SNV after identification.
+Common thresholds range from 0.1% to 1%; however, there appears to a consensus that SNV identification is unreliable below 0.1%.
+We did not apply direct frequency thresholds.
+Due to variability in the frequency measurments of our DeepSNV identified true positives we found that frequency cut offs limited sensitivity and did not improve specificity.
+However, our quality thresholds acted as an indirect frequency control.
+Before applying MapQ, Phred and read positional thresholds, no variants were found below XXX% frequency; whereas, after our thresholds the rarest false positive was found at X% frequency.
+Read depth, or coverage, is another metric that is often used in conjunciton with frequency to insure accurate SNV identification.
+Although we did not apply a direct coverage cut off, DeepSNV has been reported to require a coverage of ten times the reprical of frequency for sufficient power to call SNV.
+This means that a coverage of 1,000 is needed to detect a variant at 1% frequency.
+In our anlaysis the lowest coverage for a true positive was x (x% frequency) while the lowest coverage for a false positive variant was x (x% frequency).
+Read depth is an important variable in NGS based studies of viral diversity and it, with input concentrations determine the power to detect rare variants.
+If there are large biases in read depth or input levels, two similar popualtions may appear distinct based solely on difference in sensitivity to detect rare SNV.
+
+We observed a drastic decrease in sensitivity in our 10^3^ genomes per microliter samples mostlikely due to the lack of power discussed above.
+However, this drop in sensitivity was accompanied by reduced specificity as well.
+At lower nucleic acid concentrations, our SNV pipeline relied more heavily on RT-PCR amplification, which is an error prone process.
+RT-PCR errors were indistiguishable from true positive variants, as they do not arise from a sequencing error or a bioinformatic aritifact.
+We were able to limit these errors which are sporatic and random (cite) by processing low input samples in duplicate, and achieved comparable accuracy to what was seen at 10^5^ genomes per microliter.
+Quantifying and controling for RT-PCR errors in this way will allow us to accurately compare patient-samples that fall in this input range.
   iv) could duplicates have been pooled and sequenced as one sample?
+
+As we have shown, experimental design impacts the accuracy of SNV identifying pipelines and must be considered in NGS based studies of viral diversity.
+Target amplification, genome structure and nucleic acid concentration are unavoidable sources of error that can vary depending on the virus and sample collection method.
+The reported accuracy of SNV indentifying algorithms is not directly applicable to every experimental design, and as we have shown these diviations can have profound effects on down stream analysis.
+In studying influenza, we have benefitied from an easily accessible reverse genetics system that made for a straight forward mutagensis based approach to benchmarking.
+Under our experimental conditions we have been able to greatly improve the accuracy of DeepSNV.
+Even thouhg the accuracy remains is imperfect, we are now equiped with an understanding of the limitations of our method.
+There may be more accurate variant callers than DeepSNV, or algorithms better suited for other systems; however, what we hope to stress is the fact that all methods have their limitations and understanding these limitations is vital for accurate science.
+Next generation sequencing is powerful tool that is just begining to unlock the complex dynamics of intrahost viral populations.
+As this tool becomes more and more popular, we urge our fellow virologist to remember in NGS as in all science, controls are needed to ensure accuracy, and we would do well to use them.
+
+**Acknowledgements**
+
+We thank Pat Schloss, Mike Imperiale, and Robert Woods for helpful discussion and a critical reading of the manuscript.
+This work was supported by a Clinician Scientist Development Award from the Doris Duke Charitable Foundation and R01 AI118886, both to ASL.
+JTM was supported by the Michigan Predoctoral Training Program in Genetics (T32GM007544).
 
@@
 
-V. Consequences of errors
-  A. 
-  i) Accuracy in SNV indetification has consequences on down stream analysis.
-  ii) Diversity measurements and populations comparisions are affected.
-  iii) Differences in titer must be accounted for in population comparisions.
-
-VI. Take home message
-i) Benchmarking controls are needed to ensure accurate NGS based studies of viral populations.
-i) As influenza virologist we benefit from an easily accessible reverse genetics system that made for a straight forward way to benchmark our variant calling.
-ii) Hopefully we have convince the reader that such controls are necessisary.
-iii) While we do not recomemd a varaints calling pipeline we have provided our raw data and computational approach.
-iv) It is provides the most robust data set to data and avaialable for anyone to test their methods against.
-
-**Acknowledgements**
-
-We thank Pat Schloss, Mike Imperiale, and Robert Woods for helpful discussion and a critical reading of the manuscript.
-This work was supported by a Clinician Scientist Development Award from the Doris Duke Charitable Foundation and R01 AI118886, both to ASL.
-JTM was supported by the Michigan Predoctoral Training Program in Genetics (T32GM007544).
-
 #References

### 2016-03-22 doc/draft.Rmd
corrected figs after swivl

--- 2016-03-20 doc/draft.Rmd
@@
 
-We have shown that benchmarking is vital to accurate NGS based studies of viral diversity.
-In our opinion DeepSNV represents the most rigorously benchmarked SNV indentifier.
-We show that even under the best conditions the accuracy of DeepSNV drops when applied to whole genome sequencing of an RNA virus.
-The accuracy of SNV identification under more realistic conditions was even more sobering.
-We found input nucleic acid concentrations, which can vary greatly in patient-derived samples, had large effects on both sensitivity and specificity, and thatat moderate levels of nucleaic acid input we could greatly improve accuracy by filtering putative SNV based on quality metrics and read positions.
+We have shown that robust benchmarking is vital to accurate NGS based studies of viral diversity, and that differences in experimental design and sample preparation can lead to wide variability in the accuracy of SNV indentification.
+We found input nucleic acid concentrations, which can vary greatly in patient-derived samples, had large effects on both sensitivity and specificity.
+At moderate levels of nucleic acid input we could improve accuracy by filtering putative SNV based on quality metrics and read positions.
 We further improved our accuracy at low input levels by controlling for RT-PCR errors by processing samples in duplicates.
-While our quality cut offs may not be needed by all variant callers, our work with Lofreq indicates that experimental design is important in acurate SNV detection and few variant callers have been benchmarked under patient derived conditions .
-Finally, we show that these inaccuracies in SNV calling drastically impact down stream analysis, and that accurate conclusions require quanitifying this error.
+While our quality cut offs may not be needed by all variant callers, our work with Lofreq indicates that experimental design is important in accurate SNV detection.
+This is important given that few variant callers (if any) have been benchmarked under patient-derived conditions.
+Finally, we show that these inaccuracies in SNV calling drastically impact down stream analysis, and that accurate conclusions require quantifying and limiting this error.
 
- To our knowledge, DeepSNV is the only variant calling method that has been benchmarked on real sequencing reads in which all true positive variants and their frequencies were known *a priori* and independent of NGS.
+ To our knowledge, DeepSNV is the only variant calling method that has been benchmarked on real sequencing reads in which all true positive variants and their frequencies were known *a priori* and independent of NGS (cite Gerstung).
 Given this robust benchmarking we thought it ideal for identifying rare variants in populations of influenza virus.
 However, we found decreased sensitivity and specificity when applied to cell culture derived mixtures of WSN33 and PR8.
-The most stiking drop in sensitivity came for variants found at 0.16%.
-Under these conditions we found true variants at a rate of `r miseq.sense.16` compared to the reported 0.860 for variants at 0.1%.
+The most striking drop in sensitivity came for variants found at 0.16%.
+Under these conditions we found true variants at a rate of `r miseq.sense.16` compared the expected sensitivity of  0.860 for variants at 0.1%.
 This decrease is most likely due to the fact that we are looking for variants across `r miseq_kb`bp of the influenza genome where previous benchmarking focused on just 1,500 bp of the HIV pol gene.
-Therefore, our sequencing efforts are diluted and we expect fewer reads per variant which limits sensitivity.
-Additionally, in our analysis we aligned all reads to a PR8 reference genome which biases our detection of WSN33 variants and this would play a larger role for rare variants that have fewer reads to begin with.
-Dispite the use of bonferronni corrected p values, the relatively modest decrease in specificity is mostlikely due to multiple testing errors caused by applying the algorithm across `r miseq_kb`bp (cite).
+Therefore, our sequencing efforts were diluted across the influenza genome resulting in decreased read depth  which limits sensitivity.
+Additionally,  we aligned all reads to a PR8 reference genome which biases our detection of WSN33 variants especially rare variants, which already have few reads.
+We also observed a relatively modest decrease in specificity.
+Despite the use of bonferronni corrected p values, the drop in specificity is most likely due to multiple testing errors caused by applying the algorithm across `r miseq_kb`bp (cite).
 
-Because, our WSN3-PR8 populations were constructed from reconstituded cDNA genomes they were devoid of any biases that might accompany PCR amplification.
+ Encouraged by these results, we applied DeepSNV to experimental populations that more closely mimic the diversity and nucleic acid input levels found in patient derived samples.
+At a modest input concentration of 10^5^ genomes per microliter we found a four times as many false positives.
+It should be noted that our specificity in this case remained above 0.99 in all populations and that seemly robust specificities can result in an unacceptable number of erroreous SNV when whole genomes are analyzed.
+The reduced specificity was likely an artifact of how the populations were constructed and stresses the importance of benchmarking SNV identifying pipelines under conditions to which they will be applied.
+Our WSN3-PR8 populations were constructed from reconstituted cDNA genomes and were devoid of any biases that might accompany PCR amplification.
+However, this was not the case in our 20 mutant populations.
+To more accurately mimic the method we use to process patient samples we mixed all 20 point mutants and WSN33 WT supernatants prior to  amplification in a multiplex RT-PCR.
+This method undoubtedly introduced unavoidable PCR biases into our populations.
 Gerstung *et.
-al.* have shown that PCR amplification biases lead to overdispered data.
-We found that the dispersion parameter estimated by DeepSNV was relatively low (<5) indicating that the variability in the data was relatively close to what would be expected in a binomial distribution.
-On the other hand, the dispersions estimated for our 20 mutant populations were much higher (>100) indicating more variability in the data.
-This variability is mostlikely due to the RT-PCR based biases as has been observed in the past, and likely accounts for the drastic drop in specificity that accompanied our 20 mutant data set.
-We were able to increase specificity through a number of empirically determined quality cut offs.
-We have sequenced over 300 influenza libraries in on more that 5 Hiseq lanes and the mean MapQ and mean Phred distributions for subconsensus variants is consistant.
-We trust then that our cut offs are valid when applied to influenza sequencing.
-However, these distributions are likely influenced by the articheture of the influenza geneome which contains conserved ends of every fragment.
-We expect poorly mapping reads to map to these conserved areas and so by removing this reads we are indirectly removing variants that are being called in the poorly covered regions of the genome.
+al.* have shown that PCR amplification biases lead to increased variability in the subconsensus nucleotide counts (cite DeepSNV documentation).
+Increased bariability in the test samples increases the likelihood that an erroneous SNV will exceed the error rate estimated in the plasmid control.
+DeepSNV attempts to control for such variability in part by estimating the dispersion of a sample, which allows the algorithm to incorporate greater variance into the null model.
+In fact, our 20 mutant libraries were more dispersed than those in cell culture, indicating a need for greater variance.
+However, the dispersion estimates are limited and the variability likely exceeded what can be controled by this approach.
 
-One of the take home points of this manuscript is that accuracy is a function of the conditions under which a test is applied.
-Therefore to determine the accuracy of DeepSNV in indentifying SNV from patient derived influenza populations we
+Because RT-PCR biases reduced the efficacy of the DeepSNV algorithm, we search for alternative means of increasing accuracy.
+Out-of-the box DeepSNV is rather agnostic towards the mapping quality and base quality of a given variant, and only masks bases calls with Phred scores below a user designated threshold.
+We were able to use  quality metrics to filter putative variants and greatly increase the accuracy of SNV identification.
+However, commonly used thresholds (MapQ>20, Phred>30) did little to increase the accuracy of our method.
+Instead, we relied on our own empirically determined thresholds.
+We found the average MapQ and Phred scores of putative SNV were bimodal with true SNV found in the higher of the two distributions.
+In both the MapQ and Phred distributions there was a clear distinction between the two populations, and we applied quality thresholds to eliminate any putative SNV found in the lower of the two distributions.
+We found no correlation between MapQ and Phred (p>x), suggesting these metrics are independent and thus both thresholds eliminate distinct false positives.
+We have sequenced over 300 influenza libraries in more than 5 Hiseq runs and observe consistent quality distributions for DeepSNV identified putative SNV.
+Our group has also deeply sequenced many polio virus libraries and see the same bimodal trends.
+However, we do see shifts in the means of our MapQ distributions when comparing polio and influenza libraries.
+This is most likely due to differences in the genomic structures of the viruses.
+The terminal regions of all 8 influenza genomic segments are conserved and this conservation decreases the MapQ of reads that map to these areas.
+Therefore, while our MapQ and Phred thresholds are robust in our system, they may need to be adjusted for use in other systems.
+
+Even in the face of stringent MapQ and Phred cut-offs, we found many high quality false positive SNV were identified only at the termi of paired end reads.
+We removed these by filtering putative SNV calls based on their average position in a paired end read.
+In particular these false positives were found almost exclusively in regions of the genome that were enriched for read start sites.
+This enrichment is most likely a consequence  of sequence context, the fragmentation process, and our size selection protocol; however, it is unclear why our plasmid control would not contain similar biases.
+The cause may also have a biological foundation as previous work has shown that defective interfering  particles, which commonly arise during cell passage,contain truncated genomic segments.
+Such truncated segments would only be present in infected supernatants and not a plasmid control.
+These  "read position" false positives were reproducibly found across a number of libraries; however, because our true positives were made by overlap PCR mutagenesis we can be confident such false positives are indeed artifacts of the experimental design and/or sample preparation, and  that we have robustly controlled for such errors.
+
+It is common to filter and trim sequencing reads prior to SNV identification.
+However, we have taken a slightly different approach in our analysis.
+In the initial SNV identification step we have masked bases with Phred<30 but made no additional restrictions on the raw data.
+Only after putative SNV (those that exceed the frequency expected given a plasmid control) have been identified do we impose quality restrictions.
+Our approach treats variant nucleotides more stringently than consensus base calls, but this is of little concern for two reasons.
+Firstly, we have identified specificity as a large the problem in accurate SNV identification, and so stringently filtering potential false positives seems appropriate.
+Additional, since there are orders of magnitude more sequenced reads for a consensus base than there are for rare variant bases, and our thresholds are based on mean quality scores, it is unlikely we would remove any concensus bases from analysis.
+
+Frequency thresholds are an additional quality filter that is often applied to SNV after identification.
+Common thresholds range from 0.1% to 1%; however, there appears to a consensus that SNV identification is unreliable below 0.1%.
+We did not apply direct frequency thresholds.
+In our DeepSNV analysis no putative varaints wer found below 0.1% and we found in our case that frequency cut offs limited sensitivity and did not improve specificity.
+Read depth, or coverage, is another metric that is often used in conjunction with frequency to insure accurate SNV identification.
+Although we did not apply a direct coverage cut off, DeepSNV has been reported to require a coverage of ten times the reciprocal of frequency for sufficient power to call SNV.
+This means that a coverage of 1,000 is needed to detect a variant at 1% frequency.
+In our analysis the lowest coverage for a true positive was `r tp_cov$coverage` (`r round(tp_cov$freq.var*100,digits=1)`% frequency) while the lowest coverage for a false positive variant was `r fp_cov$coverage` (`r round(fp_cov$freq.var*100,digits=1)`% frequency).
+Read depth is an important variable in NGS based studies of viral diversity and it, with input concentrations determine the power to detect rare variants.
+If there are large biases in read depth or input levels, two similar populations may appear distinct based solely on ascertainment bias.
+
+Despite high accuracy at 10^5^ genomes per microliter we observed a decrease in sensitivity in our 10^3^ genomes per microliter samples most likely due to the lack of power discussed above.
+However, this drop in sensitivity was accompanied by reduced specificity.
+At lower nucleic acid concentrations, our SNV pipeline relied more heavily on RT-PCR amplification, which is an error prone process.
+RT-PCR errors were indistinguishable from true positive variants, as they do not arise from a sequencing errors or bioinformatic artifacts.
+We were able to limit these errors which are sporadic and random (cite) by processing low input samples in duplicate, and achieved comparable accuracy to what was seen at 10^5^ genomes per microliter.
+Quantifying and controlling for RT-PCR errors in this way will allow us to accurately compare patient-samples that fall in a wider this input.
+
+As we have shown, experimental design impacts the accuracy of SNV identifying pipelines and must be considered in NGS based studies of viral diversity.
+Target amplification, genome structure and nucleic acid concentration are unavoidable sources of error that can vary depending on the virus and sample collection method.
+The reported accuracy of SNV identifying algorithms is not directly applicable to every experimental design, and as we have shown these deviations can have profound effects on down stream analysis.
+In studying influenza, we have benefited from an easily accessible reverse genetics system that made for a straight forward mutagenesis based approach to benchmarking.
+Under our experimental conditions we have been able to greatly improve the accuracy of DeepSNV.
+Even though the accuracy remains is imperfect, we are now equipped with an understanding of the limitations of our method.
+There may be more accurate variant callers than DeepSNV, or algorithms better suited for other systems; however, our work with Lofreq stresses the fact that all methods have their limitations and understanding these limitations is vital for accurate science.
+Next generation sequencing is powerful tool that is just beginning to unlock the complex dynamics of intrahost viral populations.
+As this tool becomes more and more popular, it is vital we remember that in NGS as in all science, controls are needed to ensure accuracy.
 
@@
 
+ iv) could duplicates have been pooled and sequenced as one sample?
+
+IV. Comparision to Lofreq  and Consequences of errors
+  i) our benchmarking shows that DeepSNVs means of distinguishing true from false variants is partitially breaks down under conditions that mimic patient derived samples.
+  ii) This can be overcome with the addition of specific quality filters.
+  iii) We would not have known about these inaccuracies had we not run our benchmarking experiment.
+  iv) Do other variant callers have similar inaccuracies.
+  v) Yes Lofreq, although more accurate than our unmodified DeepSNV pipeline, had decreased specificity than had previously been reported.
+  vi) The initial Lofreq sensitivity and specificity metrics were based largely on a *in silico* data sets.
+  vii) Lofreq did not display this high accuracy in under our experimental conditions.
+  viii) While we did not robustly attempt to improve the Lofreq accuracy we did find that frequency threshold could improve its specificity.
+  ix) Experimental conditions matter, regardless of variant calling method and so accurate study requires controls to be run.
+
 #References

### 2016-03-22 scripts/figures.Rmd
corrected figs after swivl

--- 2016-03-16 scripts/figures.Rmd
@@
 author: "JT McCrone"
-date: "February 11, 2015"
+date: "March 21, 2016"
 output: pdf_document
@@
 require(knitr)
-opts_chunk$set(fig.align = "center", warning = FALSE, echo = FALSE,message=F,cache=F)
+opts_chunk$set(fig.align = "center", warning = FALSE, echo = FALSE,message=F,cache=T)
 
@@
  #plot the points and cut offs
- points.p<-ggplot(distribution,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center",dotsize = 1.5)+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks = element_blank(), axis.text.x = element_blank(),axis.title.y = element_text(vjust=1.2),legend.position=c(0.9,0.65),legend.background = element_rect(fill = alpha("white",0)))+xlab("")
+ points.p<-ggplot(distribution,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center",dotsize = 1.5)+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(),axis.title.y = element_text(vjust=1.2),legend.position=c(0.9,0.65),legend.background = element_rect(fill = alpha("white",0)))+xlab("")
 
@@
 
-# pdf("../results/figures/roc.example.pdf",width=3.4,height=4)
-# plot.ex(0.35,0.95,roc.df,example)
-# dev.off()
-# pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.ex.pdf",width = 12,height = 8)
-# plot.ex(0.35,0.95,roc.df,example)
-# dev.off()
+pdf("../results/figures/roc.example.pdf",width=3.4,height=4)
+plot.ex(0.35,0.95,roc.df,example)
+dev.off()
+pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.ex.pdf",width = 12,height = 8)
+plot.ex(0.35,0.95,roc.df,example)
+dev.off()
 
@@
 Figure 1.
-An example ROC.
-A) Hypothetical variants are stratified by p value.
-True variants are shown as filled circles, while those corresponding to RT, PCR, or sequencing errors are shown as open circles.
+An example ROC curve.
+A) Hypothetical variants are stratified by log of the p value.
+True positives variants are shown as closed circles, while false positive variants are represented as open circles.
 Potential thresholds are indicated as colored lines.
-B) An ROC made from the data shown in A.
-The colored lines indicate the points made by the thresholds in A.
+B) An ROC curve made from the hypothetical data shown in A.
+The colored lines indicate the points on the curve made by the thresholds in A.
 \pagebreak
@@
 
-m.roc.table$Frequency<-c("5.0%","2.5%","1.25%","0.63%","0.016%")
+m.roc.table$Frequency<-c("5.0%","2.5%","1.25%","0.63%","0.16%")
 m.roc.table$Sensitivity<-round(m.roc.table$Sensitivity,digits=dd)
 m.roc.table$Specificity<-round(m.roc.table$Specificity,digits=dd)
+
+miseq.sense.63<-m.roc.table$Sensitivity[m.roc.table$Frequency=="0.63%"]
+miseq.sense.16<-m.roc.table$Sensitivity[m.roc.table$Frequency=="0.16%"]
 
@@
 
-# pdf("../results/figures/miseq1.pdf",width=3.4,height=7)
-# ggdraw() +
-#   draw_plot(miseq.setup, 0.2, .66, .6, 0.33) + #x y width height
-#   draw_plot(miseq.roc.p, 0.05, 0.33, .9, 0.33) +
-#   draw_plot(tbl, 0.05, 0, .9, 0.33) +
-#   draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(0.95, 0.7, 0.35), size = 15)
-# dev.off()
-
+pdf("../results/figures/miseq1.pdf",width=3.4,height=7)
 ggdraw() +
@@
   draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(0.95, 0.7, 0.35), size = 15)
+dev.off()
+
+ggdraw() +
+  draw_plot(miseq.setup, 0.2, .66, .6, 0.33) + #x y width height
+  draw_plot(miseq.roc.p, 0.05, 0.33, .9, 0.33) +
+  draw_plot(tbl, 0.05, 0, .9, 0.33) +
+  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(0.95, 0.7, 0.35), size = 15)
 
@@
 Initial DeepSNV accuracy.
-A) RT-PCR amplified WSN33 genomes were diluted into PR8 at decreasing frequencies and sequenced on an Illumina Miseq.
-B) An ROC measuring the pipeline's ability to identify WSN33 SNVs relative to PR8 at the indicated frequencies.
-C) A summary of the DeepSNV accuracy at a p value threshold of 0.01.
+A) Reconstitued cDNA genomes of influenza strain WSN33 were diluted serially into PR8 prior to sequencing on an Illumina Miseq machine.
+B) An ROC curve measuring the accuracy of DeepSNV in identifying WSN33 variants mixed with PR8 at the indicated frequencies.
+C) A summary the data in B at a p value threshold of 0.01.
 \pagebreak
@@
 
-# pdf("../results/figures/hiseq1.pdf",width=3.4,height=7)
-# ggdraw() +
-#   draw_plot(hiseq.setup, 0.1, .57, 0.8, 0.43) + #x y width height
-#   draw_plot(Hiseq.roc, 0.05, 0.33, .8, 0.33) +
-#   draw_plot(tbl, 0.05, 0, .9, 0.33) +
-#   draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(1, 0.7, 0.35), size = 15)
-# dev.off()
-
+pdf("../results/figures/hiseq1.pdf",width=3.4,height=7)
 ggdraw() +
@@
   draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(1, 0.7, 0.35), size = 15)
+dev.off()
+
+ggdraw() +
+  draw_plot(hiseq.setup, 0.1, .57, 0.8, 0.43) + #x y width height
+  draw_plot(Hiseq.roc, 0.05, 0.33, .8, 0.33) +
+  draw_plot(tbl, 0.05, 0, .9, 0.33) +
+  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(1, 0.7, 0.35), size = 15)
 # 
@@
 Figure 3.
-DeepSNV accuracy patient sample conditions A) 20 known single point mutants were diluted into wild type at decreasing frequencies.
-These were diluted further into viral media to match the genomes concentration found in patent samples (10$^5$-10$^3$ genomes/$\mu$l) and sequenced on an Illumina Hiseq (2 x 152bp reads.
-B) An ROC indicating the ability to accurately identify the 20 known variants in the 10 $^5$ genomes/$\mu$l samples.
-C) A table of the accuracy of our variant calling for the 10 $^5$ genomes/$\mu$l samples at a p value threshold of 0.01.
+Decreased accuracy under patient-derived sample conditions A) 20 known single point mutants were diluted into WT WSN33 at decreasing frequencies.
+These populations were diluted further in basal media to match the genome concentrations found in patint-dervied samples (10$^5$-10$^3$ genomes/$\mu$l).
+B) The accuracy of DeepSNV in idenfying the 20 known true positive variants in the 10 $^5$ genomes/$\mu$l samples.
+C) A summary table of the accuracy of our method for the 10 $^5$ genomes/$\mu$l samples at a p value threshold of 0.01.
 
@@
 
-mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(size=1)+geom_hline(yintercept=20,linetype=2)+geom_vline(xintercept=30,linetype=2)+scale_x_continuous(limits=c(30,40))+scale_color_manual(name="",labels=c("False posivite","True positive"),values=c("#787878ff","#08519cff"))+ylab("Mean mapping quality")+xlab("Mean base qualtiy (Phred) ")+theme(legend.position="top")
+sig.df<-sig.df[order(sig.df$category,decreasing = F),]
+
+mapq_phred.normal<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,y=MapQ,color=category))+geom_point(size=0.8)+geom_hline(yintercept=20,linetype=2)+geom_vline(xintercept=30,linetype=2)+scale_x_continuous(limits=c(30,40))+scale_color_manual(name="",labels=c("False positives","True positives"),values=c("#b1bdc5ff","#08519cff"))+ylab("Mean MapQ")+xlab("Mean Phred ")+theme(legend.position="top")
 
@@
 
-read.pos<-ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')+scale_fill_manual(name="",labels=c("False positive","True positive"),values=c("#787878ff","#08519cff"))+xlab("Average position on read")+ylab("Number of variants")+theme(legend.position="top")
+read.pos<-ggplot(subset(sig.cut.df,category %in% c(T,F) & gc==5),aes(x=Read_pos,fill=category))+geom_histogram(position='dodge')+scale_fill_manual(name="",labels=c("False positive","True positive"),values=c("#b1bdc5ff","#08519cff"))+xlab("Average position on read")+ylab("Number of variants")+theme(legend.position="top")
 # read.pos
@@
 
-# pdf("../results/figures/figure4.pdf",width=3.4,height=9)
-# ggdraw() +
-#   draw_plot(mapq_phred.normal,0.03, 0.77, .85, 0.23) + 
-#   draw_plot(read.pos, 0.03, .53, .85, .25) +
-#   draw_plot(read.roc, 0.03, 0.23, 0.85, .3) +
-#   draw_plot(tbl,   0.025, 0, 0.95, .25) +
-#   draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(1, 0.78, 0.55,0.25), size = 15)
-# dev.off()
-
+pdf("../results/figures/figure4.pdf",width=3.4,height=9)
 ggdraw() +
@@
   draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(1, 0.78, 0.55,0.25), size = 15)
+dev.off()
+
+ggdraw() +
+  draw_plot(mapq_phred.normal,0.03, 0.77, .85, 0.23) + 
+  draw_plot(read.pos, 0.03, .53, .85, .25) +
+  draw_plot(read.roc, 0.03, 0.23, 0.85, .3) +
+  draw_plot(tbl,   0.025, 0, 0.95, .25) +
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(1, 0.78, 0.55,0.25), size = 15)
 ```
@@
 Figure 4.
-Investigating the quality of variants in the the 10 $^5$ genomes/$\mu$l samples.
-A) All called variants from the 10 $^5$ genomes/$\mu$l samples with p<0.01 stratified by the mean mapping quality of the reads containing the variant and the mean Phred scores of the variant bases.
+Accuracy can be improved through quality thresholds.
+A) We stratified all called variants from the 10 $^5$ genomes/$\mu$l samples with p<0.01 stratified by the mean mapping quality of the reads containing the variant and the mean Phred scores of the variant bases.
 Dashed lines indicate  common cutoffs of 20 and 30 for mapping quality and Phred respectively.
-B) A histogram of all called variants from the 10 $^5$ genomes/$\mu$l samples with p<0.01 binned by the average position of the variant relative to the Hiseq reads on which it is found.
-C) An ROC of the accuracy of the variant caller when the following quality cut offs are applied : mean mapping quality of 30, mean Phred of 35, and an average read position between 32 and 94 (the middle 50% of the read).
-D) A table summarizing the ROC in C at a p value threshold of 0.01.
+B) A histogram of average position on a pair end read the variants that passed our mean MapQ threshold of 30 and mean Phred threshold of 35.
+C) The accuracy of our analyis after applying the following quality cut offs : mean MapQ> 30, mean Phred> 35, and an average read position between 32 and 94 (the middle 50% of the read).
+D) A table summarizing accuracy in C at a p value threshold of 0.01.
 
@@
 
-ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab("1-Specificity (adjusted)")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position='none')+geom_text(data=x,aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2)->lofreq.roc.p
+ggplot(subset(lofreq.roc,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position='none')+geom_text(data=x,aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2)->lofreq.roc.p
 # This is so that the text is not cut off
@@
 
-# pdf("../results/figures/lofreq1.pdf",width=3.4,height=5)
-# ggdraw() +
-#   draw_plot(lofreq.roc.p, 0, 0.5, 0.9, 0.45) + 
-#   draw_plot(tbl, 0.025, 0, 0.95, .48) +
-#   draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
-# dev.off()
-
+pdf("../results/figures/lofreq1.pdf",width=3.4,height=5)
 ggdraw() +
@@
   draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
+dev.off()
+
+ggdraw() +
+  draw_plot(lofreq.roc.p, 0, 0.5, 0.9, 0.45) + 
+  draw_plot(tbl, 0.025, 0, 0.95, .48) +
+  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
 ```
@@
 Figure 5.
-The accuracy of Lofreq.
-The variant caller Lofreq was applied to the 10 $^5$ genomes/$\mu$l samples.
-A) An ROC of the accuracy of Lofreq using standard parameters.
+The accuracy of the Lofreq algorithm on our 10^5% genomes/$\mu$l populations.
+A)  The accuracy of Lofreq using standard parameters.
 The specificity of the caller was scaled to account for the same number of tests as the DeepSNV algorithm.
-Lofreq provides quality scores instead of p values.
-These were used to set cutoffs for the ROC.
 B) A table summarizing the accuracy of the ROC A using the standard cut offs applied by the algorithm.
@@
 
-# pdf("../results/figures/freq.pdf",width=3.4,height=3.4)
-# ac_freq
-# dev.off()
+pdf("../results/figures/freq.pdf",width=3.4,height=3.4)
+ac_freq
+dev.off()
 
@@
 
-# pdf("../results/figures/lower.pdf",width=3.4,height=9)
-# ggdraw() +
-#   draw_plot(roc.4,0.03, 0.69, .85, 0.25) + 
-#   draw_plot(tbl.4, 0.025, .47,  0.95, .25) +
-#   draw_plot(roc.3, 0.03, 0.22, 0.85, .25) +
-#   draw_plot(tbl.3,   0.025, 0, 0.95, .25) +
-#   draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(0.97, 0.73, 0.5,0.23), size = 15)
-# dev.off()
-
+pdf("../results/figures/lower.pdf",width=3.4,height=9)
 ggdraw() +
@@
   draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(0.97, 0.73, 0.5,0.23), size = 15)
+dev.off()
+
+ggdraw() +
+  draw_plot(roc.4,0.03, 0.69, .85, 0.25) + 
+  draw_plot(tbl.4, 0.025, .47,  0.95, .25) +
+  draw_plot(roc.3, 0.03, 0.22, 0.85, .25) +
+  draw_plot(tbl.3,   0.025, 0, 0.95, .25) +
+  draw_plot_label(c("A", "B", "C","D"), c(0, 0, 0,0), c(0.97, 0.73, 0.5,0.23), size = 15)
 ```
@@
 Figure 7.
-The accuracy of DeepSNV at lower input levels.
-A) An ROC depicting the accuracy of our method when applied to the 10 $^4$ genomes/$\mu$l samples.
+The accuracy of DeepSNV decreased when applied to samples with less input nucleic acid.
+A) An ROC curve for the 10 $^4$ genomes/$\mu$l samples.
 B) A table summarizing the accuracy at 10 $^4$ genomes/$\mu$l with a  p value threshold of 0.01.
-C) An ROC depicting the accuracy of our method when applied to the 10 $^3$ genomes/$\mu$l samples.
+C) An ROC curve depicting the accuracy of our method when applied to the 10 $^3$ genomes/$\mu$l samples.
 B) A table summarizing the accuracy at 10 $^3$ genomes/$\mu$l with a  p value threshold of 0.01
@@
 
-tbl<-make.table(dups.table,tt2) # Make table
+tbl<-make.table(dups.table,tt1) # Make table
 
@@
 
-# pdf("../results/figures/dups.roc.pdf",width=3.4,height=5)
-# ggdraw() +
-#   draw_plot(dups.roc.p, 0, 0.5, 0.9, 0.45) + 
-#   draw_plot(tbl, 0.025, 0, 0.95, .48) +
-#   draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
-# dev.off()
-
+pdf("../results/figures/dups.roc.pdf",width=3.4,height=5)
 ggdraw() +
@@
   draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
+dev.off()
+
+ggdraw() +
+  draw_plot(dups.roc.p, 0, 0.5, 0.9, 0.45) + 
+  draw_plot(tbl, 0.025, 0, 0.95, .48) +
+  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
 
@@
 Figure 8.
-The accuracy of DeepSNV at lower input levels when run in duplicate.
-A subset of the 10 $^4$ genomes/$\mu$l samples were processed in duplicate to control for RT-PCR errors.
-A) An ROC of  10 $^4$ genomes/$\mu$l samples processed in duplicate.
+The accuracy of DeepSNV at lower input levels is improved by processing samples in duplicate.
+A) An ROC curve of  the 10 $^4$ genomes/$\mu$l samples processed in duplicate.
 A variant was required to be found in both duplicates to be considered.
@@
 
-### Table 1
-
-Richness
-Inverse Simpsons D ( without replacement)
-Shannon Entropy
-For these I'm assuming all varaints are independent
-
-```{r}
+### Table 1.
+
+Table 1.
+Accurate SNV indentification is vital for accurate measure of intrahost diversity.
+We measured the richness and Shannon's Entropy of the designated 10^5^ genomes$\mu$l populations at each stage of our benchmarking procedure and compared the results with what is expected.
+Additionaly we calculated the L1-norm of the populations relative to the expected population.
+
+```{r,results="asis",cache=FALSE}
 mut_pat="(.*)_[A,T,C,G](\\d+)[A,T,C,G]"
@@
   }
-  out<-data.frame(Richness=round(rich(x),digits = 0), Entropy=format(H(x),scientific = T,digits = 3),Manhattan=round(euc(x,subset(perfect,freq.var==exp.freq)),digits=3)) #"1/D"=D(x),
+  out<-data.frame(Richness=round(rich(x),digits = 0), Entropy=format(H(x),scientific = T,digits = 3),'L1-norm'=round(euc(x,subset(perfect,freq.var==exp.freq)),digits=3)) #"1/D"=D(x),
   rownames(out)<-setting
@@
 }
+require(xtable)
 tabulate<-function(input,freq){
-get_data(subset(perfect,freq.var==freq),"Expected")->perf
-get_data(subset(sum.df,gc==input & exp.freq==freq & p.val<0.01 & category %in% c(T,F)),"DeepSNV")->deepSNV
-get_data(subset(lofreq.df,gc==input & exp.freq==freq & category %in% c(T,F)),"Lofreq")->lo
-get_data(subset(cut.read.df.all,gc==input & exp.freq==freq & p.val<0.01 & category %in% c(T,F) ),"DeepSNV Final")->deepSNV.f
-knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f))
-}
-# tabulate.4<-function(input,freq){
-# get_data(subset(perfect,freq.var==freq),"Perfect")->perf
-# get_data(subset(sum.df,gc==input & exp.freq==freq),"DeepSNV")->deepSNV
-# get_data(subset(lofreq.df,gc==input & exp.freq==freq),"Lofreq")->lo
-# get_data(subset(cut.read.sig.df,gc==input & exp.freq==freq),"DeepSNV Final")->deepSNV.f
-# get_data(subset(dups,exp.freq==freq),"DeepSNV duplicates")->deepSNV.d
-# 
-# knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f,deepSNV.d),digits = 2)
-# }
-
-```
-
-$10^5$ 5%
-
-```{r}
-tabulate(5,0.05)
-```
-
-$10^5$ 1%
-
-```{r}
-tabulate(5,0.01)
-```
-
-$10^5$ 0.5%
-
-```{r}
-tabulate(5,0.005)
-```
+x<-data.frame()
+for(i in 1:length(freq)){  
+get_data(subset(perfect,freq.var==freq[i]),"Expected")->perf
+get_data(subset(sum.df,gc==input & exp.freq==freq[i] & p.val<0.01 & category %in% c(T,F)),"DeepSNV")->deepSNV
+get_data(subset(lofreq.df,gc==input & exp.freq==freq[i] & category %in% c(T,F)),"Lofreq")->lo
+get_data(subset(cut.read.df.all,gc==input & exp.freq==freq[i] & p.val<0.01 & category %in% c(T,F) ),"DeepSNV modified")->deepSNV.f
+y<-cbind(perf,deepSNV.f,lo,deepSNV)
+x<-rbind(x,y)
+}
+names<-data.frame(Frequency=c("","5.0%","","","1.0%","","","0.5%",""),names=rep(c("Richness","Entropy","L1-norm"),times=3))
+x<-cbind(names,x)
+#x<-subset(x,select=c(names,Expected,DeepSNV,Lofreq,DeepSNV Final))
+x<-rename(x,c("names"=""))
+xtab<-xtable(x)
+align(xtab) <- rep("c", 7)
+print(xtab,include.rownames = FALSE,hline.after=c(-1,0,3,6))
+}
+
+```
+
+```{r,results="asis",cache=F}
+tabulate(5,c(0.05,0.01,0.005))
+
+#
+
+```
+
+```{r,coverage}
+cov<-mutate(subset(cut.read.df.all,p.val<0.01 & gc>2 & exp.freq>0.001),coverage=cov.tst.fw+cov.tst.bw)
+
+tp<-subset(cov,category==T)
+fp<-subset(cov,category==F)
+tp_cov<-data.frame(coverage=min(tp$coverage),freq.var=tp$freq.var[tp$coverage==min(tp$coverage)])
+fp_cov<-data.frame(coverage=min(fp$coverage),freq.var=fp$freq.var[fp$coverage==min(fp$coverage)])
+```

### 2016-03-22 doc/draft.Rmd
intro,methods,results AL

@@
 lofreq<-"Figure 6"
-Lofreq<-"figure 6"
+Lofreq<-"Figure 6"
 lower<-"figure 7"
@@
 
-miseq_vars<-format(miseq_vars,scientific=F)
-miseq_kb<-format(miseq_kb*1000,scientific = F)
+miseq_vars<-formatC(miseq_vars,format="d",big.mark = ",")
+miseq_kb<-formatC(miseq_kb*1000,format="d",big.mark = ",")
 
-hiseq_vars<-format(hiseq_vars,scientific=F)
-hiseq_kb<-format(hiseq_kb*1000,scientific = F)
+hiseq_vars<-formatC(hiseq_vars,format="d",big.mark = ",")
+hiseq_kb<-formatC(hiseq_kb*1000,format="d",big.mark = ",")
 
@@
 
-**Benchmarking and quality controls are required in sequence based studies of intrahost diversity.**
+**Validation and quality controls in sequence based studies of intrahost diversity**
 
@@
 \n
-* Corresponding author
+*Corresponding author
 
@@
 
-The advent of next generation sequencing (NGS) is revolutionizing the study of viral evolution and molecular epidemiology through high-throughput measures of intrahost diversity.
-Until recently investigations of viral diversity were limited by labor intensive sequencing approaches; however, it is now feasible to sequence patient-derived samples at sufficient read depth to detect rare single nucleotide variants (SNV), and the high-throughput, cost-effective nature of NGS has lead to an explosion of studies that quantify viral diversity within and between hosts (e.g.[@Andersen:2015jj], [@Grubaugh:2015by], [@Rogers:2015ij],[@Poon:2016gf]).
-Given the increasing contribution of NGS to the field of virology it is important to reflect on the accuracy of these methods and the controls that are needed to ensure robust conclusions.
+The advent of next generation sequencing (NGS) has enabled detailed characterization of intrahost diversity and has revolutionized studies of  viral evolution and molecular epidemiology.
+Until recently, investigations of viral diversity were limited by labor intensive sequencing approaches.
+However, it is now feasible to sequence patient-derived samples at sufficient read depth to detect rare single nucleotide variants (SNV), and the cost-effectiveness of NGS has led to an explosion of studies that quantify viral diversity within and between hosts (e.g.[@Andersen:2015jj], [@Grubaugh:2015by], [@Rogers:2015ij],[@Poon:2016gf]).
+Given the increasing contribution of NGS to the field of virology, it is important to reflect on the accuracy of these methods and the controls that are needed to ensure robust conclusions.
 
-Because next generation sequencing of viral populations often produces tens of thousands of reads per genomic position, the sensitivity of these approaches has been assumed to be quite robust (cite).
-However, patient derived samples can differ in input titer by orders of magnitude (e.g.
-[@Lau:2013ima],[@Teunis:2015iw],[@Takeyama:2015jg]), and this variability can have drastic impacts on the sensitivity of detecting rare events.
+Because next generation sequencing of viral populations often produces tens of thousands of reads per genomic position, the sensitivity of these approaches has been assumed to be quite high (cite).
+However, patient-derived samples can differ in input titer by orders of magnitude (e.g.
+[@Lau:2013ima],[@Teunis:2015iw],[@Takeyama:2015jg]), and this variability can have drastic impacts on the sensitivity of detection for rare events.
 Furthermore, some library preparations result in variable sequencing coverage both within and between samples.
@@
 The specificity of NGS is another factor to consider when designing an experiment using patient-derived samples.
-Each sequencing platform has inherent error profiles (cite), and sample collection, target amplification, and library preparation are additional processes whereby errors can be introduced and propagated.
-As we will show, the next generation sequencing of viral populations, in particular those of RNA viruses, is error prone, and these errors must be accounted for to ensure accurate secondary analysis.
+Each sequencing platform has inherent error profiles (cite), and sample collection, target amplification, and library preparation are additional processes whereby errors can be introduced and propagated.The next generation sequencing of RNA virus populations is particularly error prone, and these errors must be accounted for to ensure accurate secondary analysis.
 
-Many sample preparation pipelines have been developed to control for errors in NGS based studies of RNA virus populations; however, each approach has its own caveats that ultimately limit its application.
-For example, Cirseq is an ingenious approach in which  template RNA is sheared and circularized prior to reverse transcription  [@Acevedo:2014ej].
-Rolling circle cDNA synthesis produces tandem reads, generating a consensus sequence for each RNA fragment.
-While this method is likely to be highly sensitive for rare variant detection and can control for reverse transcription, PCR, and sequencing errors, the requirement for a large and relatively pure population of viral RNA limits its applicability to patient derived samples [@Acevedo:2014df].
-
-"Primer Id" based methods require less input and target sequencing to the viral genome.
-These methods use bar-coded primers to construct consensus sequences for each cDNA template and can control for PCR and sequencing errors  [@Jabara:2011eo].
-Primer Id based methods require each bar code be physically attached to a PCR product and are most easily applied to a small targeted regions of the genome.
+Many sample preparation protocols have been developed to control for errors in NGS-based studies of RNA virus populations.
+However, each approach has its own caveats that ultimately limit its application.
+Cirseq is an ingenious approach in which  template RNA is sheared and circularized prior to reverse transcription  [@Acevedo:2014ej].
+Subserquent rolling circle cDNA synthesis produces tandem reads, generating a consensus sequence for each RNA fragment.
+While this method is likely to be highly sensitive for rare variant detection and can control for reverse transcription, PCR, and sequencing errors, the requirement for a large and relatively pure population of viral RNA limits its applicability to patient-derived samples [@Acevedo:2014df].
+"Primer ID" based methods require less input and target sequencing to the viral genome.
+This approach relies upon barcoded primers to construct consensus sequences for each cDNA template and can control for PCR and sequencing errors  [@Jabara:2011eo].
+Because Primer ID methods require taht each bar code be physically attached to a PCR product, they  are most easily applied to a small targeted regions of the genome.
 As such, they have limited application in whole genome sequencing.
@@
 This method has been used in conjunction with statistical algorithms to control for accuracy in studies of intrahost influenza diversity (Ghedin group citations).
-However, the bar-coding reaction used in SISPA is biased and results in uneven coverage and sensitivity across the genome  [@Rosseel:2013dn].
+However, the bar-coding reaction used in SISPA is biased and can result in uneven coverage and sensitivity across the genome  [@Rosseel:2013dn].
 
 Given the above limitations, many statistical approaches have been developed to distinguish true variants from sequencing errors [@Gerstung:1js; @Isakov:2015gq; @Wilm:2012br; @Macalalad:2012fx; @Koboldt:2009dk].
-In general, these algorithms calculate base specific error rates from various metrics including but not limited to: mapping quality (MapQ), base quality (Phred), strand bias, and sequence context.
+In general, these algorithms calculate base-specific error rates from various metrics including but not limited to: mapping quality (MapQ), base quality (Phred), strand bias, and sequence context.
 True variants are identified as those with frequencies exceeding the expected error rate according to some predetermined statistical test.
@@
 ), few of these algorithms have been benchmarked using defined viral populations, and to our knowledge none have been tested under conditions that mimic those found in patient-derived samples.
-The accuracy of such algorithms in the context of NGS studies of patient-derived viral populations is not known.
+The accuracy of such algorithms in the context of NGS studies of patient-derived viral populations is largely unknown.
 
-Here, we use defined populations of Influenza A and variable input titers to indirectly determine the accuracy of rare variant detection in patient-derived samples.
-We highlight the challenges that accompany NGS based studies of viral diversity and include a few means for improved the accuracy.
+Here, we use genetically-defined populations of Influenza A virus and variable input titers to indirectly determine the accuracy of rare variant detection in patient-derived samples.
+We highlight the challenges that accompany NGS-based studies of viral diversity and include a few means for improved the accuracy.
 This work exemplifies the controls that should be run prior to any NGS based study of viral populations and provides a comprehensive data set for benchmarking other pipelines.
 
-## Methods
+# Methods
 
@@
 Influenza A/WSN/33(H1N1) virus was rescued from transfected cells using the 8 plasmid reverse genetic system containing the genomic segments (pHW181-188), a kind gift from Robert Webster (St.
-Jude's Children's Research Hospital) [@Hoffmann:2002ih].
+Jude's Children's Research Hospital) [@Hoffmann:2002ih,@Pauly:2015eq].
 A biological clone of influenza A/Puerto Rico/8/1934(H1N1) was obtained from ATCC (VR-1469) and the genomic segments were cloned into the pHW2000 reverse genetic system [@Pauly:2015eq].
@@
 
-#### PR8 WSN33 population
+#### PR8-WSN33 population
 
@@
 We then verified the sequences of the these viruses by Sanger sequencing.
-Two microliters of RNA template was used  to generate cDNA in 25 $\mu\text{l}$ segment specific one step RT-PCR with $0.2\mu\text{M}$ of the following primers :   PB2-Forward-JT	(5'-GCAGGTCAATTATATTCAATATGGAAA-3'), PB2-Reverse-JT	(5'-CAAGGTCGTTTTTAAACTATTCGACAC-3'), PB1-Forward-JT (5'-GCAGGCAAACCATTTGAATGG-3'), PB1-Reverse-JT	(5'-CAAGGCATTTTTTCATGAAGGACAAG-3'), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG-3'), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG-3'), NA-Forward-JT	5'-(GCAGGAGTTTAAATGAATCCAAACC-3'), NA-Reverse-JT	(5'-CAAGGAGTTTTTTGAACAAACTACTTG-3'), HA-Forward-JT	(5'-GCAGGGGAAAATAAAAACAACCAAAAT-3'), HA-Reverse-JT	(5'-CAAGGGTGTTTTTCCTTATATTTCTGAA-3'), NP-Forward-JT	(5'-GCAGGGTAGATAATCACTCACAG-3'), NP-Reverse-JT	(5'-CAAGGGTATTTTTCTTTAATTGTCGTACT-3'), M-Forward-JT	(5'-GCAGGTAGATATTGAAAGATGAGTC-3'), M-Reverse-JT	(5'-CAAGGTAGTTTTTTACTCCAGCTCT-3'), NS-Forward-JT	(5'-GCAGGGTGACAAAGACATAATG-3'), NS-Reverse-JT	(5'-CAAAGGGTGTTTTTTATTATTAAATAAGCTG-3').
+Two microliters of RNA template were used  to generate cDNA in eight segment-specific one-step RT-PCR with $0.2\mu\text{M}$ of the following primers :   PB2-Forward-JT	(5'-GCAGGTCAATTATATTCAATATGGAAA-3'), PB2-Reverse-JT	(5'-CAAGGTCGTTTTTAAACTATTCGACAC-3'), PB1-Forward-JT (5'-GCAGGCAAACCATTTGAATGG-3'), PB1-Reverse-JT	(5'-CAAGGCATTTTTTCATGAAGGACAAG-3'), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG-3'), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG-3'), NA-Forward-JT	5'-(GCAGGAGTTTAAATGAATCCAAACC-3'), NA-Reverse-JT	(5'-CAAGGAGTTTTTTGAACAAACTACTTG-3'), HA-Forward-JT	(5'-GCAGGGGAAAATAAAAACAACCAAAAT-3'), HA-Reverse-JT	(5'-CAAGGGTGTTTTTCCTTATATTTCTGAA-3'), NP-Forward-JT	(5'-GCAGGGTAGATAATCACTCACAG-3'), NP-Reverse-JT	(5'-CAAGGGTATTTTTCTTTAATTGTCGTACT-3'), M-Forward-JT	(5'-GCAGGTAGATATTGAAAGATGAGTC-3'), M-Reverse-JT	(5'-CAAGGTAGTTTTTTACTCCAGCTCT-3'), NS-Forward-JT	(5'-GCAGGGTGACAAAGACATAATG-3'), NS-Reverse-JT	(5'-CAAAGGGTGTTTTTTATTATTAAATAAGCTG-3').
 Reaction conditions were $50^\circ\text{C}$ (60 min), $94^\circ\text{C}$ (2 min), followed by 30 cycles of $94^\circ\text{C}$ (30 sec), $54^\circ\text{C}$ (30 sec), and $68^{\circ}\text{C}$ (3 min).
 Molar equivalents of each PCR product were pooled to generate reconstituted cDNA genomes of WSN33 and PR8.
-The WSN33 cDNA pool was then serially diluted into the PR8 cDNA pool to yield mixtures in which WSN33 made up 5, 2.5, 1.25, 0.63, and 0.16\% of the population.
-Seven hundred and fifty nanograms of purified PCR products were sheared to an average size of 300-400 bp using a Covaris S220 focused ultrasonicator with the following settings : Intensity: 4, Duty cycle: 10%, Burst/second: 200, Duration: 80 seconds.
+The WSN33 cDNA pool was then serially diluted into the PR8 cDNA pool to yield WSN33-PR8 mixtures in which WSN33 made up 5, 2.5, 1.25, 0.63, and 0.16\% of the population.
+Seven hundred and fifty nanograms of each mixture were sheared to an average size of 300-400 bp using a Covaris S220 focused ultrasonicator with the following settings : Intensity: 4, Duty cycle: 10%, Burst/second: 200, Duration: 80 seconds.
 Sequencing libraries were prepared from these fragmented products using the NEBNext Ultra DNA library prep kit (NEB E7370L), Agencourt AMPure XP beads (Beckman Coulter A63881), and NEBNext multiplex oligonucleotides for Illumina (NEB E7600S).
@@
 A clonal plasmid control library was prepared from 8 plasmids containing PR8 genomic segments.
-These plasmids were mixed equimolarly and cDNA was generated using a multiplex one step RT-PCR with the primers Uni12/Inf1 (5'-GGGGGGAGCAAAAGCAGG-3'), Uni12/Inf3 (5'-GGGGGAGCGAAAGCAGG-3'), and Uni13/Inf1 (5'-5CGGGTTATTAGTAGAAACAAGG-3') as in [@Anonymous:A25Ka2tY].
+These plasmids were mixed to equal molarity, and cDNA was generated using a multiplex one step RT-PCR with the primers Uni12/Inf1 (5'-GGGGGGAGCAAAAGCAGG-3'), Uni12/Inf3 (5'-GGGGGAGCGAAAGCAGG-3'), and Uni13/Inf1 (5'-5CGGGTTATTAGTAGAAACAAGG-3') as in [@Anonymous:A25Ka2tY].
 This library was prepared in identical fashion to the experimental populations and was sequenced in the same Miseq lane.
@@
 
-Twenty point mutants were generated in the WSN33 background using the pHW2000 reverse genetics system [@Hoffmann:2002ih] as previously described (Visher et al., manuscript in preparation).
+Twenty point mutants were generated in the WSN33 background using the pHW2000 reverse genetics system [@Hoffmann:2002ih] (Visher et al., manuscript in preparation).
 In short, we used  overlap PCR mutagenesis to introduce the following mutations: HA\_T1583G, HA\_G1006T, HA\_G542T, M\_T861G, M\_A541C, NA\_G1168T, NA\_C454T, NP\_A454C, NP\_A1160T, NS\_G227T, NS\_A809G, PA\_T964G, PA\_T237A, PA\_A1358T, PB1\_G599A, PB1\_G1764T, PB1\_T1288A, PB2\_A1854G, PB2\_A440T, PB2\_A1167T.
+Viruses were rescued from transfected cells as in [@Pauly:2015eq].
 
-We passaged the 20 WSN33 point mutants and the WSN33 WT once in MDCK cells and varified the identity of the mutants by sequencing on a Illumina Miseq as above.
-We quantified the genome copy number of each supernant using a SuperScript III Platinum One-Step qRT-PCR (Invitrogen 11732) and univeral influenza A/B primer and probe sets [@Anonymous:tc].
-Infected supernatants were mixed at equal concentrations and diluted to generate a population with each mutant present at 5% frequency and a total concentration of 10^5^ genomes per microliter.
+We passaged the 20 WSN33 point mutants and the WSN33 WT once in MDCK cells and verified the identity and relative clonality of the mutants by sequencing each on an Illumina Miseq as above.
+We quantified the genome copy number of each supernatant using a SuperScript III Platinum One-Step qRT-PCR kit (Invitrogen 11732) and universal influenza A/B primer and probe sets [@Anonymous:tc].
+Equal genome equivalents of each infected supernatant were mixed  and diluted to generate a population with each mutant present at 5% frequency and a total concentration of 10^5^ genomes per microliter.
 We diluted this mixture into WT WSN33 supernatant to create populations in which each mutant was present at 2, 1, 0.5,and 0.2% frequency all with a total concentration of 10^5^ genomes per micoliter.
@@
 We prepared libraries as before and used Quanti PicoGreen dsDNA quantification (ThermoFisher Scientific P7589) to quantify the concentration of each indexed library.
-We pooled equal quantities (by nanogram) of each indexed library, and removed adaptor dimers using gel isolation with the GeneJET Gel Extraction Kit (ThermoFisher Scientific K0691) prior to sequencing an on Illumina Hiseq 2500 with 2 x 125 paired end reads.
+We pooled equal quantities (by nanogram) of each indexed library, and removed adaptor dimers by gel isolation with the GeneJET Gel Extraction Kit (ThermoFisher Scientific K0691) prior to sequencing an on Illumina Hiseq 2500 with 2 x 125 paired end reads.
 A clonal control library was processed in an identical fashion starting from an equimolar mix of 8 plasmids containing the WSN33 genomic segments.
-
-#### Duplicate populations
 
@@
 
-The ability to reliably distinguish true from false single nucleotide variants (SNV) is integral to accurate NGS based studies of viral diversity.
+The ability to reliably identify single nucleotide variants (SNV) is integral to accurate NGS-based studies of viral diversity.
 The accuracy of any SNV calling pipeline can be described in terms of its sensitivity and specificity.
-Sensitivity is the proportion of true variants that are properly identified; while, specificity is the proportion of true negatives that are properly identified.
+Sensitivity is the proportion of true variants that are properly identified, and specificity is the proportion of true negatives that are properly identified.
 A closely related concept is the false positive rate, which is 1-specificity, and gives the ratio of false positive variants to true negatives.
-A sensitivity and specificity of 1 describe perfect accuracy in which all the true variants are found and only true variants are found.
+An assay with perfect accuracy, in which all the true variants are found and only true variants are found, has a sensitivity and specificity of 1.
 
-There is an obvious trade off between sensitivity and specificity.
-Improved sensitivity often requires less stringent criteria in variant calling which reduces specificity.
+There is an obvious trade-off between sensitivity and specificity.
+Improved sensitivity often requires less stringent criteria in variant calling, but reduces specificity.
 Conversely, increased stringency can improve specificity but often reduces sensitivity.
 This relationship can be visualized using a receiver operating characteristic (ROC) curve (`r Xamp_roc`).
-An ROC curve plots the sensitivity of an assay along the y axis and 1-specificity along the x axis.
+An ROC curve plots the sensitivity of an assay along the y-axis and 1-specificity, or the false positive rate, along the x-axis.
 A variant calling pipeline must be tested against known data in order to construct an ROC curve.
@@
 
-###Initial accuracy
+### Initial accuracy
 
@@
 Instead, we robustly benchmark one variant caller, DeepSNV, and highlight approaches for improving its accuracy.
-In doing so we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is being applied.
+In doing so we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is applied.
 
@@
 DeepSNV then applies a hierarchical binomial model at each genomic position and identifies true variants as those with frequencies significantly above the noise found in the plasmid control.
-Like many variant calling methods, the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification.
-However, the accuracy has not been tested when applied to whole genome sequencing of a viral population amplified by reverse transcription PCR (RT-PCR).
+Like many variant calling algorithms, the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification.
+However, its accuracy has not been tested when applied to whole genome sequencing of a viral population amplified by reverse transcription PCR (RT-PCR).
 
@@
 The specificity was above 0.998 in all dilutions.
-While the drop in specificity appears small, a false positive rate of 0.002 corresponds to 78 false positives when applied across the `r miseq_vars` potential variants present in the `r miseq_kb` bp of the influenza genome we analyzed.
+While this drop in specificity appears small, a false positive rate of 0.002 corresponds to 78 false positives when applied across the `r miseq_vars` potential variants present in the `r miseq_kb` bp of the influenza genome.
 
@@
 
-Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions.
-Patient-derived populations are typically less divergent than WSN33 and PR8 (cite?), and the number of viral genomes in patient samples is much lower than in cell-culture derived populations (data not shown).
+Although the initial benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions.
+Patient-derived populations are typically less divergent than WSN33 and PR8 (cite Ghedin papers), and the number of viral genomes in patient-samples is much lower than that found in cell culture.
 To mimic these conditions, we generated 20 viral clones, each with a single point mutation in the WSN33 background.
@@
 We then serially diluted these populations into basal media to obtain mixtures with lower nucleic acid input.
-The range ($10^3-10^5$ copies per microliter) matches the inputs typically found in patient-derived samples (data not shown).
+The range (10^3^-10^5^ copies per microliter) matches the inputs typically found in patient-derived influenza virus samples (data not shown).
 We sequenced the samples on the Illumina Hiseq platform, and called variants using DeepSNV.
@@
 The 20 mutations present in our initial viral mixture were the only true positives considered in our analysis.
-Any SNV that was present in both duplicates of the wild type stock or in one of the viral clones (>1.0%) was masked and  considered neither a true positive nor a true negative.
+Any SNV that was present at >1% frequency in either in both duplicates of the wild type stock or any one of the viral clones was masked excluded from the analysis and  considered neither a true positive nor a true negative.
 
 In these populations with lower diversity and input titer, we maintained greater than 0.85 sensitivity for SNV at 1% frequency or higher.
-Our sensitivity was considerably lower for variants at or below 0.5% frequency (`r Hiseq`B and C).
-We observed an average read depth of over 10,000 reads/bp in these samples.
-Therefore, the drop in sensitivity is most likely due to the 1,000-fold decrease in nucleic acid concentration compared to the first data set and the fact that library preparation requires a number of sampling steps that may limit detection.
+.
+Despite a high depth of coverage (>10,000 reads per bp), our sensitivity was considerably lower for variants at or below 0.5% frequency (`r Hiseq`B and C).
+Therefore, the drop in sensitivity is most likely due to the 1,000-fold decrease in nucleic acid concentration compared to the first data set, and the fact that library preparation requires a number of sampling steps that may limit detection.
 During library preparation, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.
 
-The specificity was greater than 0.99 in all the samples with $10^5$ genomes per microliter.
+The specificity was greater than 0.99 in all the samples with 10^5^ genomes per microliter.
 As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome.
@@
 In the above analysis we masked bases that had a Phred  score less than 30 (0.001 probability of being incorrect) and did not apply any MapQ cut-offs.
-In an initial analysis, we applied frequently employed cut-offs such as a MapQ of 20 and a Phred of 30 to our data.
-These criteria were unable to distinguish true from false positives in our $10^5$ samples (`r Qual`A) and indicate that many false positives occur on well mapped reads with high quality base calls.
+In our next analysis analysis, we applied frequently employed cut-offs such as a MapQ of 20 and a Phred of 30 to our data.
+These criteria were unable to distinguish true from false positives in our 10^5^ samples (`r Qual`A) and indicate that many false positives occur on well mapped reads with high quality base calls.
 
 We further parsed our false variant calls by locating them within individual sequencing reads.
-It is well known that sequence quality drops near the end of a read (cite?), and we found that our false positives clustered at the termini of our paired end reads (`r Qual`B).
+It is well known that sequence quality drops near the end of a read (cite), and we found that our false positives clustered at the termini of our paired end reads (`r Qual`B).
 The average Phred score of these false positives was `r false_p_phred`, further demonstrating that filtering on quality score alone is insufficient.
@@
 
-Based on these data we applied a number of empirically determined cut-offs and markedly improved our specificity to >0.999 without sacrificing sensitivity (`r Qual` C and D).
+Based on these results, we applied a number of empirically determined cut-offs and markedly improved our specificity to >0.999 without sacrificing sensitivity (`r Qual`C and 4D).
 For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean Phred score of 35, and an average read position within the middle 50\% of the read.
 Under these conditions we found 10 or fewer false positives in 4 of the 5 samples.
-Given this success we applied a number of strategies to further increase our accuracy, including but not limited to: Benjamin Hochman p-value correction,  more stringent p values (<0.01) or frequency cut-offs (>0.2%), retaining PCR duplicate reads, trimming the ends of the influenza genome, and employing multiple distributions to estimate the error rate in the control sample.
-The impact of various filtering criteria on our data can be visualized in an interactive shiny application available for download at  \emph{https://github.com/lauringlab/benchmarking_shiny.git}.
+Given this success we applied a number of other strategies to further increase our accuracy, including but not limited to: Benjamini Hochberg p-value correction,  more stringent p values (<0.01) or frequency cut-offs (>0.2%), retention of duplicate PCR reads, trimming the ends of the influenza genome, and employing alternative statistical distributions to estimate the error rate in the control sample.
+None of these approaches significanlty imporved our accuracy over teh above qualitya nd read-position criteria.
+The impact of various filtering criteria on our data can be visualized in an interactive shiny application available for download at  https://github.com/lauringlab/benchmarking_shiny.git.
 
@@
 The mean percent difference between the measured frequency and the expected was `r mean_fold`\%.
-This error is likely due to stochastic PCR biases (cite?) in the initial RT-PCR reaction and should be kept in mind when employing down stream analysis that depend on frequency measurements such as variant fitness and diversity metrics that incorporate species abundance (e.g.
-Shannon entropy).
+This error is likely due to stochastic sampling errors associated with the RT-PCR and library preparation (cite?) and should be kept in mind when employing down stream analysis that depend on frequency measurements (e.g.
+variant fitness, haplotype reconstruction, Shannon's Entropy, and other diversity metrics).
 
@@
 
-DeepSNV is one of many variant callers that use a combination of empiric and statistical approaches to model error rates and separate true from false variants.
+DeepSNV is one of many variant callers that employ a combination of empiric and statistical approaches to model error rates.
 We asked whether the decreased accuracy observed in our data set was due simply to peculiarities specific to  DeepSNV.
@@
 This increased specificity is most likely due to the fact that the LoFreq algorithm already takes MapQ and Phred scores into account when calling variants and has a stringent strand bias filter that removes many of the variants found only at one end of a paired-end read.
-However even with these additional characteristics, the specificity of LoFreq was lower than our improved DeepSNV pipeline (`r Qual`), with over 40 false positives per sample.
+However, even with these additional characteristics, the specificity of LoFreq was lower than our improved DeepSNV pipeline (compare `r Lofreq` and `r Qual`), with over 40 false positives per sample.
 It appears that higher than expected false positive rates are not specific to DeepSNV and most likely plague many variant callers applied to patient-derived viral samples.
 
-###Lower input levels
+###Accuracy at Lower input levels
 
@@
 As expected, our sensitivity was lower in populations with fewer genomes.
-For example, a variant with a 0.5\% frequency in a $10^4$ genomes per microliter sample is expected to present on only 700 genomes in the initial RT-PCR, many of these will be lost due to bottlenecks in the amplification and library preparation process.
-In addition to reduced sensitivity we found reduced specificity in samples with lower starting input.
+For example, a variant with a 0.5\% frequency in a $10^4$ genomes per microliter sample is expected to be present on only 700 genomes in the initial RT-PCR.
+Many of these will be lost due to bottlenecks in the amplification and library preparation process.
+W also found reduced specificity in samples with lower starting input.
 The increase in false positives is presumably due to a greater dependence on RT-PCR amplification and consequent propagation of errors.
@@
 If RT-PCR errors are responsible for the reduced specificity found at lower input levels, they should be easily identified as variants present in only one of two RT-PCR reactions performed on the same RNA.
-To test this hypothesis, we performed RT-PCR reactions and sequenced duplicates of the  5, 2, 1, and 0.5\% variant frequency samples from our $10^4$ genomes per microliter collection.
+To test this hypothesis, we  sequenced duplicate RT-PCR reactions of the  5, 2, 1, and 0.5\% variant frequency samples from our $10^4$ genomes per microliter collection.
 The duplicates were processed separately, but sequenced on the same lane of an Illumina Hiseq.
 We applied the previously mentioned quality cut-offs and required that a given variant be found in both duplicates.
-By running samples in duplicates we reduced the number of false positives in each sample to 10 or fewer resulting in and a specificity of 0.9998 or greater (`r Dups`).
+By analyizing samples in duplicate, we reduced the number of false positives in each sample to 10 or fewer resulting in and a specificity of >0.9998 (`r Dups`).
 This increased specificity was not accompanied by a decreased sensitivity.
-In fact, we found a slight increase in sensitivity most likely due to variability in library preparation.
+In fact, we found a slight increase in sensitivity, most likely due to variability in library preparation.
 Thus, accurate analysis of low input samples can be achieved through duplicate RT-PCR and careful benchmarking experiments.
 
-###The consequences of sub-optimal SNV identification
+###Sub-optimal SNV identification confounds diversity measurements
 
 NGS of intrahost populations is commonly used determine the interaction between host or environmental factors and viral diversity.
-However calculations of viral diversity rely entirely on SNV identified in NGS data, and are very sensitive to the accuracy of these variant calls.
+Because measurements of viral diversity rely entirely on SNV identified in NGS data, they are very sensitive to the accuracy of these variant calls.
 To illustrate this problem, we calculated the diversity of our 10^5^ genomes per microliter samples at each step of our benchmarking process using three complementary metrics (`r Table`).

### 2016-03-23 doc/draft.Rmd
work on dispersion

@@
 
-**Validation and quality controls in sequence based studies of intrahost diversity**
+**Validation and quality controls in next generation sequencing based studies of intrahost diversity**
 
@@
 In these populations with lower diversity and input titer, we maintained greater than 0.85 sensitivity for SNV at 1% frequency or higher.
-.
 Despite a high depth of coverage (>10,000 reads per bp), our sensitivity was considerably lower for variants at or below 0.5% frequency (`r Hiseq`B and C).
-Therefore, the drop in sensitivity is most likely due to the 1,000-fold decrease in nucleic acid concentration compared to the first data set, and the fact that library preparation requires a number of sampling steps that may limit detection.
+The drop in sensitivity is most likely due to the 1,000-fold decrease in nucleic acid concentration compared to the first data set, and the fact that library preparation requires a number of sampling steps that may limit detection.
 During library preparation, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.
 
-The specificity was greater than 0.99 in all the samples with 10^5^ genomes per microliter.
+In our initial analysis of these data using the same default DeepSNV settings as above, the specificity was significantly lower, 0.xxx.
+These lower input samples underwent more PCR cycles, which skews the noise distribution of variant calls in the test vs.
+plasmid control samples.
+Therefore, we utilized an alternative beta binomial model available in DeepSNV, which more appropriately models these differences (cite).
+With these settings, the specificity was greater than 0.99 in all the samples with 10^5^ genomes per microliter.
 As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome.
@@
 
-We have shown that robust benchmarking is vital to accurate NGS based studies of viral diversity, and that differences in experimental design and sample preparation can lead to wide variability in the accuracy of SNV indentification.
-We found input nucleic acid concentrations, which can vary greatly in patient-derived samples, had large effects on both sensitivity and specificity.
-At moderate levels of nucleic acid input we could improve accuracy by filtering putative SNV based on quality metrics and read positions.
-We further improved our accuracy at low input levels by controlling for RT-PCR errors by processing samples in duplicates.
-While our quality cut offs may not be needed by all variant callers, our work with Lofreq indicates that experimental design is important in accurate SNV detection.
-This is important given that few variant callers (if any) have been benchmarked under patient-derived conditions.
-Finally, we show that these inaccuracies in SNV calling drastically impact down stream analysis, and that accurate conclusions require quantifying and limiting this error.
+Robust validation is essential in NGS-based studies of viral diversity.
+Differences in experimental design and sample preparation can lead to wide variability in the accuracy of SNV indentification.
+We found that input nucleic acid concentration, which can vary greatly in patient-derived samples, had a large impact on both the sensitivity and specificity of rare variant detection.
+At moderate levels of nucleic acid input we could improve accuracy by filtering putative SNV based on quality metrics and read position.
+We further improved our accuracy at low input levels by processing these samples in duplicate.
+While our quality cut-offs may not be universally applicable to all samples and variant callers, our data sugest that experimental design is critical for accurate SNV detection.
+These findings are important as few, if any, variant callers have been benchmarked under patient-derived conditions.
+Finally, we show that these inaccuracies in SNV calling drastically impact downstream analysis.
 
- To our knowledge, DeepSNV is the only variant calling method that has been benchmarked on real sequencing reads in which all true positive variants and their frequencies were known *a priori* and independent of NGS (cite Gerstung).
-Given this robust benchmarking we thought it ideal for identifying rare variants in populations of influenza virus.
-However, we found decreased sensitivity and specificity when applied to cell culture derived mixtures of WSN33 and PR8.
+We chose DeepSNV for our studies, because it is one of the only variant callers that has been validated on real sequencing reads in which all true positive variants and their frequencies were known *a priori* and independent of NGS (cite Gerstung).
+In our experimental conditions we found decreased sensitivity and specificity when applied to cell culture derived mixtures of WSN33 and PR8.
 The most striking drop in sensitivity came for variants found at 0.16%.
@@
 
- iv) could duplicates have been pooled and sequenced as one sample?
-
-IV. Comparision to Lofreq  and Consequences of errors
-  i) our benchmarking shows that DeepSNVs means of distinguishing true from false variants is partitially breaks down under conditions that mimic patient derived samples.
-  ii) This can be overcome with the addition of specific quality filters.
-  iii) We would not have known about these inaccuracies had we not run our benchmarking experiment.
-  iv) Do other variant callers have similar inaccuracies.
-  v) Yes Lofreq, although more accurate than our unmodified DeepSNV pipeline, had decreased specificity than had previously been reported.
-  vi) The initial Lofreq sensitivity and specificity metrics were based largely on a *in silico* data sets.
-  vii) Lofreq did not display this high accuracy in under our experimental conditions.
-  viii) While we did not robustly attempt to improve the Lofreq accuracy we did find that frequency threshold could improve its specificity.
-  ix) Experimental conditions matter, regardless of variant calling method and so accurate study requires controls to be run.
-
 #References

### 2016-03-23 scripts/figures.Rmd
onesided test numbers

@@
 # knitr::kable(subset(h.table.df,gc==4))
+
+twosided.hiseq<-mean(h.table.df$Specificity)
 ```
@@
 ```
+
+```{r,oneside.Hiseq}
+sum.df<-read.csv("../data/process/2015-6-23.one.sided/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#") # reading in the variants from the pipeline output
+mfe.df<-read.csv("../data/reference/mfe.csv")
+# Getting genome lenght and segment information from the wsn33 fasta file
+reference.fasta<-"../data/reference/wsn33_wt_plasmid.fa"
+segments <- fasta.seqlengths(reference.fasta)
+regions.bed <- data.frame(chr = gsub("[ ].*","", names(segments)), start=12, stop=segments-13, row.names=NULL) # the univeral primers are 12 and 13 bp long
+regions.bed<-mutate(regions.bed,chr=as.character(chr))
+
+# The samples are named in the following format log(genome copy/ul)_expected frequency of variants We'll use that information here to get the gc# and exp.freq for each variant called
+
+pat<-"([0-9]+)_([0-9]+)" 
+mutate(sum.df, gc = sub(pat,"\\1",Id),
+       exp.freq = sub(pat,"\\2",Id)
+)-> sum.df
+mutate(sum.df,exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->sum.df
+
+sum.df<-subset(sum.df,exp.freq>0.001) # remove this to include exp.freq 0.01%
+# Id the variants as T or F 
+true_snv<-read.csv("../data/reference/mutant_id.csv",stringsAsFactor=F) # get the T variants
+sum.df<-mutate(sum.df,category=mutation %in% true_snv$mutant) # add Column for True and false variants
+wt1_mut<-subset(sum.df,Id=="WT1",select=mutation)
+wt2_mut<-subset(sum.df,Id=="WT2",select=mutation)
+
+wt_mut<-intersect(wt1_mut$mutation,wt2_mut$mutation)
+
+#length(wt_mut) How many wt muts do we find.
+sum.df$category[sum.df$mutation %in% wt_mut]<-"wt"
+sum.df<-subset(sum.df,!(grepl("WT",Id))) # remove the WT sample
+
+sum.df$category[sum.df$mutation %in% mfe.df$mutation]<-"mfe"
+
+possible_tp<-20 # there are 20 possible true positives in the mix
+regions.bed<-mutate(regions.bed,length=stop-start-1) #because the numbers in the csv are the inner most binding sites of the primers and represent the last positiions that cannot be interogated.
+
+total_possible_vars<-sum(regions.bed$length)*3 # three possible variants/ position
+
+possible_vars<-total_possible_vars-length(wt_mut)-dim(mfe.df)[1] # removing the mfe and wt_mut sites
+hiseq_vars<-possible_vars
+hiseq_kb<-sum(regions.bed$length)/1000
+
+h.roc.df<-hiseq.roc(sum.df,possible_tp,possible_vars,">")
+
+x<-ddply(h.roc.df,~samp,fill_in_plots)
+
+x<-mutate(x,frequency=find_freq.v.h(exp.freq))
+
+x<-subset(x,gc==5)
+
+x$adj.sensitivity[x$frequency=="1.0%"]<-x$adj.sensitivity[x$frequency=="1.0%"]-0.03
+#x$adj.sensitivity[x$frequency=="5.0%"]<-x$adj.sensitivity[x$frequency=="5.0%"]+0.0
+
+Hiseq.roc<-ggplot(subset(h.roc.df,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols,name="Frequency",labels=c("5.0%","2.0%","1.0%","0.5%","0.2%"))+xlab(bquote("1-Specificity (x"~10^-3~")"))+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))+scale_x_continuous(limits=c(0,0.005),breaks=c(0,0.001,0.002,0.003,0.004,0.005),labels=c(0:5))+theme(legend.position="none")+geom_text(data=subset(x,gc==5),aes(x=1-adj.specificity,y=adj.sensitivity,label=frequency),hjust=-0.2)
+
+# This is so that the text is not cut off
+Hiseq.roc <- ggplot_gtable(ggplot_build(Hiseq.roc))
+Hiseq.roc$layout$clip[Hiseq.roc$layout$name == "panel"] <- "off"
+
+h.table.df<-hiseq.roc.table(sum.df,0.01,possible_tp,possible_vars,">")
+
+#knitr::kable(subset(h.table.df,gc==5))
+
+h.table.df<-rename(h.table.df,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True\nPositives","adj.specificity"="Specificity","FP"="False\nPositives"))
+
+h.table.df<-subset(h.table.df,gc==5,select=-c(gc))
+h.table.df$Frequency<-c("5.0%","2.0%","1.0%","0.5%","0.2%")
+h.table.df$Sensitivity<-round(h.table.df$Sensitivity,digits=dd)
+h.table.df$Specificity<-round(h.table.df$Specificity,digits=dd)
+
+tbl<-make.table(h.table.df,tt1)    # Make table
+
+readPNG("../results/pictures/hiseq_setup.png")->img # read in set up png
+g <- rasterGrob(img, interpolate=TRUE)
+qplot(1:1, 1:1, geom="blank") +
+annotation_custom(g, xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
+geom_blank()+xlab("")+ylab("")+
+theme(line = element_blank(),
+text = element_blank(),
+line = element_blank(),
+title = element_blank())->hiseq.setup
+
+ggdraw() +
+  draw_plot(hiseq.setup, 0.1, .57, 0.8, 0.43) + #x y width height
+  draw_plot(Hiseq.roc, 0.05, 0.33, .8, 0.33) +
+  draw_plot(tbl, 0.05, 0, .9, 0.33) +
+  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(1, 0.7, 0.35), size = 15)
+# 
+# knitr::kable(subset(h.table.df,gc==4))
+
+onesided.hiseq<-mean(h.table.df$Specificity)
+
+```

### 2016-03-24 doc/draft.Rmd
onesided test numbers
benchmarking_v1

@@
  ####################### To do List #######################
-# Is it clear that I'm only talking about the 10^5 samples until we get to the lower input section?
-
-#Mapq exploration - concordant vs nonconcordant pairs
-
-# add mfe raw data to repo. currently pulls data from desktop directory.
 
@@
 
-# plot of read start postion vs. genome
-# do we not find the same variants
-# L1-norm of 4_5 and 5_5
 ```
@@
 \n
-*Corresponding author
+\*Corresponding author
 
@@
 Until recently, investigations of viral diversity were limited by labor intensive sequencing approaches.
-However, it is now feasible to sequence patient-derived samples at sufficient read depth to detect rare single nucleotide variants (SNV), and the cost-effectiveness of NGS has led to an explosion of studies that quantify viral diversity within and between hosts (e.g.[@Andersen:2015jj], [@Grubaugh:2015by], [@Rogers:2015ij],[@Poon:2016gf]).
+However, it is now feasible to sequence patient-derived samples at sufficient read depth to detect rare single nucleotide variants (SNV), and the cost-effectiveness of NGS has led to an explosion of studies that quantify viral diversity within and between hosts (e.g.{Andersen:2015jj, Grubaugh:2015by, Rogers:2015ij,Poon:2016gf}).
 Given the increasing contribution of NGS to the field of virology, it is important to reflect on the accuracy of these methods and the controls that are needed to ensure robust conclusions.
 
-Because next generation sequencing of viral populations often produces tens of thousands of reads per genomic position, the sensitivity of these approaches has been assumed to be quite high (cite).
+Because next generation sequencing of viral populations often yields tens of thousands of reads per genomic position, the sensitivity of these approaches has been assumed to be quite high (cite).
 However, patient-derived samples can differ in input titer by orders of magnitude (e.g.
-[@Lau:2013ima],[@Teunis:2015iw],[@Takeyama:2015jg]), and this variability can have drastic impacts on the sensitivity of detection for rare events.
+{Lau:2013ima, Teunis:2015iw , Takeyama:2015jg}), and this variability can have drastic impacts on the sensitivity of detection for rare events.
 Furthermore, some library preparations result in variable sequencing coverage both within and between samples.
@@
 
-The specificity of NGS is another factor to consider when designing an experiment using patient-derived samples.
-Each sequencing platform has inherent error profiles (cite), and sample collection, target amplification, and library preparation are additional processes whereby errors can be introduced and propagated.The next generation sequencing of RNA virus populations is particularly error prone, and these errors must be accounted for to ensure accurate secondary analysis.
+The specificity of NGS is another important factor in NGS based studies of patient-derived samples.
+Each sequencing platform has inherent error profiles (cite), and sample collection, target amplification, and library preparation are additional processes whereby errors can be introduced and propagated.
+The next generation sequencing of RNA virus populations is particularly error prone, and these errors must be accounted for to ensure accurate secondary analysis.
 
@@
 However, each approach has its own caveats that ultimately limit its application.
-Cirseq is an ingenious approach in which  template RNA is sheared and circularized prior to reverse transcription  [@Acevedo:2014ej].
-Subserquent rolling circle cDNA synthesis produces tandem reads, generating a consensus sequence for each RNA fragment.
-While this method is likely to be highly sensitive for rare variant detection and can control for reverse transcription, PCR, and sequencing errors, the requirement for a large and relatively pure population of viral RNA limits its applicability to patient-derived samples [@Acevedo:2014df].
+Cirseq is an ingenious approach in which  template RNA is sheared and circularized prior to reverse transcription  {Acevedo:2014ej].
+Subsequent rolling circle cDNA synthesis produces tandem reads, generating a consensus sequence for each RNA fragment.
+While this method is likely to be highly sensitive for rare variant detection and can control for reverse transcription, PCR, and sequencing errors, the requirement for a large and relatively pure population of viral RNA limits its applicability to patient-derived samples {Acevedo:2014df}.
 "Primer ID" based methods require less input and target sequencing to the viral genome.
-This approach relies upon barcoded primers to construct consensus sequences for each cDNA template and can control for PCR and sequencing errors  [@Jabara:2011eo].
-Because Primer ID methods require taht each bar code be physically attached to a PCR product, they  are most easily applied to a small targeted regions of the genome.
+This approach relies upon barcoded primers to construct consensus sequences for each cDNA template and can control for PCR and sequencing errors  {Jabara:2011eo}.
+Because Primer ID methods require that each bar code be physically attached to a PCR product, they  are most easily applied to a small targeted regions of the genome.
 As such, they have limited application in whole genome sequencing.
 
-An adapted form of sequence independent single primer amplification (SISPA) is an alternative approach that allows for whole genome sequencing and controls for errors propagated during library preparation [@Djikeng:2008cd].
+An adapted form of sequence independent single primer amplification (SISPA) is an alternative approach that allows for whole genome sequencing and controls for errors propagated during library preparation {Djikeng:2008cd}.
 In this method RT-PCR products are sheared and tagged with bar-coded random primers in a klenow reaction, prior to library preparation.
-SISPA controls for any errors that may arise during the library amplification such as PCR biases.
+SISPA controls for any errors that may arise during library amplification such as PCR biases.
 This method has been used in conjunction with statistical algorithms to control for accuracy in studies of intrahost influenza diversity (Ghedin group citations).
-However, the bar-coding reaction used in SISPA is biased and can result in uneven coverage and sensitivity across the genome  [@Rosseel:2013dn].
+However, the bar-coding reaction used in SISPA is biased and can result in uneven coverage and sensitivity across the genome  {Rosseel:2013dn}.
 
-Given the above limitations, many statistical approaches have been developed to distinguish true variants from sequencing errors [@Gerstung:1js; @Isakov:2015gq; @Wilm:2012br; @Macalalad:2012fx; @Koboldt:2009dk].
+Given the above limitations, many statistical approaches have been developed to distinguish true variants from sequencing errors {Gerstung:1js, Isakov:2015gq, Wilm:2012br, Macalalad:2012fx, Koboldt:2009dk}.
 In general, these algorithms calculate base-specific error rates from various metrics including but not limited to: mapping quality (MapQ), base quality (Phred), strand bias, and sequence context.
@@
 
-# Methods
+## Methods
 
@@
 Influenza A/WSN/33(H1N1) virus was rescued from transfected cells using the 8 plasmid reverse genetic system containing the genomic segments (pHW181-188), a kind gift from Robert Webster (St.
-Jude's Children's Research Hospital) [@Hoffmann:2002ih,@Pauly:2015eq].
-A biological clone of influenza A/Puerto Rico/8/1934(H1N1) was obtained from ATCC (VR-1469) and the genomic segments were cloned into the pHW2000 reverse genetic system [@Pauly:2015eq].
+Jude's Children's Research Hospital) {Hoffmann:2002ih, Pauly:2015eq}.
+A biological clone of influenza A/Puerto Rico/8/1934(H1N1) was obtained from ATCC (VR-1469) and the genomic segments were cloned into the pHW2000 reverse genetic system {Pauly:2015eq}.
 
-### Viral populations
+#### Viral populations
 
@@
 A clonal plasmid control library was prepared from 8 plasmids containing PR8 genomic segments.
-These plasmids were mixed to equal molarity, and cDNA was generated using a multiplex one step RT-PCR with the primers Uni12/Inf1 (5'-GGGGGGAGCAAAAGCAGG-3'), Uni12/Inf3 (5'-GGGGGAGCGAAAGCAGG-3'), and Uni13/Inf1 (5'-5CGGGTTATTAGTAGAAACAAGG-3') as in [@Anonymous:A25Ka2tY].
+These plasmids were mixed to equal molarity, and cDNA was generated using a multiplex one step RT-PCR with the primers Uni12/Inf1 (5'-GGGGGGAGCAAAAGCAGG-3'), Uni12/Inf3 (5'-GGGGGAGCGAAAGCAGG-3'), and Uni13/Inf1 (5'-5CGGGTTATTAGTAGAAACAAGG-3') as in {Anonymous:A25Ka2tY}.
 This library was prepared in identical fashion to the experimental populations and was sequenced in the same Miseq lane.
@@
 
-Twenty point mutants were generated in the WSN33 background using the pHW2000 reverse genetics system [@Hoffmann:2002ih] (Visher et al., manuscript in preparation).
+Twenty point mutants were generated in the WSN33 background using the pHW2000 reverse genetics system {Hoffmann:2002ih} (Visher et al., manuscript in preparation).
 In short, we used  overlap PCR mutagenesis to introduce the following mutations: HA\_T1583G, HA\_G1006T, HA\_G542T, M\_T861G, M\_A541C, NA\_G1168T, NA\_C454T, NP\_A454C, NP\_A1160T, NS\_G227T, NS\_A809G, PA\_T964G, PA\_T237A, PA\_A1358T, PB1\_G599A, PB1\_G1764T, PB1\_T1288A, PB2\_A1854G, PB2\_A440T, PB2\_A1167T.
-Viruses were rescued from transfected cells as in [@Pauly:2015eq].
+Viruses were rescued from transfected cells as in {Pauly:2015eq}.
 
 We passaged the 20 WSN33 point mutants and the WSN33 WT once in MDCK cells and verified the identity and relative clonality of the mutants by sequencing each on an Illumina Miseq as above.
-We quantified the genome copy number of each supernatant using a SuperScript III Platinum One-Step qRT-PCR kit (Invitrogen 11732) and universal influenza A/B primer and probe sets [@Anonymous:tc].
+We quantified the genome copy number of each supernatant using a SuperScript III Platinum One-Step qRT-PCR kit (Invitrogen 11732) and universal influenza A/B primer and probe sets {Anonymous:tc}.
 Equal genome equivalents of each infected supernatant were mixed  and diluted to generate a population with each mutant present at 5% frequency and a total concentration of 10^5^ genomes per microliter.
@@
 We prepared libraries as before and used Quanti PicoGreen dsDNA quantification (ThermoFisher Scientific P7589) to quantify the concentration of each indexed library.
-We pooled equal quantities (by nanogram) of each indexed library, and removed adaptor dimers by gel isolation with the GeneJET Gel Extraction Kit (ThermoFisher Scientific K0691) prior to sequencing an on Illumina Hiseq 2500 with 2 x 125 paired end reads.
+We pooled equal quantities (by nanogram) of each indexed library, and removed adapter dimers by gel isolation with the GeneJET Gel Extraction Kit (ThermoFisher Scientific K0691) prior to sequencing an on Illumina Hiseq 2500 with 2 x 125 paired end reads.
 A clonal control library was processed in an identical fashion starting from an equimolar mix of 8 plasmids containing the WSN33 genomic segments.
 
-Fresh RNA was isolated from the 5%, 2%, 1%, and 0.5% 10^4^ genomes per microliter samples and processed in duplicate as above, and these duplicates were sequenced with a plasmid control using 2 x 125 paired end reads on an Illumina Hiseq 2500 machine.
+Fresh RNA was isolated from the 5%, 2%, 1%, and 0.5% 10^4^ genomes per microliter samples and processed and seqeunced in duplicate as above.
 
@@
 
- Reads were aligned to either a PR8 or a WSN33 reference sequence using Bowtie2 [@Langmead:2012jh].
-Alignments were sorted and PCR duplicateds removed using Picard (http://broadinstitute.github.io/picard/).
-Variants were called using either DeepSNV [@Gerstung:1js] or Lofreq [@Wilm:2012br] and filtered using the Pysam module in python (https://github.com/pysam-developers/pysam) and custom R scripts.
+ Reads were aligned to either a PR8 or a WSN33 reference sequence using Bowtie2 {Langmead:2012jh}.
+Alignments were sorted and PCR duplicates removed using Picard (http://broadinstitute.github.io/picard/).
+Variants were called using either DeepSNV {Gerstung:1js} or Lofreq {Wilm:2012br} and filtered using the Pysam module in python (https://github.com/pysam-developers/pysam) and custom R scripts.
 Bases with a Phred<30 were masked in the DeepSNV analysis.
-We connected all of these steps into an analysis pipeline using bpipe [@em:2012wd] which is available for download at https://github.com/lauringlab/variant\_pipeline.
+We connected all of these steps into an analytical pipeline using bpipe {em:2012wd} which is available for download at https://github.com/lauringlab/variant\_pipeline.
 To save memory during SNV processing only variants with p<0.9 were included in our ROC curve analysis, as the vast majority of true negatives are trivial to identify and have a p=1.
@@
 
-### Initial accuracy
+## Initial accuracy
 
@@
 Instead, we robustly benchmark one variant caller, DeepSNV, and highlight approaches for improving its accuracy.
-In doing so we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is applied.
+In doing so we demonstrate the importance of validating any variant calling method under the experimental conditions to which it is applied.
 
-DeepSNV is a variant calling algorithm that uses a clonal, plasmid-derived, control to estimate local error rates across the genome [@Gerstung:1js].
+DeepSNV is a variant calling algorithm that uses a clonal, plasmid-derived, control to estimate local error rates across the genome {Gerstung:1js}.
 Because it is clonal, the sequence of the control is known with a high degree of confidence, and any nonconsensus base is indicative of an error.
@@
 We plaque purified and expanded WSN33 and PR8 populations.
-Complementary DNA from both viruses were mixed serially such that WSN33 cDNA was present at frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq`A).
+Complementary DNA from both viruses was mixed serially such that WSN33 cDNA was present at frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq`A).
 These viruses differ at 491 positions (primer sites used in RT-PCR were excluded from analysis), providing 491 true positives in each dilution.
-On plasmids subjected to limited PCR, DeepSNV identified known variants at 0.1\% frequency with a sensitivity of 0.860 and  specificity of 1.0 [@Gerstung:1js].
+On plasmids subjected to limited PCR, DeepSNV identified known variants at 0.1\% frequency with a sensitivity of 0.860 and  specificity of 1.0 {Gerstung:1js}.
 Under our experimental conditions and with a slightly more stringent p value (p=0.01 vs.
@@
 
-###An experimental intrahost population
+##An experimental intrahost population
 
@@
 We sequenced the samples on the Illumina Hiseq platform, and called variants using DeepSNV.
-We also processed and sequenced the WSN33 stock in duplicate to control for any mutations in the viral dilutent.
+We also processed and sequenced the WSN33 stock in duplicate to control for any mutations in the viral diluent.
 
 The 20 mutations present in our initial viral mixture were the only true positives considered in our analysis.
-Any SNV that was present at >1% frequency in either in both duplicates of the wild type stock or any one of the viral clones was masked excluded from the analysis and  considered neither a true positive nor a true negative.
+Any SNV that was present at >1% frequency in either in both duplicates of the wild type stock or any one of the viral clones was masked, excluded from the analysis, and  considered neither a true positive nor a true negative.
 
@@
 
-In our initial analysis of these data using the same default DeepSNV settings as above, the specificity was significantly lower, 0.xxx.
-These lower input samples underwent more PCR cycles, which skews the noise distribution of variant calls in the test vs.
-plasmid control samples.
-Therefore, we utilized an alternative beta binomial model available in DeepSNV, which more appropriately models these differences (cite).
+In our initial analysis of these data using the same default DeepSNV settings as above, the specificity was significantly lower (mean of `r onesided.hiseq` with a minimum of `r onesided.min`).
+These lower input samples underwent more PCR cycles, which have been shown to skew the error distributions in the test libraries relative to the plasmid control (cite DeepSNV doc).
+We were able to partially account for this variation, by using an alternative beta binomial model available in DeepSNV, which more appropriately fits these conditions.
 With these settings, the specificity was greater than 0.99 in all the samples with 10^5^ genomes per microliter.
@@
 The false positives outnumber the true positives by 100-fold in these populations with realistic diversity and input.
-We were able to increased our specificity  by applying a more stringent p-value cutoff.
-However, as shown by the ROC curves in `r Hiseq`, this move towards the y axis markedly reduces sensitivity.
+We were able to increase our specificity  by applying a more stringent p-value cutoff.
+However, as shown by the ROC curves in `r Hiseq`, this move towards the y-axis markedly reduces sensitivity.
 Our data demonstrate that with moderate concentrations of input nucleic acid, even statistically significant p-values from a robust variant caller are not sufficient to accurately separate true from false positive variants.
 
-###Additional filtering criteria
+##Additional filtering criteria
 
-Many next generation sequencing studies utilize mapping quality (MapQ) and/or base quality (Phred) cut-offs to ensure that only the highest caliber sequencing data is used to call variants.
+Many next generation sequencing studies utilize mapping quality (MapQ) and/or base quality (Phred) thresholds to ensure that only the highest caliber sequencing data is used to call variants.
 Mapping quality measures the probability that a given read is mapped to the correct position in the genome while base quality estimates the likelihood that the base call by the sequencer is correct.
@@
 Given this success we applied a number of other strategies to further increase our accuracy, including but not limited to: Benjamini Hochberg p-value correction,  more stringent p values (<0.01) or frequency cut-offs (>0.2%), retention of duplicate PCR reads, trimming the ends of the influenza genome, and employing alternative statistical distributions to estimate the error rate in the control sample.
-None of these approaches significanlty imporved our accuracy over teh above qualitya nd read-position criteria.
+None of these approaches significantly improved our accuracy over the above quality and read-position criteria.
 The impact of various filtering criteria on our data can be visualized in an interactive shiny application available for download at  https://github.com/lauringlab/benchmarking_shiny.git.
@@
 
-###Alternative variant callers
+##Alternative variant callers
 
@@
 We asked whether the decreased accuracy observed in our data set was due simply to peculiarities specific to  DeepSNV.
-We therefore analyzed our $10^5$ input populations using Lofreq, another variant caller that is commonly used in next generation sequencing studies, and has been reported to have perfect specificity [@Wilm:2012br].
+We analyzed our $10^5$ input populations using Lofreq, another variant caller that is commonly used in next generation sequencing studies, and has been reported to have perfect specificity {Wilm:2012br}.
 Under our experimental conditions, Lofreq had marginally reduced sensitivity compared to DeepSNV when applied to the variant frequencies $\geq$ 1.0\% but marginally increased sensitivity when applied to variant frequencies $<$ 1.0\% (`r Lofreq`).
@@
 This increased specificity is most likely due to the fact that the LoFreq algorithm already takes MapQ and Phred scores into account when calling variants and has a stringent strand bias filter that removes many of the variants found only at one end of a paired-end read.
+Because it does not compare test samples to a plasmid control, it is more robust to issues of PCR skewing than DeepSNV.
 However, even with these additional characteristics, the specificity of LoFreq was lower than our improved DeepSNV pipeline (compare `r Lofreq` and `r Qual`), with over 40 false positives per sample.
@@
 
-###Accuracy at Lower input levels
+##Accuracy at lower input levels
 
@@
 Many of these will be lost due to bottlenecks in the amplification and library preparation process.
-W also found reduced specificity in samples with lower starting input.
+We also found reduced specificity in samples with lower starting input.
 The increase in false positives is presumably due to a greater dependence on RT-PCR amplification and consequent propagation of errors.
@@
 
-###Sub-optimal SNV identification confounds diversity measurements
+##Sub-optimal SNV identification confounds diversity measurements
 
@@
 Robust validation is essential in NGS-based studies of viral diversity.
-Differences in experimental design and sample preparation can lead to wide variability in the accuracy of SNV indentification.
+Differences in experimental design and sample preparation can lead to wide variability in the accuracy of SNV identification.
 We found that input nucleic acid concentration, which can vary greatly in patient-derived samples, had a large impact on both the sensitivity and specificity of rare variant detection.
@@
 We further improved our accuracy at low input levels by processing these samples in duplicate.
-While our quality cut-offs may not be universally applicable to all samples and variant callers, our data sugest that experimental design is critical for accurate SNV detection.
+While our quality cut-offs may not be universally applicable to all samples and variant callers, our data suggest that experimental design is critical for accurate SNV detection.
 These findings are important as few, if any, variant callers have been benchmarked under patient-derived conditions.
@@
 In our experimental conditions we found decreased sensitivity and specificity when applied to cell culture derived mixtures of WSN33 and PR8.
-The most striking drop in sensitivity came for variants found at 0.16%.
-Under these conditions we found true variants at a rate of `r miseq.sense.16` compared the expected sensitivity of  0.860 for variants at 0.1%.
-This decrease is most likely due to the fact that we are looking for variants across `r miseq_kb`bp of the influenza genome where previous benchmarking focused on just 1,500 bp of the HIV pol gene.
-Therefore, our sequencing efforts were diluted across the influenza genome resulting in decreased read depth  which limits sensitivity.
-Additionally,  we aligned all reads to a PR8 reference genome which biases our detection of WSN33 variants especially rare variants, which already have few reads.
-We also observed a relatively modest decrease in specificity.
-Despite the use of bonferronni corrected p values, the drop in specificity is most likely due to multiple testing errors caused by applying the algorithm across `r miseq_kb`bp (cite).
+The decrease in our observed accuracy most likely stems from differences in our experimental system compared to what has previously been used to validate DeepSNV.
+Because we aligned all reads to the PR8 genome, many of the WSN33-derived reads may have not aligned as well, biasing the analysis against the detection of rare WSN33 SNV.
+Our data also show that extremely high specificity is required in cases of whole genome sequencing.
+There are `r miseq_vars` true negatives in our data set, and a large potential for false positive detection even when using bonferroni corrected p values.
 
- Encouraged by these results, we applied DeepSNV to experimental populations that more closely mimic the diversity and nucleic acid input levels found in patient derived samples.
-At a modest input concentration of 10^5^ genomes per microliter we found a four times as many false positives.
-It should be noted that our specificity in this case remained above 0.99 in all populations and that seemly robust specificities can result in an unacceptable number of erroreous SNV when whole genomes are analyzed.
-The reduced specificity was likely an artifact of how the populations were constructed and stresses the importance of benchmarking SNV identifying pipelines under conditions to which they will be applied.
-Our WSN3-PR8 populations were constructed from reconstituted cDNA genomes and were devoid of any biases that might accompany PCR amplification.
-However, this was not the case in our 20 mutant populations.
-To more accurately mimic the method we use to process patient samples we mixed all 20 point mutants and WSN33 WT supernatants prior to  amplification in a multiplex RT-PCR.
-This method undoubtedly introduced unavoidable PCR biases into our populations.
-Gerstung *et.
-al.* have shown that PCR amplification biases lead to increased variability in the subconsensus nucleotide counts (cite DeepSNV documentation).
-Increased bariability in the test samples increases the likelihood that an erroneous SNV will exceed the error rate estimated in the plasmid control.
-DeepSNV attempts to control for such variability in part by estimating the dispersion of a sample, which allows the algorithm to incorporate greater variance into the null model.
-In fact, our 20 mutant libraries were more dispersed than those in cell culture, indicating a need for greater variance.
-However, the dispersion estimates are limited and the variability likely exceeded what can be controled by this approach.
+ A key strength of our study is that we applied DeepSNV to experimental populations that more closely mimic the diversity and viral genome concentrations found in patient-derived samples of influenza virus.
+At a modest input of 10^5^ genomes per microliter and "default" DeepSNV settings, false positives can outnumber true positives by a factor of 4.
+It should be noted that our specificity in all cases remained above 0.99.
+When applied across an entire viral genome, a specificity of >0.9995 is required to reduce false positives to reasonable levels.
+As above, the decreased accuracy of DeepSNV under these conditions is due to small but important differences our experimental design compared to what has been previously reported, namely input nucleic acid concentration and RT-PCR amplification.
 
-Because RT-PCR biases reduced the efficacy of the DeepSNV algorithm, we search for alternative means of increasing accuracy.
-Out-of-the box DeepSNV is rather agnostic towards the mapping quality and base quality of a given variant, and only masks bases calls with Phred scores below a user designated threshold.
-We were able to use  quality metrics to filter putative variants and greatly increase the accuracy of SNV identification.
-However, commonly used thresholds (MapQ>20, Phred>30) did little to increase the accuracy of our method.
-Instead, we relied on our own empirically determined thresholds.
-We found the average MapQ and Phred scores of putative SNV were bimodal with true SNV found in the higher of the two distributions.
-In both the MapQ and Phred distributions there was a clear distinction between the two populations, and we applied quality thresholds to eliminate any putative SNV found in the lower of the two distributions.
-We found no correlation between MapQ and Phred (p>x), suggesting these metrics are independent and thus both thresholds eliminate distinct false positives.
-We have sequenced over 300 influenza libraries in more than 5 Hiseq runs and observe consistent quality distributions for DeepSNV identified putative SNV.
-Our group has also deeply sequenced many polio virus libraries and see the same bimodal trends.
-However, we do see shifts in the means of our MapQ distributions when comparing polio and influenza libraries.
-This is most likely due to differences in the genomic structures of the viruses.
-The terminal regions of all 8 influenza genomic segments are conserved and this conservation decreases the MapQ of reads that map to these areas.
-Therefore, while our MapQ and Phred thresholds are robust in our system, they may need to be adjusted for use in other systems.
+Because DeepSNV is somewhat agnostic towards the mapping quality and the base quality of a given variant, we sought to improve our accuracy by identifying thresholds that more effectively distinguished true positive SNV.
+Here we found that commonly used thresholds (MapQ>20, Phred>30) did little to improve accuracy.
+In our data sets the distributions of average MapQ and Phred scores of putative SNV were clearly bimodal with true SNV found in the higher of the two distributions.
+Our empirically determined  thresholds were chosen to eliminate any putative SNV found in the lower of the two distributions.
+Importantly we found no correlation between MapQ and Phred (adjusted R^2^=0.003), suggesting that these metrics are independent and that the two cut-offs eliminate distinct sets of false positives.
+These thresholds should be reproducible as we we have sequenced over 300 influenza libraries in 5 Hiseq runs and have observed consistent MapQ and Phred quality distributions.
+We have seen the same bimodal trend in libraries of poliovirus populations, but have applied a lower empirically determined MapQ cut-offs to these data.
+This difference is most likely due to differences in the genomic structures of the viruses.
+The terminal regions of all 8 influenza genomic segments are conserved and this conservation alters the MapQ of reads that map to these areas.
+While our MapQ and Phred thresholds are robust in our system, they may need to be adjusted for use in others.
 
-Even in the face of stringent MapQ and Phred cut-offs, we found many high quality false positive SNV were identified only at the termi of paired end reads.
-We removed these by filtering putative SNV calls based on their average position in a paired end read.
-In particular these false positives were found almost exclusively in regions of the genome that were enriched for read start sites.
-This enrichment is most likely a consequence  of sequence context, the fragmentation process, and our size selection protocol; however, it is unclear why our plasmid control would not contain similar biases.
-The cause may also have a biological foundation as previous work has shown that defective interfering  particles, which commonly arise during cell passage,contain truncated genomic segments.
-Such truncated segments would only be present in infected supernatants and not a plasmid control.
-These  "read position" false positives were reproducibly found across a number of libraries; however, because our true positives were made by overlap PCR mutagenesis we can be confident such false positives are indeed artifacts of the experimental design and/or sample preparation, and  that we have robustly controlled for such errors.
+Even in the face of stringent MapQ and Phred cut-offs, we found many high-quality false positive SNV were identified only at the termini of paired end reads.
+We removed these by filtering putative SNV based on their average position in a paired end read.
+These false positives were found almost exclusively in regions of the genome that were enriched for read start sites.
+This enrichment may be a consequence of sequence context, the fragmentation process, and our size selection protocol.
+We suggest that there might also be a biological reason for this effect, as our PCR-amplified plasmid control samples did not exhibit this bias.
+For example, defective interfering particles, which commonly arise during cell passage contain truncated genomic segments and would only be present in infected supernatants and not the plasmid control.
+We hypothesize that the the large deletions in these segments increase the number of reads that start at certain genomic positions.
+As the beginning of reads are error prone, this enrichment results false positive SNV.
+Our analysis was particularly vulnerable to this type of error as we did not trim the ends of our reads and DeepSNV, unlike other variant callers, does not directly test for strand bias or consider position on a read as a variable.
+While it is easy to diagnosis these short comings in retrospect, such errors had not been previously reported for DeepSNV, and were only elucidated through our rigorous validation.
 
-It is common to filter and trim sequencing reads prior to SNV identification.
-However, we have taken a slightly different approach in our analysis.
-In the initial SNV identification step we have masked bases with Phred<30 but made no additional restrictions on the raw data.
-Only after putative SNV (those that exceed the frequency expected given a plasmid control) have been identified do we impose quality restrictions.
-Our approach treats variant nucleotides more stringently than consensus base calls, but this is of little concern for two reasons.
-Firstly, we have identified specificity as a large the problem in accurate SNV identification, and so stringently filtering potential false positives seems appropriate.
-Additional, since there are orders of magnitude more sequenced reads for a consensus base than there are for rare variant bases, and our thresholds are based on mean quality scores, it is unlikely we would remove any concensus bases from analysis.
+While read filtering and trimming are common in NGS data sets, we have taken a slightly different approach in our analysis.
+In the initial SNV identification step we masked bases with Phred <30 but made no additional restrictions on the raw data.
+We only imposed additional quality restrictions after putative SNV - those that exceed the frequency expected given a plasmid control - were identified.
+Our approach treats variant nucleotides more stringently than consensus base calls.
+We do not think that this differential stringency introduces unnecessary bias.
+Because we identified specificity as the major problem in accurate SNV identification, stringent filtering of potential false positives seems appropriate.
+Furthermore, the vast majority of reads call a consensus base and our mean quality score thresholds would therefore not be expected to remove many consensus base calls from the analysis.
 
-Frequency thresholds are an additional quality filter that is often applied to SNV after identification.
-Common thresholds range from 0.1% to 1%; however, there appears to a consensus that SNV identification is unreliable below 0.1%.
-We did not apply direct frequency thresholds.
-In our DeepSNV analysis no putative varaints wer found below 0.1% and we found in our case that frequency cut offs limited sensitivity and did not improve specificity.
-Read depth, or coverage, is another metric that is often used in conjunction with frequency to insure accurate SNV identification.
-Although we did not apply a direct coverage cut off, DeepSNV has been reported to require a coverage of ten times the reciprocal of frequency for sufficient power to call SNV.
-This means that a coverage of 1,000 is needed to detect a variant at 1% frequency.
-In our analysis the lowest coverage for a true positive was `r tp_cov$coverage` (`r round(tp_cov$freq.var*100,digits=1)`% frequency) while the lowest coverage for a false positive variant was `r fp_cov$coverage` (`r round(fp_cov$freq.var*100,digits=1)`% frequency).
-Read depth is an important variable in NGS based studies of viral diversity and it, with input concentrations determine the power to detect rare variants.
-If there are large biases in read depth or input levels, two similar populations may appear distinct based solely on ascertainment bias.
+Frequency thresholds of 0.1 to 1% represent an additional quality filter that is often applied to SNV after identification.
+We did not apply direct frequency thresholds in our analysis, as we found that arbitrary cut-offs limited sensitivity without improving specificity.
+Read depth, or coverage, is another metric that is often used in conjunction with frequency to ensure accurate SNV identification.
+Although we did not apply a direct coverage cut-off, DeepSNV has been reported to require a coverage of ten times the reciprocal of frequency for sufficient power to call SNV.
+Therefore, a coverage of 1,000x is needed to detect a variant at 1% frequency.
+In our analysis the lowest coverage for a true positive was 1795 (4.8% frequency) while the lowest coverage for a false positive variant was 966 (8.5% frequency).
+If a given data set has variability in read depth across the genome, SNV at identical frequencies may be detected with differing sensitivity.
+The application of variant frequency or read depth thresholds would lead to severe ascertainment bias in subsequent analyses of diversity.
 
-Despite high accuracy at 10^5^ genomes per microliter we observed a decrease in sensitivity in our 10^3^ genomes per microliter samples most likely due to the lack of power discussed above.
-However, this drop in sensitivity was accompanied by reduced specificity.
-At lower nucleic acid concentrations, our SNV pipeline relied more heavily on RT-PCR amplification, which is an error prone process.
-RT-PCR errors were indistinguishable from true positive variants, as they do not arise from a sequencing errors or bioinformatic artifacts.
-We were able to limit these errors which are sporadic and random (cite) by processing low input samples in duplicate, and achieved comparable accuracy to what was seen at 10^5^ genomes per microliter.
-Quantifying and controlling for RT-PCR errors in this way will allow us to accurately compare patient-samples that fall in a wider this input.
+Few studies of intrahost diversity quantify or control for the number of genomes in a sample.
+We found that input copy number is a key factor in variant detection.
+Despite high accuracy at 10^5^ genomes per microliter we observed a decrease in sensitivity in our 10^3^ genomes per microliter samples.
+More importantly, this drop in sensitivity was accompanied by reduced specificity.
+At lower nucleic acid concentrations, our SNV pipeline relied more heavily on RT-PCR amplification, which tends to propagate errors that are otherwise indistinguishable from true positives in sequence data.
+We were able to limit these sporadic and random errors by processing low input samples in duplicate.
+Quantifying and controlling for RT-PCR errors in this way will allow us to accurately compare patient-derived samples that vary over a range of input.
 
-As we have shown, experimental design impacts the accuracy of SNV identifying pipelines and must be considered in NGS based studies of viral diversity.
-Target amplification, genome structure and nucleic acid concentration are unavoidable sources of error that can vary depending on the virus and sample collection method.
-The reported accuracy of SNV identifying algorithms is not directly applicable to every experimental design, and as we have shown these deviations can have profound effects on down stream analysis.
-In studying influenza, we have benefited from an easily accessible reverse genetics system that made for a straight forward mutagenesis based approach to benchmarking.
-Under our experimental conditions we have been able to greatly improve the accuracy of DeepSNV.
-Even though the accuracy remains is imperfect, we are now equipped with an understanding of the limitations of our method.
-There may be more accurate variant callers than DeepSNV, or algorithms better suited for other systems; however, our work with Lofreq stresses the fact that all methods have their limitations and understanding these limitations is vital for accurate science.
-Next generation sequencing is powerful tool that is just beginning to unlock the complex dynamics of intrahost viral populations.
-As this tool becomes more and more popular, it is vital we remember that in NGS as in all science, controls are needed to ensure accuracy.
+Many variant callers are benchmarked on simulated data sets, plasmids, or PCR products, and may not have comparable sensitivity and specificity when applied to viral samples.
+Our goal was not to compare the strengths and weaknesses of a few algorithms, but rather to highlight how accuracy can be experiment-specific.
+We recognize that some variant callers may perform better than DeepSNV, and that others may be better suited to other systems.
+However, our work with Lofreq suggests that all methods have inherent limitations and that understanding these limitations is essential.
+We have been able to greatly improve the accuracy of DeepSNV under our experimental conditions, and we are now equipped with an understanding of the limitations of our method.
+
+Our study highlights previously under-recognized issues in  variant calling and suggests factors that should be considered in future studies of viral diversity.
+The need for target amplification, the structure of the viral genome, and variation in input genome copy number may lead to errors specific to given experiment.
+We have shown that these errors can have profound effects on downstream analyses and are especially important in comparative studies of intrahost diversity.
+In studying influenza, we have benefited from an easily accessible reverse genetics system that made for a straightforward approach to benchmarking.
+We have made our data sets and code available to the community so that others may benchmark their own pipelines or improve on our work.
+Next generation sequencing is clearly a powerful tool that promises to help unlock the complex dynamics of intrahost viral populations.
+But no method is perfect, and the proper controls will be needed to bring this promise to fruition.
 
@@
 JTM was supported by the Michigan Predoctoral Training Program in Genetics (T32GM007544).
-
-#References

### 2016-03-24 scripts/figures.Rmd
benchmarking_v1

@@
 ---
-title: "Figures"
-author: "JT McCrone"
-date: "March 21, 2016"
+title: "Benchmarking figures"
 output: pdf_document
@@
 
-miseq.sense.63<-m.roc.table$Sensitivity[m.roc.table$Frequency=="0.63%"]
+miseq.sense.63<-format(m.roc.table$Sensitivity[m.roc.table$Frequency=="0.63%"],digits = dd)
 miseq.sense.16<-m.roc.table$Sensitivity[m.roc.table$Frequency=="0.16%"]
@@
 sum.df<-mutate(sum.df,category=mutation %in% true_snv$mutant) # add Column for True and false variants
-wt1_mut<-subset(sum.df,Id=="WT1",select=mutation)
-wt2_mut<-subset(sum.df,Id=="WT2",select=mutation)
+wt1_mut<-subset(sum.df,Id=="WT1" & freq.var>0.01,select=mutation)
+wt2_mut<-subset(sum.df,Id=="WT2"& freq.var>0.01 ,select=mutation)
 
@@
 onesided.hiseq<-mean(h.table.df$Specificity)
-
-```
+onesided.min<-min(h.table.df$Specificity)
+
+```

### 2016-03-24 scripts/figures.Rmd
jupped the gun on figures

@@
 
-rename(dups.table,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True Positives","adj.specificity"="Specificity","FP"="False Positives"))->dups.table
+rename(dups.table,c("exp.freq"="Frequency","adj.sensitivity"="Sensitivity","TP"="True\nPositives","adj.specificity"="Specificity","FP"="False\nPositives"))->dups.table
 subset(dups.table,select=-c(gc))->dups.table
@@
 
-ggdraw() +
-  draw_plot(hiseq.setup, 0.1, .57, 0.8, 0.43) + #x y width height
-  draw_plot(Hiseq.roc, 0.05, 0.33, .8, 0.33) +
-  draw_plot(tbl, 0.05, 0, .9, 0.33) +
-  draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(1, 0.7, 0.35), size = 15)
-# 
+# ggdraw() +
+#   draw_plot(hiseq.setup, 0.1, .57, 0.8, 0.43) + #x y width height
+#   draw_plot(Hiseq.roc, 0.05, 0.33, .8, 0.33) +
+#   draw_plot(tbl, 0.05, 0, .9, 0.33) +
+#   draw_plot_label(c("A", "B", "C"), c(0.01, 0.01, 0.01), c(1, 0.7, 0.35), size = 15)
+# # 
 # knitr::kable(subset(h.table.df,gc==4))

### 2016-04-28 scripts/figures.Rmd
jupped the gun on figures again
sra info

@@
 
-  mutate(roc.df, exp.freq=find_freq.v(samp))-> roc.df
+  mutate(roc.df, exp.freq=find_freq.v(samp))-> roc.df # this yeilds null in the hiseq.roc function below
 }
@@
   roc.df.adj<-roc.df.adj[order(roc.df.adj$adj.sensitivity),]
-  mutate(roc.df.adj, gc = as.numeric(sub(pat,"\\1",samp)),
+  mutate(roc.df.adj, gc = as.numeric(sub(pat,"\\1",samp)), # This fills in the null column given in the roc_df function above. line 161
        exp.freq = sub(pat,"\\2",samp)
@@
 regions.bed<-mutate(regions.bed,chr=as.character(chr))
-
 # The samples are named in the following format log(genome copy/ul)_expected frequency of variants We'll use that information here to get the gc# and exp.freq for each variant called
@@
 # Id the variants as T or F 
+sum.df<-ddply(sum.df,~chr,primer.cut) # removes any variants outside of priming sites (in this case it removes 0)
+
 true_snv<-read.csv("../data/reference/mutant_id.csv",stringsAsFactor=F) # get the T variants
@@
 wt1_mut<-subset(sum.df,Id=="WT1" & freq.var>0.01,select=mutation)
-wt2_mut<-subset(sum.df,Id=="WT2"& freq.var>0.01 ,select=mutation)
-
-wt_mut<-intersect(wt1_mut$mutation,wt2_mut$mutation)
+wt2_mut<-subset(sum.df,Id=="WT2" &freq.var>0.01 ,select=mutation)
+
+wt_mut<-intersect(wt1_mut$mutation,wt2_mut$mutation) # there aren't any of these
 
@@
 
-possible_vars<-total_possible_vars-length(wt_mut)-dim(mfe.df)[1] # removing the mfe and wt_mut sites
+possible_vars<-total_possible_vars-length(wt_mut)-dim(mfe.df)[1]-20 # removing the mfe and wt_mut sites
 hiseq_vars<-possible_vars
+print(hiseq_vars)
 hiseq_kb<-sum(regions.bed$length)/1000
@@
 
+mapq<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=MapQ,fill=category))+geom_histogram(position="dodge")+theme(legend.position="none")+scale_fill_manual(name="",labels=c("False positives","True positives"),values=c("#b1bdc5ff","#08519cff"))+ylab("SNV count")
+
+phred<-ggplot(subset(sig.df,category %in% c(T,F) & gc==5),aes(x=Phred,fill=category))+geom_histogram(position="dodge")+theme(legend.position="none")+scale_fill_manual(name="",labels=c("False positives","True positives"),values=c("#b1bdc5ff","#08519cff"))+scale_x_continuous(limits = c(30,40),breaks = c(30,32.5,35,37.5,40))+ylab("SNV count")
+
+pdf("../results/figures/mapq.pdf")
+mapq
+dev.off()
+
+pdf("../results/figures/phred.pdf")
+phred
+dev.off()
+
 #mapq_phred.normal
@@
 ### Figure 5
+
+```{r, frequency_accuracy,fig.width=3.4,fig.height=3.4}
+# Here we'll just be using
+
+## Frequency
+cut.read.sig.df<-subset(cut.read.df.all,p.val<0.01)
+medians<-ddply(subset(cut.read.sig.df,category==T ),~exp.freq+gc, summarize, median=median(freq.var))
+
+# ac_freq<-ggplot(subset(sum.df,category==T),aes(x=exp.freq,y=freq.var,fill=as.factor(exp.freq)))+facet_wrap(~gc)
+
+#cut.read.sig.df$exp.freq<-as.factor(cut.read.sig.df$exp.freq)
+#cut.read.sig.df$exp.freq <- factor(cut.read.sig.df$exp.freq, levels = rev(levels(cut.read.sig.df$exp.freq)))
+
+### here weget the R2 of the sample
+
+model<-lm(formula = freq.var~exp.freq,data = subset(cut.read.sig.df,category==T & gc=="5"))
+
+corr_coef<-summary(model)$adj.r.squared
+
+interest<-subset(cut.read.sig.df,category==T & gc=="5")
+
+mean_fold<-mean(abs(interest$freq.var-interest$exp.freq)/interest$exp.freq)
+
+print(mean_fold)
+
+ac_freq<-ggplot(subset(cut.read.sig.df,category==T & gc=="5"),aes(x=exp.freq,y=freq.var,fill=factor(exp.freq,levels=rev(levels(as.factor(exp.freq))))))
+
+ac_freq<-ac_freq+geom_dotplot(binaxis = "y", stackdir = "center",dotsize=1)+scale_fill_manual(values=rhg_cols)
+
+ac_freq<-ac_freq+scale_y_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.05,0.1))#+theme(axis.text.y = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
+ac_freq<-ac_freq+scale_x_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.050,0.1))#+theme(axis.text.x = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
+ac_freq<-ac_freq+geom_abline(intercept = 0, slope = 1,linetype=2,size=1)
+ac_freq<-ac_freq+xlab("Expected Frequency")+ylab("Measured Frequency")+theme(legend.position="none",axis.text.x = element_text(angle = 45, hjust = 1))
+
+ac_freq<-ac_freq+geom_point(data=subset(medians,gc=="5"),mapping=aes(x=exp.freq,y=median),shape=95,size=20)
+ac_freq
+
+pdf("../results/figures/freq.pdf",width=3.4,height=3.4)
+ac_freq
+dev.off()
+
+```
+
+Figure 5. The accuracy of the frequency measurements for the known variants in the 10 $^5$ genomes/$\mu$l samples. The median of each distribution is shown as a black bar.
+
+\pagebreak
+
+### Figure 6
 
@@
 
-Figure 5.
-The accuracy of the Lofreq algorithm on our 10^5% genomes/$\mu$l populations.
+Figure 6.
+The accuracy of the Lofreq algorithm on our 10^5^ genomes/$\mu$l populations.
 A)  The accuracy of Lofreq using standard parameters.
@@
 B) A table summarizing the accuracy of the ROC A using the standard cut offs applied by the algorithm.
-
-\pagebreak
-
-### Figure 6
-
-```{r, frequency_accuracy,fig.width=3.4,fig.height=3.4}
-# Here we'll just be using
-
-## Frequency
-cut.read.sig.df<-subset(cut.read.df.all,p.val<0.01)
-medians<-ddply(subset(cut.read.sig.df,category==T ),~exp.freq+gc, summarize, median=median(freq.var))
-
-# ac_freq<-ggplot(subset(sum.df,category==T),aes(x=exp.freq,y=freq.var,fill=as.factor(exp.freq)))+facet_wrap(~gc)
-
-#cut.read.sig.df$exp.freq<-as.factor(cut.read.sig.df$exp.freq)
-#cut.read.sig.df$exp.freq <- factor(cut.read.sig.df$exp.freq, levels = rev(levels(cut.read.sig.df$exp.freq)))
-
-### here weget the R2 of the sample
-
-model<-lm(formula = freq.var~exp.freq,data = subset(cut.read.sig.df,category==T & gc=="5"))
-
-corr_coef<-summary(model)$adj.r.squared
-
-interest<-subset(cut.read.sig.df,category==T & gc=="5")
-
-mean_fold<-mean(abs(interest$freq.var-interest$exp.freq)/interest$exp.freq)
-
-print(mean_fold)
-
-ac_freq<-ggplot(subset(cut.read.sig.df,category==T & gc=="5"),aes(x=exp.freq,y=freq.var,fill=factor(exp.freq,levels=rev(levels(as.factor(exp.freq))))))
-
-ac_freq<-ac_freq+geom_dotplot(binaxis = "y", stackdir = "center",dotsize=1)+scale_fill_manual(values=rhg_cols)
-
-ac_freq<-ac_freq+scale_y_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.05,0.1))#+theme(axis.text.y = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
-ac_freq<-ac_freq+scale_x_log10(limits=c(0.001,0.1),breaks=c(0.001,0.002,0.005,0.01,0.02,0.050,0.1))#+theme(axis.text.x = element_text(colour = c("red","chartreuse4","black","chocolate2","azure4","blue","black"),face="bold"))
-ac_freq<-ac_freq+geom_abline(intercept = 0, slope = 1,linetype=2,size=1)
-ac_freq<-ac_freq+xlab("Expected Frequency")+ylab("Measured Frequency")+theme(legend.position="none",axis.text.x = element_text(angle = 45, hjust = 1))
-
-ac_freq<-ac_freq+geom_point(data=subset(medians,gc=="5"),mapping=aes(x=exp.freq,y=median),shape=95,size=20)
-ac_freq
-
-pdf("../results/figures/freq.pdf",width=3.4,height=3.4)
-ac_freq
-dev.off()
-
-```
-
-Figure 6. The accuracy of the frequency measurements for the known variants in the 10 $^5$ genomes/$\mu$l samples. The median of each distribution is shown as a black bar.
 
@@
 x<-rename(x,c("names"=""))
+write.csv(x,file="../results/tables/table_1.csv")
 xtab<-xtable(x)
@@
 print(xtab,include.rownames = FALSE,hline.after=c(-1,0,3,6))
-}
+
+}
+
+#
 
@@
 tabulate(5,c(0.05,0.01,0.005))
-
 #
@@
 sum.df<-mutate(sum.df,category=mutation %in% true_snv$mutant) # add Column for True and false variants
+sum.df<-ddply(sum.df,~chr,primer.cut)
+
 wt1_mut<-subset(sum.df,Id=="WT1" & freq.var>0.01,select=mutation)
@@
 ```
+
+```{r,table2}
+data<-read.csv("../data/reference/patient.csv")
+data<-subset(data,p.val<0.01 & freq.var<0.9)
+data<-ddply(data,~chr,primer.cut) # removes sites out side of primers
+
+data.cut<-subset(data,Phred>35 & MapQ>30 & Read_pos<94 & Read_pos>32)
+
+possible_vars<-total_possible_vars
+get_data<-function(x){
+  out<-data.frame(Richness=round(rich(x),digits = 0), Entropy=format(H(x),scientific = T,digits = 3)) #"1/D"=D(x),
+  #rownames(out)<-setting
+  return(out)
+}
+get_data.cut<-function(x){
+  out<-data.frame(Richness.mod=round(rich(x),digits = 0), Entropy.mod=format(H(x),scientific = T,digits = 3)) #"1/D"=D(x),
+  #rownames(out)<-setting
+  return(out)
+}
+
+x<-ddply(data,~Id+DPI+FINAL_RESULT+season+log10(gc_ul),get_data)
+
+y<-ddply(data.cut,~Id+DPI+log10(gc_ul),get_data.cut)
+
+x<-join(x,y)
+x<-subset(x,!(Id%in% c(1240,1241)))
+x<-x[order(x$`log10(gc_ul)`,decreasing = T),]
+write.csv(x,"../results/tables/table_2.cali_vic.csv")
+ggplot(data,aes(pos,freq.var))+geom_point()+facet_wrap(~chr)
+
+ggplot(data.cut,aes(pos,freq.var))+geom_point()+facet_wrap(~chr)
+
+ggplot(data,aes(freq.var))+geom_histogram()+scale_x_log10()
+
+ggplot(data.cut,aes(freq.var))+geom_histogram()+scale_x_log10()+scale_y_log10()
+
+x<-subset(data.cut,Id %in% c(1376,1401,1405,1374,1227,1321,1229,1245))
+```

### 2016-04-28 scripts/change_names_sra.py
fix srainfo files from excel format

@@ -1,0 +1,87 @@
+#import sys
+import os
+import gzip
+import argparse
+import shutil
+import glob
+parser = argparse.ArgumentParser(description='This program takes hiseq fastq files and renames them as sample.read_direction.#.fastq and keeps a log of the change. It renames the files according the Library name in the runifo csv. This relies heavily on the change_name scripts in Variant_pipeline.
+
+-JT McCrone 2016')
+parser.add_argument('-s',action='store',dest='s',help='The sorce directory containing the original fastq files')
+parser.add_argument('-f',action='store',dest='f',help='The final directory that will hold the renamed fastq files')
+parser.add_argument('-k',action='store',dest='key',help='The sra runinfo csv that serves as the key for renaming')
+parser.add_argument('-run',action='store_true',dest='test',default=False,help='Boolean switch to run program, without this the program runs in test mode: the log is made but no files are renamed')
+
+args=parser.parse_args()
+s=args.s
+f=args.f
+key=args.key
+test=args.test
+
+#make the output dir if it doesn't exist
+if not os.path.exists(f):
+    os.makedirs(f)
+
+# add slash to final location if it was forgotten
+if f[-1] != '/':
+    f=f+'/'
+#print f
+
+if s[-1] != '/':
+    s=s+'/'
+#print s
+
+
+# input argument is the sample sheet
+junk_names = [] # The names given by the sequencing core sampleid_index
+new_names = [] # the names we want, which are in the same row as the data taht is used to make the bad names. 
+
+# so junk_names[i] should be the junk name for new_names[i] as they both come from the ith column.
+# add the bad names from the sample sheet to a list. This uses columns of the sheet to construct the names according to the capricous nature of the sequencing core.Although they may have finally settled down a bit.
+#print(key)
+names =  open(key,"r")
+next(names) # skip the header of the csv
+for line in names:
+    line = line.strip()
+    line = line.split(',')
+    junk_names.append(line[0])    
+    new=line[11]   
+    new_names.append(new)
+   # print(line)
+names.close()
+
+#print(new_names)
+#print(junk_names)
+
+if test==False:
+    print "running in test mode add option -run to run"
+    
+outfile = open(f+'renaming_log.txt','w')
+
+
+# Now to search through the fastq files or fastq.gz and copy them to the new makes
+for filename in glob.glob(s + "*.fastq"):
+    path=os.path.abspath(filename) # getteh full name and path
[…]

### 2016-04-29 results/figures.Rmd
makefile through bpipe

--- 2016-04-28 scripts/figures.Rmd
@@
 
-sum.df<-read.csv("../data/process/2014-5-30/one.sided/Variants/all.sum.csv",stringsAsFactors=F,comment.char = '#')
+sum.df<-read.csv("../data/process/2014-5-30/Variants/all.sum.csv",stringsAsFactors=F,comment.char = '#')
 
@@
 #sum.df<-read.csv("~/Box Sync/Lab_work/NGS/Benchmarking/data/2015-11-14/all.sum.csv")
-pat<-"([0-9]+)_([0-9]+)"
-mutate(dup.df, exp.freq = sub(pat,"\\1",Id),
-       dup = as.numeric(sub(pat,"\\2",Id)),
+pat<-"([0-9]+)_([0-9]+)_([0-9]+)"
+mutate(dup.df, exp.freq = sub(pat,"\\2",Id),
+       dup = as.numeric(sub(pat,"\\3",Id)),
 )-> dup.df
 
-dup.df<-subset(dup.df,Id!="PC") # remove the plasmid control that was run
+#dup.df<-subset(dup.df,Id!="PC") # remove the plasmid control that was run
 mutate(dup.df,Id=paste(4,exp.freq,sep="_"),exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->dup.df

### 2016-04-29 scripts/change_names_sra.py

@@ -4,9 +4,7 @@ import gzip
 import argparse
 import shutil
 import glob
-parser = argparse.ArgumentParser(description='This program takes hiseq fastq files and renames them as sample.read_direction.#.fastq and keeps a log of the change. It renames the files according the Library name in the runifo csv. This relies heavily on the change_name scripts in Variant_pipeline.
-
--JT McCrone 2016')
+parser = argparse.ArgumentParser(description='This program takes hiseq fastq files and renames them as sample.read_direction.#.fastq and keeps a log of the change. It renames the files according the Library name in the runifo csv. This relies heavily on the change_name scripts in Variant_pipeline. -JT McCrone 2016')
 parser.add_argument('-s',action='store',dest='s',help='The sorce directory containing the original fastq files')
 parser.add_argument('-f',action='store',dest='f',help='The final directory that will hold the renamed fastq files')
 parser.add_argument('-k',action='store',dest='key',help='The sra runinfo csv that serves as the key for renaming')
@@ -50,8 +48,6 @@ for line in names:
    # print(line)
 names.close()
 
-#print(new_names)
-#print(junk_names)
 
 if test==False:
     print "running in test mode add option -run to run"
@@ -63,11 +59,12 @@ outfile = open(f+'renaming_log.txt','w')
 for filename in glob.glob(s + "*.fastq"):
     path=os.path.abspath(filename) # getteh full name and path
     name=filename.split("_")
-    bad_name = name[0] # the bad name is the first bit
+    just_file=name[0].split("/")
+    bad_name = just_file[-1] # the bad name is the first bit
     lane_junk = name[1]
     read_number=lane_junk.split(".fastq") # info about the read 
-    fastq_number=1 #always 1 from the sra
-    read_number=read_number[0] # read direction
+    fastq_number=str(1) #always 1 from the sra
+    read_number=str(read_number[0]) # read direction
     if bad_name in junk_names: # the new name and the bad name have the same index in their respective lists
         name_index = junk_names.index(bad_name)
         better_name= new_names[name_index]

### 2016-04-29 scripts/combine.py
working on lofreq analysis

@@ -1,0 +1,38 @@
+#!/usr/bin/python
+
+# adapted from  http://stackoverflow.com/questions/3207219/how-to-list-all-files-of-a-directory-in-python/24652788#24652788
+
+# Usage input the bash style reg ex using (not including the *)  and this will combine all files together only including the header from the first one
+# This is used to combine all the output files into one concatinated file.
+
+
+ # input_dir end_pattern output_file
+import glob
+import os
+import sys
+
+os.chdir(sys.argv[1]) # The first input is the directory in which to look for files
+ending=sys.argv[2]  # The second input is the ending of the files that you want to concatenate.
+
+# Make a list of all the files in the input dir matching the description
+csv_files=[] # doesn't have to be csv files anymore
+for file in glob.glob("*"+ending): 
+    csv_files.append(file)
+print(csv_files)
+
+
+fout=open(sys.argv[3],"w") # The third input is the name of the output file
+# first file: write all the lines to the output file
+with open(csv_files[0],"r") as first_file:
+    print("working with " + csv_files[0])
+    for line in first_file:
+        fout.write(line)
+
+# now the rest:    
+for num in range(1,len(csv_files)):
+    print("working with " + csv_files[num])
+    with open(csv_files[num],"r") as next_file:
+        next_file.next() # skip the header
+        for line in next_file:
+            fout.write(line)
+fout.close()

### 2016-04-29 scripts/mapq_vcf.py

@@ -1,0 +1,83 @@
+#!/usr/bin/python
+
+import sys
+import numpy as np
+import pysam
+from collections import Counter
+import vcf
+from vcf.parser import _Info as VcfInfo
+
+
+input=sys.argv[1].split(".")[0]
+input=input.split("/")[-1]
+bam=sys.argv[2].split(".")[0]
+bam=bam.split("/")[-1]
+
+#Ensure the bam and csv file match
+if input==bam:
+        print("working with vcf: "+input + " and bam: " +bam)
+else:
+   print( "bam:"+bam +" does not match vcf: " +input)
+   sys.exit(1)
+
+
+
+
+
+
+#Put header on true false and exp csv files
+
+in_var= vcf.Reader(open(sys.argv[1], 'r'))
+## update infos ##
+in_var.infos['MapQ']=VcfInfo(id='MapQ',num=1,type='Float',desc="The average MapQ of the reads containing the called variant")#,source=None, version=None)
+in_var.infos['Read_pos']=VcfInfo(id='Read_pos',num=1,type='Float',desc="The average read cycle that called the given variant")#,source=None, version=None)
+in_var.infos['Phred']=VcfInfo(id='Phred',num=1,type='Float',desc="The average Phred score of the called variant")#,source=None, version=None)
+
+
+
+
+
+variants=list(in_var)
+with pysam.AlignmentFile(sys.argv[2], "rb") as bamfile:
+
+#with pysam.AlignmentFile('../data/5_5.removed.bam', "rb") as bamfile:
+	for record in variants:
+		#print(record)
+		mapq=[]# This will hold a list of the mapping qualtties that map to the variant
+		phred=[] #This will hold a list of the phred that map to the variant
+		Read_pos=[]  # This will hold a list of position relative to the read
+		chr=record.CHROM
+		pos=int(record.POS)
+		py_pos=pos-1
+		ref=record.REF
+		var=record.ALT[0]
+		record.ID=input
+		for pileupcolumn in bamfile.pileup(chr,py_pos,py_pos+1,truncate=True,stepper="all",max_depth=1E6):
+			if pileupcolumn.pos==py_pos:
+				for pileupread in pileupcolumn.pileups:
+					if  not pileupread.is_del and not pileupread.is_refskip:
+						called_base=pileupread.alignment.query_sequence[pileupread.query_position]
+						called_phred=pileupread.alignment.query_qualities[pileupread.query_position]
[…]

### 2016-04-29 scripts/combine.py
lofreq pipeline is a go and in makefile

@@ -22,14 +22,9 @@ print(csv_files)
 
 
 fout=open(sys.argv[3],"w") # The third input is the name of the output file
-# first file: write all the lines to the output file
-with open(csv_files[0],"r") as first_file:
-    print("working with " + csv_files[0])
-    for line in first_file:
-        fout.write(line)
 
 # now the rest:    
-for num in range(1,len(csv_files)):
+for num in range(0,len(csv_files)):
     print("working with " + csv_files[num])
     with open(csv_files[num],"r") as next_file:
         next_file.next() # skip the header

### 2016-04-29 scripts/mapq_vcf.py

@@ -8,17 +8,19 @@ import vcf
 from vcf.parser import _Info as VcfInfo
 
 
-input=sys.argv[1].split(".")[0]
-input=input.split("/")[-1]
+input=sys.argv[1].split("/")[-1]
+
+input=input.split(".")[0]
 bam=sys.argv[2].split(".")[0]
 bam=bam.split("/")[-1]
 
+#print(input)
 #Ensure the bam and csv file match
-if input==bam:
-        print("working with vcf: "+input + " and bam: " +bam)
-else:
-   print( "bam:"+bam +" does not match vcf: " +input)
-   sys.exit(1)
+#if input==bam:
+#        print("working with vcf: "+input + " and bam: " +bam)
+#else:
+#   print( "bam:"+bam +" does not match vcf: " +input)
+#   sys.exit(1)
 
 
 

### 2016-04-29 scripts/vcf_csv.py

@@ -30,18 +30,18 @@ header='CHROM,POS,ID,REF,ALT,QUAL,FILTER,'
 
 first=vcf.Reader(open(sys.argv[1], 'r'))
 first=first.next() # first varaint to get the infos that are present
-print first
+#print first
 for x in first.INFO:
     header=header+x+","
 
 out_csv.write(header+'\n')
 
 for record in in_var:
-    print record
+    #print record
     line=str(record.CHROM)+','+str(record.POS)+','+ str(record.ID)+','+str(record.REF)+','+str(record.ALT[0])+','+str(record.QUAL)+','+str(record.FILTER)+','
     for y in record.INFO:
         line=line+str(record.INFO[y]).replace(",",":")+","
-    print line
+   # print line
     out_csv.write(line+'\n')
 
 out_csv.close()

### 2016-05-03 scripts/get_fastq.sh
new get fastq approach

@@ -1,0 +1,11 @@
+#!/bin/bash
+SRA=($(cut -f1 -d ',' $1 |grep -v "Run"))
+    for (( i=0; i<${SRA[@]}; i++ )); do 
+      echo downloading ${SRA[i]}
+      prefetch ${SRA[i]}
+      sra_file=$(ls ~/ncbi/public/sra/${SRA[i]}.sra) 
+      echo converting $sra_file to fastq
+      fastq-dump  $sra_file --split-files -O $2
+      echo removing $sra_file
+      rm $sra_file
+       done    

### 2016-05-04 scripts/get_fastq.sh
from makefile

@@ -1,6 +1,6 @@
 #!/bin/bash
 SRA=($(cut -f1 -d ',' $1 |grep -v "Run"))
-    for (( i=0; i<${SRA[@]}; i++ )); do 
+    for (( i=0; i<${#SRA[@]}; i++ )); do 
       echo downloading ${SRA[i]}
       prefetch ${SRA[i]}
       sra_file=$(ls ~/ncbi/public/sra/${SRA[i]}.sra) 

### 2016-05-04 results/figures.Rmd
lofreq pipeline is a go and in makefile
work updating  Rmd for lofreq

@@
 title: "Benchmarking figures"
-output: pdf_document
+output: github_document
 ---
@@
 require(png)
+require(grid)
 theme_set(new = theme_classic())
@@
 
-lofreq.df<-read.csv("../data/process/2015-6-23/lofreq/all.removed.mapq_vcf.vcf_csv.csv",stringsAsFactors = F)
-
+lofreq.df<-read.csv("../data/process/2015-6-23/Lofreq/lofreq_csv/all.lofreq.csv",stringsAsFactors = F,comment.char="#")
+
+lofreq.df<-mutate(lofreq.df,mutation=paste0(CHROM,"_",REF,POS,ALT))
+lofreq.df<-mutate(lofreq.df,category=mutation %in% true_snv$mutant,Id=ID)
 lofreq.df$category[lofreq.df$mutation %in% wt_mut]<-"wt"
-lofreq.df$categroy[lofreq.df$mutation %in% mfe.df$mutation]<-"mfe"
-
-lofreq.df$p.val<-lofreq.df$QUAL # to make the roc based on the quality scores given by lofreq
-
+lofreq.df$category[lofreq.df$mutation %in% mfe.df$mutation]<-"mfe"
+
+lofreq.df$p.val<-as.numeric(lofreq.df$QUAL) # to make the roc based on the quality scores given by lofreq
 lofreq.roc<-hiseq.roc(lofreq.df,possible_tp,possible_vars,"<")

### 2016-05-04 results/figures.Rmd
fixes formating issues caused by SRA fastq names and such

@@
 hiseq.roc<-function(sum.df,possible_tp,possible_vars,direction){ # A function to run the roc calculations and adjustments
+  pat<-"([0-9]+)_([0-9]+)"
   roc.ls<-dlply(sum.df,~Id,sum_roc,direction)
@@
 hiseq.roc.table<-function(sum.df,cut.off,possible_tp,possible_vars,direction){
+  pat<-"([0-9]+)_([0-9]+)"
   roc.ls<-dlply(sum.df,~Id,sum_roc,direction)
@@
 lofreq.df<-mutate(lofreq.df,mutation=paste0(CHROM,"_",REF,POS,ALT))
-lofreq.df<-mutate(lofreq.df,category=mutation %in% true_snv$mutant,Id=ID)
+lofreq.df<-mutate(lofreq.df,category=mutation %in% true_snv$mutant,Id=ID,chr=CHROM,pos=POS,freq.var=AF)
 lofreq.df$category[lofreq.df$mutation %in% wt_mut]<-"wt"
 lofreq.df$category[lofreq.df$mutation %in% mfe.df$mutation]<-"mfe"
+lofreq.df<-ddply(lofreq.df,~chr,primer.cut)
 
 lofreq.df$p.val<-as.numeric(lofreq.df$QUAL) # to make the roc based on the quality scores given by lofreq
+lofreq.df<-mutate(lofreq.df, gc = sub(pat,"\\1",Id),
+       exp.freq = sub(pat,"\\2",Id)
+)
+mutate(lofreq.df,exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->lofreq.df
 lofreq.roc<-hiseq.roc(lofreq.df,possible_tp,possible_vars,"<")
@@
 #knitr::kable(subset(lofreq.table,gc==5,select=c(exp.freq,TP,FP)))
-lofreq.table<-head(subset(lofreq.table,gc==5,select=-c(gc,Id)),n=-1) # don't need the 0.1% sample
+lofreq.table<-subset(lofreq.table,gc==5,select=-c(gc,Id)) # don't need the 0.1% sample
 rename(lofreq.table,c("exp.freq"="Frequency","TP"="True\nPositives","FP"="False\nPositives"))->lofreq.table
@@
 dup.df$category[dup.df$mutation %in% wt_mut]<-"wt"
+dup.df$category[dup.df$mutation %in% mfe.df$mutation]<-"mfe"
 
@@
 
-pdf("../results/figures/dups.roc.pdf",width=3.4,height=5)
-ggdraw() +
-  draw_plot(dups.roc.p, 0, 0.5, 0.9, 0.45) +
-  draw_plot(tbl, 0.025, 0, 0.95, .48) +
-  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
-dev.off()
+# pdf("../results/figures/dups.roc.pdf",width=3.4,height=5)
+# ggdraw() +
+#   draw_plot(dups.roc.p, 0, 0.5, 0.9, 0.45) +
+#   draw_plot(tbl, 0.025, 0, 0.95, .48) +
+#   draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
+# dev.off()
 
@@
 
-```{r,results="asis",cache=F}
+```{r,cache=F}
 tabulate(5,c(0.05,0.01,0.005))
@@
 write.csv(x,"../results/tables/table_2.cali_vic.csv")
-ggplot(data,aes(pos,freq.var))+geom_point()+facet_wrap(~chr)
-
-ggplot(data.cut,aes(pos,freq.var))+geom_point()+facet_wrap(~chr)
-
-ggplot(data,aes(freq.var))+geom_histogram()+scale_x_log10()
-
-ggplot(data.cut,aes(freq.var))+geom_histogram()+scale_x_log10()+scale_y_log10()
+# ggplot(data,aes(pos,freq.var))+geom_point()+facet_wrap(~chr)
+# 
+# ggplot(data.cut,aes(pos,freq.var))+geom_point()+facet_wrap(~chr)
+# 
+# ggplot(data,aes(freq.var))+geom_histogram()+scale_x_log10()
+# 
+# ggplot(data.cut,aes(freq.var))+geom_histogram()+scale_x_log10()+scale_y_log10()
 
@@
 ```
+
+```{r}
+sessionInfo()
+```

### 2016-05-06 results/figures.Rmd
fixed bug in data handeling

@@
 ```{r,lofreq,fig.height=5,fig.width=3.4}
-
 lofreq.df<-read.csv("../data/process/2015-6-23/Lofreq/lofreq_csv/all.lofreq.csv",stringsAsFactors = F,comment.char="#")
 
-lofreq.df<-mutate(lofreq.df,mutation=paste0(CHROM,"_",REF,POS,ALT))
-lofreq.df<-mutate(lofreq.df,category=mutation %in% true_snv$mutant,Id=ID,chr=CHROM,pos=POS,freq.var=AF)
+lofreq.df<-rename(lofreq.df,c("CHROM"="chr",
+                        "POS"="pos",
+                        "ID"="Id",
+                        "REF"="ref",
+                        "ALT"="var",
+                        "AF"="freq.var"
+))
+
+lofreq.df<-mutate(lofreq.df,mutation=paste0(chr,"_",ref,pos,var))#,category=mutation %in% tp$mutant)->lofreq.df
+
+#lofreq.df$Id[lofreq.df$Id=="3_03"]<-"3_02" # correct a naming error
+
+WT<-subset(lofreq.df,grepl("WT",Id)) # Separate WT samples 
+lofreq.df<-subset(lofreq.df,!(grepl("WT",Id))& Id!="Plasmid_control")
+
+#lofreq.df<-mutate(lofreq.df,p.val=10^(QUAL/-10))
+
+pat<-"([0-9]+)_([0-9]+)"
+mutate(lofreq.df, gc = sub(pat,"\\1",Id),
+       exp.freq = sub(pat,"\\2",Id)
+)-> lofreq.df
+mutate(lofreq.df,exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->lofreq.df
+
+# TF column
+lofreq.df<-mutate(lofreq.df,category=mutation %in% true_snv$mutant) # add Column for True and false variants
+wt1_mut<-subset(WT,Id=="WT1",select=mutation)
+wt2_mut<-subset(WT,Id=="WT2",select=mutation)
+
+wt_mut<-intersect(wt1_mut$mutation,wt2_mut$mutation)
+
+length(wt_mut)
 lofreq.df$category[lofreq.df$mutation %in% wt_mut]<-"wt"
-lofreq.df$category[lofreq.df$mutation %in% mfe.df$mutation]<-"mfe"
-lofreq.df<-ddply(lofreq.df,~chr,primer.cut)
-
-lofreq.df$p.val<-as.numeric(lofreq.df$QUAL) # to make the roc based on the quality scores given by lofreq
-lofreq.df<-mutate(lofreq.df, gc = sub(pat,"\\1",Id),
-       exp.freq = sub(pat,"\\2",Id)
-)
-mutate(lofreq.df,exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->lofreq.df
+
+lofreq.df$category[lofreq.df$mutation %in% wt_mut]<-"wt"
+lofreq.df$categroy[lofreq.df$mutation %in% mfe.df$mutation]<-"mfe"
+
+lofreq.df$p.val<-lofreq.df$QUAL # to make the roc based on the quality scores given by lofreq
+
 lofreq.roc<-hiseq.roc(lofreq.df,possible_tp,possible_vars,"<")
@@
 
-#knitr::kable(subset(lofreq.table,gc==5,select=c(exp.freq,TP,FP)))
 lofreq.table<-subset(lofreq.table,gc==5,select=-c(gc,Id)) # don't need the 0.1% sample
@@
 lofreq.table<-subset(lofreq.table,select=c("Frequency","Sensitivity","True\nPositives","Specificity","False\nPositives"))
+
 tt2<-ttheme_minimal(colhead=list(fg_params=list(fontface=1,cex=0.65,fontfamily="serif")),core=list(fg_params=list(cex=0.65,fontfamily="serif")))
@@
 
-pdf("../results/figures/lofreq1.pdf",width=3.4,height=5)
-ggdraw() +
-  draw_plot(lofreq.roc.p, 0, 0.5, 0.9, 0.45) +
-  draw_plot(tbl, 0.025, 0, 0.95, .48) +
-  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
-dev.off()
+# pdf("../results/figures/lofreq1.pdf",width=3.4,height=5)
+# ggdraw() +
+#   draw_plot(lofreq.roc.p, 0, 0.5, 0.9, 0.45) +
+#   draw_plot(tbl, 0.025, 0, 0.95, .48) +
+#   draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
+# dev.off()
 

### 2016-05-10 results/figures.Rmd
rearrange figure doc

@@
 
-### Figure 1
-```{r, roc_skematic,fig.height=4,fig.width=3.4}
-set.seed(42) # make the figures reproducible
-example<-data.frame(Condition=c(rep(x = c(TRUE,FALSE),each=20)),test=c(runif(20,-10,-4),runif(20,-6,0))) # Make a dataframe of T and F positives with random p values that favors a lower distribution for the TP
-ex.roc<-roc(example$Condition~example$test) # make an ROC based on this data
-
-cord<-coords(ex.roc,x="all") # get the coordinates of the ROC
-cord.l<-melt(cord) # melt the coodinates for ease of plotting
-cord.l<- mutate(cord.l,Id=c(0,head(as.numeric(rownames(cord.l))%/%3,-1))) # Add an Id so that the specificity and sensitivity at a given threshold can be groupled together
-
-# Set an id column to group the threshold,specificity, and sensitivity together with the same number by integer division.
-I adjust with c(0, head ..., -1) since 3%/%3=1 but it should be grouped with the 0's.
-roc.df<-dcast(cord.l,Id~Var1) # dcast into columns of threshold, specificity and sensitivity
-roc.df<-roc.df[order(roc.df$sensitivity),] # So that the points connect nicely
-
-#Helper function that gets the threshold and sensitivity from and roc at a given speceificity
-helper.roc<-function(spec,roc.df){
-  above.sense.cut<-subset(roc.df,specificity>=spec,select=c(threshold,specificity,sensitivity)) # get the points where specificity is above the cutoff
-  sense.min<-min(above.sense.cut$specificity) # get the minimum of these to get as close to desired sensitivity as possible
-  sense.cut<-subset(above.sense.cut,specificity==sense.min) # get all rows where sensitivity is at this point
-  spec.max<-max(sense.cut$sensitivity) # but there can be different FPR for each TPR so here we maximize specificity
-  output<-subset(sense.cut,sensitivity==spec.max)
-}
-
-plot.ex<-function(spec1,spec2,roc.df,distribution){ 
-  cut1<-helper.roc(spec1,roc.df)#dataframe with the threshold,sensitivity and specificity at the given specificity
-  cut2<-helper.roc(spec2,roc.df)
-  output<-rbind(cut1,cut2)
-  lines<-data.frame(sense=c(output$sensitivity,output$sensitivity,0,0), # make a data frame with the coordinates for the lines whicha re to be added
-                    spec=c(0,0,1-output$specificity,1-output$specificity), # really 1-specificity
-                    group=rep(c(spec1,spec2),times=3))
-
-  #plot the roc
-  roc.p<-ggplot(roc.df,aes(x=1-specificity,y=sensitivity))+geom_step(color="black",size=1.5)+xlab("1-Specificity")+ylab("Sensitivity")+theme(axis.title.y=element_text(vjust=1))
-  roc.p<-roc.p+geom_line(data=lines,aes(x=spec,y=sense,group=as.factor(group),color=rep(c("#08519cff","#787878ff"),times=3)),size=c(1),lty=c(2))+scale_color_manual(values=c("#08519cff","#787878ff"))
-
- roc.p<-roc.p+theme(axis.title.y = element_text(vjust=1.2))+theme(axis.title.x = element_text(vjust=0))+theme(legend.position="none")
-
- #plot the points and cut offs
- points.p<-ggplot(distribution,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center",dotsize = 1.5)+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(),axis.title.y = element_text(vjust=1.2),legend.position=c(0.9,0.65),legend.background = element_rect(fill = alpha("white",0)))+xlab("")
-
- points.p<-points.p+geom_hline(yintercept=output$threshold,
-                          size=1,lty=2,color=c("#08519cff","#787878ff")) # these are the first 2 colors in set 1
-
-ggdraw() +
-  draw_plot(points.p, 0, .45, .8, 0.48) + #x y width height
-  draw_plot(roc.p, 0, 0, .8, 0.48) +
-  draw_plot_label(c("A", "B"), c(0.05, 0.05), c(1, 0.55), size = 15)
-
-}
-plot.ex(0.35,0.95,roc.df,example)
-
-pdf("../results/figures/roc.example.pdf",width=3.4,height=4)
-plot.ex(0.35,0.95,roc.df,example)
-dev.off()
-pdf("~/Box Sync/Lab_work/Talks/2016_812/figures/roc.ex.pdf",width = 12,height = 8)
-plot.ex(0.35,0.95,roc.df,example)
-dev.off()
-
-```
-
-Figure 1.
-An example ROC curve.
-A) Hypothetical variants are stratified by log of the p value.
-True positives variants are shown as closed circles, while false positive variants are represented as open circles.
-Potential thresholds are indicated as colored lines.
-B) An ROC curve made from the hypothetical data shown in A.
-The colored lines indicate the points on the curve made by the thresholds in A.
-\pagebreak
 ```{r,roc_helper_runctions}
@@
 
-Figure 2.
-Initial DeepSNV accuracy.
-A) Reconstitued cDNA genomes of influenza strain WSN33 were diluted serially into PR8 prior to sequencing on an Illumina Miseq machine.
-B) An ROC curve measuring the accuracy of DeepSNV in identifying WSN33 variants mixed with PR8 at the indicated frequencies.
-C) A summary the data in B at a p value threshold of 0.01.
 \pagebreak
-
-```{r,Miseq_read.cut,eval=F}
-### With read position cut off 63 and 188
-read.df<-subset(sum.df,Read_pos>62 & Read_pos<188)
-m.roc.df<-miseq.roc(read.df,possible_tp,possible_vars,">")
-
-miseq.roc.p<-ggplot(m.roc.df,aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_manual(values=rhg_cols)+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))#+scale_x_continuous(limits=c(0,0.005)
-
-miseq.roc.p
-
-# pdf("../results/figures/miseq_roc.one.sided.pdf")
-# miseq.roc
-# dev.off()
-
-m.roc.table<-miseq.roc.table(read.df,0.01,possible_tp,possible_vars,">")
-knitr::kable(m.roc.table)
-
-ggplot(subset(sum.df,p.val<0.01),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=1)
-
-```
 
@@
 
-Figure 3.
-Decreased accuracy under patient-derived sample conditions A) 20 known single point mutants were diluted into WT WSN33 at decreasing frequencies.
-These populations were diluted further in basal media to match the genome concentrations found in patint-dervied samples (10$^5$-10$^3$ genomes/$\mu$l).
-B) The accuracy of DeepSNV in idenfying the 20 known true positive variants in the 10 $^5$ genomes/$\mu$l samples.
-C) A summary table of the accuracy of our method for the 10 $^5$ genomes/$\mu$l samples at a p value threshold of 0.01.
-
 \pagebreak
@@
 
-Figure 4.
-Accuracy can be improved through quality thresholds.
-A) We stratified all called variants from the 10 $^5$ genomes/$\mu$l samples with p<0.01 stratified by the mean mapping quality of the reads containing the variant and the mean Phred scores of the variant bases.
-Dashed lines indicate  common cutoffs of 20 and 30 for mapping quality and Phred respectively.
-B) A histogram of average position on a pair end read the variants that passed our mean MapQ threshold of 30 and mean Phred threshold of 35.
-C) The accuracy of our analyis after applying the following quality cut offs : mean MapQ> 30, mean Phred> 35, and an average read position between 32 and 94 (the middle 50% of the read).
-D) A table summarizing accuracy in C at a p value threshold of 0.01.
-
 \pagebreak
@@
 ```
-
-Figure 5. The accuracy of the frequency measurements for the known variants in the 10 $^5$ genomes/$\mu$l samples. The median of each distribution is shown as a black bar.
 
@@
 ```
-
-Figure 6.
-The accuracy of the Lofreq algorithm on our 10^5^ genomes/$\mu$l populations.
-A)  The accuracy of Lofreq using standard parameters.
-The specificity of the caller was scaled to account for the same number of tests as the DeepSNV algorithm.
-B) A table summarizing the accuracy of the ROC A using the standard cut offs applied by the algorithm.
 
@@
 
-Figure 7.
-The accuracy of DeepSNV decreased when applied to samples with less input nucleic acid.
-A) An ROC curve for the 10 $^4$ genomes/$\mu$l samples.
-B) A table summarizing the accuracy at 10 $^4$ genomes/$\mu$l with a  p value threshold of 0.01.
-C) An ROC curve depicting the accuracy of our method when applied to the 10 $^3$ genomes/$\mu$l samples.
-B) A table summarizing the accuracy at 10 $^3$ genomes/$\mu$l with a  p value threshold of 0.01
-
 \pagebreak
@@
 
-Figure 8.
-The accuracy of DeepSNV at lower input levels is improved by processing samples in duplicate.
-A) An ROC curve of  the 10 $^4$ genomes/$\mu$l samples processed in duplicate.
-A variant was required to be found in both duplicates to be considered.
-B) A table summarizing the accuracy when applied to duplicate samples with 10 $^4$ genomes/$\mu$l input, p<0.01.
-
 \pagebreak
@@
 ### Table 1.
-
-Table 1.
-Accurate SNV indentification is vital for accurate measure of intrahost diversity.
-We measured the richness and Shannon's Entropy of the designated 10^5^ genomes$\mu$l populations at each stage of our benchmarking procedure and compared the results with what is expected.
-Additionaly we calculated the L1-norm of the populations relative to the expected population.
 
@@
 
+## Table 2
+```{r,table2}
+data<-read.csv("../data/reference/patient.csv")
+data<-subset(data,p.val<0.01 & freq.var<0.9)
+data<-ddply(data,~chr,primer.cut) # removes sites out side of primers
+
+data.cut<-subset(data,Phred>35 & MapQ>30 & Read_pos<94 & Read_pos>32)
+
+possible_vars<-total_possible_vars
+get_data<-function(x){
+  out<-data.frame(Richness=round(rich(x),digits = 0), Entropy=format(H(x),scientific = T,digits = 3)) #"1/D"=D(x),
+  #rownames(out)<-setting
+  return(out)
+}
+get_data.cut<-function(x){
+  out<-data.frame(Richness.mod=round(rich(x),digits = 0), Entropy.mod=format(H(x),scientific = T,digits = 3)) #"1/D"=D(x),
+  #rownames(out)<-setting
+  return(out)
+}
+
+x<-ddply(data,~Id+DPI+FINAL_RESULT+season+log10(gc_ul),get_data)
+
+y<-ddply(data.cut,~Id+DPI+log10(gc_ul),get_data.cut)
+
+x<-join(x,y)
+x<-subset(x,!(Id%in% c(1240,1241)))
+x<-x[order(x$`log10(gc_ul)`,decreasing = T),]
+write.csv(x,"../results/tables/table_2.cali_vic.csv")
+# ggplot(data,aes(pos,freq.var))+geom_point()+facet_wrap(~chr)
+# 
+# ggplot(data.cut,aes(pos,freq.var))+geom_point()+facet_wrap(~chr)
+# 
+# ggplot(data,aes(freq.var))+geom_histogram()+scale_x_log10()
+# 
+# ggplot(data.cut,aes(freq.var))+geom_histogram()+scale_x_log10()+scale_y_log10()
+
+x<-subset(data.cut,Id %in% c(1376,1401,1405,1374,1227,1321,1229,1245))
+```
+
+## Other analysis
+
 ```{r,coverage}
@@
 
-```{r,table2}
-data<-read.csv("../data/reference/patient.csv")
-data<-subset(data,p.val<0.01 & freq.var<0.9)
-data<-ddply(data,~chr,primer.cut) # removes sites out side of primers
-
-data.cut<-subset(data,Phred>35 & MapQ>30 & Read_pos<94 & Read_pos>32)
-
-possible_vars<-total_possible_vars
-get_data<-function(x){
-  out<-data.frame(Richness=round(rich(x),digits = 0), Entropy=format(H(x),scientific = T,digits = 3)) #"1/D"=D(x),
-  #rownames(out)<-setting
-  return(out)
-}
-get_data.cut<-function(x){
-  out<-data.frame(Richness.mod=round(rich(x),digits = 0), Entropy.mod=format(H(x),scientific = T,digits = 3)) #"1/D"=D(x),
-  #rownames(out)<-setting
-  return(out)
-}
-
-x<-ddply(data,~Id+DPI+FINAL_RESULT+season+log10(gc_ul),get_data)
-
-y<-ddply(data.cut,~Id+DPI+log10(gc_ul),get_data.cut)
-
-x<-join(x,y)
-x<-subset(x,!(Id%in% c(1240,1241)))
-x<-x[order(x$`log10(gc_ul)`,decreasing = T),]
-write.csv(x,"../results/tables/table_2.cali_vic.csv")
-# ggplot(data,aes(pos,freq.var))+geom_point()+facet_wrap(~chr)
-# 
-# ggplot(data.cut,aes(pos,freq.var))+geom_point()+facet_wrap(~chr)
-# 
-# ggplot(data,aes(freq.var))+geom_histogram()+scale_x_log10()
-# 
-# ggplot(data.cut,aes(freq.var))+geom_histogram()+scale_x_log10()+scale_y_log10()
-
-x<-subset(data.cut,Id %in% c(1376,1401,1405,1374,1227,1321,1229,1245))
-```
-
 ```{r}

### 2016-05-17 results/figures.Rmd
clean up

@@
 dup.df<-read.csv("../data/process/2015-11-14/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#")
-#sum.df<-read.csv("~/Box Sync/Lab_work/NGS/Benchmarking/data/2015-11-14/all.sum.csv")
 pat<-"([0-9]+)_([0-9]+)_([0-9]+)"
 mutate(dup.df, exp.freq = sub(pat,"\\2",Id),
-       dup = as.numeric(sub(pat,"\\3",Id)),
+       dup = as.numeric(sub(pat,"\\3",Id))
 )-> dup.df
@@
 dup.df<-mutate(dup.df,category=mutation %in% true_snv$mutant) # add Column for True and false variants
-dup.df$category[dup.df$mutation %in% wt_mut]<-"wt"
+#dup.df$category[dup.df$mutation %in% wt_mut]<-"wt" There are no wt mutants see wt_mut. For some reason this was causing an error.
 dup.df$category[dup.df$mutation %in% mfe.df$mutation]<-"mfe"
 
-cut.df<-subset(dup.df,Phred>35 & MapQ>30 & Read_pos<94 & Read_pos>32)
+cut.df<-subset(dup.df,Phred>=35 & MapQ>=30 & Read_pos<=94 & Read_pos>=32)
 
